---
title: "EGL002"
author: "Emanuelle Grody"
date: "2025-12-14"
output: html_document
---

```{r, message=FALSE, warning=FALSE}
source("~/SALVEseq/packages.R")
source("~/SALVEseq/functions.R")
```

# Individual Datasets
## GEX

### UMAP coords and cell size

```{r}
samples <- data.frame(
  datasets = c(
    "Esrp2KO_sham24_R1",
    "Normal_SHAM24_R2"
  ),
  samples = c(
    "Esrp2KO",
    "Normal"
  )
)

output.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGL002/singleCell/Seurat/UMAPcoords/"
if (!dir.exists(output.dir)) {
  dir.create(output.dir, recursive = TRUE)
}

for (i in 1:nrow(samples)) {
  sample_name <- samples$samples[i]
  dataset <- samples$datasets[i]
  folder <- "/projects/b1042/GoyalLab/egrody/extractedData/EGL002/singleCell/counts/"
  input.dir <- paste0(folder, "GRCm39_", dataset, "/outs/filtered_feature_bc_matrix/")
  
  tryCatch({
    seurat_obj <- SeuratPipeline(input.dir, sample_name, 
                         output_dir = "/projects/b1042/GoyalLab/egrody/extractedData/EGL002/singleCell/Seurat/plots/", 
                         plots = TRUE, rds = TRUE,
                         species = "mouse", remove_mac = FALSE)
    
    # Generate expression data frame without gene expression
    sample_df <- targetExpressionDF(seurat_obj, genes = "")

    # Add sample and target columns to the dataframe
    sample_df$size <- colSums(GetAssayData(seurat_obj, assay = "RNA", slot = "counts"))
    sample_df$sample <- sample_name

    write.csv(sample_df, paste0(output.dir, sample_name, "UMAP_coords.csv"))
    
    cat("Processed sample:", sample_name, "\n")
  }, error = function(e) {
    cat("Error processing sample", sample_name, ":", conditionMessage(e), "\n\n")
  })
}
```

### integrate KO and WT
```{r}
samples <- data.frame(
  datasets = c(
    "Esrp2KO_sham24_R1",
    "Normal_SHAM24_R2"
  ),
  samples = c(
    "Esrp2KO",
    "Normal"
  )
)

output.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGL002/singleCell/Seurat/integrated/"
if (!dir.exists(output.dir)) {
  dir.create(output.dir, recursive = TRUE)
}

individual <- list()
for (i in 1:nrow(samples)) {
  sample_name <- samples$samples[i]
  dataset <- samples$datasets[i]
  folder <- "/projects/b1042/GoyalLab/egrody/extractedData/EGL002/singleCell/counts/"
  input.dir <- paste0(folder, "GRCm39_", dataset, "/outs/filtered_feature_bc_matrix/")
  
  data <- Read10X(data.dir = input.dir)
  sample <- CreateSeuratObject(counts = data, project = sample_name, min.cells = 3, min.features = 200)
  sample <- Add_Mito_Ribo(sample, species = "mouse")
  sample <- subset(sample, subset = nFeature_RNA > 200 & nFeature_RNA < 3500 & 
                     nCount_RNA < 20000 & percent_mito < 5)
  individual[[sample_name]] <- sample
}

obj <- Reduce(function(x, y) merge(x, y), individual)
obj <- SCTransform(obj, vst.flavor = "v2")
obj <- RunPCA(obj, verbose = FALSE)
obj <- IntegrateLayers(obj, method = HarmonyIntegration,
                       orig.reduction = "pca",
                       new.reduction = "harmony",
                       normalization.method = "SCT",
                       verbose = F)

obj <- FindNeighbors(obj, reduction = "harmony", dims = 1:30)
obj <- FindClusters(obj, resolution = 0.6)

obj <- RunUMAP(obj, dims = 1:30, reduction = "harmony")
dplot <- DimPlot(obj, reduction = "umap", group.by = "orig.ident")
ggsave(dplot, file = paste0(output.dir, "integrated_umap_orig.ident.svg"))
dplot <- DimPlot(obj, reduction = "umap")
ggsave(dplot, file = paste0(output.dir, "integrated_umap.svg"))
saveRDS(obj, file = paste0(output.dir, "integrated.rds"))
sce <- SingleCellExperiment(
  assays = list(counts = GetAssayData(obj, layer = "counts")),
  colData = obj@meta.data
)
suppressMessages(
  writeH5AD(sce, file = paste0(output.dir, "integrated.h5ad"))
)
```

### bamsort whole gene
```{r}
input.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGL002/singleCell/bamsort/"
output.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGL002/singleCell/bamsort/reads/"

raw_cellIDs_df <-  read.csv("/projects/b1042/GoyalLab/egrody/extractedData/EGL002/singleCell/raw_cellIDs/all_raw_cellIDs.csv")
raw_cellIDs <- split(raw_cellIDs_df$cellID, raw_cellIDs_df$sample)


samples <- data.frame(
  samples = c(
    "Esrp2KO_sham24_R1",
    "Normal_SHAM24_R2"
  ),
  nicknames = c(
    "KO", "WT"
  )
)
process_bamsort_liver(samples, input.dir, output.dir, raw_cellIDs)

#minimums
input_dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGL002/singleCell/bamsort/reads/"

process_all_set_minimumsv6("liver", samples$samples, input_dir, 
                           TRUE, paste0(input_dir, "/minimum/"), 
                           min_reads_cell = 2)
```


## SALVE

### First look
From the bamsort, what are the reads and frequencies? Are the targets above background? Is there any evidence for index hopping in the pooled samples?
```{r}
bamsort_stats <- read.csv("/projects/b1042/GoyalLab/egrody/extractedData/EGL002/SALVE/bamsort/nointron/all_samples_alignment_summary.txt")
stats <- bamsort_stats %>%
  select(Sample, Target, Rows, Total_Reads) %>%
  rename(Name = Sample, Region = Target, UMI = Rows, Reads = Total_Reads) %>% 
  mutate(Sample = ifelse(grepl("WT", Name), "WT", "KO"), Target = sapply(strsplit(Name, "_"), tail, 1)) %>%
  #mutate(Target = ifelse(grepl("Slk", Target), "Slk", Target)) %>%
  select(-Name) %>% relocate(Target)

df <- stats %>%
  mutate(is_match = case_when(
    Target == Region ~ TRUE,
    str_detect(Target, fixed(Region)) ~ TRUE,
    TRUE ~ FALSE
  ))

# Plot
ggplot(df, aes(x = Target, y = UMI, fill = is_match)) +
  geom_col() +
  facet_grid(Sample ~ Region) +
  scale_fill_manual(
    values = c("TRUE" = "#E41A1C", "FALSE" = "#377EB8"),
    labels = c("TRUE" = "Matching", "FALSE" = "Non-matching"),
    name = "Target-Region Match"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.background = element_rect(fill = "grey90"),
    panel.spacing = unit(0.5, "lines")
  ) +
  labs(
    x = "Target",
    y = "UMI Count",
    title = "UMI Counts by Target and Region",
    subtitle = "Samples shown separately"
  )
```

Index hopping
```{r}
# Define combinations to test based on pooling
pools <- tribble(
  ~Sample, ~Target, ~Pool,
  "WT",    "Kras",   "WT_Pool1",
  "WT",    "Slkv3",  "WT_Pool1",
  "WT",    "Lsm14b", "WT_Pool1",
  "WT",    "Dgkd",   "WT_Pool2",
  "WT",    "Nf2",    "WT_Pool2",
  "WT",    "Slkv4",  "WT_Pool2",
  "KO",    "Kras",   "KO_Pool1",
  "KO",    "Slkv3",  "KO_Pool1",
  "KO",    "Lsm14b", "KO_Pool1",
  "KO",    "Dgkd",   "KO_Pool1",
  "KO",    "Nf2",    "KO_Pool2",
  "KO",    "Slkv4",  "KO_Pool2"
)

# define cross contamination
cross_pairs <- pools %>%
  inner_join(pools, by = c("Sample", "Pool"), suffix = c("_source", "_region")) %>%
  filter(Target_source != Target_region) %>%
  rename(Target = Target_source, Region = Target_region) %>%
  select(Sample, Pool, Target, Region)

# Slkv4 is not in df. Check and warn.
targets_in_pools <- unique(pools$Target)
targets_in_data <- unique(c(df$Target, df$Region))
missing <- setdiff(targets_in_pools, targets_in_data)
if (length(missing) > 0) {
  cat("WARNING: These targets are in pool definitions but not in data:",
      paste(missing, collapse = ", "), "\n")
  cat("Pairs involving these targets will have no data.\n\n")
}

bleed <- cross_pairs %>%
  left_join(df, by = c("Sample", "Target", "Region")) %>%
  select(Sample, Pool, Target, Region, UMI, Reads)
on_target <- df %>%
  filter(is_match == TRUE) %>%
  select(Sample, Region, on_target_UMI = UMI)

# true negatives
not_pooled_bg <- df %>%
  filter(is_match == FALSE) %>%
  left_join(pools, by = c("Sample", "Target")) %>%
  rename(source_pool = Pool) %>%
  left_join(
    pools %>% rename(Region = Target, region_pool = Pool),
    by = c("Sample", "Region")
  ) %>%
  filter(source_pool != region_pool | is.na(source_pool) | is.na(region_pool)) %>%
  group_by(Sample, Region) %>%
  summarise(
    bg_mean_UMI = mean(UMI, na.rm = TRUE),
    bg_sd_UMI = sd(UMI, na.rm = TRUE),
    bg_median_UMI = median(UMI, na.rm = TRUE),
    bg_targets = paste(Target, collapse = ","),
    n_bg = n(),
    .groups = "drop"
  )


comparison <- bleed %>%
  left_join(on_target, by = c("Sample", "Region")) %>%
  left_join(not_pooled_bg, by = c("Sample", "Region")) %>%
  mutate(
    pct_of_on_target = (UMI / on_target_UMI) * 100,
    fold_over_bg = UMI / bg_mean_UMI,
    z_score = (UMI - bg_mean_UMI) / bg_sd_UMI,
    is_elevated = UMI > (bg_mean_UMI + 2 * bg_sd_UMI)
  ) %>%
  arrange(Sample, Pool, desc(pct_of_on_target))

print(comparison)

plot_data <- comparison %>%
  filter(!is.na(UMI)) %>%
  mutate(label = paste0(Target, " -> ", Region))

ggplot(plot_data, aes(x = reorder(label, -pct_of_on_target),
                      y = pct_of_on_target,
                      fill = Pool)) +
  geom_col() +
  facet_wrap(~Sample, scales = "free_x") +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8)) +
  labs(
    x = "Source Target -> Aligned Region",
    y = "% of On-Target UMI",
    title = "Cross-Contamination: Co-Pooled Target Bleed",
    subtitle = "Each bar = UMI from source Target aligning to a different Target's Region"
  )

# Compare the SAME Target-Region pair across pools where it IS vs IS NOT co-pooled.
# If the ratio is ~1, it's off-target capture, not bleed.
off_target <- df %>%
  filter(is_match == FALSE) %>%
  left_join(pools, by = c("Sample", "Target")) %>%
  rename(source_pool = Pool) %>%
  left_join(
    pools %>% rename(Region = Target, region_pool = Pool),
    by = c("Sample", "Region")
  ) %>%
  mutate(co_pooled = !is.na(source_pool) & !is.na(region_pool) & source_pool == region_pool) %>%
  left_join(on_target, by = c("Sample", "Region")) %>%
  mutate(pct_of_on_target = (UMI / on_target_UMI) * 100)

ggplot(off_target %>% filter(!is.na(pct_of_on_target)),
       aes(x = reorder(paste0(Target, " -> ", Region), -pct_of_on_target),
           y = pct_of_on_target,
           fill = co_pooled)) +
  geom_col(position = "dodge") +
  facet_wrap(~Sample, scales = "free_x") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7)) +
  labs(
    x = "Source Target -> Aligned Region",
    y = "% of On-Target UMI",
    fill = "Co-Pooled",
    title = "Off-Target Signal: Co-Pooled vs Not Co-Pooled",
    subtitle = "If bars are similar height, signal is off-target capture, not bleed"
  )
```


### bamsort
Run this once first to get raw cellIDs:
```{r raw_cellIDs}
# Load and save raw cell IDs from 10X data
raw_cellIDs <- list()
output_dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGL002/singleCell/raw_cellIDs" 

# Create output directory if it doesn't exist
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}

samples <- data.frame(
  datasets = c(
    "Esrp2KO_sham24_R1",
    "Normal_SHAM24_R2"
  ),
  samples = c(
    "Esrp2KO",
    "Normal"
  )
)

all_cellIDs <- data.frame(cellID = character(), sample = character(), stringsAsFactors = FALSE)

for (i in 1:nrow(samples)) {
  dataset_name <- samples$datasets[i]
  sample_name <- samples$samples[i]
  folder <- "/projects/b1042/GoyalLab/egrody/extractedData/EGL002/singleCell/counts/"
  rawDataFolder <- paste0(folder, "GRCm39_", dataset_name, "/outs/raw_feature_bc_matrix/")
  
  tryCatch({
    rawdata <- Read10X(rawDataFolder)
    umi_counts <- colSums(rawdata)
    cell_ids <- names(umi_counts[umi_counts > 100]) # keep only cells with transcriptomes
    raw_cellIDs[[sample_name]] <- cell_ids
    cat("Loaded", length(cell_ids), "raw cellIDs for sample", sample_name, "\n")
    
    # Add to the combined dataframe
    sample_df <- data.frame(
      cellID = cell_ids,
      sample = rep(sample_name, length(cell_ids)),
      stringsAsFactors = FALSE
    )
    all_cellIDs <- rbind(all_cellIDs, sample_df)
    
    # Save individual sample's cell IDs to CSV
    sample_file <- file.path(output_dir, paste0(sample_name, "_raw_cellIDs.csv"))
    write.csv(data.frame(cellID = cell_ids), sample_file, row.names = FALSE)
    cat("Saved", length(cell_ids), "cell IDs to", sample_file, "\n")
    
  }, error = function(e) {
    cat("Error processing", dataset_name, ":", conditionMessage(e), "\n")
  })
  
  if (exists("rawdata")) {
    rm(rawdata)
    gc()
  }
}

# Save the combined cell IDs to a single CSV
combined_file <- file.path(output_dir, "all_raw_cellIDs.csv")
write.csv(all_cellIDs, combined_file, row.names = FALSE)
cat("Saved combined cell IDs to", combined_file, "\n")

# To easily load this data in the future:
# raw_cellIDs_df <- read.csv("path/to/output/directory/all_raw_cellIDs.csv")
# raw_cellIDs <- split(raw_cellIDs_df$cellID, raw_cellIDs_df$sample)
```

Run bamsort_combined (alignment_reads and splice) first.
Making full and minimum filtered lists:
```{r}
input.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGL002/SALVE/bamsort/nointron/"
output.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGL002/SALVE/bamsort/nointron/reads/"

raw_cellIDs_df <-  read.csv("/projects/b1042/GoyalLab/egrody/extractedData/EGL002/singleCell/raw_cellIDs/all_raw_cellIDs.csv")
raw_cellIDs <- split(raw_cellIDs_df$cellID, raw_cellIDs_df$sample)


samples <- data.frame(
  samples = c(
    "Esrp2KO",
    "Normal"
  ),
  nicknames = c(
    "KO", "WT"
  )
)
process_bamsort_liver(samples, input.dir, 
                      paste0(output.dir, "v3/"), raw_cellIDs, targets = c("Slkv3", "Dgkd", "Lsm14b", "Kras"))
process_bamsort_liver(samples, input.dir, 
                      paste0(output.dir, "v4/"), raw_cellIDs, targets = c("Slkv4", "Nf2"))

#minimums
input_dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGL002/SALVE/bamsort/nointron/reads/"

process_all_set_minimumsv6("liver", samples$samples, paste0(input_dir, "v3/"), 
                           TRUE, paste0(input_dir, "/v3/minimum/"), 
                           min_reads = 2, min_umi = 2)
process_all_set_minimumsv6("liver", samples$samples, paste0(input_dir, "v4/"), 
                           TRUE, paste0(input_dir, "/v4/minimum/"), 
                           min_reads = 2, min_umi = 2)
```



# Joint Dataset
## Joint

### Stupidity check: are there cellIDs in common?
```{r}
# GEX
output.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGL002/singleCell/Seurat/UMAPcoords/"
samples <- c(
  "Esrp2KO",
  "Normal"
)

GEX_cellids <- list()
for (i in samples) {
  test <- read.csv(paste0(output.dir, i, "UMAP_coords.csv"), row.names = "X")
  GEX_cellids[i] <- test
}

input.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGL002/SALVE/bamsort/"
all_files <- list.files(path = input.dir, 
                        pattern = "^EGL002_(KO|WT)_.*_bamsort_alignment_.*\\.csv$",
                        full.names = TRUE)
Esrp2KO.S <- c()
Normal.S <- c()

for (file in all_files) {
  filename <- basename(file)
  df <- read.csv(file, stringsAsFactors = FALSE)
  if (grepl("^EGL002_KO", filename)) {
    Esrp2KO.S <- c(Esrp2KO.S, df$cellID)
  } else if (grepl("^EGL002_WT", filename)) {
    Normal.S <- c(Normal.S, df$cellID)
  }
}

SALVE_cellids <- list("Esrp2KO" = unique(Esrp2KO.S),
                      "Normal" = unique(Normal.S))

test <- sum(SALVE_cellids$Normal %in% GEX_cellids$Normal) #none :(

output_dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGL002/singleCell/raw_cellIDs/"
GEX_cellids_raw <- list()
for (i in samples) {
  temp <- read.csv(paste0(output_dir, i, "_raw_cellIDs.csv"))
  GEX_cellids_raw[i] <- temp
}

test <- sum(SALVE_cellids$Esrp2KO %in% GEX_cellids_raw$Esrp2KO)
```


### joining
```{r}
samples <- c(
    "Esrp2KO",
    "Normal"
)
SALVE.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGL002/SALVE/bamsort/nointron/reads/"
GEX.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGL002/singleCell/Seurat/UMAPcoords/"

joint_data <- list()

for (i in 1:length(samples)) {
  sample_name <- samples[i]
  
  # Read all data files
  salve_files <- paste0(SALVE.dir, c("v3", "v4"), "/minimum/", sample_name, "_liver_filtered.csv")
  salve_files <- salve_files[file.exists(salve_files)]
  #gex_file <- paste0(GEX.dir, sample_name, "_GEX_filtered.csv")
  umap_file <- paste0(GEX.dir, sample_name, "UMAP_coords.csv")
  
  salve_list <- lapply(salve_files, function(f) {
    df <- read.csv(f)
    df$total <- NULL
    df
  })
  salve_data <- Reduce(function(x, y) merge(x, y, by = "cellID", all = TRUE), salve_list)
  salve_data[is.na(salve_data)] <- 0
  salve_data[-1] <- log1p(salve_data[-1])
  #gex_data <- if(file.exists(gex_file)) read.csv(gex_file) else NULL
  umap_data <- if(file.exists(umap_file)) read.csv(umap_file, row.names = "X") else NULL
  
  # Get cellIDs from UMAP df
  all_cells <- as.character(umap_data$cellID)
  all_cells <- unique(all_cells)
  
  # Create base dataframe
  result <- data.frame(cellID = all_cells)
  
  # Add UMAP data
  if (!is.null(umap_data)) {
    umap_processed <- umap_data %>%
      select(-V2, -sample)
    result <- left_join(result, umap_processed, by = "cellID")
  }
  
  # Add GEX data
  # if (!is.null(gex_data)) {
  #   gex_processed <- gex_data %>%
  #     mutate(cellID = as.character(cellID)) %>%
  #     select(cellID, total, US)
  #   result <- left_join(result, gex_processed, by = "cellID") %>%
  #     mutate(GEX = log1p(total)) %>%
  #     mutate(GEX.US = log1p(US)) %>%
  #     select(-total, -US)
  # } else {
  #   result$GEX <- 0
  # }
  
  # Add SALVE data
  if (!is.null(salve_data)) {
    result <- left_join(result, salve_data, by = "cellID")
  }

  result[is.na(result)] <- 0
  
  joint_data[[sample_name]] <- result
  cat("Processed", sample_name, "\n")
}
```

```{r}
output_dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGL002/joint/plots/"
#cat("\nSample\tTotal_cells\tboth\t10X_only\tSALVE_only\n")

# Process each dataset in the joint_data list
for (sample_name in names(joint_data)) {
  current_df <- joint_data[[sample_name]]

  # both <- current_df %>% filter(GEX != 0 & SALVE != 0)
  # either <- current_df %>% filter(GEX != 0 | SALVE != 0)
  # onlySingleCell <- current_df %>% 
  #   filter(GEX != 0) %>%
  #   filter(!(cellID %in% both$cellID))
  # onlySALVE <- current_df %>% 
  #   filter(SALVE != 0) %>%
  #   filter(!(cellID %in% both$cellID))
  # 
  # cat(sample_name, "\t",
  #     nrow(current_df), "\t",
  #     nrow(both), "\t",
  #     nrow(onlySingleCell), "\t",
  #     nrow(onlySALVE), "\n")

  cat(sample_name, "Maximums:\nKras: ", max(current_df$Kras),
      "\nDgkd: ", max(current_df$Dgkd),
      "\nLsm14b: ", max(current_df$Lsm14b),
      "\nSlk: ", max(current_df$Slkv3),
      "\nNf2: ", max(current_df$Nf2), "\n")

  # plotUMAP(current_df, colorby = Kras,
  #        paste0(sample_name, ": Kras VIE"),
  #        output_dir,
  #        paste0(sample_name, "_Kras.pdf"),
  #        comparison = FALSE, color = "red")
  # plotUMAP(current_df, colorby = Dgkd,
  #        paste0(sample_name, ": Dgkd VIE"),
  #        output_dir,
  #        paste0(sample_name, "_Dgkd.pdf"),
  #        comparison = FALSE, color = "red")
  # plotUMAP(current_df, colorby = Lsm14b,
  #        paste0(sample_name, ": Lsm14b VIE"),
  #        output_dir,
  #        paste0(sample_name, "_Lsm14b.pdf"),
  #        comparison = FALSE, color = "red")
  # plotUMAP(current_df, colorby = Slkv3,
  #        paste0(sample_name, ": Slk VIE"),
  #        output_dir,
  #        paste0(sample_name, "_Slkv3.pdf"),
  #        comparison = FALSE, color = "red")
  # plotUMAP(current_df, colorby = Nf2,
  #        paste0(sample_name, ": Nf2 VIE"),
  #        output_dir,
  #        paste0(sample_name, "_Nf2.pdf"),
  #        comparison = FALSE, color = "red")

  
  #Save CSVs
  # write.csv(both, paste0(output_dir, sample_name, "_both.csv"), row.names = FALSE)
  # write.csv(onlySingleCell, paste0(output_dir, sample_name, "_onlySingleCell.csv"), row.names = FALSE)
  # write.csv(onlySALVE, paste0(output_dir, sample_name, "_onlySALVE.csv"), row.names = FALSE)
  # write.csv(current_df, paste0(output_dir, sample_name, "_all.csv"), row.names = FALSE)
}
```

```{r}
plot_sc_expression <- function(joint_data, columns,
                                labels = c("Group 1", "Group 2"),
                                zero_threshold = 0) {
  # labels[1] matches joint_data[[1]], labels[2] matches joint_data[[2]]
  # Plot order: reverse so labels[2] (WT) appears first on x-axis
  plot_order <- rev(labels)
  palette <- setNames(c("#DD8452", "#4C72B0"), labels)

  combined <- bind_rows(
    joint_data[[1]] %>% select(all_of(columns)) %>% mutate(Group = labels[1]),
    joint_data[[2]] %>% select(all_of(columns)) %>% mutate(Group = labels[2])
  ) %>%
    pivot_longer(cols = all_of(columns), names_to = "Gene", values_to = "Value") %>%
    mutate(
      Gene = factor(Gene, levels = columns),
      Group = factor(Group, levels = plot_order),
      Expressing = Value > zero_threshold
    )

  prop_data <- combined %>%
    group_by(Gene, Group) %>%
    summarise(
      n_total = n(),
      n_expr = sum(Expressing),
      pct = 100 * n_expr / n_total,
      .groups = "drop"
    )

  p_bar <- ggplot(prop_data, aes(x = Group, y = pct, fill = Group)) +
    geom_col(width = 0.6, alpha = 0.7) +
    geom_text(aes(label = paste0(n_expr, "/", n_total)),
              vjust = -0.5, size = 3) +
    facet_wrap(~ Gene, nrow = 1) +
    scale_fill_manual(values = palette) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
    theme_minimal(base_size = 12) +
    theme(
      legend.position = "none",
      strip.text = element_text(face = "italic", size = 11),
      axis.title.x = element_blank(),
      panel.grid.major.x = element_blank()
    ) +
    labs(y = "% Expressing")

  expr_only <- combined %>% filter(Expressing)

  p_violin <- ggplot(expr_only, aes(x = Group, y = Value, fill = Group)) +
    geom_violin(scale = "width", alpha = 0.3, trim = TRUE) +
    geom_boxplot(width = 0.15, alpha = 0.6, outlier.shape = NA) +
    geom_jitter(aes(color = Group), width = 0.1, size = 0.8, alpha = 0.4) +
    facet_wrap(~ Gene, nrow = 1, scales = "free_y") +
    scale_y_log10() +
    scale_fill_manual(values = palette) +
    scale_color_manual(values = palette) +
    theme_minimal(base_size = 12) +
    theme(
      legend.position = "bottom",
      strip.text = element_blank(),
      axis.title.x = element_blank(),
      panel.grid.major.x = element_blank()
    ) +
    labs(y = "Expression (log scale)", fill = NULL, color = NULL)

  p_bar / p_violin + plot_layout(heights = c(1, 2.5))
}

columns <- c("Kras", "Dgkd", "Lsm14b", "Nf2", "Slkv3")
p <- plot_sc_expression(joint_data, columns, labels = c("Esrp2KO", "WT"))
print(p)
```

### Pseudobulk
Let's look at counts as if this were bulk.
We expect the following trends when comparing KO vs WT:
Kras down
Nf2 down
Dgkd up
Slk down
Lsm14b up

```{r}
WT <- joint_data[["Normal"]]
KO <- joint_data[["Esrp2KO"]]
targets <- c("Kras", "Nf2", "Dgkd", "Slkv3", "Lsm14b")
KO[targets] <- expm1(KO[targets])
WT[targets] <- expm1(WT[targets])
cat("KO vs WT:\n")
for (target in targets) {
  compare <- ifelse(sum(KO[[target]]) > sum(WT[[target]]), "up", "down")
  cat(target, "\t", compare, "\t", sum(KO[[target]]), "vs", sum(WT[[target]]), "\n")
}
cat("\nKO vs WT (normalized by cell count):\n")
for (target in targets) {
  norm_KO <- sum(KO[[target]]) / nrow(KO)
  norm_WT <- sum(WT[[target]]) / nrow(WT)
  compare <- ifelse(norm_KO > norm_WT, "up", "down")
  cat(target, "\t", compare, "\t", norm_KO, "vs", norm_WT, "\n")
}
```

Normalizing for GEX counts per gene
```{r load}
samples <- c(
    "Esrp2KO",
    "Normal"
)
SALVE.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGL002/SALVE/bamsort/nointron/reads/"
GEX.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGL002/singleCell/bamsort/reads/minimum/"
umap.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGL002/singleCell/Seurat/UMAPcoords/"

joint_data <- list()

for (i in 1:length(samples)) {
  sample_name <- samples[i]
  
  # Read all data files
  salve_files <- paste0(SALVE.dir, c("v3", "v4"), "/minimum/", sample_name, "_liver_filtered.csv")
  salve_files <- salve_files[file.exists(salve_files)]
  gex_file <- paste0(GEX.dir, sample_name, "*_liver_filtered.csv")
  umap_file <- paste0(umap.dir, sample_name, "UMAP_coords.csv")
  
  salve_list <- lapply(salve_files, function(f) {
    df <- read.csv(f)
    df$total <- NULL
    df
  })
  salve_data <- Reduce(function(x, y) merge(x, y, by = "cellID", all = TRUE), salve_list)
  salve_data[is.na(salve_data)] <- 0
  #salve_data[-1] <- log1p(salve_data[-1])
  gex_matches <- Sys.glob(gex_file)
  gex_data <- if (length(gex_matches) > 0) read.csv(gex_matches[1]) else NULL
  #gex_data[-1] <- log1p(gex_data[-1])
  umap_data <- if(file.exists(umap_file)) read.csv(umap_file, row.names = "X") else NULL
  
  # Get cellIDs from UMAP df
  all_cells <- as.character(umap_data$cellID)
  all_cells <- unique(all_cells)
  
  # Create base dataframe
  result <- data.frame(cellID = all_cells)
  
  # Add UMAP data
  if (!is.null(umap_data)) {
    umap_processed <- umap_data %>%
      select(-V2, -sample)
    result <- left_join(result, umap_processed, by = "cellID")
  }
  
  # Add GEX data
  if (!is.null(gex_data)) {
    gex_processed <- gex_data %>%
      mutate(cellID = as.character(cellID)) %>%
      select(-total) %>%
      rename_with(~ paste0(.x, "_GEX"), -cellID)
    result <- left_join(result, gex_processed, by = "cellID")
  } else {
    result$GEX <- 0
  }
  
  # Add SALVE data
  if (!is.null(salve_data)) {
    salve_processed <- salve_data %>%
      rename_with(~ paste0(.x, "_SALVE"), -cellID)
    result <- left_join(result, salve_processed, by = "cellID")
  }

  result[is.na(result)] <- 0
  
  joint_data[[sample_name]] <- result
  cat("Processed", sample_name, "\n")
}
```

```{r}
WT <- joint_data[["Normal"]]
KO <- joint_data[["Esrp2KO"]]
targets <- c("Kras", "Nf2", "Dgkd", "Slk", "Lsm14b")
gex_cols <- paste0(targets, "_GEX")
#KO[gex_cols] <- expm1(KO[gex_cols])
#WT[gex_cols] <- expm1(WT[gex_cols])

cat("GEX\nKO vs WT:\n")
for (i in seq_along(targets)) {
  col <- gex_cols[i]
  compare <- ifelse(sum(KO[[col]]) > sum(WT[[col]]), "up", "down")
  cat(targets[i], "\t", compare, "\t", sum(KO[[col]]), "vs", sum(WT[[col]]), "\n")
}

cat("\nKO vs WT (normalized by cell count):\n")
for (i in seq_along(targets)) {
  col <- gex_cols[i]
  norm_KO <- sum(KO[[col]]) / nrow(KO)
  norm_WT <- sum(WT[[col]]) / nrow(WT)
  compare <- ifelse(norm_KO > norm_WT, "up", "down")
  cat(targets[i], "\t", compare, "\t", norm_KO, "vs", norm_WT, "\n")
}

#now normalize SALVE by GEX
salve_cols <- paste0(targets, "_SALVE")
salve_cols[targets == "Slk"] <- "Slkv3_SALVE"
#KO[salve_cols] <- expm1(KO[salve_cols])
#WT[salve_cols] <- expm1(WT[salve_cols])
cat("\n\nSALVE\nKO vs WT (normalized by GEX):\n")
for (i in seq_along(targets)) {
  gex_col <- gex_cols[i]
  salve_col <- salve_cols[i]
  norm_KO <- sum(KO[[salve_col]]) / sum(KO[[gex_col]])
  norm_WT <- sum(WT[[salve_col]]) / sum(WT[[gex_col]])
  compare <- ifelse(norm_KO > norm_WT, "up", "down")
  cat(targets[i], "\t", compare, "\t", norm_KO, "vs", norm_WT, "\n")
}
```
### VIE trends
```{r}
plot_composition <- function(joint_data, targets, expected,
                              labels = c("Group 1", "Group 2"),
                              zero_threshold = 0) {
  
  gex_cols   <- paste0(targets, "_GEX")
  salve_cols <- paste0(targets, "_SALVE")
  salve_cols[targets == "Slk"] <- "Slkv3_SALVE"
  
  plot_order <- rev(labels)
  palette <- setNames(c("#DD8452", "#4C72B0"), labels)
  
  target_labels <- setNames(
    paste0(targets, "\n(Expected: ", expected[targets], ")"),
    targets
  )
  target_labeller <- as_labeller(target_labels)
  
  build_long <- function(df, label) {
    rows <- list()
    for (i in seq_along(targets)) {
      rows[[i]] <- data.frame(
        Target = targets[i],
        Group  = label,
        GEX    = df[[gex_cols[i]]],
        SALVE  = df[[salve_cols[i]]],
        stringsAsFactors = FALSE
      )
    }
    bind_rows(rows)
  }
  
  combined <- bind_rows(
    build_long(joint_data[[1]], labels[1]),
    build_long(joint_data[[2]], labels[2])
  ) %>%
    mutate(
      Target = factor(Target, levels = targets),
      Group  = factor(Group, levels = plot_order)
    )
  
  # --- Row 1: Mean counts per cell (GEX and SALVE) ---
  mean_data <- combined %>%
    group_by(Target, Group) %>%
    summarise(
      salve_per_cell = sum(SALVE) / n(),
      gex_per_cell   = sum(GEX)   / n(),
      .groups = "drop"
    ) %>%
    pivot_longer(
      cols = c(gex_per_cell, salve_per_cell),
      names_to = "Metric", values_to = "Value"
    ) %>%
    mutate(Metric = factor(Metric,
      levels = c("gex_per_cell", "salve_per_cell"),
      labels = c("GEX", "SALVE")
    ))
  
  p_mean <- ggplot(mean_data, aes(x = Group, y = Value, fill = Group)) +
    geom_col(width = 0.6, alpha = 0.7) +
    geom_text(aes(label = sprintf("%.2f", Value)), vjust = -0.3, size = 2.5) +
    facet_grid(Metric ~ Target, scales = "free_y",
               labeller = labeller(Target = target_labeller)) +
    scale_fill_manual(values = palette) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +
    theme_minimal(base_size = 11) +
    theme(
      legend.position = "none",
      strip.text.x = element_text(face = "italic", size = 10, lineheight = 1.2),
      axis.title.x = element_blank(),
      panel.grid.major.x = element_blank()
    ) +
    labs(y = "Mean counts\nper cell")
  
  # --- Row 2: SALVE / GEX ratio ---
  ratio_data <- combined %>%
    group_by(Target, Group) %>%
    summarise(
      ratio = 100 * sum(SALVE) / sum(GEX),
      .groups = "drop"
    )
  
  p_ratio <- ggplot(ratio_data, aes(x = Group, y = ratio, fill = Group)) +
    geom_col(width = 0.6, alpha = 0.7) +
    geom_text(aes(label = sprintf("%.1f%%", ratio)), vjust = -0.3, size = 2.5) +
    facet_wrap(~ Target, nrow = 1) +
    scale_fill_manual(values = palette) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +
    theme_minimal(base_size = 11) +
    theme(
      legend.position = "none",
      strip.text = element_blank(),
      axis.title.x = element_blank(),
      panel.grid.major.x = element_blank()
    ) +
    labs(y = "SALVE / GEX (%)")
  
  # --- Row 3: % cells expressing SALVE ---
  cell_pct <- combined %>%
    mutate(Expressing = SALVE > zero_threshold) %>%
    group_by(Target, Group) %>%
    summarise(
      n_total = n(),
      n_expr  = sum(Expressing),
      pct     = 100 * n_expr / n_total,
      .groups = "drop"
    )
  
  p_bar <- ggplot(cell_pct, aes(x = Group, y = pct, fill = Group)) +
    geom_col(width = 0.6, alpha = 0.7) +
    geom_text(aes(label = paste0(n_expr, "/", n_total)), vjust = -0.5, size = 2.8) +
    facet_wrap(~ Target, nrow = 1) +
    scale_fill_manual(values = palette) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
    theme_minimal(base_size = 11) +
    theme(
      legend.position = "none",
      strip.text = element_blank(),
      axis.title.x = element_blank(),
      panel.grid.major.x = element_blank()
    ) +
    labs(y = "% Cells SALVE")
  
  # --- Row 4: Violin of SALVE counts in expressing cells ---
  expr_only <- combined %>% filter(SALVE > zero_threshold)
  
  p_violin <- ggplot(expr_only, aes(x = Group, y = SALVE, fill = Group)) +
    geom_violin(scale = "width", alpha = 0.3, trim = TRUE) +
    geom_boxplot(width = 0.15, alpha = 0.6, outlier.shape = NA) +
    geom_jitter(aes(color = Group), width = 0.1, size = 0.8, alpha = 0.4) +
    facet_wrap(~ Target, nrow = 1, scales = "free_y") +
    scale_y_log10() +
    scale_fill_manual(values = palette) +
    scale_color_manual(values = palette) +
    theme_minimal(base_size = 11) +
    theme(
      legend.position = "bottom",
      strip.text = element_blank(),
      axis.title.x = element_blank(),
      panel.grid.major.x = element_blank()
    ) +
    labs(y = "SALVE (log scale)", fill = NULL, color = NULL)
  
  p_mean / p_ratio / p_bar / p_violin + plot_layout(heights = c(2, 1, 1, 2.5))
}

expected <- c(Kras = "down", Nf2 = "down", Dgkd = "up", Slk = "down", Lsm14b = "up")

p <- plot_composition(joint_data,
                       targets  = c("Kras", "Nf2", "Dgkd", "Slk", "Lsm14b"),
                       expected = expected,
                       labels   = c("Esrp2KO", "WT"))
print(p)
```


## Subset
### DEG analysis

```{r, warning=FALSE}
input.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGL002/singleCell/Seurat/plots/"
output.base <- "/projects/b1042/GoyalLab/egrody/extractedData/EGL002/joint/subset/"

samples <- c("Esrp2KO",
             "Normal")
targets <- c("Kras", "Nf2", "Dgkd", "Slkv3", "Lsm14b")

for (i in 1:length(samples)) {
  sample_name <- samples[i]
  input_file <- paste0(input.dir, sample_name, ".rds")
  
  tryCatch({
    seurat_obj <- readRDS(input_file)
    
    salve <- joint_data[[sample_name]]
    for (target in targets) {
      list_of_cells1 <- salve %>% filter(.data[[target]] > 0) %>%
      select(cellID) %>% unlist()
    # list_of_cells2 <- salve %>% filter(total != 0) %>%
    #   select(cellID) %>% filter(!(cellID %in% list_of_cells1)) %>% unlist()
    results <- analyze_cell_subset(seurat_obj, list_of_cells1#, 
                                   #list_of_cells2, group_names = c("MS", "noMS")
                                   )
    
    output.dir <- paste0(output.base, sample_name, "/")
    if (!dir.exists(output.dir)) {
      dir.create(output.dir, recursive = TRUE)
    }
    #save_subset_analysis(results, output_dir = output.dir)
    plot_subset_analysis(seurat_obj, results, list_of_cells1, output_dir = paste0(output.dir, "plots/"))

    cat("Processed target:", target, "\n")
    }
    cat("Processed sample:", sample_name, "\n\n")
  }, error = function(e) {
    cat("Error processing sample", sample_name, ":", conditionMessage(e), "\n\n")
  })
}

```

Rewrote some functions
```{r}
library(gprofiler2)

perform_deg_pathway_analysis <- function(seurat_obj,
                                          cells_group1,
                                          cells_group2,
                                          assay = "RNA",
                                          test_use = "wilcox",
                                          fc_threshold = 0.5,
                                          pval_threshold = 0.01,
                                          output_prefix = "DEG_analysis") {
  
  # Validate inputs
  if (!inherits(seurat_obj, "Seurat")) {
    stop("seurat_obj must be a Seurat object")
  }
  
  all_cells <- colnames(seurat_obj)
  missing_cells1 <- setdiff(cells_group1, all_cells)
  missing_cells2 <- setdiff(cells_group2, all_cells)
  
  if (length(missing_cells1) > 0) {
    warning(paste(length(missing_cells1), "cells from group1 not found in Seurat object"))
    cells_group1 <- intersect(cells_group1, all_cells)
  }
  
  if (length(missing_cells2) > 0) {
    warning(paste(length(missing_cells2), "cells from group2 not found in Seurat object"))
    cells_group2 <- intersect(cells_group2, all_cells)
  }
  
  if (length(cells_group1) == 0 || length(cells_group2) == 0) {
    stop("One or both groups have no valid cells after filtering")
  }
  
  cat(sprintf("Group 1: %d cells\n", length(cells_group1)))
  cat(sprintf("Group 2: %d cells\n", length(cells_group2)))
  
  # Create temporary identity for DEG analysis
  seurat_obj$temp_ident <- "other"
  seurat_obj$temp_ident[colnames(seurat_obj) %in% cells_group1] <- "group1"
  seurat_obj$temp_ident[colnames(seurat_obj) %in% cells_group2] <- "group2"
  
  # Subset Seurat object to only include cells from both groups
  seurat_subset <- subset(seurat_obj, cells = c(cells_group1, cells_group2))
  Idents(seurat_subset) <- "temp_ident"
  
  cat("Performing differential expression analysis...\n")
  
  # Perform DEG analysis
  # group1 vs group2: positive log2FC means higher in group1
  deg_results <- FindMarkers(
    object = seurat_subset,
    ident.1 = "group1",
    ident.2 = "group2",
    assay = assay,
    test.use = test_use,
    logfc.threshold = 0,  # Get all genes, filter later
    min.pct = 0.1,  # Gene must be expressed in at least 10% of cells in one group
    verbose = TRUE
  )
  
  # Add gene names as column
  deg_results$gene <- rownames(deg_results)
  
  # Reorder columns
  deg_results <- deg_results[, c("gene", "avg_log2FC", "pct.1", "pct.2", 
                                  "p_val", "p_val_adj")]
  
  # Sort by adjusted p-value
  deg_results <- deg_results[order(deg_results$p_val_adj), ]
  
  # Save full results
  write.csv(deg_results, 
            file = paste0(output_prefix, "_all_genes.csv"),
            row.names = FALSE)
  cat(sprintf("Saved full results: %s_all_genes.csv\n", output_prefix))
  
  # Filter for significant DEGs
  deg_filtered <- deg_results %>%
    filter(abs(avg_log2FC) > fc_threshold & p_val_adj < pval_threshold)
  
  cat(sprintf("Significant DEGs: %d (|log2FC| > %.2f, adj p-val < %.2f)\n", 
              nrow(deg_filtered), fc_threshold, pval_threshold))
  cat(sprintf("  Upregulated in group1: %d\n", sum(deg_filtered$avg_log2FC > 0)))
  cat(sprintf("  Upregulated in group2: %d\n", sum(deg_filtered$avg_log2FC < 0)))
  
  # Save filtered results
  write.csv(deg_filtered,
            file = paste0(output_prefix, "_significant_genes.csv"),
            row.names = FALSE)
  cat(sprintf("Saved significant DEGs: %s_significant_genes.csv\n", output_prefix))
  
  # Pathway enrichment analysis using g:Profiler
  pathway_results <- NULL
  if (nrow(deg_filtered) > 0) {
    cat("\nPerforming pathway enrichment analysis with g:Profiler...\n")
    
    # Split into upregulated and downregulated genes
    up_genes <- deg_filtered %>% filter(avg_log2FC > 0) %>% pull(gene)
    down_genes <- deg_filtered %>% filter(avg_log2FC < 0) %>% pull(gene)
    
    pathway_results <- list()
    
    # Enrichment for upregulated genes (group1)
    if (length(up_genes) > 0) {
      cat(sprintf("  Analyzing %d upregulated genes (group1)...\n", length(up_genes)))
      gost_up <- gost(
        query = up_genes,
        organism = "mmulatta",  # Change to "mmusculus" for mouse
        significant = TRUE,
        correction_method = "g_SCS",  # g:Profiler's custom method
        sources = c("GO:BP", "GO:MF", "GO:CC", "KEGG", "REAC", "WP")
      )
      
      if (!is.null(gost_up$result)) {
        # Select available columns
        base_cols <- c("term_id", "term_name", "source", "p_value", "term_size", 
                       "query_size", "intersection_size")
        optional_cols <- c("intersection")
        available_cols <- base_cols[base_cols %in% colnames(gost_up$result)]
        optional_available <- optional_cols[optional_cols %in% colnames(gost_up$result)]
        
        pathway_results$upregulated <- gost_up$result %>%
          select(all_of(c(available_cols, optional_available))) %>%
          arrange(p_value)
        
        write.csv(pathway_results$upregulated,
                  file = paste0(output_prefix, "_pathways_group1_up.csv"),
                  row.names = FALSE)
        cat(sprintf("  Found %d enriched pathways for group1\n", 
                    nrow(pathway_results$upregulated)))
      } else {
        cat("  No significant pathways found for upregulated genes\n")
      }
    }
    
    # Enrichment for downregulated genes (group2)
    if (length(down_genes) > 0) {
      cat(sprintf("  Analyzing %d downregulated genes (group2)...\n", length(down_genes)))
      gost_down <- gost(
        query = down_genes,
        organism = "hsapiens",
        significant = TRUE,
        correction_method = "g_SCS",
        sources = c("GO:BP", "GO:MF", "GO:CC", "KEGG", "REAC", "WP")
      )
      
      if (!is.null(gost_down$result)) {
        # Select available columns
        base_cols <- c("term_id", "term_name", "source", "p_value", "term_size",
                       "query_size", "intersection_size")
        optional_cols <- c("intersection")
        available_cols <- base_cols[base_cols %in% colnames(gost_down$result)]
        optional_available <- optional_cols[optional_cols %in% colnames(gost_down$result)]
        
        pathway_results$downregulated <- gost_down$result %>%
          select(all_of(c(available_cols, optional_available))) %>%
          arrange(p_value)
        
        write.csv(pathway_results$downregulated,
                  file = paste0(output_prefix, "_pathways_group2_up.csv"),
                  row.names = FALSE)
        cat(sprintf("  Found %d enriched pathways for group2\n",
                    nrow(pathway_results$downregulated)))
      } else {
        cat("  No significant pathways found for downregulated genes\n")
      }
    }
  } else {
    cat("No significant DEGs found - skipping pathway analysis\n")
  }
  
  # Create volcano plot
  cat("\nGenerating volcano plot...\n")
  
  # Add significance category
  deg_results$significance <- "Not significant"
  deg_results$significance[deg_results$avg_log2FC > fc_threshold & 
                            deg_results$p_val_adj < pval_threshold] <- "Group1 enriched"
  deg_results$significance[deg_results$avg_log2FC < -fc_threshold & 
                            deg_results$p_val_adj < pval_threshold] <- "Group2 enriched"
  
  # Label top genes
  all_genes <- deg_results %>%
    filter(significance != "Not significant") %>%
    arrange(p_val_adj)
  
  volcano_plot <- ggplot(deg_results, aes(x = avg_log2FC, y = -log10(p_val_adj))) +
    geom_point(aes(color = significance), alpha = 0.6, size = 1.5) +
    scale_color_manual(
      values = c("Group1 enriched" = "#E64B35", 
                 "Not significant" = "grey70",
                 "Group2 enriched" = "#4DBBD5")
    ) +
    geom_vline(xintercept = c(-fc_threshold, fc_threshold), 
               linetype = "dashed", color = "grey30", linewidth = 0.5) +
    geom_hline(yintercept = -log10(pval_threshold), 
               linetype = "dashed", color = "grey30", linewidth = 0.5) +
    geom_text_repel(
      data = all_genes,
      aes(label = gene),
      size = 3,
      max.overlaps = 20,
      box.padding = 0.5,
      segment.size = 0.2
    ) +
    labs(
      title = "Differential Gene Expression: Group1 vs Group2",
      x = "log2 Fold Change",
      y = "-log10(Adjusted P-value)",
      color = "Significance"
    ) +
    theme_classic(base_size = 12) +
    theme(
      legend.position = "right",
      plot.title = element_text(hjust = 0.5, face = "bold"),
      axis.text = element_text(color = "black"),
      axis.title = element_text(face = "bold")
    )
  
  # Save plot
  ggsave(
    filename = paste0(output_prefix, "_volcano_plot.pdf"),
    plot = volcano_plot,
    width = 10,
    height = 8,
    units = "in"
  )
  
  ggsave(
    filename = paste0(output_prefix, "_volcano_plot.png"),
    plot = volcano_plot,
    width = 10,
    height = 8,
    units = "in",
    dpi = 300
  )
  
  cat(sprintf("Saved volcano plots: %s_volcano_plot.pdf/png\n", output_prefix))
  
  # Return results
  return(list(
    all_results = deg_results,
    significant_degs = deg_filtered,
    pathway_results = pathway_results,
    volcano_plot = volcano_plot,
    n_group1 = length(cells_group1),
    n_group2 = length(cells_group2),
    parameters = list(
      fc_threshold = fc_threshold,
      pval_threshold = pval_threshold,
      test_method = test_use
    )
  ))
}

```

```{r}
input.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS016/singleCell/Seurat/RDS/"
output.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/joint/subset/new/"

samples <- c("Invitro",
             "D195",
             "Pacute",
             "W2")
for (i in 1:length(samples)) {
  sample_name <- samples[i]
  input_file <- paste0(input.dir, sample_name, ".rds")
  
  tryCatch({
    seurat_obj <- readRDS(input_file)
    genes_to_exclude <- c("CD4", "CD8A", "CD8B")  # remove from DEG analysis
    genes_to_keep <- rownames(seurat_obj)[!rownames(seurat_obj) %in% genes_to_exclude]
    seurat_obj <- subset(seurat_obj, features = genes_to_keep)

    if (!dir.exists(output.dir)) {
      dir.create(output.dir, recursive = TRUE)
    }
    
    salve <- combined_data[[sample_name]]
    available_cols <- intersect(c("MS", "SS-MS", "SS"), colnames(salve))
    list_of_cells1 <- salve %>%
      filter(if_any(all_of(available_cols), ~ . != 0)) %>%
      select(cellID) %>%
      unlist()
    list_of_cells2 <- salve %>% filter(total != 0) %>%
      select(cellID) %>% filter(!(cellID %in% list_of_cells1)) %>% unlist()

    result <- analyze_cell_subset(
      seurat_obj,
      list_of_cells1,
      list_of_cells2,
      group_names = c("SIV_spliced", "SIV_other")
    )
    save_subset_analysis(result, paste0(output.dir, "combined_splicedvnot/", sample_name, "/"))
    
    
    cat("Processed sample:", sample_name, "\n")
  }, error = function(e) {
    cat("Error processing sample", sample_name, ":", conditionMessage(e), "\n\n")
  })
}
```


### cNMF


```{r}
# =============================================================================
# cNMF Pattern Annotation by Biological Program
# =============================================================================
# Annotates cNMF patterns using:
#   1. Preranked GSEA on gene_spectra_score (Z-scores) -- primary method
#   2. ORA on top 50 genes per pattern -- supplementary method
#
# Preranked GSEA is preferred because cNMF produces continuous gene loadings
# across all genes, and GSEA leverages the full ranked structure without an
# arbitrary top-N cutoff (Kotliar et al., eLife 2019).
# Top 50 genes for ORA follows the community standard established by
# Gavish et al., Nature 2023 for cross-study program comparison.
# =============================================================================

# =============================================================================
# Configuration
# =============================================================================
load_dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS016/singleCell/Seurat/RDS/"
output_dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS016/joint/plots/cNMF/"
run_name <- "invitro"
selected_k <- 7

usage_file <- file.path(load_dir, paste0(run_name, "_usage_K", selected_k, ".csv"))
spectra_file <- file.path(load_dir, paste0(run_name, "_spectra_K", selected_k, ".csv"))
spectra_tpm_file <- file.path(load_dir, paste0(run_name, "_spectra_tpm_K", selected_k, ".csv"))
top_genes_file <- file.path(load_dir, paste0(run_name, "_top_genes_K", selected_k, ".csv"))

organism <- "Macaca mulatta"
n_top_genes_ora <- 50
gsea_pval_cutoff <- 0.05
ora_pval_cutoff <- 0.05

# =============================================================================
# 1. Load cNMF Outputs
# =============================================================================
usage <- read.csv(usage_file, row.names = 1, check.names = FALSE)
spectra <- read.csv(spectra_file, row.names = 1, check.names = FALSE)
spectra_tpm <- read.csv(spectra_tpm_file, row.names = 1, check.names = FALSE)
top_genes <- read.csv(top_genes_file, row.names = 1, check.names = FALSE)

cat("Loaded cNMF outputs:\n")
cat("  Usage matrix:", nrow(usage), "cells x", ncol(usage), "patterns\n")
cat("  Spectra (Z-score) matrix:", nrow(spectra), "genes x", ncol(spectra), "patterns\n")
cat("  Spectra (TPM) matrix:", nrow(spectra_tpm), "genes x", ncol(spectra_tpm), "patterns\n")

cat("\nTop 20 genes per pattern (by Z-score loading):\n")
for (pat in colnames(spectra)) {
  pat_scores <- spectra[[pat]]
  names(pat_scores) <- rownames(spectra)
  top_20 <- names(sort(pat_scores, decreasing = TRUE))[1:20]
  cat(paste0("  ", pat, ": ", paste(top_20, collapse = ", "), "\n"))
}

# =============================================================================
# 2. Build Gene Set Collections from MSigDB
# =============================================================================
hallmark_sets <- msigdbr(species = organism, collection = "H") %>%
  split(x = .$gene_symbol, f = .$gs_name)

reactome_sets <- msigdbr(species = organism, collection = "C2", subcollection = "CP:REACTOME") %>%
  split(x = .$gene_symbol, f = .$gs_name)

gobp_sets <- msigdbr(species = organism, collection = "C5", subcollection = "GO:BP") %>%
  split(x = .$gene_symbol, f = .$gs_name)

cat("\nGene set collections loaded:\n")
cat("  Hallmark:", length(hallmark_sets), "sets\n")
cat("  Reactome:", length(reactome_sets), "sets\n")
cat("  GO:BP:", length(gobp_sets), "sets\n")

# =============================================================================
# 3. Preranked GSEA per Pattern
# =============================================================================
# gene_spectra_score Z-scores reflect expected change in gene expression per
# unit increase in pattern usage. fgsea uses the full ranked vector per pattern
# rather than a fixed top-N cutoff (Kotliar et al., eLife 2019, Methods).

run_preranked_gsea <- function(spectra_scores, gene_sets, pval_cutoff = 0.05) {
  patterns <- colnames(spectra_scores)
  all_results <- list()

  for (pat in patterns) {
    ranks <- spectra_scores[[pat]]
    names(ranks) <- rownames(spectra_scores)
    ranks <- ranks[is.finite(ranks)]
    ranks <- sort(ranks, decreasing = TRUE)

    res <- fgsea(
      pathways = gene_sets,
      stats = ranks,
      minSize = 10,
      maxSize = 500,
      nPermSimple = 10000,
      eps = 0
    )
    res$pattern <- pat
    all_results[[pat]] <- res
  }

  combined <- rbindlist(all_results)
  combined <- combined[combined$padj < pval_cutoff, ]
  combined <- combined[order(combined$pattern, combined$padj), ]
  return(as.data.frame(combined))
}

cat("\nRunning preranked GSEA (Hallmark)...\n")
gsea_hallmark <- run_preranked_gsea(spectra, hallmark_sets, gsea_pval_cutoff)

cat("Running preranked GSEA (Reactome)...\n")
gsea_reactome <- run_preranked_gsea(spectra, reactome_sets, gsea_pval_cutoff)

cat("Running preranked GSEA (GO:BP)...\n")
gsea_gobp <- run_preranked_gsea(spectra, gobp_sets, gsea_pval_cutoff)

cat("\nTop 5 Hallmark terms per pattern (by NES):\n")
for (pat in colnames(spectra)) {
  pat_res <- gsea_hallmark[gsea_hallmark$pattern == pat, ]
  if (nrow(pat_res) == 0) {
    cat(paste0("  ", pat, ": no significant terms\n"))
    next
  }
  pat_res <- pat_res[order(abs(pat_res$NES), decreasing = TRUE), ]
  top_terms <- head(pat_res, 5)
  for (j in seq_len(nrow(top_terms))) {
    direction <- ifelse(top_terms$NES[j] > 0, "+", "-")
    term <- gsub("^HALLMARK_", "", top_terms$pathway[j])
    cat(sprintf("  %s: [%s] %s (NES=%.2f, padj=%.1e)\n",
                pat, direction, term, top_terms$NES[j], top_terms$padj[j]))
  }
}

# =============================================================================
# 4. ORA per Pattern (Top 50 Genes) -- Hypergeometric Test
# =============================================================================

get_top_n_genes <- function(spectra_scores, n = 50) {
  top_list <- list()
  for (pat in colnames(spectra_scores)) {
    scores <- spectra_scores[[pat]]
    names(scores) <- rownames(spectra_scores)
    top_list[[pat]] <- names(sort(scores, decreasing = TRUE))[1:n]
  }
  return(top_list)
}

run_ora_hypergeometric <- function(query_genes, gene_sets, background_genes,
                                   min_size = 10, max_size = 500,
                                   pval_cutoff = 0.05) {
  # One-sided Fisher/hypergeometric test for overrepresentation
  bg <- unique(background_genes)
  query <- intersect(query_genes, bg)
  N <- length(bg)
  n_query <- length(query)

  results <- list()
  for (set_name in names(gene_sets)) {
    set_genes <- intersect(gene_sets[[set_name]], bg)
    K <- length(set_genes)
    if (K < min_size || K > max_size) next

    overlap <- intersect(query, set_genes)
    k <- length(overlap)
    if (k == 0) next

    pval <- phyper(k - 1, K, N - K, n_query, lower.tail = FALSE)

    results[[length(results) + 1]] <- data.frame(
      pathway = set_name,
      pval = pval,
      overlap = k,
      set_size = K,
      query_size = n_query,
      background_size = N,
      overlap_genes = paste(overlap, collapse = ";"),
      stringsAsFactors = FALSE
    )
  }

  if (length(results) == 0) {
    return(data.frame(pathway = character(), pval = numeric(), padj = numeric(),
                      overlap = integer(), set_size = integer(),
                      query_size = integer(), background_size = integer(),
                      overlap_genes = character(), stringsAsFactors = FALSE))
  }

  res_df <- do.call(rbind, results)
  res_df$padj <- p.adjust(res_df$pval, method = "BH")
  res_df <- res_df[res_df$padj < pval_cutoff, ]
  res_df <- res_df[order(res_df$padj), ]
  return(res_df)
}

top50_per_pattern <- get_top_n_genes(spectra, n = n_top_genes_ora)
background <- rownames(spectra)

cat("\nRunning ORA on top", n_top_genes_ora, "genes per pattern (GO:BP)...\n")
ora_results <- list()
for (pat in names(top50_per_pattern)) {
  ora_results[[pat]] <- run_ora_hypergeometric(
    query_genes = top50_per_pattern[[pat]],
    gene_sets = gobp_sets,
    background_genes = background,
    pval_cutoff = ora_pval_cutoff
  )
}

cat("\nTop 5 ORA GO:BP terms per pattern:\n")
for (pat in names(ora_results)) {
  res <- ora_results[[pat]]
  if (nrow(res) == 0) {
    cat(paste0("  ", pat, ": no significant terms\n"))
    next
  }
  top_terms <- head(res, 5)
  for (j in seq_len(nrow(top_terms))) {
    cat(sprintf("  %s: %s (padj=%.1e, %d/%d genes)\n",
                pat, gsub("^GOBP_", "", top_terms$pathway[j]),
                top_terms$padj[j], top_terms$overlap[j], n_top_genes_ora))
  }
}

# =============================================================================
# 5. Visualization
# =============================================================================

# --- 5a. GSEA Hallmark NES heatmap ---
if (nrow(gsea_hallmark) > 0) {
  nes_matrix <- gsea_hallmark %>%
    select(pattern, pathway, NES) %>%
    mutate(pathway = gsub("^HALLMARK_", "", pathway)) %>%
    pivot_wider(names_from = pattern, values_from = NES, values_fill = 0) %>%
    tibble::column_to_rownames("pathway")

  nes_mat <- as.matrix(nes_matrix)
  max_abs <- max(abs(nes_mat), na.rm = TRUE)

  p_hallmark <- pheatmap::pheatmap(
    nes_mat,
    color = colorRampPalette(c("blue", "white", "red"))(100),
    breaks = seq(-max_abs, max_abs, length.out = 101),
    cluster_cols = TRUE,
    cluster_rows = TRUE,
    fontsize_row = 7,
    fontsize_col = 10,
    main = "GSEA Hallmark NES per cNMF Pattern",
    silent = TRUE
  )

  ggsave(
    file.path(output_dir, "cNMF_gsea_hallmark_heatmap.pdf"),
    plot = p_hallmark,
    width = 10,
    height = max(6, nrow(nes_mat) * 0.25)
  )
  cat("\nSaved Hallmark GSEA heatmap.\n")
}

# --- 5b. Top genes dot plot per pattern ---
n_show <- 20
plot_data <- list()
for (pat in colnames(spectra)) {
  scores <- spectra[[pat]]
  names(scores) <- rownames(spectra)
  top_g <- names(sort(scores, decreasing = TRUE))[1:n_show]

  plot_data[[pat]] <- data.frame(
    gene = factor(top_g, levels = rev(top_g)),
    pattern = pat,
    z_score = spectra[top_g, pat],
    tpm = spectra_tpm[top_g, pat],
    stringsAsFactors = FALSE
  )
}
plot_df <- do.call(rbind, plot_data)

p_dotplot <- ggplot(plot_df, aes(x = pattern, y = gene, color = z_score, size = tpm)) +
  geom_point() +
  scale_color_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0,
                        name = "Z-score") +
  scale_size_continuous(name = "TPM loading", range = c(0.5, 4)) +
  facet_wrap(~pattern, scales = "free_y", nrow = 1) +
  theme_minimal(base_size = 9) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    strip.text = element_text(size = 10, face = "bold"),
    panel.grid.major.x = element_blank()
  ) +
  labs(x = NULL, y = NULL, title = "Top 20 genes per cNMF pattern")

ggsave(
  file.path(output_dir, "cNMF_top_genes_dotplot.pdf"),
  plot = p_dotplot,
  width = 3 * selected_k,
  height = 8
)
cat("Saved top genes dot plot.\n")

# -- plot GSEA --

n_top_gsea_per_collection <- 5
for (pat in colnames(spectra)) {
  
  combined_list <- list()
  
  for (collection_name in c("Hallmark", "Reactome", "GO:BP")) {
    
    gsea_df <- switch(collection_name,
      "Hallmark" = gsea_hallmark,
      "Reactome" = gsea_reactome,
      "GO:BP"    = gsea_gobp
    )
    
    if (is.null(gsea_df) || nrow(gsea_df) == 0) next
    
    pat_df <- gsea_df[gsea_df$pattern == pat, ]
    if (nrow(pat_df) == 0) next
    
    prefix_pattern <- switch(collection_name,
      "Hallmark" = "^HALLMARK_",
      "Reactome" = "^REACTOME_",
      "GO:BP"    = "^GOBP_"
    )
    
    pat_df$pathway_clean <- gsub(prefix_pattern, "", pat_df$pathway)
    pat_df$pathway_clean <- gsub("_", " ", pat_df$pathway_clean)
    pat_df$pathway_clean <- tolower(pat_df$pathway_clean)
    pat_df$pathway_clean <- paste0(toupper(substring(pat_df$pathway_clean, 1, 1)),
                                    substring(pat_df$pathway_clean, 2))
    pat_df$pathway_clean <- ifelse(nchar(pat_df$pathway_clean) > 50,
                                    paste0(substr(pat_df$pathway_clean, 1, 47), "..."),
                                    pat_df$pathway_clean)
    
    # Top N by |NES| per collection
    pat_df <- pat_df[order(abs(pat_df$NES), decreasing = TRUE), ]
    pat_df <- head(pat_df, n_top_gsea_per_collection)
    pat_df$collection <- collection_name
    
    combined_list[[collection_name]] <- pat_df
  }
  
  if (length(combined_list) == 0) next
  combined_df <- do.call(rbind, combined_list)
  
  combined_df$neg_log10_padj <- -log10(combined_df$padj)
  
  # Set collection as ordered factor for grouping
  combined_df$collection <- factor(combined_df$collection,
                                    levels = c("Hallmark", "Reactome", "GO:BP"))
  
  # Within each collection, order by NES
  combined_df <- combined_df[order(combined_df$collection, combined_df$NES), ]
  
  # Create unique y labels (in case of duplicate pathway names across collections)
  combined_df$y_label <- make.unique(combined_df$pathway_clean, sep = " ")
  combined_df$y_label <- factor(combined_df$y_label, levels = combined_df$y_label)
  
  # Symmetric NES axis
  max_abs_nes <- max(abs(combined_df$NES)) * 1.05
  
  p <- ggplot(combined_df, aes(x = NES, y = y_label, fill = NES, size = neg_log10_padj)) +
    geom_point(shape = 21, stroke = 0.3) +
    scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B", midpoint = 0,
                         name = "NES") +
    scale_size_continuous(name = expression(-log[10]~p[adj]), range = c(2, 7)) +
    scale_x_continuous(limits = c(-max_abs_nes, max_abs_nes)) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey40", linewidth = 0.3) +
    facet_grid(collection ~ ., scales = "free_y", space = "free_y", switch = "y") +
    theme_minimal(base_size = 10) +
    theme(
      panel.grid.major.y = element_blank(),
      panel.grid.minor = element_blank(),
      axis.text.y = element_text(size = 9),
      strip.text.y.left = element_text(size = 10, face = "bold", angle = 0),
      strip.placement = "outside",
      plot.title = element_text(size = 11, face = "bold"),
      legend.position = "right"
    ) +
    labs(
      x = "Normalized Enrichment Score (NES)",
      y = NULL,
      title = paste0(pat, " \u2014 GSEA")
    )
  
  n_terms <- nrow(combined_df)
  ggsave(
    file.path(output_dir, paste0("cNMF_gsea_combined_", pat, ".pdf")),
    plot = p,
    width = 9,
    height = max(4, n_terms * 0.32 + 2)
  )
}
cat("Saved combined GSEA dot plots per pattern.\n")

# Plot ORA
n_top_ora <- 10

for (pat in names(ora_results)) {
  res <- ora_results[[pat]]
  if (is.null(res) || nrow(res) == 0) next
  
  res$pathway_clean <- gsub("^GOBP_", "", res$pathway)
  res$pathway_clean <- gsub("_", " ", res$pathway_clean)
  res$pathway_clean <- tolower(res$pathway_clean)
  res$pathway_clean <- paste0(toupper(substring(res$pathway_clean, 1, 1)),
                               substring(res$pathway_clean, 2))
  res$pathway_clean <- ifelse(nchar(res$pathway_clean) > 50,
                               paste0(substr(res$pathway_clean, 1, 47), "..."),
                               res$pathway_clean)
  
  res$gene_ratio <- res$overlap / res$query_size
  res$neg_log10_padj <- -log10(res$padj)
  
  # Top N by padj
  res <- head(res[order(res$padj), ], n_top_ora)
  
  # Order by gene ratio for plotting
  res <- res[order(res$gene_ratio), ]
  res$pathway_clean <- factor(res$pathway_clean, levels = res$pathway_clean)
  
  p <- ggplot(res, aes(x = gene_ratio, y = pathway_clean, size = overlap, color = neg_log10_padj)) +
    geom_point() +
    scale_color_viridis_c(option = "inferno", direction = -1, end = 0.85,
                          name = expression(-log[10]~p[adj])) +
    scale_size_continuous(name = "Overlap", range = c(2, 7)) +
    theme_minimal(base_size = 10) +
    theme(
      panel.grid.major.y = element_blank(),
      panel.grid.minor = element_blank(),
      axis.text.y = element_text(size = 9),
      plot.title = element_text(size = 11, face = "bold"),
      legend.position = "right"
    ) +
    labs(
      x = "Gene ratio (overlap / query size)",
      y = NULL,
      title = paste0(pat, " \u2014 ORA GO:BP (top ", n_top_genes_ora, " genes)")
    )
  
  ggsave(
    file.path(output_dir, paste0("cNMF_ora_gobp_", pat, ".pdf")),
    plot = p,
    width = 8,
    height = max(3, nrow(res) * 0.35 + 1)
  )
}
cat("Saved ORA GO:BP dot plots per pattern.\n")

# --- 5c. Save results ---
collapse_leading_edge <- function(df) {
  if ("leadingEdge" %in% colnames(df) && is.list(df$leadingEdge)) {
    df$leadingEdge <- vapply(df$leadingEdge, paste, collapse = ";", FUN.VALUE = character(1))
  }
  df
}

write.csv(collapse_leading_edge(gsea_hallmark), file.path(output_dir, "gsea_hallmark_results.csv"), row.names = FALSE)
write.csv(collapse_leading_edge(gsea_reactome), file.path(output_dir, "gsea_reactome_results.csv"), row.names = FALSE)
write.csv(collapse_leading_edge(gsea_gobp), file.path(output_dir, "gsea_gobp_results.csv"), row.names = FALSE)

ora_combined <- do.call(rbind, lapply(names(ora_results), function(pat) {
  df <- ora_results[[pat]]
  if (nrow(df) > 0) df$pattern <- pat
  df
}))
if (!is.null(ora_combined) && nrow(ora_combined) > 0) {
  write.csv(ora_combined, file.path(output_dir, "ora_gobp_results.csv"), row.names = FALSE)
}

top50_df <- data.frame(
  lapply(top50_per_pattern, function(x) { length(x) <- n_top_genes_ora; x }),
  check.names = FALSE
)
write.csv(top50_df, file.path(output_dir, "top50_genes_per_pattern.csv"), row.names = FALSE)

cat("\nAll annotation outputs saved to:", output_dir, "\n")
```

SIV+ vs SIV-
```{r}
# =============================================================================
# Pattern Marker Identification
# =============================================================================
find_pattern_markers <- function(spectra_tpm,
                                  spectra_score = NULL,
                                  n_markers = 50,
                                  method = "weighted_zscore",
                                  min_score = 0,
                                  exclude_shared = FALSE,
                                  shared_threshold = 1.5) {

  stopifnot(is.matrix(spectra_tpm) || is.data.frame(spectra_tpm))
  spectra_tpm <- as.matrix(spectra_tpm)
  stopifnot(all(spectra_tpm >= 0))

  patterns <- colnames(spectra_tpm)
  genes <- rownames(spectra_tpm)
  n_patterns <- ncol(spectra_tpm)

  if (is.null(patterns)) patterns <- paste0("pattern_", seq_len(n_patterns))
  if (is.null(genes)) stop("spectra_tpm must have rownames (gene names).")

  # Compute Z-scores per gene if not provided
  if (is.null(spectra_score)) {
    row_means <- rowMeans(spectra_tpm)
    row_sds <- apply(spectra_tpm, 1, sd)
    row_sds[row_sds == 0] <- 1  # avoid division by zero for constant rows
    spectra_score <- (spectra_tpm - row_means) / row_sds
  } else {
    spectra_score <- as.matrix(spectra_score)
    stopifnot(identical(dim(spectra_score), dim(spectra_tpm)))
  }

  # Build per-gene, per-pattern stats (vectorized)
  # For each gene and each pattern, compute the ratio to the best competing pattern
  # and other useful quantities.
  stats_list <- vector("list", n_patterns)

  for (i in seq_len(n_patterns)) {
    tpm_this <- spectra_tpm[, i]
    zscore_this <- spectra_score[, i]

    # Max loading across all OTHER patterns for each gene
    if (n_patterns == 2) {
      max_other <- spectra_tpm[, -i]
    } else {
      max_other <- apply(spectra_tpm[, -i, drop = FALSE], 1, max)
    }

    ratio <- tpm_this / (max_other + 1e-10)

    stats_list[[i]] <- data.frame(
      gene = genes,
      pattern = patterns[i],
      tpm = tpm_this,
      zscore = zscore_this,
      max_other_tpm = max_other,
      ratio = ratio,
      stringsAsFactors = FALSE
    )
  }

  marker_df <- do.call(rbind, stats_list)
  rownames(marker_df) <- NULL

  # Apply minimum score filter
  if (min_score > 0) {
    marker_df <- marker_df[marker_df$tpm >= min_score, ]
  }

  # Apply shared-gene exclusion filter
  if (exclude_shared) {
    marker_df <- marker_df[marker_df$ratio >= shared_threshold, ]
  }

  # Compute the composite marker_score based on method
  marker_df$marker_score <- switch(method,

    "weighted_zscore" = {
      # Z-score captures specificity (how unusually high this gene loads on
      # this pattern relative to its own distribution). Multiplying by
      # log1p(TPM) upweights genes that are also highly expressed.
      # log1p prevents a handful of very-high-TPM housekeeping genes from
      # dominating; the log compression keeps magnitude informative but
      # bounded.
      marker_df$zscore * log1p(marker_df$tpm)
    },

    "softmax_weighted" = {
      # For each gene, compute softmax across patterns on TPM values.
      # softmax(x_i) = exp(x_i) / sum(exp(x_j)) gives a specificity
      # proportion that amplifies differences more than a linear proportion
      # would. Multiply by TPM to weight by magnitude.
      # Vectorized: subtract row max for numerical stability, then compute.
      row_max <- apply(spectra_tpm, 1, max)
      scaled <- spectra_tpm - row_max  # subtract max per row
      exp_scaled <- exp(scaled)
      softmax_mat <- exp_scaled / rowSums(exp_scaled)
      score_mat <- softmax_mat * spectra_tpm
      # Map back to the long-format marker_df
      as.vector(score_mat[cbind(
        match(marker_df$gene, genes),
        match(marker_df$pattern, patterns)
      )])
    },

    "rank_product" = {
      # Non-parametric approach: for each pattern, rank genes by
      # (1) specificity (ratio to max_other) and (2) magnitude (tpm).
      # Combine as reciprocal of the product of ranks. This is robust
      # to skewed distributions and outliers. Genes that rank well on
      # BOTH dimensions get the highest score.
      rp_scores <- numeric(nrow(marker_df))
      for (pat in patterns) {
        idx <- which(marker_df$pattern == pat)
        pat_data <- marker_df[idx, ]
        n_genes_pat <- nrow(pat_data)
        rank_spec <- rank(-pat_data$ratio, ties.method = "average")
        rank_mag <- rank(-pat_data$tpm, ties.method = "average")
        rp_scores[idx] <- 1 / (rank_spec * rank_mag)
      }
      rp_scores
    },

    "ratio" = {
      # Pure specificity baseline: fold-enrichment over best competitor.
      # No magnitude weighting. Useful for comparison but tends to select
      # low-expression genes that happen to be very pattern-specific.
      marker_df$ratio
    },

    stop("Unknown method: ", method,
         ". Choose from: weighted_zscore, softmax_weighted, rank_product, ratio")
  )

  # Select top markers per pattern
  markers <- list()
  for (pat in patterns) {
    pat_df <- marker_df[marker_df$pattern == pat, ]
    pat_df <- pat_df[order(pat_df$marker_score, decreasing = TRUE), ]
    markers[[pat]] <- head(pat_df$gene, n_markers)
  }

  return(list(
    markers = markers,
    marker_stats = marker_df,
    method = method
  ))
}


# Step 1: Pattern markers weighted for druggability
result <- find_pattern_markers(
  spectra_tpm = spectra_tpm,
  spectra_score = spectra,
  n_markers = 100,
  method = "rank_product",
  min_score = 5,             # require meaningful expression
  exclude_shared = TRUE,     # remove multi-pattern genes
  shared_threshold = 2.0     # require 2x enrichment over next-best
)

# Step 2: Filter against a druggability annotation
# (e.g., DGIdb categories, surface proteins, kinases)
# druggable_genes <- read.csv("dgidb_druggable_genome.csv")$gene
# for (pat in names(result$markers)) {
#   targetable <- intersect(result$markers[[pat]], druggable_genes)
#   cat(sprintf("  %s: %d druggable markers out of %d\n",
#               pat, length(targetable), length(result$markers[[pat]])))
# }
#
# Print top 20 markers per pattern:
markers <- result$markers
for (pat in names(markers)) {
  cat(sprintf("  %s: %s\n", pat, paste(head(markers[[pat]], 20), collapse = ", ")))
}


# =============================================================================
# 3. Infected vs Uninfected Comparison
# =============================================================================

compare_pattern_usage <- function(usage, infected_cellids, group_labels = c("infected", "uninfected")) {
  # Get cell IDs
  all_cellids <- rownames(usage)
  uninfected_cellids <- setdiff(all_cellids, infected_cellids)
  
  cat("Cell counts:\n")
  cat("  Infected:", length(intersect(infected_cellids, all_cellids)), "\n")
  cat("  Uninfected:", length(uninfected_cellids), "\n")
  
  # Subset pattern matrices
  infected_patterns <- usage[rownames(usage) %in% infected_cellids, , drop = FALSE]
  uninfected_patterns <- usage[rownames(usage) %in% uninfected_cellids, , drop = FALSE]
  
  # Calculate summary statistics
  infected_means <- colMeans(infected_patterns)
  uninfected_means <- colMeans(uninfected_patterns)
  
  comparison_df <- data.frame(
    pattern = colnames(usage),
    infected_mean = infected_means,
    uninfected_mean = uninfected_means,
    log2_fold_change = log2((infected_means + 1e-10) / (uninfected_means + 1e-10))
  )
  
  cat("\nPattern usage comparison (means):\n")
  print(comparison_df)
  
  # Statistical testing with Wilcoxon rank-sum test
  stat_results <- lapply(colnames(usage), function(pat) {
    test <- wilcox.test(
      infected_patterns[, pat],
      uninfected_patterns[, pat]
    )
    data.frame(
      pattern = pat,
      p_value = test$p.value,
      infected_median = median(infected_patterns[, pat]),
      uninfected_median = median(uninfected_patterns[, pat])
    )
  })
  
  stat_df <- do.call(rbind, stat_results)
  stat_df$p_adj <- p.adjust(stat_df$p_value, method = "BH")
  
  cat("\nStatistical test results:\n")
  print(stat_df)
  
  # Prepare data for plotting
  plot_data <- rbind(
    data.frame(
      cellID = rownames(infected_patterns),
      group = group_labels[1],
      as.data.frame(infected_patterns),
      check.names = FALSE
    ),
    data.frame(
      cellID = rownames(uninfected_patterns),
      group = group_labels[2],
      as.data.frame(uninfected_patterns),
      check.names = FALSE
    )
  )
  
  return(list(
    comparison = comparison_df,
    stats = stat_df,
    plot_data = plot_data,
    infected_patterns = infected_patterns,
    uninfected_patterns = uninfected_patterns
  ))
}


joint_dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/joint/dataframes/"
invitro_both <- read.csv(paste0(joint_dir, "Invitro_both.csv"))
invitro_GEXonly <- read.csv(paste0(joint_dir, "Invitro_onlySingleCell.csv"))
invitro_SALVEonly <- read.csv(paste0(joint_dir, "Invitro_onlySALVE.csv"))
invitro_all <- read.csv(paste0(joint_dir, "Invitro_all.csv"))

infected <- rbind(invitro_SALVEonly, invitro_both)
infected_cellids <- infected$cellID
comparison <- compare_pattern_usage(usage, infected_cellids)

# =============================================================================
# 4. Visualization Functions
# =============================================================================

# 4a. Violin plot of pattern weights by group
plot_pattern_violins <- function(comparison_result, output_path = NULL) {
  plot_data <- comparison_result$plot_data
  stat_df <- comparison_result$stats
  
  # Convert to long format
  plot_long <- pivot_longer(
    plot_data,
    cols = -c(cellID, group),
    names_to = "pattern",
    values_to = "weight"
  )
  
  # Calculate max weights for significance bar positioning
  max_weights <- plot_long %>%
    group_by(pattern) %>%
    summarise(ymax = max(weight), .groups = "drop")
  
  # Merge with stats and create significance labels
  sig_data <- merge(stat_df, max_weights, by = "pattern")
  sig_data$y_pos <- sig_data$ymax * 1.05
  sig_data$label <- case_when(
    sig_data$p_adj < 0.001 ~ "***",
    sig_data$p_adj < 0.01 ~ "**",
    sig_data$p_adj < 0.05 ~ "*",
    TRUE ~ "ns"
  )
  
  # Create plot
  p <- ggplot(plot_long, aes(x = pattern, y = weight, fill = group)) +
    geom_violin(position = position_dodge(0.8), scale = "width", trim = TRUE) +
    geom_boxplot(
      position = position_dodge(0.8),
      width = 0.1,
      outlier.shape = NA,
      alpha = 0.7
    ) +
    geom_text(
      data = sig_data,
      aes(x = pattern, y = y_pos, label = label),
      inherit.aes = FALSE,
      size = 4
    ) +
    scale_fill_manual(values = c("#E41A1C", "#377EB8")) +
    labs(
      x = "Pattern",
      y = "Pattern Weight",
      fill = "Group",
      title = "cNMF Pattern Usage by Infection Status"
    ) +
    theme_classic() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "bottom"
    )
  
  if (!is.null(output_path)) {
    ggsave(output_path, p, width = 10, height = 6)
    cat("Saved:", output_path, "\n")
  }
  
  return(p)
}

# 4b. Heatmap of pattern usage
plot_usage_heatmap <- function(usage, metadata = NULL, output_path = NULL) {
  # Normalize usage per cell (row-wise)
  usage_norm <- usage / rowSums(usage)
  
  # Set up annotation if metadata provided
  if (!is.null(metadata)) {
    # Ensure rownames match
    common_cells <- intersect(rownames(usage_norm), rownames(metadata))
    usage_norm <- usage_norm[common_cells, ]
    annotation_row <- metadata[common_cells, , drop = FALSE]
  } else {
    annotation_row <- NA
  }
  
  # Create heatmap
  p <- pheatmap(
    t(usage_norm),
    cluster_rows = TRUE,
    cluster_cols = TRUE,
    show_colnames = FALSE,
    color = viridis(100),
    main = "cNMF Pattern Usage (normalized)",
    annotation_col = if (!is.null(metadata)) annotation_row else NA,
    silent = TRUE
  )
  
  if (!is.null(output_path)) {
    pdf(output_path, width = 12, height = 6)
    print(p)
    dev.off()
    cat("Saved:", output_path, "\n")
  }
  
  return(p)
}

# 4c. Top genes heatmap per pattern
plot_marker_heatmap <- function(spectra, markers, n_genes = 20, output_path = NULL) {
  # Get top n genes for each pattern
  all_marker_genes <- unique(unlist(lapply(markers, head, n_genes)))
  
  # Subset spectra to marker genes
  spectra_subset <- spectra[all_marker_genes, ]
  
  # Z-score for visualization
  spectra_scaled <- t(scale(t(spectra_subset)))
  
  # Create annotation for which pattern each gene belongs to
  gene_pattern <- data.frame(
    pattern = sapply(all_marker_genes, function(g) {
      for (pat in names(markers)) {
        if (g %in% head(markers[[pat]], n_genes)) return(pat)
      }
      return(NA)
    }),
    row.names = all_marker_genes
  )
  
  p <- pheatmap(
    spectra_scaled,
    cluster_rows = FALSE,
    cluster_cols = FALSE,
    show_rownames = nrow(spectra_scaled) <= 50,
    color = colorRampPalette(c("blue", "white", "red"))(100),
    main = paste("Top", n_genes, "Marker Genes per Pattern"),
    annotation_row = gene_pattern,
    silent = TRUE
  )
  
  if (!is.null(output_path)) {
    pdf(output_path, width = 8, height = 12)
    print(p)
    dev.off()
    cat("Saved:", output_path, "\n")
  }
  
  return(p)
}

# 4d. Log2 fold change bar plot
plot_log2fc <- function(comparison_result, output_path = NULL) {
  comparison_df <- comparison_result$comparison
  stat_df <- comparison_result$stats
  
  # Merge data
  plot_df <- merge(comparison_df, stat_df[, c("pattern", "p_adj")], by = "pattern")
  plot_df$significant <- plot_df$p_adj < 0.05
  
  p <- ggplot(plot_df, aes(x = reorder(pattern, log2_fold_change), y = log2_fold_change, fill = significant)) +
    geom_bar(stat = "identity") +
    geom_hline(yintercept = 0, linetype = "dashed") +
    scale_fill_manual(values = c("grey60", "#E41A1C"), labels = c("ns", "p < 0.05")) +
    coord_flip() +
    labs(
      x = "Pattern",
      y = "Log2 Fold Change (infected / uninfected)",
      fill = "Significance",
      title = "Pattern Enrichment in Infected Cells"
    ) +
    theme_classic()
  
  if (!is.null(output_path)) {
    ggsave(output_path, p, width = 8, height = 6)
    cat("Saved:", output_path, "\n")
  }
  
  return(p)
}

# =============================================================================
# 5. Add cNMF Results to Seurat Object
# =============================================================================

add_cnmf_to_seurat <- function(seurat_obj, usage) {
  # Align cell barcodes
  common_cells <- intersect(Cells(seurat_obj), rownames(usage))
  cat("Matching cells:", length(common_cells), "of", ncol(seurat_obj), "\n")
  
  # Add pattern weights to metadata
  for (pat in colnames(usage)) {
    seurat_obj[[pat]] <- NA
    seurat_obj[[pat]][common_cells, ] <- usage[common_cells, pat]
  }
  
  return(seurat_obj)
}

# =============================================================================
# Example Complete Analysis
# =============================================================================
# Uncomment and run with your data:

# # Load Seurat object
invitro <- readRDS("/projects/b1042/GoyalLab/egrody/extractedData/EGS016/singleCell/Seurat/RDS/Invitro.rds")
seurat_obj <- invitro
#
# # Add cNMF patterns to Seurat
seurat_obj <- add_cnmf_to_seurat(seurat_obj, usage)
#
# # Define infected cells
# infected <- rbind(invitro_SALVEonly, invitro_both)
# infected_cellids <- infected$cellID
#
# # Run comparison
# comparison <- compare_pattern_usage(usage, infected_cellids)
#
# # Generate plots
# plot_pattern_violins(comparison, file.path(output_dir, "cnmf_pattern_violins.pdf"))
# plot_log2fc(comparison, file.path(output_dir, "cnmf_log2fc.pdf"))
# plot_marker_heatmap(spectra, markers, n_genes = 20, file.path(output_dir, "cnmf_marker_heatmap.pdf"))
#
# # Feature plots in Seurat
FeaturePlot(seurat_obj, features = paste0("Pattern_", 1:selected_k), ncol = 3)


```


