---
title: "EGS002scRNAseq"
author: "Emanuelle Grody"
date: "2024-01-30"
output: html_document
---

This file is for processing SALVEseq reads


```{r, echo = FALSE}
library(dplyr)
library(stringr)
library(patchwork)
library(scCustomize)
library(ggplot2)
library(reshape2)
library(svglite)
library(stringdist)
#library(data.table)
```

# env Analysis
## Previous analysis
```{r, eval = FALSE}
#update the file in this chunk and two chunks down
input.dir = "/projects/b1042/GoyalLab/egrody/20240116_VISER_SALVEseq/analysis/dualRead/"
output.dir = "/projects/b1042/GoyalLab/egrody/20240116_VISER_SALVEseq/analysis/forBLAST"
readsdirectory <- paste0(input.dir, "PL_env/") #let's start with just one for now; we will try separately for each sample

# Get list of CSV files in the directory
file_name <- list.files(readsdirectory, pattern = "*_shavedReads.txt", full.names = TRUE)

# let's read in these reads for now
temp_data <- read.csv(file_name, header = TRUE) %>% select(target) %>% unique()

```
PL_env only has 3401 targets

### Pairwise LV Dist Histogram
The first step is to see what the pairwise LV histogram looks like to see if we have more similar or dissimilar sequences.
```{r}
set.seed(2059)
subsample1 = sample(temp_data$target,3000)
subsample2 = sample(temp_data$target,3000)
subsample3 = sample(temp_data$target,3000)
BarcodesLv1 = as.integer(stringdistmatrix(subsample1, method = "lv"))
BarcodesLv2 = as.integer(stringdistmatrix(subsample2, method = "lv"))
BarcodesLv3 = as.integer(stringdistmatrix(subsample3, method = "lv"))
lBarcodesLv = length(BarcodesLv1)

BarcodesLv = tibble(
  lvdist = c(BarcodesLv1, BarcodesLv2, BarcodesLv3),
  subsamNum = c(rep("subsamping1", lBarcodesLv), rep("subsamping2", lBarcodesLv), rep("subsamping3", lBarcodesLv)))

BarcodesLvHist <- BarcodesLv %>% group_by(subsamNum, lvdist) %>% summarise(length(lvdist)) %>%
  group_by(subsamNum) %>% mutate(totalNum = sum(`length(lvdist)`), fracLvDist = `length(lvdist)`/totalNum)


BarcodesLvHistPlot <- ggplot(BarcodesLvHist, aes(lvdist, fracLvDist)) +
  geom_bar(width = 0.5, stat = 'identity') +
  facet_wrap(facets = vars(subsamNum)) +
  theme_classic()
BarcodesLvHistPlot
ggsave(plot = BarcodesLvHistPlot, file = paste0(readsdirectory, 'PL_env_LVhist.svg'))
```

### BLAST
This is to extract non-redundant and relevant (commented-out) sequences from a subsample to submit for BLAST.
```{r}
# Custom function to compute Levenshtein distance and save the strings being compared
custom_lv_distance <- function(x, y) {
  lv_distance <- stringdistmatrix(x, y, method = "lv")
  row_names <- rep(x, each = length(y))
  col_names <- rep(y, length(x))
  df <- data.frame(dist = as.vector(lv_distance), string1 = row_names, string2 = col_names)
  return(df)
}

pairwise1 <- custom_lv_distance(subsample1, subsample1)
pairwise2 <- custom_lv_distance(subsample2, subsample2)
pairwise3 <- custom_lv_distance(subsample3, subsample3)

# taking the most similar sequences
#subLVdist1 <- pairwise1 %>% filter(dist <= 8 & dist > 0) %>% select(string1) %>% unique()
#subLVdist2 <- pairwise2 %>% filter(dist <= 8 & dist > 0) %>% select(string1) %>% unique()
#subLVdist3 <- pairwise3 %>% filter(dist <= 8 & dist > 0) %>% select(string1) %>% unique()

# triangularizing my matrix

library(reshape2)

pairwise1_long <- melt(pairwise1, varnames = c("string1", "string2"), value.name = "dist")
pairwise1_long <- pairwise1_long[pairwise1_long$string1 != pairwise1_long$string2, ]
pairwise1_triangular <- dcast(pairwise1_long, string1 ~ string2, value.var = "dist")
pairwise1_triangular[upper.tri(pairwise1_triangular, diag = FALSE)] <- NA

pairwise2_long <- melt(pairwise2, varnames = c("string1", "string2"), value.name = "dist")
pairwise2_long <- pairwise2_long[pairwise2_long$string1 != pairwise2_long$string2, ]
pairwise2_triangular <- dcast(pairwise2_long, string1 ~ string2, value.var = "dist")
pairwise2_triangular[upper.tri(pairwise2_triangular, diag = FALSE)] <- NA

pairwise3_long <- melt(pairwise3, varnames = c("string1", "string2"), value.name = "dist")
pairwise3_long <- pairwise3_long[pairwise3_long$string1 != pairwise3_long$string2, ]
pairwise3_triangular <- dcast(pairwise3_long, string1 ~ string2, value.var = "dist")
pairwise3_triangular[upper.tri(pairwise3_triangular, diag = FALSE)] <- NA

# Converting it back
pairwise1_clean <- melt(pairwise1_triangular, varnames = c("string2", "string1"), value.name = "dist")
pairwise2_clean <- melt(pairwise2_triangular, varnames = c("string2", "string1"), value.name = "dist")
pairwise3_clean <- melt(pairwise3_triangular, varnames = c("string2", "string1"), value.name = "dist")

colnames(pairwise1_clean) <- c("string1", "string2", "dist")
colnames(pairwise2_clean) <- c("string1", "string2", "dist")
colnames(pairwise3_clean) <- c("string1", "string2", "dist")

# taking the most similar sequences again
subLVdist1_clean <- pairwise1_clean %>% filter(dist <= 6 & dist > 0) %>% select(string1) %>% unique()
subLVdist2_clean <- pairwise2_clean %>% filter(dist <= 6 & dist > 0) %>% select(string1) %>% unique()
subLVdist3_clean <- pairwise3_clean %>% filter(dist <= 6 & dist > 0) %>% select(string1) %>% unique()
allinone <- rbind(subLVdist1_clean, subLVdist2_clean, subLVdist3_clean)

file_conn <- file(paste0(output.dir, "W2forBLAST.fsa"), "w")
sequences <- unlist(allinone)
#envprimer = "CCAGCAGACCCATATCCAACAGG"

# Loop through each sequence and write it to the file with a custom header
for (i in seq_along(sequences)) {
  header <- sprintf(">sequence_%05d", i)  # Generate the header line
  cat(header, "\n", paste0(sequences[i]), "\n", file = file_conn)  # Write header and sequence to file; paste0 to add envprimer before the sequence
}

# Close the file connection
close(file_conn)
```

### starcode
```{r}
#update the file in this chunk and two chunks down
input.dir = "/projects/b1042/GoyalLab/egrody/20240116_VISER_SALVEseq/analysis/dualRead/"
sample = "PL_env/"
readsdirectory <- paste0(input.dir, sample) #let's start with just one for now
outsdirectory = paste0(output.dir, sample)

# Get list of CSV files in the directory
file_name <- list.files(readsdirectory, pattern = "*_shavedReads.txt", full.names = TRUE)

# output for starcode
#manually looped through each of the file output combinations
VISER_out <- read.csv(file_name, header = TRUE) %>% select(target)
VISER_out$target <- substr(VISER_out$target, 1, 30)
VISER_out <- VISER_out %>% unique()
write.table(VISER_out$target, paste0(readsdirectory, "PL_env_shavedReadsList_30.txt"), sep = ",", quote = FALSE, row.names = FALSE, col.names = FALSE)
```
I saved three different lengths of target sequences: 18bp, 24bp, and 30bp ("full length"), because starcode has a maximum LV dist of 8. I ran starcode in Quest using the starcodeRun.py script over these files. From that analysis, I got a list of the most popular consensus sequences for each sample for each target length. I used those consensus sequences to identify a cutoff for target reads instead of using the reference.

### UTR Read1
```{r, eval = FALSE}
#update the file in this chunk and two chunks down
input.dir = "/projects/b1042/GoyalLab/egrody/20230929_Goyal_P1_BarcodeseqVISER/analysis/dualRead/"
output.dir = "/projects/b1042/GoyalLab/egrody/20230929_Goyal_P1_BarcodeseqVISER/analysis/forBLAST/"
readsdirectory <- paste0(input.dir, "W2/") #let's start with just one for now; we will try separately for each sample

# Get list of CSV files in the directory
file_name <- list.files(readsdirectory, pattern = "*_shavedUtrReads.txt", full.names = TRUE)

# let's read in these reads for now
temp_data <- read.csv(file_name, header = TRUE) %>% select(target) %>% unique()

```
### Pairwise LV Dist Histogram
The first step is to see what the pairwise LV histogram looks like to see if we have more similar or dissimilar sequences.
```{r}
set.seed(2059)
subsample1 = sample(temp_data$target,10000) #can lower to 5k to reduce run time
subsample2 = sample(temp_data$target,10000)
subsample3 = sample(temp_data$target,10000)
BarcodesLv1 = as.integer(stringdistmatrix(subsample1, method = "lv"))
BarcodesLv2 = as.integer(stringdistmatrix(subsample2, method = "lv"))
BarcodesLv3 = as.integer(stringdistmatrix(subsample3, method = "lv"))
lBarcodesLv = length(BarcodesLv1)

BarcodesLv = tibble(
  lvdist = c(BarcodesLv1, BarcodesLv2, BarcodesLv3),
  subsamNum = c(rep("subsamping1", lBarcodesLv), rep("subsamping2", lBarcodesLv), rep("subsamping3", lBarcodesLv)))

BarcodesLvHist <- BarcodesLv %>% group_by(subsamNum, lvdist) %>% summarise(length(lvdist)) %>%
  group_by(subsamNum) %>% mutate(totalNum = sum(`length(lvdist)`), fracLvDist = `length(lvdist)`/totalNum)

BarcodesLvHistPlot <- ggplot(BarcodesLvHist, aes(lvdist, fracLvDist)) +
  geom_bar(width = 0.5, stat = 'identity') +
  facet_wrap(facets = vars(subsamNum)) +
  theme_classic()
BarcodesLvHistPlot
#ggsave(plot = BarcodesLvHistPlot, file = paste0(readsdirectory, 'invitro_LVhist.svg'))
```



### Target cutoff
```{r, eval = FALSE}
#update the file in this chunk and two chunks down
input.dir = "/projects/b1042/GoyalLab/egrody/20240116_VISER_SALVEseq/analysis/dualRead/"
sample = "PL_tat1/"
readsdirectory <- paste0(input.dir, sample) #let's start with just one for now

# Get list of CSV files in the directory
file_name <- list.files(readsdirectory, pattern = "*_shavedReads.txt", full.names = TRUE)

# read in these reads
SALVE_PL_tat1 <- read.csv(file_name, header = TRUE) %>% select(target) %>% unique()

#reference sequences
envprimer = "CCAGCAGACCCATATCCAACAGG"
tat1primer = "GCTAATACATCTTCTGCCCATATCCAACAGGACC"
envReferenceTarget = "CCCGGCACTGCCAACCAGAGAAGGCAGAGA" #30bp, from starcode
tat1ReferenceTarget = "CGGCACTGCCAACCAGAGAAGGCAAGAAGA"

#Cutoff by LV
SALVE_PL_tat1$dist <- NA
for (i in 1:nrow(SALVE_PL_tat1)) {
  # Calculate the Levenshtein distance between the current string and the constant
  dist <- adist(substr(SALVE_PL_tat1$target[i], 1, 30), tat1ReferenceTarget)
  # Assign the distance to the 'dist' column
  SALVE_PL_tat1$dist[i] <- dist
}

hist(SALVE_PL_tat1$dist, breaks = 50, xlab = "Levenshtein Distance: recovered to expected")
```
This histogram has two peaks: one centered at 3-4 and one centered at 17.

What are these more dissimilar sequences? Are they divergent or are they similar to each other?
```{r}
dissimilar_env <- VISER_w2_env %>% filter(dist > 15)
similar_env <- VISER_w2_env %>% filter(dist < 16)
VISER_w2_env_30 <- VISER_w2_env
VISER_w2_env_30$target <- substr(VISER_w2_env_30$target, 1, 18)

set.seed(2059)
subsample1 = sample(VISER_w2_env_30$target,5000)
subsample2 = sample(VISER_w2_env_30$target,5000)
subsample3 = sample(VISER_w2_env_30$target,5000)
BarcodesLv1 = as.integer(stringdistmatrix(subsample1, method = "lv"))
BarcodesLv2 = as.integer(stringdistmatrix(subsample2, method = "lv"))
BarcodesLv3 = as.integer(stringdistmatrix(subsample3, method = "lv"))
lBarcodesLv = length(BarcodesLv1)

BarcodesLv = tibble(
  lvdist = c(BarcodesLv1, BarcodesLv2, BarcodesLv3),
  subsamNum = c(rep("subsamping1", lBarcodesLv), rep("subsamping2", lBarcodesLv), rep("subsamping3", lBarcodesLv)))

BarcodesLvHist <- BarcodesLv %>% group_by(subsamNum, lvdist) %>% summarise(length(lvdist)) %>%
  group_by(subsamNum) %>% mutate(totalNum = sum(`length(lvdist)`), fracLvDist = `length(lvdist)`/totalNum)

BarcodesLvHistPlot <- ggplot(BarcodesLvHist, aes(lvdist, fracLvDist)) +
  geom_bar(width = 0.5, stat = 'identity') +
  facet_wrap(facets = vars(subsamNum)) +
  theme_classic()
BarcodesLvHistPlot
```
Even when just looking at the more dissimilar sequences, we see that they have some sequences that are more similar and some that are divergent. So just setting a cutoff based on the reference genome sequence is not helping us distinguish between real target sequences and PCR/sequencing error.

I ran starcode on all the sequences. Read in all starcode consensus sequences: UPDATE
```{r}
main_folder <- "/projects/b1042/GoyalLab/egrody/20240116_VISER_SALVEseq/analysis/starcode"
lengths = c("18", "24", "30")
samples = c("PL_env", "PL_tat1")
subfolders <- file.path(main_folder, samples)

starcodeTarget <- matrix(NA, nrow = length(samples), ncol = length(lengths))
rownames(starcodeTarget) <- samples
colnames(starcodeTarget) <- lengths


# Loop through the subfolders and files
for (folder_path in subfolders) {
  for (bp in lengths) {
    # List all files in the subfolder with a specific file name structure
    file_list <- list.files(path = folder_path, pattern = paste0("*",bp,"_d8.txt"), full.names = TRUE)
    # Check if any files were found
    if (length(file_list) > 0) {
      # Read the first entry from each file and add it to the dataframe
      first_entry_data <- lapply(file_list, function(file) {
        first_entry <- read.table(file, header = FALSE, sep = "\t", stringsAsFactors = FALSE, nrows = 1) %>% select(V1)  # Change nrows if you need more than one row
        return(first_entry)
      })
  
      # Add the data to the combined dataframe with a unique name
      split <- unlist(strsplit(folder_path, "/"))
      sample <- split[length(split)]
      starcodeTarget[sample, bp] <- do.call(rbind, first_entry_data)[[1]]
      
    }
  }
}

# let's look at the distance of these things to the reference. Just looking at the full length sequences here
for (row in c(7,8,9)) {
  print(starcodeTarget[row])
  print(adist(starcodeTarget[row], referenceTarget))
}

```

Now, let's compare the LV dist of all the reads to these target sequences. This next chunk will take a good amount of time to run, about 10 minutes
```{r, eval = FALSE}
# reading in the data
main_folder <- "/projects/b1042/GoyalLab/egrody/20240116_VISER_SALVEseq/analysis/dualRead/"
lengths = c("18", "24", "30")
samples = c("PL_env", "PL_tat1")
subfolders <- file.path(main_folder, samples)

allSALVE <- list()


# Loop through the subfolders and files
for (folder_path in subfolders) {
  # List all files in the subfolder with a specific file name structure
  file_list <- list.files(path = folder_path, pattern = paste0("*shavedReads.txt"), full.names = TRUE)
  # Check if any files were found
  if (length(file_list) > 0) {
    # Read the first entry from each file and add it to the dataframe
    entry_data <- lapply(file_list, function(file) {
      entry <- read.table(file, header = TRUE, sep = ",") %>% unique() #%>% select(target) %>% unique()
      return(entry)
    })

    # Add the data to the combined dataframe with a unique name
    split <- unlist(strsplit(folder_path, "/"))
    sample <- split[length(split)]
    allSALVE[[sample]] <- entry_data
  }
}

# to index just one column, do this: allVISER$invitro[[1]]$cellID


# comparing to our starcode results
allSALVE_dist <- list()
# Loop through each data frame in allVISER
for (sample in names(allSALVE)) {
  SALVE_PL_env <- allSALVE[[sample]][[1]]
  SALVE_PL_env$dist <- NA
  consensusTarget <- starcodeTarget[sample, "30"]
  for (i in 1:nrow(SALVE_PL_env)) {
    dist <- adist(SALVE_PL_env$target[i], consensusTarget)
    SALVE_PL_env$dist[i] <- dist
  }
  allSALVE_dist[[sample]] <- SALVE_PL_env

  # Plot the histogram for each data frame (optional)
  #hist(VISER_w2_env$dist, breaks = 50, xlab = "Levenshtein Distance: recovered to expected", main = sample)
}

for (sample in names(allSALVE_dist)) {
  plotme <- allSALVE_dist[[sample]]$dist
  hist(plotme, breaks = 50, xlab = "Levenshtein Distance: recovered to consensus", main = sample)
}
```

Based on the histograms, I will filter all sequences that have LV dist greater than 9 from the consensus sequence. I will also save the resulting dataframes to make loading easier.
```{r}
for (sample in names(allVISER_dist)) {
  allVISER_dist[[sample]] <- allVISER_dist[[sample]] %>% filter(dist < 10)
  write.csv(allVISER_dist[[sample]], paste0(main_folder, "/R/", sample, "_shavedReadsFiltered.csv"), row.names=FALSE)
}
```

### Linking to 10X
Read back in VISER sequences.
```{r}
output.dir <- "/projects/b1042/GoyalLab/egrody/20231017_VISER/scViralQuant/"
SingleCell_invitro_umap <- read.csv(paste0(output.dir, "invitro_SingleCellumap_env.csv")) #%>% filter(SingleCellcount > 0)
SingleCell_w2_umap <- read.csv(paste0(output.dir, "w2_SingleCellumap_env.csv")) #%>% filter(SingleCellcount > 0)
SingleCell_invitro_umap_10X <- SingleCell_invitro_umap %>% mutate(log1pSingleCell = log1p(SingleCellcount))
SingleCell_w2_umap_10X <- SingleCell_w2_umap %>% mutate(log1pSingleCell = log1p(SingleCellcount))
```

Updated the variables to be the preclustered 10X UMAP coordinates
```{r}
VISER_invitro <- allVISER_dist[["invitro"]] %>% select(-dist, -target) %>% unique() %>% count(cellID)
VISER_w2 <- allVISER_dist[["W2"]] %>% select(-dist, -target) %>% unique() %>% count(cellID)

invitro_paint_umap = full_join(SingleCell_invitro_umap_10X, VISER_invitro, by = "cellID") %>% rename(VISERcount = n) %>% 
                    mutate(log2ViserCount = log2(VISERcount))
invitro_paint_umap[is.na(invitro_paint_umap)] <- 0
w2_paint_umap = full_join(SingleCell_w2_umap_10X, VISER_w2, by = "cellID") %>% rename(VISERcount = n) %>% 
                    mutate(log2ViserCount = log2(VISERcount))
w2_paint_umap[is.na(w2_paint_umap)] <- 0

# how many cells from each method?
totalvirus_10X <- sum(invitro_paint_umap$log1pSingleCell != 0, na.rm = TRUE)
totalvirus_10X <- sum(w2_paint_umap$log1pSingleCell != 0, na.rm = TRUE)
#totalvirus_VISER <- invitro_paint_umap %>% select(log2SingleCell) %>% filter(log2SingleCell > 0) %>% nrow()
totalvirus_VISER <- sum(invitro_paint_umap$log2ViserCount != 0, na.rm = TRUE)
totalvirus_VISER <- sum(w2_paint_umap$log2ViserCount != 0, na.rm = TRUE)
#totalvirus_VISER <- invitro_paint_umap %>% select(log2ViserCount) %>% filter(log2ViserCount > 0) %>% nrow()
# did it two different ways because I was surprised that this was the outcome; 10X seems low and VISER seems high
```

Not sure why I started using full join here instead of left join (which retains the UMAP'ed cells) or inner join (which keeps only cells in common). Let me see if it makes a difference.
```{r}
test_invitro_paint_umap = left_join(SingleCell_invitro_umap_10X, VISER_invitro, by = "cellID") %>% rename(VISERcount = n) %>% 
                    mutate(log2ViserCount = log2(VISERcount))
test_invitro_paint_umap[is.na(test_invitro_paint_umap)] <- 0
test_w2_paint_umap = left_join(SingleCell_w2_umap_10X, VISER_w2, by = "cellID") %>% rename(VISERcount = n) %>% 
                    mutate(log2ViserCount = log2(VISERcount))
test_w2_paint_umap[is.na(test_w2_paint_umap)] <- 0


test_invitro_paint_umap = left_join(SingleCell_invitro_umap, VISER_invitro, by = "cellID") %>% rename(VISERcount = n) %>% 
                    mutate(log2ViserCount = log2(VISERcount))
test_invitro_paint_umap[is.na(test_invitro_paint_umap)] <- 0
test_w2_paint_umap = left_join(SingleCell_w2_umap, VISER_w2, by = "cellID") %>% rename(VISERcount = n) %>% 
                    mutate(log2ViserCount = log2(VISERcount))
test_w2_paint_umap[is.na(test_w2_paint_umap)] <- 0

totalvirus_10X <- sum(test_invitro_paint_umap$log1pSingleCell != 0, na.rm = TRUE)
totalvirus_10X <- sum(test_w2_paint_umap$log1pSingleCell != 0, na.rm = TRUE)
#totalvirus_VISER <- invitro_paint_umap %>% select(log2SingleCell) %>% filter(log2SingleCell > 0) %>% nrow()
totalvirus_VISER <- sum(test_invitro_paint_umap$log2ViserCount != 0, na.rm = TRUE)
totalvirus_VISER <- sum(test_w2_paint_umap$log2ViserCount != 0, na.rm = TRUE)
#totalvirus_VISER <- invitro_paint_umap %>% select(log2ViserCount) %>% filter(log2ViserCount > 0) %>% nrow()

paint_identities_umap <- test_invitro_paint_umap %>% mutate(identity = case_when(
    log2ViserCount > 0 & log1pSingleCell > 0 ~ "both",
    log2ViserCount > 0 ~ "Viser",
    log1pSingleCell > 0 ~ "SingleCell",
    log2ViserCount == 0 & log1pSingleCell == 0 ~ "neither"
))
paint_identities_umap$identity <- as.factor(paint_identities_umap$identity)

barplot <- ggplot(paint_identities_umap, aes(x = identity)) + 
  geom_bar() + 
  theme_classic()
barplot

testpaintplot <- ggplot(paint_identities_umap, aes(x = UMAP1, y = UMAP2, color = identity)) +
  geom_point(position = "identity") +
  scale_color_manual(values = c("purple", "gray93", "blue", "red")) +
  theme_classic() +
  theme(legend.position = "bottom",
        legend.title = element_text(size = rel(0.6)),
        legend.text = element_text(size = rel(0.6), angle = 30),
        axis.text = element_blank(),
        axis.ticks = element_blank()) +
  labs(title =  "env between methods", color = "")
testpaintplot
```


```{r}
# VISER plots
output.dir <- "/projects/b1042/GoyalLab/egrody/20230929_Goyal_P1_BarcodeseqVISER/analysis/jointUMAP/"
paintplot <- ggplot(invitro_paint_umap, aes(x = UMAP1, y = UMAP2)) +
  geom_point(aes_string(color = invitro_paint_umap$log2ViserCount), size = 1, shape = 16) +
  scale_color_gradient(low = "gray93", high = "darkblue") +
  theme_classic() +
  theme(legend.position = "bottom",
        legend.title = element_text(size = rel(0.6)),
        legend.text = element_text(size = rel(0.6), angle = 30),
        axis.text = element_blank(),
        axis.ticks = element_blank()) +
  labs(title =  "mac239 from VISER, log counts", color = "")
paintplot
ggsave(paintplot, file = paste0(output.dir, "umap_invitro_preclustered_VISER.svg"), width = 7, height = 7)

paintplot <- ggplot(w2_paint_umap, aes(x = UMAP1, y = UMAP2)) +
  geom_point(aes_string(color = w2_paint_umap$log2ViserCount), size = 1, shape = 16) +
  scale_color_gradient(low = "gray93", high = "darkblue") +
  theme_classic() +
  theme(legend.position = "bottom",
        legend.title = element_text(size = rel(0.6)),
        legend.text = element_text(size = rel(0.6), angle = 30),
        axis.text = element_blank(),
        axis.ticks = element_blank()) +
  labs(title =  "mac239 from VISER, log counts", color = "")
paintplot
ggsave(paintplot, file = paste0(output.dir, "umap_w2_preclustered_VISER.svg"), width = 7, height = 7)

# 10X plots, VARIABLES NOT UPDATED!
paintplot <- ggplot(invitro_paint_umap, aes(x = UMAP1, y = UMAP2)) +
  geom_point(aes_string(color = invitro_paint_umap$log2SingleCell), size = 1, shape = 16) +
  scale_color_gradient(low = "gray93", high = "darkblue") +
  theme_classic() +
  theme(legend.position = "bottom",
        legend.title = element_text(size = rel(0.6)),
        legend.text = element_text(size = rel(0.6), angle = 30),
        axis.text = element_blank(),
        axis.ticks = element_blank()) +
  labs(title =  "mac239 from 10X, log counts", color = "")
paintplot
#ggsave(paintplot, file = paste0(output.dir, "logUMAP_invitro_mac239_10X.svg"))

paintplot <- ggplot(w2_paint_umap, aes(x = UMAP1, y = UMAP2)) +
  geom_point(aes_string(color = w2_paint_umap$log2SingleCell), size = 1, shape = 16) +
  scale_color_gradient(low = "gray93", high = "darkblue") +
  theme_classic() +
  theme(legend.position = "bottom",
        legend.title = element_text(size = rel(0.6)),
        legend.text = element_text(size = rel(0.6), angle = 30),
        axis.text = element_blank(),
        axis.ticks = element_blank()) +
  labs(title =  "mac239 from 10X, log counts", color = "")
paintplot
#ggsave(paintplot, file = paste0(output.dir, "logUMAP_w2_mac239_10X.svg"))
```

Where are the shared ones? Where are the not shared ones?
```{r}
#invitro
paint_identities_umap <- invitro_paint_umap %>% mutate(identity = case_when(
    log2ViserCount > 0 & log1pSingleCell > 0 ~ "both",
    log2ViserCount > 0 ~ "Viser",
    log1pSingleCell > 0 ~ "SingleCell",
    log2ViserCount == 0 & log1pSingleCell == 0 ~ "neither"
))
paint_identities_umap$identity <- as.factor(paint_identities_umap$identity)

paintplot <- ggplot(paint_identities_umap, aes(x = UMAP1, y = UMAP2, color = identity)) +
  geom_point(position = "identity") +
  scale_color_manual(values = c("purple", "gray93", "blue", "red")) +
  theme_classic() +
  theme(legend.position = "bottom",
        legend.title = element_text(size = rel(0.6)),
        legend.text = element_text(size = rel(0.6), angle = 30),
        axis.text = element_blank(),
        axis.ticks = element_blank()) +
  labs(title =  "env between methods", color = "")
paintplot
ggsave(paintplot, file = paste0(output.dir, "invitro_betweenmethods.svg"))

#w2
paint_identities_umap <- w2_paint_umap %>% mutate(identity = case_when(
    log2ViserCount > 0 & log1pSingleCell > 0 ~ "both",
    log2ViserCount > 0 ~ "Viser",
    log1pSingleCell > 0 ~ "SingleCell",
    log2ViserCount == 0 & log1pSingleCell == 0 ~ "neither"
))
paint_identities_umap$identity <- as.factor(paint_identities_umap$identity)

paintplot <- ggplot(paint_identities_umap, aes(x = UMAP1, y = UMAP2, color = identity)) +
  geom_point(position = "identity") +
  scale_color_manual(values = c("purple", "gray93", "blue", "red")) +
  theme_classic() +
  theme(legend.position = "bottom",
        legend.title = element_text(size = rel(0.6)),
        legend.text = element_text(size = rel(0.6), angle = 30),
        axis.text = element_blank(),
        axis.ticks = element_blank()) +
  labs(title =  "env between methods", color = "")
paintplot
ggsave(paintplot, file = paste0(output.dir, "w2_betweenmethods.svg"))

```

### Correlation plots
```{r}
corVISER_invitro <- allVISER_dist[["invitro"]] %>% select(cellID, UMI)
corVISER_w2 <- allVISER_dist[["W2"]] %>% select(cellID, UMI)

corVISER_invitro %>% select(-UMI) %>% unique() %>% nrow() #10k unique cellIDs
corVISER_invitro %>% select(-cellID) %>% unique() %>% nrow() #15k unique UMIs
corVISER_w2 %>% select(-UMI) %>% unique() %>% nrow() #7.5k unique cellIDs
corVISER_w2 %>% select(-cellID) %>% unique() %>% nrow() #10k unique UMIs

corUMIcount <- corVISER_invitro %>% group_by(cellID,UMI) %>% count(name = "count") %>% filter(count > 1) #10k UMIs that are > 1
corUMIcount_cellID <- corUMIcount$cellID %>% unique() #3k cellIDs with UMI=1 filtered

corUMIcount_cellID_noN <- corUMIcount_cellID[!(grepl("N", corUMIcount_cellID))] %>% unique() #only 9 cellIDs have N's
corV <- data.frame(cellID = corUMIcount_cellID_noN)

# N correction... why does this need to happen?? why isn't corV enough?
VISER_invitro_noN <- corVISER_invitro %>% filter(!grepl("N", UMI)) %>% filter(!grepl("N", cellID))
VISER_invitro_noN <- VISER_invitro_noN %>% unique() %>% group_by(cellID) %>% summarise(count = length(UMI))
VISER_w2_noN <- corVISER_w2 %>% filter(!grepl("N", UMI)) %>% filter(!grepl("N", cellID))
VISER_w2_noN <- VISER_w2_noN %>% unique() %>% group_by(cellID) %>% summarise(count = length(UMI))

#  now let's repeat the correlation analysis above
VISER_w2_filter <- inner_join(VISER_w2_noN, corV, by = "cellID")
jointTableFilter = inner_join(VISER_w2_filter, SingleCell_w2, by = "cellID") %>% rename(VISERcount = count, SingleCellcount = Count) %>% 
  mutate(log1pViserCount = log1p(VISERcount), log2ViserCount = log2(VISERcount), lnViserCount = log(VISERcount), normalizedViserCount = 1000*VISERcount/sum(VISERcount), logNormViser = log1p(normalizedViserCount), log1pSingleCell = log1p(SingleCellcount))
#cat("Correlation between 10X data and log1p VISER data:", cor(jointTableFilter$SingleCellcount, jointTableFilter$log1pViserCount),
#    "\nCorrelation between 10X data and logNorm VISER data:", cor(jointTableFilter$SingleCellcount, jointTableFilter$logNormViser),
#    "\nCorrelation between 10X data and log2 VISER data:", cor(jointTableFilter$SingleCellcount, jointTableFilter$log2ViserCount),
#    "\nCorrelation between 10X data and ln VISER data:", cor(jointTableFilter$SingleCellcount, jointTableFilter$lnViserCount),
#    "\nCorrelation between 10X data and non-transformed VISER data:", cor(jointTableFilter$SingleCellcount, jointTableFilter$VISERcount),
#    "\nCorrelation between 10X data and non-transformed normalized VISER:", cor(jointTableFilter$SingleCellcount, jointTableFilter$normalizedViserCount))

cat("Correlation between 10X data and log1p VISER data:", cor(jointTableFilter$log1pSingleCell, jointTableFilter$log1pViserCount),
    "\nCorrelation between 10X data and logNorm VISER data:", cor(jointTableFilter$log1pSingleCell, jointTableFilter$logNormViser),
    "\nCorrelation between 10X data and log2 VISER data:", cor(jointTableFilter$log1pSingleCell, jointTableFilter$log2ViserCount),
    "\nCorrelation between 10X data and ln VISER data:", cor(jointTableFilter$log1pSingleCell, jointTableFilter$lnViserCount),
    "\nCorrelation between 10X data and non-transformed VISER data:", cor(jointTableFilter$log1pSingleCell, jointTableFilter$VISERcount),
    "\nCorrelation between 10X data and non-transformed normalized VISER:", cor(jointTableFilter$log1pSingleCell, jointTableFilter$normalizedViserCount))

corrplot <- ggplot(data = jointTableFilter, aes(x=log1pViserCount, y=log1pSingleCell)) +
  geom_point()
ggsave(corrplot, file = paste0(output.dir, "w2_correlationplot.svg"))


VISER_invitro_filter <- inner_join(VISER_invitro_noN, corV, by = "cellID")
jointTableFilter = inner_join(VISER_invitro_filter, SingleCell_invitro, by = "cellID") %>% rename(VISERcount = count, SingleCellcount = Count) %>% 
  mutate(log1pViserCount = log1p(VISERcount), log2ViserCount = log2(VISERcount), lnViserCount = log(VISERcount), normalizedViserCount = 1000*VISERcount/sum(VISERcount), logNormViser = log1p(normalizedViserCount), log1pSingleCell = log1p(SingleCellcount))
#cat("Correlation between 10X data and log1p VISER data:", cor(jointTableFilter$SingleCellcount, jointTableFilter$log1pViserCount),
#    "\nCorrelation between 10X data and logNorm VISER data:", cor(jointTableFilter$SingleCellcount, jointTableFilter$logNormViser),
#    "\nCorrelation between 10X data and log2 VISER data:", cor(jointTableFilter$SingleCellcount, jointTableFilter$log2ViserCount),
#    "\nCorrelation between 10X data and ln VISER data:", cor(jointTableFilter$SingleCellcount, jointTableFilter$lnViserCount),
#    "\nCorrelation between 10X data and non-transformed VISER data:", cor(jointTableFilter$SingleCellcount, jointTableFilter$VISERcount),
#    "\nCorrelation between 10X data and non-transformed normalized VISER:", cor(jointTableFilter$SingleCellcount, jointTableFilter$normalizedViserCount))

cat("Correlation between 10X data and log1p VISER data:", cor(jointTableFilter$log1pSingleCell, jointTableFilter$log1pViserCount),
    "\nCorrelation between 10X data and logNorm VISER data:", cor(jointTableFilter$log1pSingleCell, jointTableFilter$logNormViser),
    "\nCorrelation between 10X data and log2 VISER data:", cor(jointTableFilter$log1pSingleCell, jointTableFilter$log2ViserCount),
    "\nCorrelation between 10X data and ln VISER data:", cor(jointTableFilter$log1pSingleCell, jointTableFilter$lnViserCount),
    "\nCorrelation between 10X data and non-transformed VISER data:", cor(jointTableFilter$log1pSingleCell, jointTableFilter$VISERcount),
    "\nCorrelation between 10X data and non-transformed normalized VISER:", cor(jointTableFilter$log1pSingleCell, jointTableFilter$normalizedViserCount))

corrplot <- ggplot(data = jointTableFilter, aes(x=log1pViserCount, y=log1pSingleCell)) +
  geom_point()
ggsave(corrplot, file = paste0(output.dir, "invitro_correlationplot.svg"))
```


### Exploring the filtered cells

The objective was to figure out what was the origin of all the additional cellIDs we were seeing in VISER versus 10X. I didn't find any evidence for dead cells accounting for these cellIDs. Anyway, many of these cellIDs went away after I filtered using the 10X 3'v3.1 whitesheet (see Current Analysis).

Reading raw counts matrices and no QC filters. Reading in scViserQuant output.
When I take away the filter on the CreateSeuratObject, the Seurat objects become huge. So I just lowered the minimum thresholds instead of getting rid of them.
```{r scViralQuant thru Seurat}
# Reading in supplemented feature matrices
w2.data <- Read10X(data.dir = "/projects/b1042/GoyalLab/egrody/20231017_VISER/counts/Mmul_10_only/run_count_W2/outs/raw_feature_bc_matrix_bbmap/")
w0.data <- Read10X(data.dir = "/projects/b1042/GoyalLab/egrody/20231017_VISER/counts/Mmul_10_only/run_count_W0/outs/raw_feature_bc_matrix_bbmap/")
invitro.data <- Read10X(data.dir = "/projects/b1042/GoyalLab/egrody/20231017_VISER/counts/Mmul_10_only/run_count_invitro/outs/raw_feature_bc_matrix_bbmap/")

# Initialize the Seurat object with the raw (non-normalized data).
w2 <- CreateSeuratObject(counts = w2.data, project = "w2", min.cells = 2, min.features = 100)
w0 <- CreateSeuratObject(counts = w0.data, project = "w0", min.cells = 2, min.features = 100)
invitro <- CreateSeuratObject(counts = invitro.data, project = "invitro", min.cells = 2, min.features = 100)

# QC
w2 <- Add_Mito_Ribo_Seurat(w2, species = "macaque")
w0 <- Add_Mito_Ribo_Seurat(w0, species = "macaque")
invitro <- Add_Mito_Ribo_Seurat(invitro, species = "macaque")
#vplot <- VlnPlot(invitro, features = c("nFeature_RNA", "nCount_RNA", "percent_mito"), ncol = 3)
#ggsave(vplot, file = paste0(output.dir, "invitro_preQC_violinplot.svg"))
#w2 <- subset(w2, subset = nFeature_RNA > 200 & nFeature_RNA < 3500 & nCount_RNA < 20000 & percent_mito < 5)
#w0 <- subset(w0, subset = nFeature_RNA > 200 & nFeature_RNA < 3500 & nCount_RNA < 20000 & percent_mito < 5)
#invitro <- subset(invitro, subset = nFeature_RNA > 200 & nFeature_RNA < 3500 & nCount_RNA < 20000 & percent_mito < 5)
#vplot <- VlnPlot(invitro, features = c("nFeature_RNA", "nCount_RNA", "percent_mito"), ncol = 3)
#ggsave(vplot, file = paste0(output.dir, "invitro_postQC_violinplot.svg"))

# Normalizing
w2 <- NormalizeData(w2, verbose = FALSE)
w0 <- NormalizeData(w0, verbose = FALSE)
invitro <- NormalizeData(invitro, verbose = FALSE)

# Feature selection
w2 <- FindVariableFeatures(w2, selection.method = "vst", nfeatures = 2000, verbose = FALSE)
w0 <- FindVariableFeatures(w0, selection.method = "vst", nfeatures = 2000, verbose = FALSE)
invitro <- FindVariableFeatures(invitro, selection.method = "vst", nfeatures = 2000, verbose = FALSE)

# Scaling
w2.genes <- rownames(w2)
w0.genes <- rownames(w0)
invitro.genes <- rownames(invitro)
w2 <- ScaleData(w2, features = w2.genes, verbose = FALSE)
w0 <- ScaleData(w0, features = w0.genes, verbose = FALSE)
invitro <- ScaleData(invitro, features = invitro.genes, verbose = FALSE)

# Linear dimensional reduction
w2 <- RunPCA(w2, features = VariableFeatures(object = w2), verbose = FALSE)
w0 <- RunPCA(w0, features = VariableFeatures(object = w0), verbose = FALSE)
invitro <- RunPCA(invitro, features = VariableFeatures(object = invitro), verbose = FALSE)
#eplot <- ElbowPlot(invitro, ndims = 50)
#ggsave(eplot, file = paste0(output.dir, "invitro_elbow.svg"))

# Clustering
w2 <- FindNeighbors(w2, dims = 1:30)
w2 <- FindClusters(w2, resolution = 0.5, verbose = FALSE)
w0 <- FindNeighbors(w0, dims = 1:30)
w0 <- FindClusters(w0, resolution = 0.5, verbose = FALSE)
invitro <- FindNeighbors(invitro, dims = 1:30)
invitro <- FindClusters(invitro, resolution = 0.5, verbose = FALSE)

# Dimension reduction
w2 <- RunUMAP(w2, dims = 1:30, verbose = FALSE)#, umap.method = "umap-learn", metric = "correlation")
w0 <- RunUMAP(w0, dims = 1:30, verbose = FALSE)#, umap.method = "umap-learn", metric = "correlation")
invitro <- RunUMAP(invitro, dims = 1:30, verbose = FALSE)#, umap.method = "umap-learn", metric = "correlation")
#dplot <- DimPlot(w2, reduction = "umap")
#ggsave(dplot, file = paste0(output.dir, "w2_umap.svg"))
```

Code to look at the scViralQuant counts: raw, not normalized. W0 did not have any viral reads.
```{r outputs}
# scViralQuant raw counts
invitro_raw = GetAssayData(object = invitro, assay = "RNA", slot = "counts")["env",]
w2_raw = GetAssayData(object = w2, assay = "RNA", slot = "counts")["env",]
SingleCell_invitro <- data.frame(cellID = names(invitro_raw), Count = unname(invitro_raw), stringsAsFactors = FALSE) %>%
  mutate(cellID = str_sub(cellID, end = -3))
SingleCell_w2 <- data.frame(cellID = names(w2_raw), Count = unname(w2_raw), stringsAsFactors = FALSE) %>%
  mutate(cellID = str_sub(cellID, end = -3))

output.dir <- "/projects/b1042/GoyalLab/egrody/20231017_VISER/noQC/"
write.csv(SingleCell_w2, paste0(output.dir, "w2_scViralQuantexpression.csv"))
write.csv(SingleCell_invitro, paste0(output.dir, "invitro_scViralQuantexpression.csv"))

# scViralQuant counts and UMAP coords
w2.umap.coord <- as.data.frame(w2[["umap"]]@cell.embeddings)
invitro.umap.coord <- as.data.frame(invitro[["umap"]]@cell.embeddings)

#SingleCell_umapTest <- data.frame(cellID = names(KLRB1_expression), UMAP1 = test$UMAP_1, UMAP2 = test$UMAP_2, SingleCell_KLRB1_normalizedCounts = unname(KLRB1_expression), stringsAsFactors = FALSE) %>% mutate(cellID = str_sub(cellID, end = -3))

# UMAP coordinates joined to mac239 expression
output.dir = "/projects/b1042/GoyalLab/egrody/20231017_VISER/noQC/"
SingleCell_w2_umap <- data.frame(cellID = rownames(w2.umap.coord), UMAP1 = w2.umap.coord$UMAP_1, UMAP2 = w2.umap.coord$UMAP_2) %>%
  mutate(cellID = str_sub(cellID, end = -3))
SingleCell_w2_umap <- inner_join(SingleCell_w2_umap, SingleCell_w2, by = "cellID") %>% rename(SingleCellcount = Count)
write.csv(SingleCell_w2_umap, paste0(output.dir, "w2_SingleCellumap_env.csv"))

SingleCell_invitro_umap <- data.frame(cellID = rownames(invitro.umap.coord), UMAP1 = invitro.umap.coord$UMAP_1, UMAP2 = invitro.umap.coord$UMAP_2) %>%
  mutate(cellID = str_sub(cellID, end = -3))
SingleCell_invitro_umap <- inner_join(SingleCell_invitro_umap, SingleCell_invitro, by = "cellID") %>% rename(SingleCellcount = Count)
write.csv(SingleCell_invitro_umap, paste0(output.dir, "invitro_SingleCellumap_env.csv"))
```

How many UMIs in the cells that are env+ but got filtered by QC? Consider setting a minimum UMI threshold of 3.
```{r}
noQC_w2_nozero <- SingleCell_w2 %>% filter(Count > 0)
noQC_invitro_nozero <- SingleCell_invitro %>% filter(Count > 0)

output.dir <- "/projects/b1042/GoyalLab/egrody/20231017_VISER/scViralQuant/"
QC_invitro_nozero <- read.csv(paste0(output.dir, "invitro_scViralQuantexpression.csv")) %>% filter(Count > 0)
QC_w2_nozero <- read.csv(paste0(output.dir, "w2_scViralQuantexpression.csv")) %>% filter(Count > 0)

lowquality_invitro <- anti_join(noQC_invitro_nozero, QC_invitro_nozero, by = "cellID") %>% mutate(cells = paste0(cellID, "-1"))
lowquality_w2 <- anti_join(noQC_w2_nozero, QC_w2_nozero, by = "cellID") %>% mutate(cells = paste0(cellID, "-1"))
env_i <- QC_invitro_nozero %>% mutate(cells = paste0(cellID, "-1"))
env_2 <- QC_w2_nozero %>% mutate(cells = paste0(cellID, "-1"))
lowQ_invitro <- subset(invitro, cells = lowquality_invitro$cells)
lowQ_w2 <- subset(w2, cells = lowquality_w2$cells)
env_invitro <- subset(invitro, cells = env_i$cells)
env_w2 <- subset(w2, cells = env_2$cells)

# total UMIs
hist(lowQ_invitro$nCount_RNA, breaks = 50)
hist(invitro$nCount_RNA, breaks = 50)
hist(lowQ_w2$nCount_RNA, breaks = 50)
hist(w2$nCount_RNA, breaks = 50)

#viral UMIs
mean(lowQ_invitro@assays$RNA@counts["env",])
mean(lowQ_w2@assays$RNA@counts["env",])
mean(env_invitro@assays$RNA@counts["env",])
mean(env_w2@assays$RNA@counts["env",])
hist(lowQ_invitro@assays$RNA@counts["env",], breaks = 50, main = "env expression in QC filtered out env+ cells", xlab = "env counts")
hist(env_invitro@assays$RNA@counts["env",], breaks = 50, main = "env expression in all env+ cells", xlab = "env counts")
```

How about the cells that are scViralQuant+ but VISER-?
```{r}
VQonly_invitro <- left_join(SingleCell_invitro_umap_10X, VISER_invitro_full, by = "cellID") %>% rename(VISERcount = n) %>% 
                    mutate(logViserCount = log2(VISERcount))
VQonly_invitro[is.na(VQonly_invitro)] <- 0
VQonly_invitro <- VQonly_invitro %>% filter(logViserCount == 0) %>% filter(log1pSingleCell > 0)

VQonly_w2 <- left_join(SingleCell_w2_umap_10X, VISER_W2_full, by = "cellID") %>% rename(VISERcount = n) %>% 
                    mutate(logViserCount = log2(VISERcount))
VQonly_w2[is.na(VQonly_w2)] <- 0
VQonly_w2 <- VQonly_w2 %>% filter(logViserCount == 0) %>% filter(log1pSingleCell > 0)
compare_invitro <- SingleCell_invitro_umap_10X %>% filter(log1pSingleCell > 0)
compare_w2 <- SingleCell_w2_umap_10X %>% filter(log1pSingleCell > 0)

cat(
  "Mean scViralQuant expression\nInvitro  VQonly: ", mean(VQonly_invitro$log1pSingleCell),
  "   All: ", mean(compare_invitro$log1pSingleCell),
  "\nW2       VQonly: ", mean(VQonly_w2$log1pSingleCell), "   All: ", mean(compare_w2$log1pSingleCell)
)
```


You can just run this chunk now to reload all the variables needed:
```{r}
output.dir <- "/projects/b1042/GoyalLab/egrody/20230929_Goyal_P1_BarcodeseqVISER/analysis/dualRead/splits/"
VISER_invitro_full <- read.csv(paste0(output.dir, "invitro_fullVISER_uniqued.csv"))
VISER_W2_full <- read.csv(paste0(output.dir, "w2_fullVISER_uniqued.csv"))
#VISER_W0_full <- read.csv(paste0(output.dir, "w0_fullVISER_uniqued.csv"))

output.dir <- "/projects/b1042/GoyalLab/egrody/20231017_VISER/noQC/"
SingleCell_invitro_umap <- read.csv(paste0(output.dir, "invitro_SingleCellumap_env.csv")) #%>% filter(SingleCellcount > 0)
SingleCell_w2_umap <- read.csv(paste0(output.dir, "w2_SingleCellumap_env.csv")) #%>% filter(SingleCellcount > 0)
SingleCell_invitro_umap_10X <- SingleCell_invitro_umap %>% mutate(log1pSingleCell = log1p(SingleCellcount))
SingleCell_w2_umap_10X <- SingleCell_w2_umap %>% mutate(log1pSingleCell = log1p(SingleCellcount))

#joint UMAP
test_invitro_paint_umap = left_join(SingleCell_invitro_umap_10X, VISER_invitro_full, by = "cellID") %>% rename(VISERcount = n) %>% 
                    mutate(log2ViserCount = log2(VISERcount))
test_invitro_paint_umap[is.na(test_invitro_paint_umap)] <- 0
test_w2_paint_umap = left_join(SingleCell_w2_umap_10X, VISER_W2_full, by = "cellID") %>% rename(VISERcount = n) %>% 
                    mutate(log2ViserCount = log2(VISERcount))
test_w2_paint_umap[is.na(test_w2_paint_umap)] <- 0

cat(
  "Invitro  scViralQuant: ", sum(test_invitro_paint_umap$log1pSingleCell != 0, na.rm = TRUE), "   VISER: ", 
  sum(test_invitro_paint_umap$log2ViserCount != 0, na.rm = TRUE),
  "\nW2       scViralQuant: ", sum(test_w2_paint_umap$log1pSingleCell != 0, na.rm = TRUE), "   VISER: ", 
  sum(test_w2_paint_umap$log2ViserCount != 0, na.rm = TRUE)
)

```




How about the mitochondrial content? Is it higher than usual?
```{r}
Idents(lowQ_invitro) <- "orig.ident"
vplot <- VlnPlot(lowQ_invitro, features = c("nFeature_RNA", "nCount_RNA", "percent_mito"), ncol = 3)
Idents(lowQ_w2) <- "orig.ident"
vplot <- VlnPlot(lowQ_w2, features = c("nFeature_RNA", "nCount_RNA", "percent_mito"), ncol = 3)
vplot
```
So the low quality cells in W2 do have high mitochondrial reads (but not the invitro). Let's isolate them and look at their viral reads.

```{r}
lowQ_w2_highmito <- subset(lowQ_w2, subset = percent_mito > 5)
cat(
  "Of the", ncol(lowQ_w2), "cells that have low quality,", ncol(lowQ_w2_highmito), "of them have a mitochondrial percentage > 5"
)

lowQ_w2_highmito_env = GetAssayData(object = lowQ_w2_highmito, assay = "RNA", slot = "counts")["env",]
lowQ_w2_highmito_env
mean(lowQ_w2_highmito@assays$RNA@counts["env",])
mean(lowQ_w2@assays$RNA@counts["env",])

lowQ_w2_env = GetAssayData(object = lowQ_w2, assay = "RNA", slot = "counts")["env",]
lowQ_w2_env

mito_vs_env <- data.frame(cells = names(GetAssayData(object = lowQ_w2, assay = "RNA", slot = "counts")["env",]),
                env = unname(GetAssayData(object = lowQ_w2, assay = "RNA", slot = "counts")["env",]),
                mito = lowQ_w2@meta.data$percent_mito)
mito_vs_env_UMI <- mito_vs_env %>% filter(env > 2)
ggplot(mito_vs_env, aes(x = mito, y = env)) + 
  geom_point() + 
  theme_classic() + 
  labs(x = "% mitochondrial reads", y = "env expression", title = "W2 filtered out cells")
ggplot(mito_vs_env_UMI, aes(x = mito, y = env)) + 
  geom_point(size = 3) + 
  theme_classic() + 
  labs(x = "% mitochondrial reads", y = "env expression", title = "W2 filtered out cells, for env > 2")
```



### Filter gag only
Looking at the possible sources of gag reads, we realized that gag transcripts could be read with an erroneous cellID sequence. To avoid this from adding more scViralQuant+ cells to our data, I will filter these cells out (alternatively, I could have only added env to the reference genome instead of the whole SIV genome). The easiest way to do this is to only look at env reads.
```{r}
#How many cells have 5pUTR or gag or pol expression?
fivePrime <- SingleCell_allviralgenes_w2_filtered %>% filter("5pUTR" != 0 | gag != 0 | pol != 0) #564
#How many cells have env or nef expression?
threePrime <- SingleCell_allviralgenes_w2_filtered%>% filter(env != 0 | nef != 0) #277
#How many have only env or nef and not 5pUTR or gag or pol?
threePrimeOnly <- SingleCell_allviralgenes_w2_filtered%>% filter(env != 0 | nef != 0 & "5pUTR" == 0 & gag == 0 & pol == 0) #265
```

## Current analysis

The sample FASTQ files were divided into multiple "splits" or "chunks" (used interchangeably) to improve processing speed thru the upstream seqIO analysis and avoid memory issues. Now we read all the splits in to reconstruct the full dataset.

### Loading in splits
```{r}
main_folder <- "/projects/b1042/GoyalLab/egrody/20230929_VISER/analysis/dualRead/"
samples = c("PL_env", "PL_tat1", "PL_tat2")
subfolders <- file.path(main_folder, samples)

SALVE_PLenv_full <- data.frame()
SALVE_PLtat1_full <- data.frame()
SALVE_PLtat2_full <- data.frame()


# Read in shaved reads

```

First, we ensure that our cell barcodes are present in the 10X 3'v3.1 whitelist:
```{r}
whitelist_location <- "/home/egy2296/packages/cellranger-7.2.0/lib/python/cellranger/barcodes/3M-february-2018.txt.gz"
whitelist <- as.list(fread(whitelist_location, sep = "\n", header = FALSE, data.table = FALSE)$V1)

SALVE_PLenv_whitelist <- SALVE_PLenv_full %>% filter(cellID %in% whitelist)
SALVE_PLtat1_whitelist <- SALVE_PLtat1_full %>% filter(cellID %in% whitelist)
SALVE_PLtat2_whitelist <- SALVE_PLtat2_full %>% filter(cellID %in% whitelist)

```
For these datasets, we filter out the following percentages (rounded) of cellIDs which are likely artifacts: 23% from invitro, 31% from W2, and 71% from W0.


Next, we trim divergent env sequences. We start by loading in the most populous sequences from our starcode analysis. For reference, these sequences are as follows:
invitro: ACACTGCACTGCGAACCAGAGAAGGCAAA
W2: ACACTGCACTGCCACCCAGAGAGGGCAAAG
W0: CTGAAGATCGGAAGAGCGTCGTGTAGG
```{r}
# trim by distance to starcode results

main_folder <- "/projects/b1042/GoyalLab/egrody/20230929_VISER/analysis/starcode"
subfolders <- file.path(main_folder, samples)

starcodeTarget <- matrix(nrow = length(samples), ncol = 1)
rownames(starcodeTarget) <- samples
colnames(starcodeTarget) <- "30"


# Loop through the subfolders and files
for (folder_path in subfolders) {
    # List all files in the subfolder with a specific file name structure
    file_list <- list.files(path = folder_path, pattern = paste0("*30_d8.txt"), full.names = TRUE)
    # Check if any files were found
    if (length(file_list) > 0) {
      # Read the first entry from each file and add it to the dataframe
      first_entry_data <- lapply(file_list, function(file) {
        first_entry <- read.table(file, header = FALSE, sep = "\t", stringsAsFactors = FALSE, nrows = 1) %>% select(V1)  # Change nrows if you need more than one row
        return(first_entry)
      })
  
      # Add the data to the combined dataframe with a unique name
      split <- unlist(strsplit(folder_path, "/"))
      sample <- split[length(split)]
      starcodeTarget[sample] <- do.call(rbind, first_entry_data)[[1]]
  }
}
starcodeTarget <- na.omit(starcodeTarget)

# Add LV dist column

VISER_invitro_whitelist$dist <- adist(VISER_invitro_whitelist$target, starcodeTarget["invitro"])
VISER_W2_whitelist$dist <- adist(VISER_W2_whitelist$target, starcodeTarget["W2"])
VISER_W0_whitelist$dist <- adist(VISER_W0_whitelist$target, starcodeTarget["W0"])

# Plotting
plotme <- VISER_W2_whitelist$dist
hist(plotme, breaks = 50, xlab = "Levenshtein Distance: recovered to consensus", main = "W2")
plotme <- VISER_W0_whitelist$dist
hist(plotme, breaks = 50, xlab = "Levenshtein Distance: recovered to consensus", main = "W0")
plotme <- VISER_invitro_whitelist$dist
hist(plotme, breaks = 50, xlab = "Levenshtein Distance: recovered to consensus", main = "invitro")


# Trimming at 10
VISER_invitro_whitelist <- VISER_invitro_whitelist %>% filter(dist < 10) %>% select(-dist, -target) %>% unique() %>% count(cellID)
VISER_W2_whitelist <- VISER_W2_whitelist %>% filter(dist < 10) %>% select(-dist, -target) %>% unique() %>% count(cellID)
VISER_W0_whitelist <- VISER_W0_whitelist %>% filter(dist < 10) %>% select(-dist, -target) %>% unique() %>% count(cellID)

output.dir <- "/projects/b1042/GoyalLab/egrody/20230929_VISER/analysis/seqIOdualRead/splits/cleaned/"

write.csv(VISER_invitro_whitelist, paste0(output.dir, "invitro_VISER_whitelist_uniqued.csv"))
write.csv(VISER_W2_whitelist, paste0(output.dir, "w2_VISER_whitelist_uniqued.csv"))
write.csv(VISER_W0_whitelist, paste0(output.dir, "w0_VISER_whitelist_uniqued.csv"))
```

Finally, let's filter out cellIDs with UMIs = 1 as we can't confirm that these are real:
```{r}
VISER_invitro_whitelist <- VISER_invitro_whitelist %>% filter(n > 1)
VISER_W2_whitelist <- VISER_W2_whitelist %>% filter(n > 1)
VISER_W0_whitelist <- VISER_W0_whitelist %>% filter(n > 1)
```


### Linking to 10X
Read back in 10X data with env counts and UMAP coordinates. See EGS002scRNAseq.Rm for the analysis that generated these datasets.
```{r}
output.dir <- "/projects/b1042/GoyalLab/egrody/20231017_VISER/analysis/scViralQuant/outputs/"
SingleCell_invitro_umap <- read.csv(paste0(output.dir, "invitro_SingleCellumap_env.csv")) #%>% filter(SingleCellcount > 0)
SingleCell_w2_umap <- read.csv(paste0(output.dir, "w2_SingleCellumap_env.csv")) #%>% filter(SingleCellcount > 0)
SingleCell_invitro_umap_10X <- SingleCell_invitro_umap %>% mutate(log1pSingleCell = log1p(SingleCellcount))
SingleCell_w2_umap_10X <- SingleCell_w2_umap %>% mutate(log1pSingleCell = log1p(SingleCellcount))

#output.dir <- "/projects/b1042/GoyalLab/egrody/20230929_VISER/analysis/seqIOdualRead/splits/cleaned/"
#VISER_invitro_full <- read.csv(paste0(output.dir, "invitro_VISER_whitelist_uniqued.csv"))
#VISER_W2_full <- read.csv(paste0(output.dir, "w2_VISER_whitelist_uniqued.csv"))
#when reading in, remember to filter UMIs = 1 if desired
```

Updated the variables to be the preclustered 10X UMAP coordinates.
```{r}
invitro_paint_umap = full_join(SingleCell_invitro_umap_10X, VISER_invitro_whitelist, by = "cellID") %>% rename(VISERcount = n) %>% 
                    mutate(log2ViserCount = log2(VISERcount))
invitro_paint_umap[is.na(invitro_paint_umap)] <- 0
w2_paint_umap = full_join(SingleCell_w2_umap_10X, VISER_W2_whitelist, by = "cellID") %>% rename(VISERcount = n) %>% 
                    mutate(log2ViserCount = log2(VISERcount))
w2_paint_umap[is.na(w2_paint_umap)] <- 0

# how many env+ cells from each method?
totalvirus_10X <- sum(invitro_paint_umap$log1pSingleCell != 0, na.rm = TRUE)
totalvirus_10X <- sum(w2_paint_umap$log1pSingleCell != 0, na.rm = TRUE)
#totalvirus_VISER <- invitro_paint_umap %>% select(log2SingleCell) %>% filter(log2SingleCell > 0) %>% nrow()
totalvirus_VISER <- sum(invitro_paint_umap$log2ViserCount != 0, na.rm = TRUE)
totalvirus_VISER <- sum(w2_paint_umap$log2ViserCount != 0, na.rm = TRUE)
#totalvirus_VISER <- invitro_paint_umap %>% select(log2ViserCount) %>% filter(log2ViserCount > 0) %>% nrow()
# did it two different ways because I was surprised that this was the outcome; 10X seems low and VISER seems high
```

To see what are in common, now we use left_join to look at only the cells with 10X anchors:
```{r}
test_invitro_paint_umap = left_join(SingleCell_invitro_umap_10X, VISER_invitro_whitelist, by = "cellID") %>% rename(VISERcount = n) %>% 
                    mutate(log2ViserCount = log2(VISERcount))
test_invitro_paint_umap[is.na(test_invitro_paint_umap)] <- 0
test_w2_paint_umap = left_join(SingleCell_w2_umap_10X, VISER_W2_whitelist, by = "cellID") %>% rename(VISERcount = n) %>% 
                    mutate(log2ViserCount = log2(VISERcount))
test_w2_paint_umap[is.na(test_w2_paint_umap)] <- 0


totalvirus_10X <- sum(test_invitro_paint_umap$log1pSingleCell != 0, na.rm = TRUE)
totalvirus_10X <- sum(test_w2_paint_umap$log1pSingleCell != 0, na.rm = TRUE)
#totalvirus_VISER <- invitro_paint_umap %>% select(log2SingleCell) %>% filter(log2SingleCell > 0) %>% nrow()
totalvirus_VISER <- sum(test_invitro_paint_umap$log2ViserCount != 0, na.rm = TRUE)
totalvirus_VISER <- sum(test_w2_paint_umap$log2ViserCount != 0, na.rm = TRUE)
#totalvirus_VISER <- invitro_paint_umap %>% select(log2ViserCount) %>% filter(log2ViserCount > 0) %>% nrow()

```

I'm not seeing a marked increase in the number of joint cells, which shouldn't be a surprise given that these are extensively PCR'ed, well-mixed samples.
^Later: Not sure what this comment means

### Correlation plots
```{r}
corVISER_invitro <- VISER_invitro_whitelist
corVISER_w2 <- VISER_W2_whitelist

corVISER_invitro %>% select(-UMI) %>% unique() %>% nrow() #10k unique cellIDs
corVISER_invitro %>% select(-cellID) %>% unique() %>% nrow() #15k unique UMIs
corVISER_w2 %>% select(-UMI) %>% unique() %>% nrow() #7.5k unique cellIDs
corVISER_w2 %>% select(-cellID) %>% unique() %>% nrow() #10k unique UMIs

corUMIcount <- corVISER_invitro %>% group_by(cellID,UMI) %>% count(name = "count") %>% filter(count > 1) #10k UMIs that are > 1
corUMIcount_cellID <- corUMIcount$cellID %>% unique() #3k cellIDs with UMI=1 filtered

corUMIcount_cellID_noN <- corUMIcount_cellID[!(grepl("N", corUMIcount_cellID))] %>% unique() #only 9 cellIDs have N's
corV <- data.frame(cellID = corUMIcount_cellID_noN)

# N correction... why does this need to happen?? why isn't corV enough?
VISER_invitro_noN <- corVISER_invitro %>% filter(!grepl("N", UMI)) %>% filter(!grepl("N", cellID))
VISER_invitro_noN <- VISER_invitro_noN %>% unique() %>% group_by(cellID) %>% summarise(count = length(UMI))
VISER_w2_noN <- corVISER_w2 %>% filter(!grepl("N", UMI)) %>% filter(!grepl("N", cellID))
VISER_w2_noN <- VISER_w2_noN %>% unique() %>% group_by(cellID) %>% summarise(count = length(UMI))

#  now let's repeat the correlation analysis above
VISER_w2_filter <- inner_join(VISER_w2_noN, corV, by = "cellID")
jointTableFilter = inner_join(VISER_w2_filter, SingleCell_w2, by = "cellID") %>% rename(VISERcount = count, SingleCellcount = Count) %>% 
  mutate(log1pViserCount = log1p(VISERcount), log2ViserCount = log2(VISERcount), lnViserCount = log(VISERcount), normalizedViserCount = 1000*VISERcount/sum(VISERcount), logNormViser = log1p(normalizedViserCount), log1pSingleCell = log1p(SingleCellcount))
#cat("Correlation between 10X data and log1p VISER data:", cor(jointTableFilter$SingleCellcount, jointTableFilter$log1pViserCount),
#    "\nCorrelation between 10X data and logNorm VISER data:", cor(jointTableFilter$SingleCellcount, jointTableFilter$logNormViser),
#    "\nCorrelation between 10X data and log2 VISER data:", cor(jointTableFilter$SingleCellcount, jointTableFilter$log2ViserCount),
#    "\nCorrelation between 10X data and ln VISER data:", cor(jointTableFilter$SingleCellcount, jointTableFilter$lnViserCount),
#    "\nCorrelation between 10X data and non-transformed VISER data:", cor(jointTableFilter$SingleCellcount, jointTableFilter$VISERcount),
#    "\nCorrelation between 10X data and non-transformed normalized VISER:", cor(jointTableFilter$SingleCellcount, jointTableFilter$normalizedViserCount))

cat("Correlation between 10X data and log1p VISER data:", cor(jointTableFilter$log1pSingleCell, jointTableFilter$log1pViserCount),
    "\nCorrelation between 10X data and logNorm VISER data:", cor(jointTableFilter$log1pSingleCell, jointTableFilter$logNormViser),
    "\nCorrelation between 10X data and log2 VISER data:", cor(jointTableFilter$log1pSingleCell, jointTableFilter$log2ViserCount),
    "\nCorrelation between 10X data and ln VISER data:", cor(jointTableFilter$log1pSingleCell, jointTableFilter$lnViserCount),
    "\nCorrelation between 10X data and non-transformed VISER data:", cor(jointTableFilter$log1pSingleCell, jointTableFilter$VISERcount),
    "\nCorrelation between 10X data and non-transformed normalized VISER:", cor(jointTableFilter$log1pSingleCell, jointTableFilter$normalizedViserCount))

corrplot <- ggplot(data = jointTableFilter, aes(x=log1pViserCount, y=log1pSingleCell)) +
  geom_point()
ggsave(corrplot, file = paste0(output.dir, "w2_correlationplot.svg"))


VISER_invitro_filter <- inner_join(VISER_invitro_noN, corV, by = "cellID")
jointTableFilter = inner_join(VISER_invitro_filter, SingleCell_invitro, by = "cellID") %>% rename(VISERcount = count, SingleCellcount = Count) %>% 
  mutate(log1pViserCount = log1p(VISERcount), log2ViserCount = log2(VISERcount), lnViserCount = log(VISERcount), normalizedViserCount = 1000*VISERcount/sum(VISERcount), logNormViser = log1p(normalizedViserCount), log1pSingleCell = log1p(SingleCellcount))
#cat("Correlation between 10X data and log1p VISER data:", cor(jointTableFilter$SingleCellcount, jointTableFilter$log1pViserCount),
#    "\nCorrelation between 10X data and logNorm VISER data:", cor(jointTableFilter$SingleCellcount, jointTableFilter$logNormViser),
#    "\nCorrelation between 10X data and log2 VISER data:", cor(jointTableFilter$SingleCellcount, jointTableFilter$log2ViserCount),
#    "\nCorrelation between 10X data and ln VISER data:", cor(jointTableFilter$SingleCellcount, jointTableFilter$lnViserCount),
#    "\nCorrelation between 10X data and non-transformed VISER data:", cor(jointTableFilter$SingleCellcount, jointTableFilter$VISERcount),
#    "\nCorrelation between 10X data and non-transformed normalized VISER:", cor(jointTableFilter$SingleCellcount, jointTableFilter$normalizedViserCount))

cat("Correlation between 10X data and log1p VISER data:", cor(jointTableFilter$log1pSingleCell, jointTableFilter$log1pViserCount),
    "\nCorrelation between 10X data and logNorm VISER data:", cor(jointTableFilter$log1pSingleCell, jointTableFilter$logNormViser),
    "\nCorrelation between 10X data and log2 VISER data:", cor(jointTableFilter$log1pSingleCell, jointTableFilter$log2ViserCount),
    "\nCorrelation between 10X data and ln VISER data:", cor(jointTableFilter$log1pSingleCell, jointTableFilter$lnViserCount),
    "\nCorrelation between 10X data and non-transformed VISER data:", cor(jointTableFilter$log1pSingleCell, jointTableFilter$VISERcount),
    "\nCorrelation between 10X data and non-transformed normalized VISER:", cor(jointTableFilter$log1pSingleCell, jointTableFilter$normalizedViserCount))

corrplot <- ggplot(data = jointTableFilter, aes(x=log1pViserCount, y=log1pSingleCell)) +
  geom_point()
ggsave(corrplot, file = paste0(output.dir, "invitro_correlationplot.svg"))
```



### LV dist of cellIDs
I want to look at the cellsIDs that are...
1. VISER+ and scViralQuant-
2. scViralQuant+ and first VISER+ and then VISER-.
```{r}
#VISER+ and scViralQuant-
VISERonly_invitro <- VISER_invitro_whitelist %>% filter(!(cellID %in% test_invitro_paint_umap$cellID))
VISERonly_W2 <- VISER_W2_whitelist %>% filter(!(cellID %in% test_w2_paint_umap$cellID))
VISERonly_invitro_cellIDLv = as.integer(stringdistmatrix(VISERonly_invitro$cellID, method = "lv"))
VISERonly_W2_cellIDLv = as.integer(stringdistmatrix(VISERonly_W2$cellID, method = "lv"))

hist(VISERonly_invitro_cellIDLv, breaks = 50,
     main = "Histogram of pairwise LV dist between cellIDs: invitro",
    xlab = "cellIDs env+ in VISER only")
hist(VISERonly_W2_cellIDLv, breaks = 50,
     main = "Histogram of pairwise LV dist between cellIDs: W2",
    xlab = "cellIDs env+ in VISER only")


#VISER+ and scViralQuant+
VISER10X_invitro <- VISER_invitro_whitelist %>% filter(cellID %in% test_invitro_paint_umap$cellID)
VISER10X_W2 <- VISER_W2_whitelist %>% filter(cellID %in% test_w2_paint_umap$cellID)
VISER10X_invitro_cellIDLv = as.integer(stringdistmatrix(VISER10X_invitro$cellID, method = "lv"))
VISER10X_W2_cellIDLv = as.integer(stringdistmatrix(VISER10X_W2$cellID, method = "lv"))

hist(VISER10X_invitro_cellIDLv, breaks = 50,
     main = "Histogram of pairwise LV dist between cellIDs: invitro",
    xlab = "cellIDs env+ in VISER and 10X")
hist(VISER10X_W2_cellIDLv, breaks = 50,
     main = "Histogram of pairwise LV dist between cellIDs: W2",
    xlab = "cellIDs env+ in VISER and 10X")

#ggsave(plot = BarcodesLvHistPlot, file = paste0(readsdirectory, 'invitro_LVhist.svg'))

```


### UMI cutoff >2
Let's take a quick peek at a UMI cutoff of 3
```{r}
VISER_invitro_UMI3 <- VISER_invitro_whitelist %>% filter(n > 2)
VISER_W2_UMI3 <- VISER_W2_whitelist %>% filter(n > 2)
VISER_W0_UMI3 <- VISER_W0_whitelist %>% filter(n > 2)
```

To see what are in common, now we use left_join to look at only the cells with 10X anchors:
```{r}
UMI3_invitro_paint_umap = left_join(SingleCell_invitro_umap_10X, VISER_invitro_UMI3, by = "cellID") %>% rename(VISERcount = n) %>% 
                    mutate(log2ViserCount = log2(VISERcount))
UMI3_invitro_paint_umap[is.na(test_invitro_paint_umap)] <- 0
UMI3_w2_paint_umap = left_join(SingleCell_w2_umap_10X, VISER_W2_UMI3, by = "cellID") %>% rename(VISERcount = n) %>% 
                    mutate(log2ViserCount = log2(VISERcount))
UMI3_w2_paint_umap[is.na(test_w2_paint_umap)] <- 0


totalvirus_10X <- sum(UMI3_invitro_paint_umap$log1pSingleCell != 0, na.rm = TRUE)
totalvirus_10X <- sum(UMI3_w2_paint_umap$log1pSingleCell != 0, na.rm = TRUE)
#totalvirus_VISER <- invitro_paint_umap %>% select(log2SingleCell) %>% filter(log2SingleCell > 0) %>% nrow()
totalvirus_VISER <- sum(UMI3_invitro_paint_umap$log2ViserCount != 0, na.rm = TRUE)
totalvirus_VISER <- sum(UMI3_w2_paint_umap$log2ViserCount != 0, na.rm = TRUE)
#totalvirus_VISER <- invitro_paint_umap %>% select(log2ViserCount) %>% filter(log2ViserCount > 0) %>% nrow()

```


