---
title: "EGS009scRNAseq"
author: "Emanuelle Grody"
date: "2024-12-03"
output: html_document
---


```{r, echo = FALSE}
source("~/SALVEseq/packages.R")
source("~/SALVEseq/functions.R")
```


# Single Cell Data

We take as an input the count matrix(es). To get from base calls to count matrices, run cellranger with custom reference Mmu (no viral genes).

## Functions
```{r}

# Function to analyze gene expression in multiple Seurat objects
analyze_seurat_genes <- function(seurat_list, genes_of_interest, min_count = 0) {
  # Initialize results list
  results <- list()
  
  # Initialize empty dataframe for counts
  count_df <- data.frame()
  
  # Loop through each Seurat object
  for (i in seq_along(seurat_list)) {
    seurat_obj <- seurat_list[[i]]
    obj_name <- names(seurat_list)[i]
    
    # Check if genes exist in the object
    genes_present <- genes_of_interest %in% rownames(GetAssayData(seurat_obj))
    results[[obj_name]] <- data.frame(
      gene = genes_of_interest,
      present = genes_present
    )
    
    # Get cell IDs and counts for present genes
    if (any(genes_present)) {
      counts <- GetAssayData(seurat_obj, slot = "counts")
      present_genes <- genes_of_interest[genes_present]
      
      for (gene in present_genes) {
        gene_counts <- counts[gene, ]
        temp_df <- data.frame(
          object = obj_name,
          cell_id = colnames(seurat_obj),
          gene = gene,
          count = as.numeric(gene_counts)
        )
        # Filter for counts above minimum threshold
        temp_df <- temp_df[temp_df$count > min_count, ]
        count_df <- rbind(count_df, temp_df)
      }
    }
  }
  
  return(list(
    gene_presence = results,
    expression_data = count_df
  ))
}

# For exporting vRNA+ cells
reshape_gene_data <- function(df) {
  wide_df <- df %>%
    pivot_wider(
      id_cols = cell_id,
      names_from = gene,
      values_from = count,
      values_fill = 0
    )
  
  expected_cols <- c("pol", "env", "nef")
  missing_cols <- setdiff(expected_cols, names(wide_df))
  
  if (length(missing_cols) > 0) {
    for (col in missing_cols) {
      wide_df[[col]] <- 0
    }
  }
  
  wide_df <- wide_df %>%
    select(cell_id, all_of(expected_cols))
  
  return(wide_df)
}
```


## Current analysis
### Seurat loading

```{r}
output_dir = "/projects/b1042/GoyalLab/egrody/20241127_EGS009/analysis/Seurat/"
input_root = "/projects/b1042/GoyalLab/egrody/20241127_EGS009/analysis/counts/"

objects <- list()
tissues <- c("P","L","G")
conditions <- c("acute", "ART")
for (i in tissues) {
  for (j in conditions) {
    sample_name <- paste(i, j, "GEX", sep = "_")
    input_dir <- paste0(input_root, "run_count_", sample_name, "/outs/filtered_feature_bc_matrix/")
    if (sample_name != "G_acute_GEX") {
      objects[[sample_name]] <- SeuratPipeline(input_dir, sample_name = sample_name, output_dir = output_dir, plots = "TRUE")
    }
  }
}
```

### Mmulmac alignment analysis
```{r}
output_dir = "/projects/b1042/GoyalLab/egrody/20241127_EGS009/analysis/Seurat/GEX_Mmulmac/"
input_root = "/projects/b1042/GoyalLab/egrody/20241127_EGS009/analysis/counts/GEX_Mmulmac/"

objects <- list()
tissues <- c("P","L","G")
conditions <- c("acute", "ART")
for (i in tissues) {
  for (j in conditions) {
    sample_name <- paste(i, j, "GEX", sep = "_")
    input_dir <- paste0(input_root, "run_count_", sample_name, "/outs/filtered_feature_bc_matrix/")
    if (sample_name != "G_acute_GEX") {
      objects[[sample_name]] <- SeuratPipeline(input_dir, sample_name = sample_name, output_dir = output_dir, plots = "TRUE")
    }
  }
}
```

```{r}
# not sure about this list
viral_genes <- c("5'U3",
"5'R",
"5'U5",
"5'UTR",
"gag",
"pol",
"vif",
"vpx",
"vpr",
"tat",
"rev",
"env",
"nef",
"3'U3",
"3'R",
"3'U5")

results <- analyze_seurat_genes(objects, viral_genes, min_count = 1)

# vRNA+ cells:
uniquecells <- results$expression_data %>% select(cell_id) %>%
  unique()
cat("Number of cells vRNA+: ", nrow(uniquecells))

comparecells_GEX <- results$expression_data %>% filter(gene == "pol" | gene == "env" | gene =="nef")

comparecells_GEX <- reshape_gene_data(comparecells_GEX)

output_dir <- "/projects/b1042/GoyalLab/egrody/20241127_EGS009/analysis/joint/"
file_path <- paste0(output_dir, "comparecells_GEX_Mmulmac.csv")
write.csv(comparecells_GEX, file_path)
```



## Reuse code

### Without virus UMAP coords
If we are loading in the filtered data without viral gene annotation, we join the scPathoQuant output (not normalized?) to the cellIDs. 

```{r}
virusfreeUMAPscPathoQuant <- function(sample, input.directory, seurat_object, gene_list = NULL) {
  expression_file <- paste0(input.directory, sample, "/pathogen_al_gene_counts_mac239.csv")
  if (!file.exists(expression_file)) {
    stop("scPathoQuant output file does not exist: ", expression_file)
  }
  
  expression <- tryCatch({
    read.csv(expression_file)
  }, error = function(e) {
    stop("Error reading CSV file: ", e$message)
  })
  
  if (!is.data.frame(expression)) {
    stop("Expression data is not a dataframe")
  }
  
  if (ncol(expression) < 3) {
    stop("Expression data should have at least 3 columns: cell ID, gene, and count")
  }
  
  colnames(expression)[1:3] <- c("cellID", "Gene", "Count")
  if (!is.null(gene_list)) {
    expression <- expression[expression$Gene %in% gene_list, ]
    if (nrow(expression) == 0) {
      stop("No genes from the provided gene_list found in the expression data")
    }
  } else {
    gene_list <- unique(expression$Gene)
  }
  if (!requireNamespace("Seurat", quietly = TRUE)) {
    stop("Seurat package is required but not installed")
  }
  
  if (!"umap" %in% names(seurat_object@reductions)) {
    stop("UMAP reduction not found in Seurat object")
  }
  
  cell_metadata <- data.frame(
    cellID = colnames(seurat_object),
    UMAP1 = seurat_object@reductions$umap@cell.embeddings[,1],
    UMAP2 = seurat_object@reductions$umap@cell.embeddings[,2]
  )
  
  expression_wide <- reshape2::dcast(expression, cellID ~ Gene, value.var = "Count", fill = 0)
  combined_data <- merge(cell_metadata, expression_wide, by = "cellID", all.x = TRUE)
  
  combined_data[is.na(combined_data)] <- 0
  missing_genes <- setdiff(gene_list, colnames(combined_data))
  if (length(missing_genes) > 0) {
    combined_data[, missing_genes] <- 0
  }
  
  gene_cols <- setdiff(colnames(combined_data), c("cellID", "UMAP1", "UMAP2"))
  combined_data <- combined_data[, c("cellID", "UMAP1", "UMAP2", gene_cols)]
  
  return(combined_data)
}

output.dir <- "/projects/b1042/GoyalLab/egrody/20240116_VISER_SALVEseq/EGS004/scRNAseq/"
input.dir <- "/projects/b1042/GoyalLab/egrody/20240116_VISER_SALVEseq/EGS004/scRNAseq/scPathoQuant/"
gene_list <- c("5pU3", "5pR", "5pU5", "5pUTR",
               "gag", "pol", "vif", "vpx",
               "vpr", "tat", "rev", "env",
               "nef", "3pU3", "3pR", "3pU5")
combined_d13 <- virusfreeUMAPscPathoQuant("D13", input.dir, d13, gene_list)
write.csv(combined_d13, paste0(output.dir, "d13_virusfreeUMAPscPathoQuant.csv"))
combined_d83 <- virusfreeUMAPscPathoQuant("D83", input.dir, d83, gene_list)
write.csv(combined_d83, paste0(output.dir, "d83_virusfreeUMAPscPathoQuant.csv"))
combined_d195 <- virusfreeUMAPscPathoQuant("D195", input.dir, d195, gene_list)
write.csv(combined_d195, paste0(output.dir, "d195_virusfreeUMAPscPathoQuant.csv"))
```

### scPathoQuant UMAP coords
If we are loading in the scPathoQuant data, we next isolate the viral counts from the Seurat objects.
```{r output viral expression .csvs}
objects <- list(d13 = d13, d83 = d83, d195 = d195)
genes <- c("env", "tat", "mac239")
output_dir <- "/projects/b1042/GoyalLab/egrody/20240116_VISER_SALVEseq/EGS004/scRNAseq/"
save_scPQ_expression(
  objects, 
  genes, 
  output_dir = output_dir)

```

### Mmul10 UMAP painted with scViralQuant (skip)

I use the scViralQuant alignment information from the previous section, run Seurat on the non-scViralQuant original matrices, and join the two datasets by cellID.

```{r}
w2_painted <- SeuratPipeline("/projects/b1042/GoyalLab/egrody/20231017_VISER/counts/Mmul_10_only/run_count_W2/outs/raw_feature_bc_matrix/", "w2")

SingleCell_w2_painted <- scViralCountCellID(w2_painted, "env")
SingleCell_w2_painted_umap <- scUMAPViralCellID(w2_painted, SingleCell_w2)
```

Plotting.
```{r}
plotUMAP(data = SingleCell_w2_painted_umap,
         colorby = log1pSingleCell,
         title = "scViralQuant onto preclustered 10X, log counts",
         output_dir = output_dir,
         saveas = "scviralquant_preclustered10X_w2")
```



