---
title: "EGS024"
author: "Emanuelle Grody"
date: "2026-02-26"
output: html_document
---

```{r, message=FALSE, warning=FALSE}
source("~/SALVEseq/extractionScripts/R/packages.R")
source("~/SALVEseq/extractionScripts/R/functions.R")
```


# Individual Datasets

## SALVE
### bamsort
Run this once first to get raw cellIDs:
```{r raw_cellIDs}
# Load and save raw cell IDs from 10X data
raw_cellIDs <- list()
input.dir = "/projects/b1042/GoyalLab/egrody/extractedData/EGS024/singleCell/counts/concat/"
output_dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS024/singleCell/raw_cellIDs" 

# Create output directory if it doesn't exist
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}

all_cellIDs <- data.frame(cellID = character(), sample = character(), stringsAsFactors = FALSE)

dataset_name <- "GFP_GEX"
sample_name <- "GFP_invitro"
rawDataFolder <- paste0(input.dir, "Mmul_10_mac239_GFP_", 
                        dataset_name, "/outs/raw_feature_bc_matrix/")

tryCatch({
  rawdata <- Read10X(rawDataFolder)
  umi_counts <- colSums(rawdata)
  cell_ids <- names(umi_counts[umi_counts > 100]) # keep only cells with transcriptomes
  raw_cellIDs[[sample_name]] <- cell_ids
  cat("Loaded", length(cell_ids), "raw cellIDs for sample", sample_name, "\n")
  
  sample_df <- data.frame(
    cellID = cell_ids,
    sample = rep(sample_name, length(cell_ids)),
    stringsAsFactors = FALSE
  )
  all_cellIDs <- rbind(all_cellIDs, sample_df)
  
  # Save individual sample's cell IDs to CSV
  sample_file <- file.path(output_dir, paste0(sample_name, "_raw_cellIDs.csv"))
  write.csv(data.frame(cellID = cell_ids), sample_file, row.names = FALSE)
  cat("Saved", length(cell_ids), "cell IDs to", sample_file, "\n")
  
}, error = function(e) {
  cat("Error processing", dataset_name, ":", conditionMessage(e), "\n")
})

if (exists("rawdata")) {
  rm(rawdata)
  gc()
}


# To easily load this data in the future:
# raw_cellIDs_df <- read.csv("path/to/output/directory/GFP_invitro_raw_cellIDs.csv")
# raw_cellIDs <- split(raw_cellIDs_df$cellID, raw_cellIDs_df$sample)
```

Run bamsort_combined (alignment_reads and splice) first.
Making full and minimum filtered lists:
```{r}
input.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS024/SALVE/bamsort/alignment/"
output.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS024/SALVE/bamsort/reads/"

samples <- c("GFP_invitro")
raw_cellIDs <- read.csv("/projects/b1042/GoyalLab/egrody/extractedData/EGS024/singleCell/raw_cellIDs/GFP_invitro_raw_cellIDs.csv")

process_bamsortv6(samples, input.dir, output.dir, raw_cellIDs,
      correction_dir = "/projects/b1042/GoyalLab/egrody/extractedData/EGS024/SALVE/bamsort/splice/")

# minimums
input_dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS024/SALVE/bamsort/reads/"
output_dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS024/SALVE/bamsort/reads/minimum/" 


process_all_set_minimumsv6("SALVE", samples, input_dir, TRUE, output_dir, 
                           min_reads = 2, min_umi = 2)

```


## GEX
### bamsort
```{r}
samples = c(
  "GFP_invitro"
)

# Process
input.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS024/singleCell/bamsort/alignment/"
output.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS024/singleCell/bamsort/reads/"
raw_cellIDs <- read.csv("/projects/b1042/GoyalLab/egrody/extractedData/EGS024/singleCell/raw_cellIDs/GFP_invitro_raw_cellIDs.csv")

process_bamsortv6(samples, input.dir, output.dir, raw_cellIDs,
        correction_dir = "/projects/b1042/GoyalLab/egrody/extractedData/EGS024/singleCell/bamsort/splice/")

# Minimums
input_dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS024/singleCell/bamsort/reads/"
output_dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS024/singleCell/bamsort/reads/minimum/"

process_all_set_minimumsv6("GEX", samples, input_dir, TRUE, output_dir, min_reads_cell = 1)

```

### Seurat: UMAP coords and cell size
```{r}
samples <- data.frame(
  datasets = c(
    "GFP_GEX"
  ),
  samples = c(
    "GFP_invitro"
  ),
  folders = c(
    "/projects/b1042/GoyalLab/egrody/extractedData/EGS024/singleCell/counts/concat/"
  )
)

output.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS024/singleCell/Seurat/"

for (i in 1:nrow(samples)) {
  sample_name <- samples$samples[i]
  dataset <- samples$datasets[i]
  folder <- samples$folders[i]
  input.dir <- paste0(folder, "Mmul_10_mac239_GFP_", dataset, "/outs/filtered_feature_bc_matrix/")
  
  tryCatch({
    seurat_obj <- SeuratPipeline(input.dir, sample_name, 
                       output_dir = output.dir, 
                       plots = TRUE, rds = TRUE)
    
    # Generate expression data frame without gene expression
    sample_df <- targetExpressionDF(seurat_obj, genes = "")

    # Add sample and target columns to the dataframe
    sample_df$size <- colSums(GetAssayData(seurat_obj, assay = "RNA", layer = "counts"))
    sample_df$sample <- sample_name

    outUMAP.dir <- paste0(output.dir, "UMAPcoords/")
    if (!dir.exists(outUMAP.dir)) {
      dir.create(outUMAP.dir, recursive = TRUE)
    }
    write.csv(sample_df, paste0(outUMAP.dir, sample_name, "UMAP_coords.csv"))
    
    cat("Processed sample:", sample_name, "\n")
  }, error = function(e) {
    cat("Error processing sample", sample_name, ":", conditionMessage(e), "\n\n")
  })
}


```

### Seurat: CD4 and cell type markers
```{r}
genes <- c(
  # CD4+ T cell
  "CD4",
  # CD3 components
  "CD3D", "CD3E", "CD3G",
  # CD8
  "CD8A", "CD8B",
  # CD4+ T cell markers with higher transcript abundance
  "IL7R", "LCK", "LEF1", "TCF7", "CCR7",
  # CD4+ T helper subsets
  "GATA3", "TBX21", "RORC", "FOXP3", "IL2RA",
  # Contaminant markers
  "NKG7", "KLRD1",   # NK cells
  "CD14", "CD68",     # monocytes
  "MS4A1", "CD79A"    # B cells
)

input.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS024/singleCell/Seurat/RDS/"
output.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS024/singleCell/Seurat/expressionDF/"
if (!dir.exists(output.dir)) {
  dir.create(output.dir, recursive = TRUE)
}

sample_name <- "GFP_invitro"
input_file <- paste0(input.dir, sample_name, ".rds")
seurat_obj <- readRDS(input_file)

# Generate expression data frame
sample_df <- targetExpressionDF(seurat_obj, genes = genes)
sample_df$sample <- sample_name

write.csv(sample_df, paste0(output.dir, sample_name, "_CD4T.csv"))

coex <- sample_df %>%
  # T cell identity
  mutate(CD3 = ifelse(CD3D > 0 | CD3E > 0 | CD3G > 0, 1, 0)) %>%
  mutate(CD4_transcript = ifelse(CD4 > 0, 1, 0)) %>%
  mutate(CD4_signature = ifelse(IL7R > 0 | LCK > 0 | LEF1 > 0 | TCF7 > 0, 1, 0)) %>%
  mutate(CD8 = ifelse(CD8A > 0 | CD8B > 0, 1, 0)) %>%
  # CD4+ T helper subsets (require CD4 signature context)
  mutate(Th1 = ifelse(CD4_signature == 1 & TBX21 > 0, 1, 0)) %>%
  mutate(Th2 = ifelse(CD4_signature == 1 & GATA3 > 0, 1, 0)) %>%
  mutate(Th17 = ifelse(CD4_signature == 1 & RORC > 0, 1, 0)) %>%
  mutate(Treg = ifelse(CD4_signature == 1 & FOXP3 > 0, 1, 0)) %>%
  mutate(activated = ifelse(IL2RA > 0, 1, 0)) %>%
  # Contaminants -- use markers not shared with activated T cells
  mutate(NK = ifelse(KLRD1 > 0 & CD3 == 0, 1, 0)) %>%
  mutate(monocyte = ifelse(CD14 > 0 | CD68 > 0, 1, 0)) %>%
  mutate(B_cell = ifelse(MS4A1 > 0 | CD79A > 0, 1, 0)) %>%
  select(cellID, CD3, CD4_transcript, CD4_signature, CD8,
         Th1, Th2, Th17, Treg, activated, NK, monocyte, B_cell)

cat("Total cells:\t", nrow(coex),
    "\nCD3+:\t\t", sum(coex$CD3 == 1),
    "\nCD4+:\t\t", sum(coex$CD4_transcript == 1),
    "\nCD4T signature:\t", sum(coex$CD4_signature == 1),
    "\nCD8+:\t\t", sum(coex$CD8 == 1),
    "\nTh1:\t\t", sum(coex$Th1 == 1),
    "\nTh2:\t\t", sum(coex$Th2 == 1),
    "\nTh17:\t\t", sum(coex$Th17 == 1),
    "\nTreg:\t\t", sum(coex$Treg == 1),
    "\nactivated:\t", sum(coex$activated == 1),
    "\nNK:\t\t", sum(coex$NK == 1),
    "\nMonocyte:\t", sum(coex$monocyte == 1),
    "\nB cell:\t\t", sum(coex$B_cell == 1))
```
## Barcode
```{r}
umap.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS024/singleCell/Seurat/UMAPcoords/"
UMAP_data <- read.csv(paste0(umap.dir, "GFP_invitroUMAP_coords.csv"), row.names = "X")
UMAP_data <- UMAP_data %>% select(cellID, UMAP1, UMAP2, cluster, size)

singlet.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS024/barcode/singletCode/"

side_reactions <- c("GFP_barcode_1", "GFP_barcode_9")
base.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS024/barcode/stepThree/"

barcodeData <- list()
filtered_cells <- tibble()

for (i in side_reactions) {
  cat(i, "\n")

  singlet_data <- read.csv(paste0(singlet.dir, i, "__singlets_all.txt"), col.names = "cellID")
  singlet_data <- singlet_data %>% mutate(cellID = paste0(cellID, "-1"))

  temp <- read.delim(file.path(base.dir, i, "stepThreeStarcodeShavedReads.txt"))
  barcodeData[[i]] <- temp %>% mutate(cellID = paste0(cellID, "-1")) %>%
    select(cellID, UMI, BC50StarcodeD8) %>%
    filter(cellID %in% singlet_data$cellID) %>%
    filter(cellID %in% UMAP_data$cellID) %>%
    count(cellID, BC50StarcodeD8, name = "count") %>%
    filter(count > 4)

  test <- barcodeData[[i]]
  
  # number of barcodes per cell
  test2 <- test %>%
    select(cellID, BC50StarcodeD8) %>%
    distinct(cellID, BC50StarcodeD8) %>%
    count(cellID, name = "barcodes", sort = TRUE)

  cells1bc <- test2 %>% filter(barcodes == 1)
  cells2bc <- test2 %>% filter(barcodes == 2)
  cell_combos <- test %>%
    filter(cellID %in% cells2bc$cellID) %>%
    arrange(cellID, BC50StarcodeD8) %>%
    group_by(cellID) %>%
    summarise(combo = paste(BC50StarcodeD8, collapse = "_")) %>%
    ungroup()
  valid_combos <- cell_combos %>%
    count(combo) %>%
    filter(n > 1)
  cell_combos <- semi_join(cell_combos, valid_combos, by = "combo")
  cat("Filtered cells: \n1 bc\t\t", nrow(cells1bc), "\n2 bc in clones\t", nrow(cell_combos), 
      "\n2 bc combos\t", nrow(valid_combos), "\n\n")

  fc_1bc <- test %>%
    filter(cellID %in% cells1bc$cellID) %>%
    distinct(cellID, BC50StarcodeD8) %>%
    select(cellID, barcode = BC50StarcodeD8)
  fc_2bc <- cell_combos %>%
    select(cellID, barcode = combo)
  filtered_cells <- bind_rows(filtered_cells, fc_1bc, fc_2bc) %>%
    distinct(cellID, .keep_all = TRUE)
}

# removing problematic barcodes
barcode_counts <- filtered_cells %>%
  count(barcode, name = "count", sort = TRUE)
combo <- barcode_counts %>% filter(grepl("_", barcode))

partner_check <- combo %>%
  rowwise() %>%
  mutate(bc_list = list(strsplit(barcode, "_")[[1]])) %>%
  unnest(bc_list) %>%
  group_by(bc_list) %>%
  summarise(
    n_unique_partners = n_distinct(barcode),
    .groups = "drop"
  )

# Blacklist: barcodes appearing in >3 distinct combos
blacklist <- partner_check %>%
  filter(n_unique_partners > 3) %>%
  pull(bc_list)

cat("Blacklisted barcodes:", length(blacklist))

blacklist_pattern <- paste(blacklist, collapse = "|")
filtered_clean_cells <- filtered_cells %>%
  filter(!grepl(blacklist_pattern, barcode))

output.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS024/barcode/"
write.csv(filtered_clean_cells, paste0(output.dir, "filtered3_barcoded_cells.csv"))

```
Comparing barcodes in library_1 and library_9
```{r}
# overlap between GFP_barcode_1 and GFP_barcode_9
results <- list()

for (threshold in 0:5) {
  barcodeData <- list()
  for (i in side_reactions) {
    temp <- read.csv(file.path(base.dir, i, "stepThreeStarcodeShavedReads.txt"), 
                     sep = "\t")
    barcodeData[[i]] <- temp %>% mutate(cellID = paste0(cellID, "-1")) %>%
      select(cellID, UMI, BC50StarcodeD8) %>%
      filter(cellID %in% singlet_data$cellID) %>%
      filter(cellID %in% UMAP_data$cellID) %>%
      count(cellID, BC50StarcodeD8, name = "count") %>%
      filter(count > threshold)
  }
  
  bc1 <- unique(barcodeData[[1]]$BC50StarcodeD8)
  bc2 <- unique(barcodeData[[2]]$BC50StarcodeD8)
  
  shared <- length(intersect(bc1, bc2))
  only1 <- length(setdiff(bc1, bc2))
  only2 <- length(setdiff(bc2, bc1))
  
  results[[as.character(threshold)]] <- data.frame(
    threshold = threshold,
    barcodes_1 = length(bc1),
    barcodes_2 = length(bc2),
    shared = shared,
    only_in_1 = only1,
    only_in_2 = only2,
    jaccard = shared / length(union(bc1, bc2))
  )
}

overlap_df <- bind_rows(results)
print(overlap_df)
```

### Check top barcodes dominance
Are my top barcodes the same top barcodes as Nitu got during her diversity check sequencing? If so, there may be a bias in the barcode diversity
```{r}
input.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS024/barcode/NK2_barcodeDiversity/analyzed/Multiple_Samples/separated/"
diversity <- c(barcode = character(), count = integer())
for (i in 1:3) {
  div_part <- read.csv(paste0(input.dir, "NK2_old_", i, "_diversity_final50_8.txt"),
                       sep = "\t", header = FALSE, col.names = c("BC50", "count"))
  diversity <- rbind(diversity, div_part)
}

diversity <- diversity %>%
  group_by(BC50) %>%
  summarise(count = sum(count)) %>%
  ungroup() %>% arrange(desc(count))

input.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS024/barcode/"
barcoded_cells <- read.csv(paste0(input.dir, "filtered_barcoded_cells.csv"), row.names = "X")
barcode_counts <- barcoded_cells %>%
  count(barcode, name = "count", sort = TRUE)

barcodes <- barcode_counts %>%
  pull(barcode) %>%
  strsplit("_") %>%
  unlist() %>%
  unique()

barcodes <- substr(barcodes, 1, 50)
# results <- lapply(barcodes, function(query) {
#   dists <- stringdist(query, diversity$BC50, method = "lv")
#   hits <- which(dists <= 8)
#   if (length(hits) == 0) return(NULL)
#   data.frame(
#     query   = query,
#     target  = diversity$barcode[hits],
#     lv_dist = dists[hits]
#   )
# }) %>% bind_rows()

cat("Diversity Check # barcodes:\t", nrow(diversity), "\n")
cat("Emmie Experiment # barcodes:\t", length(barcodes), "\n")
cat("# barcodes shared:\t", sum(barcodes %in% diversity$BC50), "\n")
#cat("# barcodes within LV dist 8:\t", nrow(results))

```


### two-barcode combos
```{r}
input.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS024/barcode/"
barcoded_cells <- read.csv(paste0(input.dir, "filtered3_barcoded_cells.csv"), row.names = "X")
barcode_counts <- barcoded_cells %>%
  count(barcode, name = "count", sort = TRUE)

single <- barcode_counts %>% filter(!grepl("_", barcode))
combo  <- barcode_counts %>% filter(grepl("_", barcode))

cat("Single barcodes:", nrow(single), "\n")
cat("Combo barcodes:", nrow(combo), "\n")

combo_individual <- combo %>%
  pull(barcode) %>%
  strsplit("_") %>%
  unlist() %>%
  unique()

overlap <- combo_individual[combo_individual %in% single$barcode]
cat("Combo barcodes also found as singles:", length(overlap), "/", length(combo_individual), "\n")


```

Diagnostic plots
```{r}
# loading data
# ============================================================
umap.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS024/singleCell/Seurat/UMAPcoords/"
UMAP_data <- read.csv(paste0(umap.dir, "GFP_invitroUMAP_coords.csv"), row.names = "X")
UMAP_data <- UMAP_data %>% select(cellID, UMAP1, UMAP2, cluster, size)

singlet.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS024/singletCode/"
singlet_data <- read.csv(paste0(singlet.dir, "GFP_barcode_1__singlets_all.txt"), col.names = "cellID")
singlet_data <- singlet_data %>% mutate(cellID = paste0(cellID, "-1"))

side_reactions <- c("GFP_barcode_1", "GFP_barcode_9")
base.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS024/barcode/stepThree/"
# Collect known combos from the threshold-4 run for reference
known_combos <- combo %>% pull(barcode) %>% sort() %>% unique()

threshold_results <- list()

for (thresh in 1:4) {
  recovered <- tibble()
  
  for (i in side_reactions) {
    temp <- read.csv(file.path(base.dir, i, "stepThreeStarcodeShavedReads.txt"),
                     sep = "\t")
    raw <- temp %>%
      mutate(cellID = paste0(cellID, "-1")) %>%
      select(cellID, UMI, BC50StarcodeD8) %>%
      filter(cellID %in% singlet_data$cellID,
             cellID %in% UMAP_data$cellID) %>%
      count(cellID, BC50StarcodeD8, name = "count") %>%
      filter(count > thresh)
    
    # Cells that were singles at threshold 5: they have exactly one barcode
    # in the overlap set. Check if they now gain a second barcode.
    single_cellIDs <- filtered_cells %>%
      filter(barcode %in% overlap) %>%  # singles whose barcode appears in combos
      pull(cellID)
    
    rescued <- raw %>%
      filter(cellID %in% single_cellIDs) %>%
      arrange(cellID, BC50StarcodeD8) %>%
      group_by(cellID) %>%
      filter(n() == 2) %>%  # now has exactly 2 barcodes at this threshold
      summarise(
        new_combo = paste(BC50StarcodeD8, collapse = "_"),
        min_umi   = min(count),
        .groups   = "drop"
      ) %>%
      mutate(
        matches_known_combo = new_combo %in% known_combos,
        threshold = thresh,
        sample = i
      )
    
    recovered <- bind_rows(recovered, rescued)
  }
  
  threshold_results[[thresh]] <- recovered
}

all_results <- bind_rows(threshold_results)


summary_table <- all_results %>%
  group_by(threshold) %>%
  summarise(
    cells_recovered    = n(),
    match_known_combo  = sum(matches_known_combo),
    novel_combo        = sum(!matches_known_combo),
    median_min_umi     = median(min_umi),
    .groups = "drop"
  )

#print(summary_table)

combo_recovery <- all_results %>%
  filter(matches_known_combo) %>%
  count(new_combo, name = "times_recovered", sort = TRUE)

#print(combo_recovery)

partner_check <- combo %>%
  rowwise() %>%
  mutate(bc_list = list(strsplit(barcode, "_")[[1]])) %>%
  unnest(bc_list) %>%
  group_by(bc_list) %>%
  summarise(
    n_unique_partners = n_distinct(barcode),
    .groups = "drop"
  )

barcoded_cells <- filtered_cells
barcode_counts <- barcoded_cells %>%
  count(barcode, name = "count", sort = TRUE)
single <- barcode_counts %>% filter(!grepl("_", barcode))
combo  <- barcode_counts %>% filter(grepl("_", barcode))
combo_individual <- combo %>%
  pull(barcode) %>% strsplit("_") %>% unlist() %>% unique()
overlap <- combo_individual[combo_individual %in% single$barcode]
# ============================================================

output.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS024/barcode/plots/"
# ============================================================
# PLOT 1: Threshold sweep — supports Hypothesis 1
# Shows that lowering UMI threshold recovers real combos
# ============================================================

sweep_data <- summary_table %>%
  pivot_longer(cols = c(match_known_combo, novel_combo),
               names_to = "type", values_to = "count")

p1 <- ggplot(sweep_data, aes(x = factor(threshold), y = count, fill = type)) +
  geom_col(position = "dodge") +
  geom_hline(yintercept = 0) +
  labs(
    title = "Lowering UMI threshold reclassifies singles as known combos",
    subtitle = "Newly upgraded combos also seen at UMI > 5 exceed novel combos at UMI > 2",
    x = "UMI threshold (UMI > X)",
    y = "# cells upgraded from singles to combos",
    fill = NULL
  ) +
  scale_fill_manual(
    values = c(match_known_combo = "#4D4D4D", novel_combo = "#B0B0B0"),
    labels = c("Matches UMI > 5 combo", "Novel combo")
  ) +
  theme_minimal(base_size = 13)

ggsave(paste0(output.dir, "plot1_threshold_sweep.pdf"), p1, width = 7, height = 5)

# ============================================================
# PLOT 2: Partner consistency — supports Hypothesis 1
# Most overlap barcodes have 1 consistent partner
# ============================================================

p2 <- ggplot(partner_check, aes(x = n_unique_partners)) +
  geom_histogram(binwidth = 1, fill = "#4D4D4D", color = "white",
                 boundary = 0.4) +
  geom_vline(xintercept = 5, linetype = "dashed", color = "red") +
  annotate("text", x = 6, y = max(table(partner_check$n_unique_partners)) * 0.9,
           label = "Could set\ncontamination\nthreshold\nhere", hjust = 0, color = "red", size = 3.5) +
  labs(
    title = "Partner consistency per overlap barcode",
    subtitle = "Most barcodes have 1 partner (H1); outliers have 50+ (H2)",
    x = "Number of unique combo partners",
    y = "Number of barcodes"
  ) +
  theme_minimal(base_size = 13)

ggsave(paste0(output.dir, "plot2_partner_consistency.pdf"), p2, width = 7, height = 5)

# ============================================================
# PLOT 3: Single count vs combo appearances — both hypotheses
# Correlation expected under both, but outliers indicate H2
# ============================================================

overlap_stats <- barcode_counts %>%
  filter(barcode %in% overlap) %>%
  rename(single_count = count) %>%
  left_join(
    combo %>%
      separate_rows(barcode, sep = "_") %>%
      filter(barcode %in% overlap) %>%
      count(barcode, name = "combo_appearances"),
    by = "barcode"
  ) %>%
  mutate(
    ratio = single_count / combo_appearances,
    pct_single = single_count / (single_count + combo_appearances) * 100
  ) %>%
  arrange(ratio) %>%
  left_join(partner_check, by = c("barcode" = "bc_list"))

p3 <- ggplot(overlap_stats, aes(x = single_count, y = combo_appearances)) +
  geom_point(aes(color = n_unique_partners > 3), size = 2, alpha = 0.7) +
  scale_color_manual(values = c("FALSE" = "#2166AC", "TRUE" = "#B2182B"),
                     labels = c("Few partners (H1)", "Promiscuous (H2)"),
                     name = NULL) +
  labs(
    title = "Frequency as sole barcode vs combo",
    subtitle = "Most barcodes have few combo partners (H1); 
    outliers likely contaminants (H2)",
    x = "# cells with this barcode as sole barcode",
    y = "# distinct combos containing this barcode"
  ) +
  theme_minimal(base_size = 13)

ggsave(paste0(output.dir, "plot3_single_vs_combo.pdf"), p3, width = 7, height = 5)

# ============================================================
# PLOT 4: UMI distribution of the second barcode in combos
# Supports Hypothesis 1 — weak second barcodes cluster near threshold
# ============================================================

# For each combo cell, get the UMI counts of both barcodes
combo_umi <- tibble()
for (i in side_reactions) {
  raw <- barcodeData[[i]]  # already filtered at current threshold
  
  combo_cells <- filtered_cells %>%
    filter(grepl("_", barcode))
  
  umi_data <- raw %>%
    filter(cellID %in% combo_cells$cellID) %>%
    group_by(cellID) %>%
    mutate(rank = rank(count, ties.method = "first")) %>%
    ungroup() %>%
    mutate(bc_rank = ifelse(rank == 1, "weaker", "stronger"))
  
  combo_umi <- bind_rows(combo_umi, umi_data)
}

p4 <- ggplot(combo_umi, aes(x = count, fill = bc_rank)) +
  geom_histogram(binwidth = 1, position = "identity", alpha = 0.6) +
  geom_vline(xintercept = 5, linetype = "dashed", color = "red") +
  annotate("text", x = 6, y = Inf, label = "UMI>4\nthreshold",
           vjust = 2, hjust = 0, color = "red", size = 3.5) +
  labs(
    title = "UMI counts of combo barcodes",
    subtitle = "Weaker barcodes cluster near threshold",
    x = "UMI count",
    y = "Number of barcode-cell pairs",
    fill = NULL
  ) +
  scale_fill_manual(values = c(weaker = "#4D4D4D", stronger = "#B0B0B0")) +
  theme_minimal(base_size = 13)

ggsave(paste0(output.dir, "plot4_combo_umi_distribution.pdf"), p4, width = 7, height = 5)

# ============================================================
# PLOT 5: Per-sample consistency
# Check if the same barcodes behave the same across replicates
# Supports H1 if overlap barcodes are consistent across samples
# ============================================================

sample_overlap <- tibble()
for (i in side_reactions) {
  raw <- barcodeData[[i]]
  
  singles_i <- raw %>%
    group_by(cellID) %>%
    filter(n() == 1) %>%
    ungroup() %>%
    distinct(BC50StarcodeD8) %>%
    pull(BC50StarcodeD8)
  
  combos_i <- raw %>%
    group_by(cellID) %>%
    filter(n() == 2) %>%
    ungroup() %>%
    distinct(BC50StarcodeD8) %>%
    pull(BC50StarcodeD8)
  
  overlap_i <- intersect(singles_i, combos_i)
  
  sample_overlap <- bind_rows(sample_overlap, tibble(
    sample = i,
    n_singles = length(singles_i),
    n_combo_barcodes = length(combos_i),
    n_overlap = length(overlap_i),
    pct_overlap = length(overlap_i) / length(combos_i) * 100
  ))
}

print(sample_overlap)
# If overlap percentages are similar across samples, the effect is systematic
# (threshold-driven, H1) rather than stochastic (contamination, H2)
```


### Barcode abundance plot
```{r}
#plotting barcode abundance
lib1 <- barcodeData[["GFP_barcode_1"]] %>%
  count(BC50StarcodeD8, name = "lib1_abundance")

lib2 <- barcodeData[["GFP_barcode_9"]] %>%
  count(BC50StarcodeD8, name = "lib2_abundance")

abundance <- inner_join(lib1, lib2, by = "BC50StarcodeD8") %>%
  replace_na(list(lib1_abundance = 0, lib2_abundance = 0))

cor_val <- round(cor(abundance$lib1_abundance, abundance$lib2_abundance), 3)

p <- ggplot(abundance, aes(x = lib2_abundance, y = lib1_abundance)) +
  geom_bin2d(binwidth = c(200, 200)) +
  scale_fill_gradientn(colors = c("black", "blue", "purple", "red", "yellow"),
                       name = "Number of\nBarcodes",
                       trans = "log10") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
  annotate("text", x = max(abundance$lib2_abundance) * 0.3,
           y = max(abundance$lib1_abundance) * 0.95,
           label = paste("R =", cor_val), size = 4) +
  labs(x = "Barcode abundance\nGFP_barcode_9",
       y = "Barcode abundance\nGFP_barcode_1",
       title = "Correlation between two barcode libraries") +
  coord_fixed(ratio = 1) +
  theme_classic()

ggsave(file.path(base.dir, "barcode_abundance_correlation_reads.pdf"),
       p, width = 6, height = 6)
```


### Check GFP and barcode co-expression
```{r GFP expression}
sample_name <- "GFP_invitro"
dataset <- "GFP_GEX"
folder <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS024/singleCell/Seurat/RDS/"
output.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS024/singleCell/Seurat/"
input.dir <- paste0(folder, sample_name, ".rds")

tryCatch({
  seurat_obj <- readRDS(input.dir)
  
  # Generate expression data frame without gene expression
  sample_df <- targetExpressionDF(seurat_obj, genes = "GFP")

  # Add sample and target columns to the dataframe
  sample_df$size <- colSums(GetAssayData(seurat_obj, assay = "RNA", layer = "counts"))
  sample_df$sample <- sample_name

  outUMAP.dir <- paste0(output.dir, "UMAPcoords/")
  if (!dir.exists(outUMAP.dir)) {
    dir.create(outUMAP.dir, recursive = TRUE)
  }
  write.csv(sample_df, paste0(outUMAP.dir, sample_name, "_GFP.csv"))
  
  cat("Processed sample:", sample_name, "\n")
}, error = function(e) {
  cat("Error processing sample", sample_name, ":", conditionMessage(e), "\n\n")
})
```

```{r}
input.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS024/singleCell/Seurat/UMAPcoords/"
GFP_invitro <- read.csv(paste0(input.dir, "GFP_invitro_GFP.csv"), row.names = "X")

joint <- left_join(GFP_invitro, test2, by = "cellID")
joint[is.na(joint)] <- 0
p <- ggplot(data = joint, aes(x = log1p(barcodes), y = GFP)) +
  geom_jitter(alpha = 0.6, size = 1.5, width = 0.1, height = 0.1) +
  labs(
    x = "barcodes per cell (log counts)",
    y = "GFP (log counts)",
    title = paste0("Barcode - GFP correlation")
  ) +
  theme_minimal()# +
  # coord_fixed(ratio = 1) +
  # xlim(0, 9.1) +
  # ylim(0, 9.1)
print(p)


joint <- left_join(GFP_invitro, test3, by = "cellID")
joint[is.na(joint)] <- 0
p <- ggplot(data = joint, aes(x = log1p(barcode_reads), y = GFP)) +
  geom_jitter(alpha = 0.6, size = 1.5, width = 0.1, height = 0.1) +
  labs(
    x = "barcode reads per cell (log counts)",
    y = "GFP (log counts)",
    title = paste0("Barcode - GFP correlation")
  ) +
  theme_minimal()# +
  # coord_fixed(ratio = 1) +
  # xlim(0, 9.1) +
  # ylim(0, 9.1)
print(p)


df <- joint %>%
  mutate(barcode_status = ifelse(barcode_reads > 0, "Barcoded", "Non-barcoded"))

ggplot(df, aes(x = barcode_status, y = GFP, fill = barcode_status)) +
  geom_violin(trim = FALSE) +
  geom_boxplot(width = 0.1, outlier.shape = NA) +
  labs(x = NULL, y = "GFP Expression (log1p)") +
  theme_minimal() +
  theme(legend.position = "none")
```


# Joint Dataset
## Joining

### bamsort
```{r}
# Loading
SALVE.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS024/SALVE/bamsort/reads/minimum/"
GEX.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS024/singleCell/bamsort/reads/minimum/"
barcode.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS024/barcode/"

sample_name <- "GFP_invitro"

# Read all data files
salve_file <- paste0(SALVE.dir, sample_name, "_SALVE_filtered.csv")
gex_file <- paste0(GEX.dir, sample_name, "_GEX_filtered.csv")
umap_file <- paste0(
        "/projects/b1042/GoyalLab/egrody/extractedData/EGS024/singleCell/Seurat/UMAPcoords/", 
        sample_name, "UMAP_coords.csv")
cd4_file <- paste0(
        "/projects/b1042/GoyalLab/egrody/extractedData/EGS024/singleCell/Seurat/expressionDF/", 
         sample_name, "_CD4T.csv")
barcode_file <- paste0(barcode.dir, "filtered3_barcoded_cells.csv")

salve_data <- if(file.exists(salve_file)) read.csv(salve_file) else NULL
gex_data <- if(file.exists(gex_file)) read.csv(gex_file) else NULL
umap_data <- if(file.exists(umap_file)) read.csv(umap_file, row.names = "X") else NULL
cd4_data <- if(file.exists(cd4_file)) read.csv(cd4_file, row.names = "X") else NULL
barcode_data <- if(file.exists(barcode_file)) read.csv(barcode_file, row.names = "X") else NULL

# Get cellIDs from Mmul_10 only df
all_cells <- as.character(umap_data$cellID)
all_cells <- unique(all_cells)

# Create base dataframe
result <- data.frame(cellID = all_cells)

# Add UMAP data
umap_processed <- umap_data %>%
    select(-V2, -sample)
result <- left_join(result, umap_processed, by = "cellID")

# Add barcode
result <- left_join(result, barcode_data, by = "cellID")
result$barcode[is.na(result$barcode)] <- "none"

# Add GEX data
if (!is.null(gex_data)) {
  gex_processed <- gex_data %>%
    mutate(cellID = as.character(cellID)) %>%
    select(cellID, total, US)
  result <- left_join(result, gex_processed, by = "cellID") %>%
    mutate(GEX = log1p(total)) %>%
    mutate(GEX.US = log1p(US)) %>%
    select(-total, -US)
} else {
  result$GEX <- 0
}

# Add SALVE data
if (!is.null(salve_data)) {
  salve_processed <- salve_data %>%
    #group_by(cellID) %>%
    rename(SALVE = total) %>%
    mutate(cellID = as.character(cellID)) %>%
    select(cellID, SALVE, US, SS, MS)
  result <- left_join(result, salve_processed, by = "cellID") %>%
    mutate(SALVE = log1p(SALVE)) %>%
    mutate(USprop = US / (US+MS+SS)) %>%
    mutate(SSprop = SS / (US+MS+SS)) %>%
    mutate(MSprop = MS / (US+MS+SS)) %>%
    mutate(US = log1p(US)) %>%
    mutate(SS = log1p(SS)) %>%
    mutate(MS = log1p(MS))
} else {
  result$SALVE <- 0
}

# Replace NAs with 0
result[is.na(result)] <- 0

# Cell type filtering
cd4_data <- cd4_data %>%
  # T cell identity
  mutate(CD3 = ifelse(CD3D > 0 | CD3E > 0 | CD3G > 0, 1, 0)) %>%
  mutate(CD4_transcript = ifelse(CD4 > 0, 1, 0)) %>%
  mutate(CD4_signature = ifelse(IL7R > 0 | LCK > 0 | LEF1 > 0 | TCF7 > 0, 1, 0)) %>%
  mutate(CD8 = ifelse(CD8A > 0 | CD8B > 0, 1, 0)) %>%
  # # CD4+ T helper subsets (require CD4 signature context)
  # mutate(Th1 = ifelse(CD4_signature == 1 & TBX21 > 0, 1, 0)) %>%
  # mutate(Th2 = ifelse(CD4_signature == 1 & GATA3 > 0, 1, 0)) %>%
  # mutate(Th17 = ifelse(CD4_signature == 1 & RORC > 0, 1, 0)) %>%
  # mutate(Treg = ifelse(CD4_signature == 1 & FOXP3 > 0, 1, 0)) %>%
  # mutate(activated = ifelse(IL2RA > 0, 1, 0)) %>%
  # # Contaminants -- use markers not shared with activated T cells
  # mutate(NK = ifelse(KLRD1 > 0 & CD3 == 0, 1, 0)) %>%
  # mutate(monocyte = ifelse(CD14 > 0 | CD68 > 0, 1, 0)) %>%
  # mutate(B_cell = ifelse(MS4A1 > 0 | CD79A > 0, 1, 0)) %>%
  select(cellID, CD3, CD4_transcript, CD4_signature, CD8)
result <- left_join(result, cd4_data, by = "cellID")
result <- result %>%
  # mutate(across(any_of(c("GEX", "GEX.US")), 
  #           ~ ifelse((CD4_signature > 0 | CD4_transcript > 0) & CD3 > 0, .x, 0))) %>% 
  # mutate(across(any_of(c("SALVE", "US", "MS", "SS")), 
  #           ~ ifelse((CD4_signature > 0 | CD4_transcript > 0) & CD3 > 0, .x, 0))) %>%
  filter(CD4_signature == 0 | CD8 == 0) %>% # exclude cells CD4+ and CD8+
  select(-CD8, -CD3, -CD4_transcript, -CD4_signature)

joint_data <- result

cat("Processed", sample_name, "\n")

```

Saving
```{r}
output_dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS024/joint/dataframes/"

cat("\nSample\tTotal_cells\tboth\t10X_only\tSALVE_only\n")

current_df <- joint_data

both <- current_df %>% filter(GEX != 0 & SALVE != 0)
either <- current_df %>% filter(GEX != 0 | SALVE != 0)
onlySingleCell <- current_df %>% 
  filter(GEX != 0) %>%
  filter(!(cellID %in% both$cellID))
onlySALVE <- current_df %>% 
  filter(SALVE != 0) %>%
  filter(!(cellID %in% both$cellID))

cat(sample_name, "\t",
    nrow(current_df), "\t",
    nrow(both), "\t",
    nrow(onlySingleCell), "\t",
    nrow(onlySALVE), "\n\n")
cat("Barcoded cells:\t\t", sum(joint_data$barcode != "none"), 
    "\nBarcoded infected:\t", sum(either$barcode != "none"),
    "\nTotal infected:\t\t", nrow(either))

# cat(sample_name, "\tCorrelation:", cor(current_df$GEX, current_df$SALVE), "\n",
#     "\tMax SALVE:", max(current_df$SALVE), "\n",
#     "\tMax GEX:", max(current_df$GEX), "\n\n")


# p <- ggplot(data = current_df, aes(x = SALVE, y = GEX)) +
#   geom_jitter(alpha = 0.6, size = 1.5, width = 0.1, height = 0.1) +
#   labs(
#     x = "SALVE",
#     y = "GEX",
#     title = paste0(sample_name, " SIV total Correlation")
#   ) +
#   theme_minimal() +
#   coord_fixed(ratio = 1) +
#   xlim(0, 9.1) +
#   ylim(0, 9.1)
# print(p)
# p_exclude_zero <- ggplot(data = subset(current_df, SALVE > 0 | GEX > 0), aes(x = SALVE, y = GEX)) +
#   #geom_abline(intercept = 0, slope = 1, color = "red", linewidth = 1) +
#   geom_smooth(method = "lm", se = TRUE, color = "white", alpha = 0.7) +
#   geom_bin2d(bins = 30, alpha = 0.8) +
#   labs(
#     x = "SALVE",
#     y = "GEX",
#     title = paste0(sample_name, " SIV total 2D Binning (Excluding 0,0)")
#   ) +
#   theme_minimal() +
#   coord_cartesian(xlim = c(0, 9.1), ylim = c(0, 9.1)) +
#   scale_fill_viridis_c(name = "Count") +
#   theme(aspect.ratio = 1) +
#   guides(fill = guide_colorbar(raster = TRUE))
# print(p_exclude_zero)
# output.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/joint/plots/"
# ggsave(paste0(output.dir, sample_name, "_CD4_correlation_jitter.svg"),
#        plot = p)
# ggsave(paste0(output.dir, sample_name, "_CD4_correlation_2Dbinning.svg"),
#        plot = p_exclude_zero)


output.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/joint/plots/"
plotUMAP(current_df, colorby = GEX,
       paste0(sample_name, " joint SIV total: GEX (bamsort)"),
       output.dir,
       paste0(sample_name, "_CD4_GEX_bamsort.pdf"),
       comparison = FALSE, color = "red")
plotUMAP(current_df, colorby = SALVE,
       paste0(sample_name, " joint SIV total: SALVE (bamsort)"),
       output.dir,
       paste0(sample_name, "_CD4_SALVE_bamsort.pdf"),
       comparison = FALSE, color = "red")



#Save CSVs
write.csv(both, paste0(output_dir, sample_name, "_both.csv"), row.names = FALSE)
write.csv(onlySingleCell, paste0(output_dir, sample_name, "_onlySingleCell.csv"), row.names = FALSE)
write.csv(onlySALVE, paste0(output_dir, sample_name, "_onlySALVE.csv"), row.names = FALSE)
write.csv(current_df, paste0(output_dir, sample_name, "_all.csv"), row.names = FALSE)


```

### Seurat and bamsort
First getting counts from GEX Seurat
```{r}
folder <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS024/singleCell/counts/concat/"
input.dir <- paste0(folder, "Mmul_10_mac239_GFP_GFP_GEX/outs/filtered_feature_bc_matrix/")

seurat_obj <- SeuratPipeline(input.dir, "GFP_invitro", 
                   output_dir = output.dir, 
                   plots = FALSE, rds = FALSE, remove_mac = FALSE)
sample_df <- targetExpressionDF(seurat_obj, genes = "mac239")

```

```{r}
# Loading
SALVE.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS024/SALVE/bamsort/reads/minimum/"
sample_name <- "GFP_invitro"

# Read all data files
salve_file <- paste0(SALVE.dir, sample_name, "_SALVE_filtered.csv")
cd4_file <- paste0(
        "/projects/b1042/GoyalLab/egrody/extractedData/EGS024/singleCell/Seurat/expressionDF/", 
         sample_name, "_CD4T.csv")

salve_data <- if(file.exists(salve_file)) read.csv(salve_file) else NULL
cd4_data <- if(file.exists(cd4_file)) read.csv(cd4_file, row.names = "X") else NULL

# Get cellIDs from Mmul_10 only df
all_cells <- as.character(umap_data$cellID)
all_cells <- unique(all_cells)

# Create base dataframe
result <- data.frame(cellID = all_cells)

# Add UMAP data
if (!is.null(umap_data)) {
  umap_processed <- umap_data %>%
    select(-V2, -sample)
  result <- left_join(result, umap_processed, by = "cellID")
}

# Add GEX data
gex_processed <- sample_df %>%
  mutate(cellID = as.character(cellID)) %>%
  select(cellID, mac239)
result <- left_join(result, gex_processed, by = "cellID") %>%
  mutate(GEX = log1p(mac239)) %>%
  select(-mac239)

# Add SALVE data
if (!is.null(salve_data)) {
  salve_processed <- salve_data %>%
    #group_by(cellID) %>%
    rename(SALVE = total) %>%
    mutate(cellID = as.character(cellID)) %>%
    select(cellID, SALVE, US, SS, MS)
  result <- left_join(result, salve_processed, by = "cellID") %>%
    mutate(SALVE = log1p(SALVE)) %>%
    mutate(USprop = US / (US+MS+SS)) %>%
    mutate(SSprop = SS / (US+MS+SS)) %>%
    mutate(MSprop = MS / (US+MS+SS)) %>%
    mutate(US = log1p(US)) %>%
    mutate(SS = log1p(SS)) %>%
    mutate(MS = log1p(MS))
} else {
  result$SALVE <- 0
}

# Replace NAs with 0
result[is.na(result)] <- 0

# Cell type filtering
cd4_data <- cd4_data %>%
  # T cell identity
  mutate(CD3 = ifelse(CD3D > 0 | CD3E > 0 | CD3G > 0, 1, 0)) %>%
  mutate(CD4_transcript = ifelse(CD4 > 0, 1, 0)) %>%
  mutate(CD4_signature = ifelse(IL7R > 0 | LCK > 0 | LEF1 > 0 | TCF7 > 0, 1, 0)) %>%
  mutate(CD8 = ifelse(CD8A > 0 | CD8B > 0, 1, 0)) %>%
  # # CD4+ T helper subsets (require CD4 signature context)
  # mutate(Th1 = ifelse(CD4_signature == 1 & TBX21 > 0, 1, 0)) %>%
  # mutate(Th2 = ifelse(CD4_signature == 1 & GATA3 > 0, 1, 0)) %>%
  # mutate(Th17 = ifelse(CD4_signature == 1 & RORC > 0, 1, 0)) %>%
  # mutate(Treg = ifelse(CD4_signature == 1 & FOXP3 > 0, 1, 0)) %>%
  # mutate(activated = ifelse(IL2RA > 0, 1, 0)) %>%
  # # Contaminants -- use markers not shared with activated T cells
  # mutate(NK = ifelse(KLRD1 > 0 & CD3 == 0, 1, 0)) %>%
  # mutate(monocyte = ifelse(CD14 > 0 | CD68 > 0, 1, 0)) %>%
  # mutate(B_cell = ifelse(MS4A1 > 0 | CD79A > 0, 1, 0)) %>%
  select(cellID, CD3, CD4_transcript, CD4_signature, CD8)
result <- left_join(result, cd4_data, by = "cellID")
result <- result %>%
  mutate(across(any_of(c("GEX", "GEX.US")),
            ~ ifelse((CD4_signature > 0 | CD4_transcript > 0) & CD3 > 0, .x, 0))) %>%
  mutate(across(any_of(c("SALVE", "US", "MS", "SS")),
            ~ ifelse((CD4_signature > 0 | CD4_transcript > 0) & CD3 > 0, .x, 0))) %>%
  filter(CD4_signature == 0 | CD8 == 0) %>% # exclude cells CD4+ and CD8+
  select(-CD8, -CD3, -CD4_transcript, -CD4_signature)

joint_data <- result
cat("Processed", sample_name, "\n")

```

### combined counts
```{r}
# Loading
SALVE.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS024/SALVE/bamsort/reads/"
GEX.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS024/singleCell/bamsort/reads/"
barcode.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS024/barcode/"
Seurat.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS024/singleCell/Seurat/"
sample_name <- "GFP_invitro"

combined_data <- list()

# Read all data files
salve_file <- paste0(SALVE.dir, sample_name, "_UMI_read_counts_full.csv")
gex_file <- paste0(GEX.dir, sample_name, "_UMI_read_counts_full.csv")
umap_file <- paste0(Seurat.dir, "UMAPcoords/",
  sample_name, "UMAP_coords.csv")
cd4_file <- paste0(Seurat.dir, "expressionDF/",
  sample_name, "_CD4T.csv")
barcode_file <- paste0(barcode.dir, "filtered3_barcoded_cells.csv")
salve_umi_file <- paste0(SALVE.dir, "minimum/valid_UMI/", sample_name, "_SALVE_valid_cellID_UMI.csv")
gex_umi_file <- paste0(GEX.dir, "minimum/valid_UMI/", sample_name, "_GEX_valid_cellID_UMI.csv")

salve_data <- if(file.exists(salve_file)) read.csv(salve_file) else NULL
gex_data <- if(file.exists(gex_file)) read.csv(gex_file) else NULL
salve_umi <- if(file.exists(salve_umi_file)) read.csv(salve_umi_file) else NULL
gex_umi <- if(file.exists(gex_umi_file)) read.csv(gex_umi_file) else NULL
umap_data <- if(file.exists(umap_file)) read.csv(umap_file, row.names = "X") else NULL
cd4_data <- if(file.exists(cd4_file)) read.csv(cd4_file, row.names = "X") else NULL
barcode_data <- if(file.exists(barcode_file)) read.csv(barcode_file, row.names = "X") else NULL

full_preprocess <- NULL

# Join GEX and SALVE and filter by minimums-applied UMI
if (!is.null(gex_data) && !is.null(gex_umi)) {
  full_preprocess <- gex_data %>%
    mutate(cellID = as.character(cellID)) %>%
    semi_join(gex_umi, by = c("cellID", "UMI"))
}
if (!is.null(salve_data) && !is.null(salve_umi)) {
  salve_processed <- salve_data %>%
    mutate(cellID = as.character(cellID)) %>%
    semi_join(salve_umi, by = c("cellID", "UMI"))
  full_preprocess <- full_join(full_preprocess, salve_processed, by = c("cellID", "UMI"))
}

full_processed <- resolve_multimap_gs(full_preprocess)
full_processed <- full_processed %>%
  count(cellID, category) %>%
  pivot_wider(names_from = category, values_from = n, values_fill = 0) %>%
  mutate(total = log1p(rowSums(select(., -cellID), na.rm = TRUE)))

# Get cellIDs from Mmul_10 only df
all_cells <- as.character(umap_data$cellID)
all_cells <- unique(all_cells)

# Create base dataframe
result <- data.frame(cellID = all_cells)

# Add UMAP data
if (!is.null(umap_data)) {
  umap_processed <- umap_data %>%
    select(-V2, -sample)
  result <- left_join(result, umap_processed, by = "cellID")
}

# Add barcode data
if (!is.null(barcode_data)) {
  result <- left_join(result, barcode_data, by = "cellID")
  result$barcode[is.na(result$barcode)] <- "none"
}

# Add combined GEX + SALVE data
result <- left_join(result, full_processed, by = "cellID")
result[is.na(result)] <- 0

# Cell type filtering
cd4_data <- cd4_data %>%
  mutate(CD3 = ifelse(CD3D > 0 | CD3E > 0 | CD3G > 0, 1, 0)) %>%
  mutate(CD4_transcript = ifelse(CD4 > 0, 1, 0)) %>%
  mutate(CD4_signature = ifelse(IL7R > 0 | LCK > 0 | LEF1 > 0 | TCF7 > 0, 1, 0)) %>%
  mutate(CD8 = ifelse(CD8A > 0 | CD8B > 0, 1, 0)) %>%
  select(cellID, CD3, CD4_transcript, CD4_signature, CD8)
result <- left_join(result, cd4_data, by = "cellID")
result <- result %>%
  filter(CD4_signature == 0 | CD8 == 0) %>%
  select(-CD8, -CD3, -CD4_transcript, -CD4_signature)

combined_data <- result
cat("Processed", sample_name, "\n")
```

```{r}
output_dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS024/combined/dataframes/"
dir.create(output_dir, recursive = TRUE)
cat("\nSample\tTotal_cells\tcombined\n")

current_df <- combined_data
both <- current_df %>% filter(total != 0)

cat(sample_name, "\t",
    nrow(current_df), "\t",
    nrow(both), "\n")

# cat(sample_name, "\tCorrelation:", cor(current_df$GEX, current_df$SALVE), "\n",
#     "\tMax SALVE:", max(current_df$SALVE), "\n",
#     "\tMax GEX:", max(current_df$GEX), "\n\n")

output.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS024/combined/plots/"
plotUMAP(combined_data, colorby = total,
       paste0(sample_name, " - SIV combined"),
       output.dir,
       paste0("UMAP_SIV.pdf"),
       comparison = FALSE, color = "red")


#Save CSVs
write.csv(both, paste0(output_dir, sample_name, "_both.csv"), row.names = FALSE)
write.csv(current_df, paste0(output_dir, sample_name, "_all.csv"), row.names = FALSE)
```



## Splicing, not updated
###Splicing proportions
```{r proportions}
output_dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/joint/plots/proportions/"
for (sample_name in names(joint_data)) {
  df <- joint_data[[sample_name]]
  
  df_sorted <- df %>%
    filter(SALVE > 0) %>%
    arrange(desc(USprop), desc(SSprop)) %>%
    mutate(cellID_ordered = factor(cellID, levels = cellID))
  df_long <- df_sorted %>%
    pivot_longer(cols = contains("prop"), 
                 names_to = "component", 
                 values_to = "proportion")
  
  p1 <- ggplot(df_long, aes(x = cellID_ordered, y = proportion, fill = component)) +
    geom_bar(stat = "identity", width = 1) +
    scale_y_continuous(expand = c(0, 0)) +
    labs(title = paste0(sample_name, " Isoforms: Stacked Bar Chart"),
         x = "Cell ID",
         y = "Proportion",
         fill = "Component") +
    theme_minimal() +
    theme(axis.text.x = element_blank(),  # Remove x-axis labels for clarity
          axis.ticks.x = element_blank(),
          panel.grid.major.x = element_blank())
  
  ggsave(paste0(output_dir, sample_name, "_proportions_stacked.pdf"), p1, width = 10, height = 6, dpi = 300)
  
  
  
  p4 <- ggplot(df_long, aes(x = component, y = proportion, fill = component)) +
    geom_violin(alpha = 0.6) +
    geom_boxplot(width = 0.3, alpha = 0.8, outlier.alpha = 0.5) +
    labs(title = paste0(sample_name, " Isoforms: Violins"),
         x = "Isoform Proportion",
         y = "Proportion",
         fill = "Component") +
    theme_minimal() +
    theme(legend.position = "none")
  
  ggsave(paste0(output_dir, sample_name, "_proportions_violin.pdf"), p4, width = 8, height = 6, dpi = 300)
}

```

###Splicing UMAPS w expression
```{r}
## for lightning talk figures
  output.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS016/joint/plots/v5/subset/"
  plotUMAP(current_df, colorby = US,
         paste0(sample_name, " joint SALVE US"),
         output.dir,
         paste0(sample_name, "_umap_SALVE_US.svg"),
         comparison = FALSE, color = "royalblue1")
  plotUMAP(current_df, colorby = spliced,
         paste0(sample_name, " joint SALVE spliced"),
         output.dir,
         paste0(sample_name, "_umap_SALVE_spliced.svg"),
         comparison = FALSE, color = "red")
  plotUMAP(current_df, colorby = SALVE,
         paste0(sample_name, " joint SALVE"),
         output.dir,
         paste0(sample_name, "_umap_SALVE.svg"),
         comparison = FALSE, color = "purple4")
    plotUMAP(current_df, colorby = GEX,
         paste0(sample_name, " joint GEX"),
         output.dir,
         paste0(sample_name, "_umap_GEX.svg"),
         comparison = FALSE, color = "purple4")
```

###Splicing UMAPS w/o expression (identity only)
```{r}
output.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/joint/plots/identity/"
col1s <- c("infected", "splicers")
col2s <- c("uninfected", "nonsplicers")

for (i in 1:length(samples)) {
  sample_name <- samples[i]
  df <- joint_data[[sample_name]]
  
  df <- df %>%
    mutate(infected = ifelse(SALVE > 0, 1, 0), uninfected = ifelse(SALVE == 0, 1, 0),
           splicers = ifelse(SS > 0 | MS > 0, 1, 0)) %>%
    mutate(nonsplicers = ifelse(infected == 1 & splicers == 0, 1, 0))
  
  cat(sample_name, "\nTotal:\t", nrow(df),
        "\nInfected:\t", sum(df$infected),
        "\nUninfected:\t", sum(df$uninfected),
        "\nSplice+:\t", sum(df$splicers),
        "\nSplice-:\t", sum(df$nonsplicers),
        "\n\n")
  
  for (j in 1:length(col1s)) {
    col1 <- col1s[j]
    col2 <- col2s[j]
    
    create_identity_umap(df, col1, col2,
              paste0("Identity: ", col1, " vs ", col2),
              paste0(sample_name, "_identity_", col1, "_", col2, ".pdf"),
              output_dir = output.dir)
  }
}

```


## Barcode mapping

### UMAPs
```{r}
sample_name <- "GFP_invitro"
combined.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS024/combined/"
combined_data <- read.csv(paste0(combined.dir, "dataframes/", sample_name, "_all.csv"))

plotBarcodeUMAP(combined_data, "GFP_invitro clones", min_count = 2, 
                output_dir = paste0(combined.dir, "plots/"),
                saveas = paste0("UMAP_barcode_clones.pdf"))
plotBarcodeUMAP(combined_data, "GFP_invitro singletons", max_count = 1, 
                output_dir = paste0(combined.dir, "plots/"),
                saveas = paste0("UMAP_barcode_singletons.pdf"))
plotBarcodeUMAP(combined_data, "GFP_invitro clones sized > 50", min_count = 50, 
                output_dir = paste0(combined.dir, "plots/"),
                saveas = paste0("UMAP_barcode_clonesGreaterThan50.pdf"))


# identity UMAP: barcoded and infected
plot_data <- combined_data
plot_data$highlight <- ifelse(plot_data$barcode != "none" & plot_data$total > 0, "yes", "no")
plot_data <- arrange(plot_data, highlight)

p <- ggplot(plot_data, aes(x = UMAP1, y = UMAP2, color = highlight)) +
  geom_point(aes(size = highlight)) +
  scale_color_manual(values = c("no" = "gray93", "yes" = "black")) +
  scale_size_manual(values = c("no" = 0.5, "yes" = 0.8)) +
  theme_classic() +
  theme(aspect.ratio = 1) +
  guides(size = "none", color = "none") +
  ggtitle("Barcoded cells with nonzero total expression")

print(p)
ggsave(p, file = paste0(combined.dir, "plots/UMAP_barcoded_infected.pdf"),
       device = "pdf")

# clone fate: infected proportion of clone
plot_data <- combined_data
plot_data <- plot_data %>%
  filter(barcode != "none") %>%
  mutate(infected = ifelse(total > 0, "infected", "uninfected"))
clone_counts <- plot_data %>%
  group_by(barcode) %>%
  summarise(
    clone_size = n(),
    n_infected = sum(infected == "infected"),
    n_uninfected = sum(infected == "uninfected"),
    pct_infected = round(100 * n_infected / clone_size, 1),
    .groups = "drop"
  ) %>%
  arrange(desc(n_infected), desc(clone_size))

cat("Total infected cells:", sum(clone_counts$n_infected), "\n")
cat("Total clones with >= 1 infected cell:",
    sum(clone_counts$n_infected > 0), "of", nrow(clone_counts), "\n\n")

cat("Per-clone breakdown:\n")
for (i in seq_len(nrow(clone_counts))) {
  row <- clone_counts[i, ]
  cat(sprintf("  %s | infected: %d | uninfected: %d | %s%%\n",
              row$barcode, row$n_infected, row$n_uninfected, row$pct_infected))
}

# Plot: only clones with >= 1 infected cell
infected_clones <- clone_counts %>% filter(n_infected > 0)

if (nrow(infected_clones) > 0) {
  infected_clones$label <- substr(infected_clones$barcode, 1, 10)
  infected_clones$label <- factor(infected_clones$label,
                                   levels = rev(infected_clones$label))

  infected_long <- infected_clones %>%
    tidyr::pivot_longer(cols = c(n_infected, n_uninfected),
                        names_to = "status", values_to = "count") %>%
    mutate(status = ifelse(status == "n_infected", "infected", "uninfected"))

  p <- ggplot(infected_long, aes(x = label, y = count, fill = status)) +
    geom_col() +
    scale_fill_manual(values = c("infected" = "black", "uninfected" = "gray80")) +
    coord_flip() +
    theme_classic() +
    labs(x = "", y = "Number of cells in clone", fill = "",
         title = "Clones with infected cells")

  print(p)
}
```

### Clone similarity
```{r}
# Barcode Similarity Analysis
#
# Metrics:
#   1. E-distance: energy statistic between clone and random sample (Szekely & Rizzo, 2004)
#   2. KNN mixing coefficient: fraction of k nearest neighbors sharing same barcode
#   3. Pairwise E-distance between barcodes
#   4. Phenotypic volume: log-determinant of covariance matrix in embedding space
#   5. Shannon equitability index (Pielou's J): evenness of clone across clusters
#
# Each metric is compared against a null distribution built from
# size-matched random cell samples to assess whether clonemates are
# more similar in UMAP/PC space than expected by chance.

library(FNN)      # get.knn(): fast exact k-nearest neighbor search

# ---------------------------------------------------------------------------
# 0. Setup: load Seurat object and barcode data
# ---------------------------------------------------------------------------
sample_name <- "GFP_invitro"
rds.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS024/singleCell/Seurat/RDS/"
rds_file <- paste0(rds.dir, sample_name, ".rds")
barcode.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS024/barcode/"
barcode_file <- paste0(barcode.dir, "filtered3_barcoded_cells.csv")

seurat_obj <- readRDS(rds_file)
barcode_data <- read.csv(barcode_file, row.names = "X")

output.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS024/barcode/similarity/"

# Build data frame: UMAP coords, PCA embeddings, cluster, barcode
umap_coords <- as.data.frame(Embeddings(seurat_obj, "umap"))
colnames(umap_coords) <- c("UMAP1", "UMAP2")
umap_coords$cellID <- rownames(umap_coords)

pca_coords <- as.data.frame(Embeddings(seurat_obj, "pca"))
pca_coords$cellID <- rownames(pca_coords)

meta <- seurat_obj@meta.data
meta$cellID <- rownames(meta)
# Use seurat_clusters if available; otherwise first cluster column found
cluster_col <- if ("seurat_clusters" %in% colnames(meta)) {
  "seurat_clusters"
} else {
  grep("cluster", colnames(meta), ignore.case = TRUE, value = TRUE)[1]
}

data <- umap_coords %>%
  left_join(pca_coords, by = "cellID") %>%
  left_join(meta %>% select(cellID, all_of(cluster_col)), by = "cellID") %>%
  left_join(barcode_data, by = "cellID")
data$barcode[is.na(data$barcode)] <- "none"
colnames(data)[colnames(data) == cluster_col] <- "cluster"

cat("Loaded", nrow(data), "cells,",
    sum(data$barcode != "none"), "barcoded\n")

# Coordinate columns for distance-based metrics.
# PCA embeddings preserve true distances better than UMAP, which applies
# nonlinear warping that distorts inter-point distances. Using the first
# 20 PCs captures the majority of variance while keeping computation fast.
pc_cols <- grep("^PC_", colnames(data), value = TRUE)
pc_cols <- pc_cols[1:min(20, length(pc_cols))]
# UMAP columns used only for plotting
umap_cols <- c("UMAP1", "UMAP2")
coord_cols <- pc_cols

# Filter to barcoded cells (clones) for analysis
barcode_counts <- table(data$barcode)
clone_barcodes <- names(barcode_counts[barcode_counts >= 2 & names(barcode_counts) != "none"])

n_permutations <- 100   # number of random samples for null distributions
k_neighbors    <- 20    # k for KNN mixing coefficient

set.seed(42)

# ---------------------------------------------------------------------------
# 1. E-distance: clone vs size-matched random sample
# ---------------------------------------------------------------------------
# For each clone, compute the E-distance between its cells and a random
# sample of equal size drawn from all cells. Repeat n_permutations times
# for the random-vs-random null. A clone whose E-distance exceeds the null
# occupies a distinct region of the embedding.
#
# E-distance formula (Szekely & Rizzo, 2005):
#   e(C_i, C_j) = (n_i * n_j)/(n_i + n_j) * [2*M_ij - M_ii - M_jj]
# where M_ij = mean(||X_ip - X_jq||) over all cross-pairs.

all_coords <- as.matrix(data[, coord_cols])

compute_edist <- function(coords_a, coords_b) {
  # Manual E-distance (Szekely & Rizzo, 2005) without the energy package.
  # e(A,B) = (n_a * n_b)/(n_a + n_b) * [2*M_ab - M_aa - M_bb]
  # where M_ab = mean of all pairwise distances between A and B.
  # Uses dist() for within-sample and a cross-distance matrix for between.
  n_a <- nrow(coords_a)
  n_b <- nrow(coords_b)
  M_ab <- mean(as.matrix(dist(rbind(coords_a, coords_b)))[1:n_a, (n_a + 1):(n_a + n_b)])
  M_aa <- if (n_a > 1) mean(dist(coords_a)) else 0
  M_bb <- if (n_b > 1) mean(dist(coords_b)) else 0
  (n_a * n_b) / (n_a + n_b) * (2 * M_ab - M_aa - M_bb)
}

edist_results <- data.frame(
  barcode = character(),
  clone_size = integer(),
  edist_observed = numeric(),
  edist_null_mean = numeric(),
  edist_null_sd = numeric(),
  edist_z = numeric(),
  stringsAsFactors = FALSE
)

cat("Computing E-distances for", length(clone_barcodes), "clones\n")

for (bc in clone_barcodes) {
  clone_idx <- which(data$barcode == bc)
  clone_coords <- all_coords[clone_idx, , drop = FALSE]
  n_clone <- nrow(clone_coords)

  # Observed: clone vs random sample of same size
  rand_idx <- sample(nrow(all_coords), n_clone)
  obs_edist <- compute_edist(clone_coords, all_coords[rand_idx, , drop = FALSE])

  # Null: random-vs-random of same size
  null_edists <- replicate(n_permutations, {
    r1 <- sample(nrow(all_coords), n_clone)
    r2 <- sample(nrow(all_coords), n_clone)
    compute_edist(all_coords[r1, , drop = FALSE], all_coords[r2, , drop = FALSE])
  })

  null_mean <- mean(null_edists)
  null_sd <- sd(null_edists)
  z_score <- if (null_sd > 0) (obs_edist - null_mean) / null_sd else NA

  edist_results <- rbind(edist_results, data.frame(
    barcode = bc,
    clone_size = n_clone,
    edist_observed = obs_edist,
    edist_null_mean = null_mean,
    edist_null_sd = null_sd,
    edist_z = z_score,
    stringsAsFactors = FALSE
  ))
}

cat("E-distance computation complete\n")

# Plot: observed E-distance vs null distribution z-scores
p_edist <- ggplot(edist_results, aes(x = clone_size, y = edist_z)) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  theme_classic() +
  theme(aspect.ratio = 1) +
  labs(
    x = "Clone size (number of cells)",
    y = "E-distance z-score (vs random null)",
    title = "Clone spatial coherence (E-distance)"
  )
print(p_edist)
ggsave(p_edist, file = paste0(output.dir, "edist_dotplot_clonesize_vs_zscore.pdf"),
         device = "pdf")

# Plot: distribution of observed vs null E-distances
p_edist_hist <- ggplot() +
  geom_density(data = edist_results, aes(x = edist_observed, fill = "Observed clones"),
               alpha = 0.5) +
  geom_density(data = edist_results, aes(x = edist_null_mean, fill = "Random null (mean)"),
               alpha = 0.5) +
  scale_fill_manual(values = c("Observed clones" = "steelblue", "Random null (mean)" = "gray70")) +
  theme_classic() +
  labs(x = "E-distance", y = "Density", fill = "", title = "E-distance: observed vs null")
print(p_edist_hist)
ggsave(p_edist_hist, file = paste0(output.dir, "edist_density_clones_vs_random.pdf"),
         device = "pdf")

# ---------------------------------------------------------------------------
# 2. Pairwise Mixing Coefficient
# ---------------------------------------------------------------------------
# For each pair of clones of similar size, compute:
#   mixing coefficient = (non-self neighbors) / (self neighbors)
# where neighbors are found in 50-dimensional PC space. For each cell in
# clone A, its neighbors are classified as "self" (from clone A) or
# "non-self" (from clone B), and vice versa. The neighbor count k is
# min(10, size of smaller clone). A mixing coefficient of 1 means perfect
# mixing (equal self and non-self neighbors); 0 means no mixing (clones
# occupy entirely separate regions).
#
# To keep comparisons fair, only clones within a similar size range are
# compared. Within each pair, neighbor counts are normalized by clone size.

cat("Computing pairwise mixing coefficients\n")

# Use first 50 PCs as specified in the methods
pc_cols_50 <- grep("^PC_", colnames(data), value = TRUE)
pc_cols_50 <- pc_cols_50[1:min(50, length(pc_cols_50))]

# Size bins for pairing clones of similar size (fold-change <= 2)
clone_sizes <- data.frame(
  barcode = clone_barcodes,
  size = as.integer(barcode_counts[clone_barcodes]),
  stringsAsFactors = FALSE
)

compute_mixing_coeff <- function(coords_a, coords_b) {
  # Pool both clones
  n_a <- nrow(coords_a)
  n_b <- nrow(coords_b)
  pooled <- rbind(coords_a, coords_b)
  labels <- c(rep("A", n_a), rep("B", n_b))

  # k = min(10, smaller clone size), as per Goyal et al. methods
  k <- min(10, min(n_a, n_b))
  if (k < 1) return(NA)

  knn <- FNN::get.knn(pooled, k = k)
  nn_idx <- knn$nn.index

  total_self <- 0
  total_nonself <- 0

  for (i in seq_len(nrow(pooled))) {
    neighbor_labels <- labels[nn_idx[i, ]]
    self_count <- sum(neighbor_labels == labels[i])
    nonself_count <- sum(neighbor_labels != labels[i])

    # Normalize by clone size to account for size differences:
    # weight non-self by (own_size / other_size) so that a larger clone
    # does not dominate the neighbor counts
    if (labels[i] == "A") {
      nonself_count <- nonself_count * (n_a / n_b)
    } else {
      nonself_count <- nonself_count * (n_b / n_a)
    }

    total_self <- total_self + self_count
    total_nonself <- total_nonself + nonself_count
  }

  if (total_self == 0) return(NA)
  return(total_nonself / total_self)
}

# Generate all pairs within similar size (ratio <= 2)
mixing_pairs <- data.frame(
  barcode_a = character(), barcode_b = character(),
  size_a = integer(), size_b = integer(),
  mixing_coeff = numeric(), stringsAsFactors = FALSE
)

for (i in 1:(nrow(clone_sizes) - 1)) {
  for (j in (i + 1):nrow(clone_sizes)) {
    sa <- clone_sizes$size[i]
    sb <- clone_sizes$size[j]
    ratio <- max(sa, sb) / max(min(sa, sb), 1)
    if (ratio > 2) next  # skip pairs with >2-fold size difference

    ca <- as.matrix(data[data$barcode == clone_sizes$barcode[i], pc_cols_50])
    cb <- as.matrix(data[data$barcode == clone_sizes$barcode[j], pc_cols_50])
    mc <- compute_mixing_coeff(ca, cb)

    mixing_pairs <- rbind(mixing_pairs, data.frame(
      barcode_a = clone_sizes$barcode[i],
      barcode_b = clone_sizes$barcode[j],
      size_a = sa, size_b = sb,
      mixing_coeff = mc, stringsAsFactors = FALSE
    ))
  }
}

# Null: for each pair, shuffle cells to create two random "clones"
# of the same sizes and compute mixing
null_mixing_pairs <- replicate(min(n_permutations, 50), {
  mc_null <- sapply(seq_len(nrow(mixing_pairs)), function(p) {
    sa <- mixing_pairs$size_a[p]
    sb <- mixing_pairs$size_b[p]
    idx <- sample(nrow(all_coords), sa + sb)
    ra <- as.matrix(data[idx[1:sa], pc_cols_50])
    rb <- as.matrix(data[idx[(sa + 1):(sa + sb)], pc_cols_50])
    compute_mixing_coeff(ra, rb)
  })
  mean(mc_null, na.rm = TRUE)
})
null_mix_mean <- mean(null_mixing_pairs, na.rm = TRUE)

cat("Pairwise mixing coefficient computation complete:",
    nrow(mixing_pairs), "pairs\n")

# Plot: distribution of pairwise mixing coefficients
p_mix <- ggplot(mixing_pairs, aes(x = mixing_coeff)) +
  geom_histogram(fill = "steelblue", alpha = 0.6, bins = 40) +
  geom_vline(xintercept = null_mix_mean, linetype = "dashed", color = "red") +
  geom_vline(xintercept = 1, linetype = "dotted", color = "gray40") +
  annotate("text", x = 1.02, y = Inf, vjust = 2,
           label = "perfect mixing", color = "gray40", size = 3, hjust = 0) +
  annotate("text", x = null_mix_mean + 0.02, y = Inf, vjust = 4,
           label = "random null", color = "red", size = 3, hjust = 0) +
  theme_classic() +
  labs(
    x = "Mixing coefficient (non-self / self neighbors)",
    y = "Count (clone pairs)",
    title = "Pairwise Mixing Coefficient"
  )
print(p_mix)
ggsave(p_mix, file = paste0(output.dir, "mixingcoef_histogram_pairwise.pdf"),
       device = "pdf")

# Plot: mixing coefficient vs mean clone pair size
mixing_pairs$mean_size <- (mixing_pairs$size_a + mixing_pairs$size_b) / 2
p_mix_size <- ggplot(mixing_pairs, aes(x = mean_size, y = mixing_coeff)) +
  geom_point(size = 1.5, alpha = 0.6) +
  geom_hline(yintercept = null_mix_mean, linetype = "dashed", color = "red") +
  theme_classic() +
  theme(aspect.ratio = 1) +
  labs(
    x = "Mean clone pair size",
    y = "Mixing coefficient",
    title = "Pairwise Mixing vs Clone Size"
  )
print(p_mix_size)
ggsave(p_mix_size, file = paste0(output.dir, "mixingcoef_dotplot_pairwise.pdf"),
       device = "pdf")

# ---------------------------------------------------------------------------
# 3. Pairwise E-distance Between Clones
# ---------------------------------------------------------------------------
# Compute E-distance between every pair of clones. Clones that are more
# distant from each other in the embedding adopt more divergent fates.
# Only computed for clones with >= min_cells cells to keep runtime feasible.

min_cells_pairwise <- 10
large_clones <- names(barcode_counts[barcode_counts >= min_cells_pairwise &
                                       names(barcode_counts) != "none"])
n_large <- length(large_clones)

if (n_large >= 2) {
  cat("Computing pairwise E-distances for", n_large, "clones (>=", min_cells_pairwise, "cells)\n")

  pairwise_edist <- matrix(0, nrow = n_large, ncol = n_large,
                           dimnames = list(large_clones, large_clones))

  for (i in 1:(n_large - 1)) {
    ci <- all_coords[data$barcode == large_clones[i], , drop = FALSE]
    for (j in (i + 1):n_large) {
      cj <- all_coords[data$barcode == large_clones[j], , drop = FALSE]
      pairwise_edist[i, j] <- compute_edist(ci, cj)
      pairwise_edist[j, i] <- pairwise_edist[i, j]
    }
  }

  # Hierarchical clustering on pairwise E-distances
  hc <- hclust(as.dist(pairwise_edist), method = "ward.D2")

  # Truncate barcode labels to first 8 characters for readability.
  # Full barcode sequences are too long to display without overlap.
  short_labels <- substr(large_clones, 1, 8)

  if (n_large <= 40) {
    # Few enough clones to show truncated labels
    p_dend <- plot(hc, labels = short_labels, hang = -1,
         main = "Hierarchical clustering of clones by E-distance",
         xlab = "", sub = "", cex = 0.7)
    print(p_dend)
    ggsave(p_dend, file = paste0(output.dir, "edist_dendrogram.pdf"),
           device = "pdf")
  } else {
    # Too many clones: drop labels entirely, show structure only
    p_dend <- plot(hc, labels = FALSE, hang = -1,
         main = paste0("Hierarchical clustering of clones by E-distance (n=", n_large, ")"),
         xlab = "", sub = "")
    print(p_dend)
    ggsave(p_dend, file = paste0(output.dir, "edist_dendrogram.pdf"),
           device = "pdf")
  }

  cat("Pairwise E-distance computation complete\n")
} else {
  cat("Fewer than 2 clones with >=", min_cells_pairwise, "cells; skipping pairwise analysis\n")
}
# ---------------------------------------------------------------------------
# 4. Phenotypic Volume
# ---------------------------------------------------------------------------
# Phenotypic volume quantifies how much of the embedding space a clone
# occupies. Computed as the log-determinant of the covariance matrix of
# a clone's coordinates. The log-determinant of
# a p-dimensional covariance matrix equals the log of the generalized
# variance, which is proportional to the squared hypervolume of the
# point cloud's principal ellipsoid. Using log avoids numerical overflow
# and makes the metric additive across dimensions.
#
# Compared to a convex hull, the covariance determinant is more robust
# to outliers and does not require the geometry package, which can fail
# in degenerate configurations (e.g., collinear points).
#
# Null: random samples of the same size drawn from all cells.

library(geometry)  # convhulln() for convex hull volume as alternative

cat("Computing phenotypic volume for", length(clone_barcodes), "clones\n")

phenovol_results <- data.frame(
  barcode = character(),
  clone_size = integer(),
  log_det_cov = numeric(),
  hull_volume = numeric(),
  null_logdet_mean = numeric(),
  null_logdet_sd = numeric(),
  logdet_z = numeric(),
  stringsAsFactors = FALSE
)

for (bc in clone_barcodes) {
  clone_idx <- which(data$barcode == bc)
  clone_coords <- all_coords[clone_idx, , drop = FALSE]
  n_clone <- nrow(clone_coords)

  # Log-determinant of covariance matrix
  cov_mat <- cov(clone_coords)
  obs_logdet <- log(max(det(cov_mat), .Machine$double.xmin))

  # Convex hull volume (area in 2D)
  obs_hull <- tryCatch({
    if (n_clone > ncol(clone_coords)) {
      ch <- convhulln(clone_coords, options = "FA")
      ch$vol
    } else { NA }
  }, error = function(e) NA)

  # Null distribution
  null_logdets <- replicate(n_permutations, {
    ri <- sample(nrow(all_coords), n_clone)
    rc <- cov(all_coords[ri, , drop = FALSE])
    log(max(det(rc), .Machine$double.xmin))
  })

  null_mean <- mean(null_logdets)
  null_sd <- sd(null_logdets)
  z <- if (null_sd > 0) (obs_logdet - null_mean) / null_sd else NA

  phenovol_results <- rbind(phenovol_results, data.frame(
    barcode = bc, clone_size = n_clone,
    log_det_cov = obs_logdet, hull_volume = obs_hull,
    null_logdet_mean = null_mean, null_logdet_sd = null_sd,
    logdet_z = z, stringsAsFactors = FALSE
  ))
}

cat("Phenotypic volume computation complete\n")

# Plot: phenotypic volume z-score vs clone size
p_pvol <- ggplot(phenovol_results, aes(x = clone_size, y = logdet_z)) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  theme_classic() +
  theme(aspect.ratio = 1) +
  labs(
    x = "Clone size (number of cells)",
    y = "Phenotypic volume z-score (log|cov| vs random)",
    title = "Phenotypic Volume by Clone"
  )
print(p_pvol)

# Plot: observed log-det vs null
p_pvol_hist <- ggplot() +
  geom_density(data = phenovol_results,
               aes(x = log_det_cov, fill = "Observed clones"), alpha = 0.5) +
  geom_density(data = phenovol_results,
               aes(x = null_logdet_mean, fill = "Random null (mean)"), alpha = 0.5) +
  scale_fill_manual(values = c("Observed clones" = "steelblue",
                                "Random null (mean)" = "gray70")) +
  theme_classic() +
  labs(x = "log(det(Cov))", y = "Density", fill = "",
       title = "Phenotypic volume: observed vs null")
print(p_pvol_hist)

# ---------------------------------------------------------------------------
# 5. Shannon's Equitability Index (Pielou's J)
# ---------------------------------------------------------------------------
# Measures how evenly a clone's cells distribute across transcriptomic
# clusters. Defined as J = H / H_max, where:
#   H     = -sum(p_i * ln(p_i))   (Shannon entropy)
#   H_max = ln(S)                 (S = number of clusters occupied)
#   p_i   = proportion of the clone's cells in cluster i
#
# J = 1 means cells are perfectly evenly spread; J -> 0 means cells
# concentrate in a single cluster. This captures fate diversity: clones
# with high J adopt multiple transcriptomic states, while low J clones
# converge to one state (Goyal et al., 2023; Zhang, Melzer et al., 2024).
#
# Requires a cluster column. The Seurat cluster assignments were loaded
# in setup. If the column is missing, fall back to k-means on PCs.

if (!"cluster" %in% colnames(data) || all(is.na(data$cluster))) {
  n_clusters <- 10
  km <- kmeans(all_coords, centers = n_clusters, nstart = 25)
  data$cluster <- as.character(km$cluster)
  cat("No Seurat cluster column found; assigned", n_clusters, "k-means clusters on PCs\n")
} else {
  data$cluster <- as.character(data$cluster)
  cat("Using Seurat cluster assignments:", length(unique(data$cluster)), "clusters\n")
}

# Shannon equitability per clone
shannon_equitability <- function(cluster_vec) {
  tab <- table(cluster_vec)
  props <- as.numeric(tab) / sum(tab)
  S <- length(tab)              # number of occupied clusters
  if (S <= 1) return(0)         # all cells in one cluster
  H <- -sum(props * log(props))
  H_max <- log(S)
  J <- H / H_max
  return(J)
}

shannon_results <- data.frame(
  barcode = character(),
  clone_size = integer(),
  n_clusters_occupied = integer(),
  shannon_H = numeric(),
  equitability_J = numeric(),
  null_J_mean = numeric(),
  null_J_sd = numeric(),
  J_z = numeric(),
  stringsAsFactors = FALSE
)

cat("Computing Shannon equitability for", length(clone_barcodes), "clones\n")

cluster_vec_all <- data$cluster

for (bc in clone_barcodes) {
  clone_idx <- which(data$barcode == bc)
  clone_clusters <- cluster_vec_all[clone_idx]
  n_clone <- length(clone_idx)

  tab <- table(clone_clusters)
  props <- as.numeric(tab) / sum(tab)
  S <- length(tab)
  H <- if (S > 1) -sum(props * log(props)) else 0
  J <- shannon_equitability(clone_clusters)

  # Null: random cells of same size
  null_Js <- replicate(n_permutations, {
    ri <- sample(length(cluster_vec_all), n_clone)
    shannon_equitability(cluster_vec_all[ri])
  })

  null_mean <- mean(null_Js)
  null_sd <- sd(null_Js)
  z <- if (null_sd > 0) (J - null_mean) / null_sd else NA

  shannon_results <- rbind(shannon_results, data.frame(
    barcode = bc, clone_size = n_clone,
    n_clusters_occupied = S,
    shannon_H = H, equitability_J = J,
    null_J_mean = null_mean, null_J_sd = null_sd,
    J_z = z, stringsAsFactors = FALSE
  ))
}

cat("Shannon equitability computation complete\n")

# Plot: equitability vs clone size
p_shan <- ggplot(shannon_results, aes(x = clone_size, y = equitability_J)) +
  geom_point(size = 1.5) +
  geom_hline(aes(yintercept = mean(null_J_mean)),
             linetype = "dashed", color = "red", data = shannon_results) +
  theme_classic() +
  theme(aspect.ratio = 1) +
  ylim(0, 1) +
  labs(
    x = "Clone size (number of cells)",
    y = "Shannon equitability (J)",
    title = "Fate Diversity by Clone (Shannon Equitability)"
  )
print(p_shan)

# Plot: equitability z-scores
p_shan_z <- ggplot(shannon_results, aes(x = clone_size, y = J_z)) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  theme_classic() +
  theme(aspect.ratio = 1) +
  labs(
    x = "Clone size (number of cells)",
    y = "Equitability z-score (vs random null)",
    title = "Shannon Equitability Z-Score"
  )
print(p_shan_z)

# Plot: distribution of J observed vs null
p_shan_hist <- ggplot() +
  geom_histogram(data = shannon_results,
                 aes(x = equitability_J, fill = "Observed"), alpha = 0.6, bins = 30) +
  geom_vline(xintercept = mean(shannon_results$null_J_mean),
             linetype = "dashed", color = "red") +
  scale_fill_manual(values = c("Observed" = "steelblue")) +
  theme_classic() +
  labs(x = "Shannon equitability (J)", y = "Count", fill = "",
       title = "Per-clone equitability vs random null mean")
print(p_shan_hist)

# ---------------------------------------------------------------------------
# 6. Summary table
# ---------------------------------------------------------------------------
summary_table <- edist_results %>%
  left_join(mixing_per_clone, by = c("barcode", "clone_size")) %>%
  left_join(phenovol_results %>% select(barcode, log_det_cov, hull_volume, logdet_z),
            by = "barcode") %>%
  left_join(shannon_results %>% select(barcode, n_clusters_occupied, equitability_J, J_z),
            by = "barcode")
cat("Summary of barcode similarity metrics:\n")
print(summary_table)
```


## Other Joint Analyses

### looking at patterns, cell cycle
```{r}
rds_dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS024/singleCell/Seurat/RDS/"
seurat_obj <- readRDS(paste0(rds_dir, "GFP_invitro.rds"))
combined.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS024/combined/"
combined_data <- read.csv(paste0(combined.dir, "dataframes/GFP_invitro_all.csv"))


s_genes <- cc.genes.updated.2019$s.genes
g2m_genes <- cc.genes.updated.2019$g2m.genes

seurat_obj <- CellCycleScoring(seurat_obj,
                        s.features = s_genes,
                        g2m.features = g2m_genes,
                        set.ident = FALSE)

# Add barcode info from combined_data to Seurat metadata
barcode_map <- combined_data %>%
  select(cellID, barcode) %>%
  distinct()
seurat_obj@meta.data$cellID <- rownames(seurat_obj@meta.data)
seurat_obj@meta.data <- seurat_obj@meta.data %>%
  left_join(barcode_map, by = "cellID")
rownames(seurat_obj@meta.data) <- seurat_obj@meta.data$cellID

# Now compute clone size
barcode_counts <- seurat_obj@meta.data %>%
  filter(barcode != "none") %>%
  count(barcode, name = "clone_size")
seurat_obj@meta.data <- seurat_obj@meta.data %>%
  left_join(barcode_counts, by = "barcode") %>%
  mutate(
    clone_size = ifelse(barcode == "none", NA_integer_, clone_size),
    is_singleton = clone_size == 1
  )
rownames(seurat_obj@meta.data) <- seurat_obj@meta.data$cellID


# obj@meta.data now has: S.Score, G2M.Score, Phase
table(seurat_obj@meta.data$Phase[seurat_obj@meta.data$clone_size == 1])
table(seurat_obj@meta.data$Phase[seurat_obj@meta.data$clone_size > 1])


Idents(seurat_obj) <- "seurat_clusters"
markers_7 <- FindMarkers(seurat_obj, ident.1 = "7", 
                         only.pos = TRUE, 
                         min.pct = 0.25, 
                         logfc.threshold = 0.5)
head(markers_7, 30)

Idents(seurat_obj) <- "seurat_clusters"
markers_6 <- FindMarkers(seurat_obj, ident.1 = "6", 
                         only.pos = TRUE, 
                         min.pct = 0.25, 
                         logfc.threshold = 0.5)
head(markers_6, 30)





# Pattern 2 weight by barcode status
meta <- seurat_obj@meta.data

meta$barcode_status <- case_when(
  meta$barcode == "none" ~ "unbarcoded",
  meta$clone_size == 1   ~ "singleton",
  meta$clone_size > 1    ~ "clone"
)

load_dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS024/singleCell/cNMF/"
usage_file <- file.path(load_dir, paste0("GFP_invitro_usage_K6.csv"))
usage <- read.csv(usage_file, row.names = 1, check.names = FALSE)
seurat_obj <- add_cnmf_to_seurat(seurat_obj, usage)

tapply(meta$Pattern_2, meta$barcode_status, mean, na.rm = TRUE)



pattern_cols <- paste0("Pattern_", 1:6)

for (p in pattern_cols) {
  cat(p, "\n")
  
  # Overall test across all three groups
  kw <- kruskal.test(meta[[p]] ~ meta$barcode_status)
  cat("  Kruskal-Wallis p =", format(kw$p.value, digits = 3), "\n")
  
  # Pairwise comparisons
  pw <- pairwise.wilcox.test(meta[[p]], meta$barcode_status, 
                             p.adjust.method = "BH")
  print(pw$p.value)
  cat("\n")
}

library(effectsize)

for (p in pattern_cols) {
  cat(p, "\n")
  r <- rank_biserial(meta[[p]][meta$barcode_status == "singleton"],
                     meta[[p]][meta$barcode_status == "clone"])
  cat("  Singleton vs Clone rank-biserial r =", round(r$r_rank_biserial, 3), "\n\n")
}



# Cluster composition by barcode status
cluster_by_status <- table(meta$seurat_clusters, meta$barcode_status)
print(round(prop.table(cluster_by_status, margin = 2), 3))

# Also show raw counts
print(cluster_by_status)

# Enrichment test per cluster: are singletons over- or under-represented?
for (cl in sort(unique(meta$seurat_clusters))) {
  in_cluster <- meta$seurat_clusters == cl
  tbl <- table(in_cluster, meta$barcode_status)
  
  # Skip if any column is all zero
  if (ncol(tbl) < 3) next
  
  # Fisher test: singleton vs clone enrichment within this cluster
  barcoded <- meta$barcode_status %in% c("singleton", "clone")
  if (sum(in_cluster & barcoded) == 0) next
  
  ft <- fisher.test(table(
    in_cluster[barcoded],
    meta$barcode_status[barcoded] == "singleton"
  ))
  
  cat("Cluster", cl, 
      "- Singleton OR:", round(ft$estimate, 2),
      "- p:", format(ft$p.value, digits = 3), "\n")
}

# Compare CCR5 across barcode status, excluding cluster 7
non7 <- subset(seurat_obj, seurat_clusters != 7)
non7_meta <- non7@meta.data

# Check if CCR5 is in the data (may be named differently in macaque)
grep("CCR5", rownames(seurat_obj), value = TRUE)

# If present:
non7_meta$CCR5 <- FetchData(non7, vars = "CCR5")[[1]]
non7_meta$barcode_status <- case_when(
  non7_meta$barcode == "none" ~ "unbarcoded",
  non7_meta$clone_size == 1   ~ "singleton",
  non7_meta$clone_size > 1    ~ "clone"
)

tapply(non7_meta$CCR5, non7_meta$barcode_status, mean, na.rm = TRUE)
wilcox.test(CCR5 ~ is_singleton, data = non7_meta[non7_meta$barcode_status != "unbarcoded",])



# Within cluster 1 (largest mixed cluster), DE between singletons and clones
seurat_obj@meta.data$barcode_status <- case_when(
  seurat_obj@meta.data$barcode == "none" ~ "unbarcoded",
  seurat_obj@meta.data$clone_size == 1   ~ "singleton",
  seurat_obj@meta.data$clone_size > 1    ~ "clone"
)

cluster1 <- subset(seurat_obj, seurat_clusters == 1 & 
                     barcode_status %in% c("singleton", "clone"))
Idents(cluster1) <- "is_singleton"
de_within_c1 <- FindMarkers(cluster1, ident.1 = TRUE, ident.2 = FALSE,
                             logfc.threshold = 0.25, min.pct = 0.1)
cat("Significant DE genes (padj < 0.05):", sum(de_within_c1$p_val_adj < 0.05), "\n")
head(de_within_c1[order(de_within_c1$p_val_adj),], 10)
```

Difference between clones and singletons in clusters
```{r}
for (cl in c(1, 2, 3, 4, 5, 6, 7, 8)) {
  cl_obj <- subset(seurat_obj, seurat_clusters == cl & 
                     barcode_status %in% c("singleton", "clone"))
  
  n_sing <- sum(cl_obj@meta.data$is_singleton)
  n_clone <- sum(!cl_obj@meta.data$is_singleton)
  
  if (n_sing < 3 || n_clone < 3) {
    cat("Cluster", cl, "- skipped (singleton:", n_sing, "clone:", n_clone, ")\n")
    next
  }
  
  Idents(cl_obj) <- "is_singleton"
  de <- FindMarkers(cl_obj, ident.1 = TRUE, ident.2 = FALSE,
                     logfc.threshold = 0.25, min.pct = 0.1)
  n_sig <- sum(de$p_val_adj < 0.05)
  cat("Cluster", cl, "- Singletons:", n_sing, "Clones:", n_clone,
      "- Sig DE genes:", n_sig, "\n")
}
```

Distance between clones and singletons in cluster 1
```{r}
#cluster1 <- subset(seurat_obj, 
#                     barcode_status %in% c("singleton", "clone"))
cluster1 <- subset(seurat_obj, seurat_clusters == 1 & 
                     barcode_status %in% c("singleton", "clone"))
pca <- Embeddings(cluster1, "pca")[, 1:30]
meta_c1 <- cluster1@meta.data
is_sing <- meta_c1$is_singleton

sing_idx <- which(is_sing)
clone_idx <- which(!is_sing)

set.seed(42)
n_pairs <- 2000  # consistent across all comparisons

pair_dist <- function(pairs, pca) {
  sqrt(rowSums((pca[pairs[,1],] - pca[pairs[,2],])^2))
}

# 1. Within-clone: both cells share the same barcode
clone_barcodes <- meta_c1$barcode[clone_idx]
# Only barcodes with 2+ cells in this cluster
bc_counts <- table(clone_barcodes)
eligible_bcs <- names(bc_counts[bc_counts >= 2])

within_pairs <- matrix(NA, nrow = n_pairs, ncol = 2)
for (k in seq_len(n_pairs)) {
  bc <- sample(eligible_bcs, 1)
  members <- clone_idx[clone_barcodes == bc]
  picked <- sample(members, 2, replace = FALSE)
  within_pairs[k,] <- picked
}
d_within <- pair_dist(within_pairs, pca)

# 2. Between-clone: two clone cells with different barcodes
between_pairs <- matrix(NA, nrow = n_pairs, ncol = 2)
for (k in seq_len(n_pairs)) {
  i <- sample(clone_idx, 1)
  j <- sample(clone_idx, 1)
  while (meta_c1$barcode[j] == meta_c1$barcode[i]) {
    j <- sample(clone_idx, 1)
  }
  between_pairs[k,] <- c(i, j)
}
d_between <- pair_dist(between_pairs, pca)

# 3. Singleton-Singleton
ss_pairs <- matrix(NA, nrow = n_pairs, ncol = 2)
for (k in seq_len(n_pairs)) {
  picked <- sample(sing_idx, 2, replace = TRUE)
  while (picked[1] == picked[2]) {
    picked <- sample(sing_idx, 2, replace = TRUE)
  }
  ss_pairs[k,] <- picked
}
d_ss <- pair_dist(ss_pairs, pca)

# 4. Clone-Singleton
cs_pairs <- cbind(
  sample(clone_idx, n_pairs, replace = TRUE),
  sample(sing_idx, n_pairs, replace = TRUE)
)
d_cs <- pair_dist(cs_pairs, pca)

# 5. Random (any two cells in cluster 1)
all_idx <- seq_len(nrow(pca))
rand_pairs <- matrix(NA, nrow = n_pairs, ncol = 2)
for (k in seq_len(n_pairs)) {
  picked <- sample(all_idx, 2)
  rand_pairs[k,] <- picked
}
d_rand <- pair_dist(rand_pairs, pca)

# Summary
cat("Within-clone:       ", round(mean(d_within), 3), "+/-", round(sd(d_within), 3), "\n")
cat("Between-clone:      ", round(mean(d_between), 3), "+/-", round(sd(d_between), 3), "\n")
cat("Singleton-Singleton:", round(mean(d_ss), 3), "+/-", round(sd(d_ss), 3), "\n")
cat("Clone-Singleton:    ", round(mean(d_cs), 3), "+/-", round(sd(d_cs), 3), "\n")
cat("Random:             ", round(mean(d_rand), 3), "+/-", round(sd(d_rand), 3), "\n")

# Plot
dist_df <- data.frame(
  distance = c(d_within, d_between, d_ss, d_cs, d_rand),
  comparison = rep(c("Within-clone", "Between-clone", 
                      "Singleton-Singleton", "Clone-Singleton", "Random"),
                   each = n_pairs)
)
dist_df$comparison <- factor(dist_df$comparison,
                              levels = c("Within-clone", "Between-clone",
                                         "Singleton-Singleton", 
                                         "Clone-Singleton", "Random"))


library(ggpubr)

comparisons_list <- list(
  c("Within-clone", "Between-clone"),
  c("Within-clone", "Singleton-Singleton"),
  c("Within-clone", "Clone-Singleton"),
  c("Within-clone", "Random"),
  c("Between-clone", "Singleton-Singleton"),
  c("Between-clone", "Clone-Singleton"),
  c("Singleton-Singleton", "Clone-Singleton"),
  c("Singleton-Singleton", "Random")
)

p <- ggplot(dist_df, aes(x = comparison, y = distance)) +
  geom_violin(fill = "gray80") +
  geom_boxplot(width = 0.1) +
  stat_compare_means(comparisons = comparisons_list, 
                     method = "wilcox.test",
                     label = "p.signif",
                     step.increase = 0.05) +
  theme_classic() +
  labs(x = "", y = "Euclidean distance (PCA space)") +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
output_dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS024/joint/plots/"
ggsave(paste0(output_dir, "barcode_edist_violins_cluster1.pdf"), p)
```



### Host factors

```{r restrict and depend}
factors_restrict <- c("APOBEC3G", "APOBEC3H", "TRIM5", "SAMHD1", 
                      "BST2", "SERINC5", "SERINC3", "MX2", "IFITM2")
factors_depend <- c("CD4", "CCR5", "PPIA", "PSIP1", "TNPO3", "CPSF6", "NUP153", "RANBP2", 
                    "TSG101", "PDCD6IP", "SP1", "NFKB1", "RELA", "CDK9", "CCNT1", "TNPO1",
                    "XPO1", "DDX3X", "RAN", "RANBP1", "RANGAP1")
cytotox <- c("TBX21", "EOMES", "RUNX3", "IFNG-AS1", "GZMK", "GZMA", "CX3CR1", "PRF1", 
                    "GZMB", "GZMH", 'GNLY', "ZNF683")

# Load all RDS
input.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS016/singleCell/Seurat/RDS/"
objects <- list()
for (sample_name in names(joint_data)) {
  seurat_obj <- readRDS(paste0(input.dir, sample_name, ".rds"))
  objects[[sample_name]] <- seurat_obj
}



# Run the enhanced sparse analysis
results_sparse <- host_virus_correlation_sparse(
    joint_salve_data = joint_data,
    joint_rds = objects,
    genes_list = factors_restrict,
    viral_col = "SALVE",
    min_coexpress = 10,
    n_permutations = 1000,
    output.dir = "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/joint/hostVirus/new_restrictionFactors/",
    outputs = TRUE
)

host_virus_correlation(joint_data, objects, factors_restrict, viral_col = "SALVE", 
            output.dir = "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/joint/hostVirus/new_restrictionFactors/")
host_virus_correlation(joint_data, objects, factors_restrict, viral_col = "USprop", 
            output.dir = "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/joint/hostVirus/restrictionFactors/")
host_virus_correlation(joint_data, objects, factors_restrict, viral_col = "SSprop", 
            output.dir = "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/joint/hostVirus/restrictionFactors/")
host_virus_correlation(joint_data, objects, factors_restrict, viral_col = "MSprop", 
            output.dir = "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/joint/hostVirus/restrictionFactors/")
host_virus_correlation(joint_data, objects, factors_restrict, viral_col = "GEX", 
            output.dir = "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/joint/hostVirus/restrictionFactors/")

host_virus_correlation(joint_data, objects, factors_depend, viral_col = "SALVE", 
            output.dir = "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/joint/hostVirus/dependencyFactors/")
host_virus_correlation(joint_data, objects, factors_depend, viral_col = "USprop", 
            output.dir = "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/joint/hostVirus/dependencyFactors/")
host_virus_correlation(joint_data, objects, factors_depend, viral_col = "SSprop", 
            output.dir = "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/joint/hostVirus/dependencyFactors/")
host_virus_correlation(joint_data, objects, factors_depend, viral_col = "MSprop", 
            output.dir = "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/joint/hostVirus/dependencyFactors/")
host_virus_correlation(joint_data, objects, factors_depend, viral_col = "GEX", 
            output.dir = "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/joint/hostVirus/dependencyFactors/")

# are the trends the same between SALVE and GEX? And the isoforms?
input.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/joint/hostVirus/dependencyFactors/"
for (sample_name in names(joint_data)) {
  gex_file <- paste0(input.dir, sample_name, "_GEX_correlations.csv")
  salve_file <- paste0(input.dir, sample_name, "_SALVE_correlations.csv")
  
  # Check if both files exist
  if (!file.exists(gex_file) || !file.exists(salve_file)) {
    cat("Skipping", sample_name, "- correlation files not found\n")
    next
  }
  
  file_GEX <- read.csv(gex_file) %>% 
    rename(GEXcor = correlation, GEXpadj = p_adj)
  file_SALVE <- read.csv(salve_file) %>% 
    rename(SALVEcor = correlation, SALVEpadj = p_adj)
  joint_corr <- full_join(file_GEX, file_SALVE, by = "gene") %>%
    select(gene, GEXcor, GEXpadj, SALVEcor, SALVEpadj)
  p1 <- ggplot(joint_corr, aes(x = GEXcor, y = SALVEcor)) +
        geom_point(alpha = 0.3, size = 2) +
        geom_text_repel(aes(label = gene)) +
        labs(title = paste0(sample_name, " GEX vs SALVE Correlations: Dependency Factors"),
             x = "GEX",
             y = "SALVE") +
        theme_minimal() +
        theme(plot.title = element_text(face = "bold", size = 14),
              axis.text = element_text(size = 10))
  ggsave(p1, file = paste0(input.dir, sample_name, "_compare_correlations.pdf"),
         device = "pdf")
}

```

```{r memory cytotox}
# Define cell type markers
memory_cd4_markers <- list(
  CD4 = "CD4",
  memory = c("IL7R", "CCR7", "SELL"),  # Central memory
  exclude_naive = c("CCR7", "SELL"),    # High in naive
  exclude_effector = c("GZMB", "PRF1")  # High in effector
)


subset_memory_cd4 <- function(seurat_obj, min_cd4_expr = 0.5) {
  cd4_expr <- GetAssayData(seurat_obj, slot = "data")["CD4", ]
  il7r_expr <- GetAssayData(seurat_obj, slot = "data")["IL7R", ]
  
  memory_cd4_cells <- names(cd4_expr)[cd4_expr > min_cd4_expr & il7r_expr > 0]
  seurat_subset <- subset(seurat_obj, cells = memory_cd4_cells)
  
  return(seurat_subset)
}

joint_data_memory <- list()
joint_rds_memory <- list()
# Subset both RDS and SALVE data
for (sample_name in names(objects)) {
  seurat_subset <- subset_memory_cd4(objects[[sample_name]])
  cells_keep <- colnames(seurat_subset)
  
  joint_rds_memory[[sample_name]] <- seurat_subset
  joint_data_memory[[sample_name]] <- joint_data[[sample_name]] %>%
    filter(cellID %in% cells_keep)
  
  cat("Sample:", sample_name, 
      "- Original cells:", ncol(seurat_obj),
      "- Memory CD4 cells:", ncol(joint_rds_memory[[sample_name]]), "\n")
}

# Run correlation on subset
host_virus_correlation(
  joint_salve_data = joint_data_memory,
  joint_rds = joint_rds_memory, 
  genes_list = cytotox,
  output.dir = "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/joint/hostVirus/memoryCytotoxic/",
  viral_col = "SALVE"
)
```

```{r metabolic}

calculate_metabolic_scores_macaque <- function(seurat_obj) {
  
  # Check if percent_mito already exists
  if (!"percent_mito" %in% colnames(seurat_obj@meta.data)) {
    stop("percent_mito not found. Run Add_Mito_Ribo() first.")
  }
  
  cat("Using existing percent_mito column\n")
  
  # Define metabolic genes (human names)
  glycolysis_human <- c("HK2", "PFKP", "ALDOA", "GAPDH", "PGK1", 
                        "PGAM1", "ENO1", "PKM")
  oxphos_human <- c("NDUFA1", "NDUFB1", "SDHB", "UQCRB", "COX5A", 
                    "ATP5A1", "ATP5B", "ATP5F1A", "ATP5PB")
  
  # Convert to match your data
  glycolysis_genes <- convert_gene_list(seurat_obj, glycolysis_human)
  oxphos_genes <- convert_gene_list(seurat_obj, oxphos_human)
  
  # Calculate scores
  if (length(glycolysis_genes) > 2) {
    seurat_obj <- AddModuleScore(seurat_obj,
                                 features = list(glycolysis_genes),
                                 name = "glycolysis_score")
    cat("Calculated glycolysis score with", length(glycolysis_genes), "genes\n")
  }
  
  if (length(oxphos_genes) > 2) {
    seurat_obj <- AddModuleScore(seurat_obj,
                                 features = list(oxphos_genes),
                                 name = "oxphos_score")
    cat("Calculated OXPHOS score with", length(oxphos_genes), "genes\n")
  }
  
  # Ribosomal genes (check if percent_ribo also exists from Add_Mito_Ribo)
  if (!"percent_ribo" %in% colnames(seurat_obj@meta.data)) {
    ribo_genes <- unique(c(
      grep("^RPL|^RPS", rownames(seurat_obj), value = TRUE),
      grep("^Rpl|^Rps", rownames(seurat_obj), value = TRUE)
    ))
    
    if (length(ribo_genes) > 10) {
      seurat_obj <- AddModuleScore(seurat_obj,
                                   features = list(ribo_genes),
                                   name = "translation_score")
      cat("Calculated translation score with", length(ribo_genes), "genes\n")
    }
  } else {
    cat("Using existing percent_ribo column\n")
  }
  
  return(seurat_obj)
}
# Function to test all metabolic correlations with viral load
test_metabolic_correlations <- function(joint_salve_data, joint_rds, 
                                       viral_col = "SALVE",
                                       output.dir = NULL) {
  
  all_results <- list()
  
  for (sample_name in names(joint_salve_data)) {
    cat("\n=== SAMPLE:", sample_name, "===\n")
    
    current_df <- joint_salve_data[[sample_name]]
    seurat_obj <- joint_rds[[sample_name]]
    
    # Extract viral measure
    if (!viral_col %in% colnames(current_df)) {
      cat("Column", viral_col, "not found. Skipping.\n")
      next
    }
    viral_measure <- current_df[[viral_col]]
    
    if (sd(viral_measure) == 0) {
      cat(viral_col, "has zero variance. Skipping.\n")
      next
    }
    
    # Calculate metabolic scores if not already done
    if (!"glycolysis_score1" %in% colnames(seurat_obj@meta.data)) {
      seurat_obj <- calculate_metabolic_scores_macaque(seurat_obj)
    }
    
    # Add total UMI/genes if not present
    if (!"total_UMI" %in% colnames(seurat_obj@meta.data)) {
      seurat_obj$total_UMI <- colSums(GetAssayData(seurat_obj, slot = "counts"))
      seurat_obj$total_genes <- colSums(GetAssayData(seurat_obj, slot = "counts") > 0)
    }
    
    # Extract all metabolic metrics
    metabolic_metrics <- c("percent_mito", "total_UMI", "total_genes",
                          "glycolysis_score1", "oxphos_score1")
    
    # Add translation score if exists
    if ("translation_score1" %in% colnames(seurat_obj@meta.data)) {
      metabolic_metrics <- c(metabolic_metrics, "translation_score1")
    }
    
    # Add ribosomal if exists
    if ("percent_ribo" %in% colnames(seurat_obj@meta.data)) {
      metabolic_metrics <- c(metabolic_metrics, "percent_ribo")
    }
    
    # Test correlations
    cat("\n=== METABOLIC CORRELATIONS WITH", viral_col, "===\n")
    cat(sprintf("%-25s %10s %12s %10s %8s\n", 
                "Metric", "Corr", "P-value", "N_cells", "Sig"))
    cat(strrep("-", 70), "\n")
    
    results <- data.frame()
    
    for (metric in metabolic_metrics) {
      if (metric %in% colnames(seurat_obj@meta.data)) {
        metric_values <- seurat_obj@meta.data[[metric]]
        
        # Filter out NA values
        valid_cells <- !is.na(metric_values) & !is.na(viral_measure)
        
        if (sum(valid_cells) > 10) {
          test <- cor.test(metric_values[valid_cells], 
                          viral_measure[valid_cells],
                          method = "spearman", 
                          exact = FALSE)
          
          sig <- if (test$p.value < 0.001) "***" else 
                 if (test$p.value < 0.01) "**" else 
                 if (test$p.value < 0.05) "*" else "ns"
          
          cat(sprintf("%-25s %10.4f %12.2e %10d %8s\n",
                     metric, test$estimate, test$p.value, sum(valid_cells), sig))
          
          results <- rbind(results, data.frame(
            sample = sample_name,
            metric = metric,
            correlation = as.numeric(test$estimate),
            p_value = test$p.value,
            n_cells = sum(valid_cells)
          ))
        }
      }
    }
    
    results$p_adj <- p.adjust(results$p_value, method = "BH")
    all_results[[sample_name]] <- results
    
    cat(strrep("-", 70), "\n\n")
  }
  
  # Combine all results
  combined_results <- do.call(rbind, all_results)
  
  # Save if output directory provided
  if (!is.null(output.dir)) {
    if (!dir.exists(output.dir)) {
      dir.create(output.dir, recursive = TRUE)
    }
    write.csv(combined_results, 
              file.path(output.dir, paste0(viral_col, "_metabolic_correlations.csv")),
              row.names = FALSE)
    cat("Saved results to:", output.dir, "\n")
  }
  
  return(combined_results)
}

# Run analysis
metabolic_results <- test_metabolic_correlations(
  joint_salve_data = joint_data,
  joint_rds = objects,
  viral_col = "SALVE",
  output.dir = "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/joint/hostVirus/metabolic/"
)

# View summary
metabolic_results %>%
  group_by(metric) %>%
  summarise(
    mean_corr = mean(correlation),
    median_corr = median(correlation),
    n_samples = n(),
    n_sig = sum(p_adj < 0.05)
  ) %>%
  arrange(desc(abs(mean_corr)))


```



## Subset analysis

To load data, run first chunk of Joint bamsort (SALVE vs GEX) or Joint combined

### DEG analysis

```{r, warning=FALSE}
input.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS016/singleCell/Seurat/RDS/"
output.base <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/joint/subset/SALVE_new/"

samples <- c("Invitro",
             "D195",
             "Pacute",
             "W2")
for (i in 1:length(samples)) {
  sample_name <- samples[i]
  input_file <- paste0(input.dir, sample_name, ".rds")
  
  tryCatch({
    seurat_obj <- readRDS(input_file)
    genes_to_exclude <- c("CD4", "CD8A", "CD8B")  # remove from DEG analysis
    genes_to_keep <- rownames(seurat_obj)[!rownames(seurat_obj) %in% genes_to_exclude]
    seurat_obj <- subset(seurat_obj, features = genes_to_keep)

    
    salve <- joint_data[[sample_name]]
    list_of_cells1 <- salve %>% filter(SALVE != 0) %>% #update as desired
      select(cellID) %>% unlist()
    # list_of_cells2 <- salve %>% filter(total != 0) %>%
    #   select(cellID) %>% filter(!(cellID %in% list_of_cells1)) %>% unlist()
    results <- analyze_cell_subset(seurat_obj, list_of_cells1#, 
                                   #list_of_cells2, group_names = c("MS", "noMS")
                                   )
    
    output.dir <- paste0(output.base, sample_name, "/")
    if (!dir.exists(output.dir)) {
      dir.create(output.dir, recursive = TRUE)
    }
    #save_subset_analysis(results, output_dir = output.dir)
    plot_subset_analysis(seurat_obj, results, list_of_cells1, output_dir = paste0(output.dir, "plots/"))
    
    # results_uc <- analyze_cell_subset_unclustered(seurat_obj, list_of_cells1#, 
    #                                               #list_of_cells2, group_names = c("MS", "noMS")
    #                                               )
    # save_subset_analysis(results_uc, output_dir = output.dir)
    
    cat("Processed sample:", sample_name, "\n")
  }, error = function(e) {
    cat("Error processing sample", sample_name, ":", conditionMessage(e), "\n\n")
  })
}

```

Rewrote some functions
```{r}
library(gprofiler2)

#' Perform DEG analysis between two cell populations
#'
#' @param seurat_obj Seurat object containing single-cell data
#' @param cells_group1 Character vector of cell IDs for group 1 (numerator)
#' @param cells_group2 Character vector of cell IDs for group 2 (denominator)
#' @param assay Character, assay to use (default: "RNA")
#' @param test_use Character, statistical test method (default: "wilcox")
#' @param fc_threshold Numeric, log2FC threshold for filtering (default: 0.5)
#' @param pval_threshold Numeric, adjusted p-value threshold (default: 0.01)
#' @param output_prefix Character, prefix for output files (default: "DEG_analysis")
#' 
#' @return List containing DEG results, filtered DEGs, pathway results, and plot
#' 
#' @details
#' Statistical Test: Wilcoxon rank-sum test is recommended for single-cell data
#' due to its non-parametric nature and robustness to the zero-inflated, 
#' non-normal distributions typical of scRNA-seq data. Alternative tests include
#' "bimod" (likelihood-ratio test), "t" (Student's t-test), or "MAST" 
#' (Model-based Analysis of Single-cell Transcriptomics).
#' 
#' Multiple Testing Correction: Bonferroni correction is applied by default in
#' Seurat's FindMarkers function to control family-wise error rate.
#' 
#' Pathway Analysis: g:Profiler uses a hypergeometric test with g:SCS multiple
#' testing correction to identify significantly enriched pathways from GO, KEGG,
#' Reactome, and other databases.
perform_deg_pathway_analysis <- function(seurat_obj,
                                          cells_group1,
                                          cells_group2,
                                          assay = "RNA",
                                          test_use = "wilcox",
                                          fc_threshold = 0.5,
                                          pval_threshold = 0.01,
                                          output_prefix = "DEG_analysis") {
  
  # Validate inputs
  if (!inherits(seurat_obj, "Seurat")) {
    stop("seurat_obj must be a Seurat object")
  }
  
  all_cells <- colnames(seurat_obj)
  missing_cells1 <- setdiff(cells_group1, all_cells)
  missing_cells2 <- setdiff(cells_group2, all_cells)
  
  if (length(missing_cells1) > 0) {
    warning(paste(length(missing_cells1), "cells from group1 not found in Seurat object"))
    cells_group1 <- intersect(cells_group1, all_cells)
  }
  
  if (length(missing_cells2) > 0) {
    warning(paste(length(missing_cells2), "cells from group2 not found in Seurat object"))
    cells_group2 <- intersect(cells_group2, all_cells)
  }
  
  if (length(cells_group1) == 0 || length(cells_group2) == 0) {
    stop("One or both groups have no valid cells after filtering")
  }
  
  cat(sprintf("Group 1: %d cells\n", length(cells_group1)))
  cat(sprintf("Group 2: %d cells\n", length(cells_group2)))
  
  # Create temporary identity for DEG analysis
  seurat_obj$temp_ident <- "other"
  seurat_obj$temp_ident[colnames(seurat_obj) %in% cells_group1] <- "group1"
  seurat_obj$temp_ident[colnames(seurat_obj) %in% cells_group2] <- "group2"
  
  # Subset Seurat object to only include cells from both groups
  seurat_subset <- subset(seurat_obj, cells = c(cells_group1, cells_group2))
  Idents(seurat_subset) <- "temp_ident"
  
  cat("Performing differential expression analysis...\n")
  
  # Perform DEG analysis
  # group1 vs group2: positive log2FC means higher in group1
  deg_results <- FindMarkers(
    object = seurat_subset,
    ident.1 = "group1",
    ident.2 = "group2",
    assay = assay,
    test.use = test_use,
    logfc.threshold = 0,  # Get all genes, filter later
    min.pct = 0.1,  # Gene must be expressed in at least 10% of cells in one group
    verbose = TRUE
  )
  
  # Add gene names as column
  deg_results$gene <- rownames(deg_results)
  
  # Reorder columns
  deg_results <- deg_results[, c("gene", "avg_log2FC", "pct.1", "pct.2", 
                                  "p_val", "p_val_adj")]
  
  # Sort by adjusted p-value
  deg_results <- deg_results[order(deg_results$p_val_adj), ]
  
  # Save full results
  write.csv(deg_results, 
            file = paste0(output_prefix, "_all_genes.csv"),
            row.names = FALSE)
  cat(sprintf("Saved full results: %s_all_genes.csv\n", output_prefix))
  
  # Filter for significant DEGs
  deg_filtered <- deg_results %>%
    filter(abs(avg_log2FC) > fc_threshold & p_val_adj < pval_threshold)
  
  cat(sprintf("Significant DEGs: %d (|log2FC| > %.2f, adj p-val < %.2f)\n", 
              nrow(deg_filtered), fc_threshold, pval_threshold))
  cat(sprintf("  Upregulated in group1: %d\n", sum(deg_filtered$avg_log2FC > 0)))
  cat(sprintf("  Upregulated in group2: %d\n", sum(deg_filtered$avg_log2FC < 0)))
  
  # Save filtered results
  write.csv(deg_filtered,
            file = paste0(output_prefix, "_significant_genes.csv"),
            row.names = FALSE)
  cat(sprintf("Saved significant DEGs: %s_significant_genes.csv\n", output_prefix))
  
  # Pathway enrichment analysis using g:Profiler
  pathway_results <- NULL
  if (nrow(deg_filtered) > 0) {
    cat("\nPerforming pathway enrichment analysis with g:Profiler...\n")
    
    # Split into upregulated and downregulated genes
    up_genes <- deg_filtered %>% filter(avg_log2FC > 0) %>% pull(gene)
    down_genes <- deg_filtered %>% filter(avg_log2FC < 0) %>% pull(gene)
    
    pathway_results <- list()
    
    # Enrichment for upregulated genes (group1)
    if (length(up_genes) > 0) {
      cat(sprintf("  Analyzing %d upregulated genes (group1)...\n", length(up_genes)))
      gost_up <- gost(
        query = up_genes,
        organism = "mmulatta",  # Change to "mmusculus" for mouse
        significant = TRUE,
        correction_method = "g_SCS",  # g:Profiler's custom method
        sources = c("GO:BP", "GO:MF", "GO:CC", "KEGG", "REAC", "WP")
      )
      
      if (!is.null(gost_up$result)) {
        # Select available columns
        base_cols <- c("term_id", "term_name", "source", "p_value", "term_size", 
                       "query_size", "intersection_size")
        optional_cols <- c("intersection")
        available_cols <- base_cols[base_cols %in% colnames(gost_up$result)]
        optional_available <- optional_cols[optional_cols %in% colnames(gost_up$result)]
        
        pathway_results$upregulated <- gost_up$result %>%
          select(all_of(c(available_cols, optional_available))) %>%
          arrange(p_value)
        
        write.csv(pathway_results$upregulated,
                  file = paste0(output_prefix, "_pathways_group1_up.csv"),
                  row.names = FALSE)
        cat(sprintf("  Found %d enriched pathways for group1\n", 
                    nrow(pathway_results$upregulated)))
      } else {
        cat("  No significant pathways found for upregulated genes\n")
      }
    }
    
    # Enrichment for downregulated genes (group2)
    if (length(down_genes) > 0) {
      cat(sprintf("  Analyzing %d downregulated genes (group2)...\n", length(down_genes)))
      gost_down <- gost(
        query = down_genes,
        organism = "hsapiens",
        significant = TRUE,
        correction_method = "g_SCS",
        sources = c("GO:BP", "GO:MF", "GO:CC", "KEGG", "REAC", "WP")
      )
      
      if (!is.null(gost_down$result)) {
        # Select available columns
        base_cols <- c("term_id", "term_name", "source", "p_value", "term_size",
                       "query_size", "intersection_size")
        optional_cols <- c("intersection")
        available_cols <- base_cols[base_cols %in% colnames(gost_down$result)]
        optional_available <- optional_cols[optional_cols %in% colnames(gost_down$result)]
        
        pathway_results$downregulated <- gost_down$result %>%
          select(all_of(c(available_cols, optional_available))) %>%
          arrange(p_value)
        
        write.csv(pathway_results$downregulated,
                  file = paste0(output_prefix, "_pathways_group2_up.csv"),
                  row.names = FALSE)
        cat(sprintf("  Found %d enriched pathways for group2\n",
                    nrow(pathway_results$downregulated)))
      } else {
        cat("  No significant pathways found for downregulated genes\n")
      }
    }
  } else {
    cat("No significant DEGs found - skipping pathway analysis\n")
  }
  
  # Create volcano plot
  cat("\nGenerating volcano plot...\n")
  
  # Add significance category
  deg_results$significance <- "Not significant"
  deg_results$significance[deg_results$avg_log2FC > fc_threshold & 
                            deg_results$p_val_adj < pval_threshold] <- "Group1 enriched"
  deg_results$significance[deg_results$avg_log2FC < -fc_threshold & 
                            deg_results$p_val_adj < pval_threshold] <- "Group2 enriched"
  
  # Label top genes
  all_genes <- deg_results %>%
    filter(significance != "Not significant") %>%
    arrange(p_val_adj)
  
  volcano_plot <- ggplot(deg_results, aes(x = avg_log2FC, y = -log10(p_val_adj))) +
    geom_point(aes(color = significance), alpha = 0.6, size = 1.5) +
    scale_color_manual(
      values = c("Group1 enriched" = "#E64B35", 
                 "Not significant" = "grey70",
                 "Group2 enriched" = "#4DBBD5")
    ) +
    geom_vline(xintercept = c(-fc_threshold, fc_threshold), 
               linetype = "dashed", color = "grey30", linewidth = 0.5) +
    geom_hline(yintercept = -log10(pval_threshold), 
               linetype = "dashed", color = "grey30", linewidth = 0.5) +
    geom_text_repel(
      data = all_genes,
      aes(label = gene),
      size = 3,
      max.overlaps = 20,
      box.padding = 0.5,
      segment.size = 0.2
    ) +
    labs(
      title = "Differential Gene Expression: Group1 vs Group2",
      x = "log2 Fold Change",
      y = "-log10(Adjusted P-value)",
      color = "Significance"
    ) +
    theme_classic(base_size = 12) +
    theme(
      legend.position = "right",
      plot.title = element_text(hjust = 0.5, face = "bold"),
      axis.text = element_text(color = "black"),
      axis.title = element_text(face = "bold")
    )
  
  # Save plot
  ggsave(
    filename = paste0(output_prefix, "_volcano_plot.pdf"),
    plot = volcano_plot,
    width = 10,
    height = 8,
    units = "in"
  )
  
  ggsave(
    filename = paste0(output_prefix, "_volcano_plot.png"),
    plot = volcano_plot,
    width = 10,
    height = 8,
    units = "in",
    dpi = 300
  )
  
  cat(sprintf("Saved volcano plots: %s_volcano_plot.pdf/png\n", output_prefix))
  
  # Return results
  return(list(
    all_results = deg_results,
    significant_degs = deg_filtered,
    pathway_results = pathway_results,
    volcano_plot = volcano_plot,
    n_group1 = length(cells_group1),
    n_group2 = length(cells_group2),
    parameters = list(
      fc_threshold = fc_threshold,
      pval_threshold = pval_threshold,
      test_method = test_use
    )
  ))
}

```

```{r}
input.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS016/singleCell/Seurat/RDS/"
output.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/joint/subset/new/"

samples <- c("Invitro",
             "D195",
             "Pacute",
             "W2")
for (i in 1:length(samples)) {
  sample_name <- samples[i]
  input_file <- paste0(input.dir, sample_name, ".rds")
  
  tryCatch({
    seurat_obj <- readRDS(input_file)
    genes_to_exclude <- c("CD4", "CD8A", "CD8B")  # remove from DEG analysis
    genes_to_keep <- rownames(seurat_obj)[!rownames(seurat_obj) %in% genes_to_exclude]
    seurat_obj <- subset(seurat_obj, features = genes_to_keep)

    if (!dir.exists(output.dir)) {
      dir.create(output.dir, recursive = TRUE)
    }
    
    salve <- combined_data[[sample_name]]
    available_cols <- intersect(c("MS", "SS-MS", "SS"), colnames(salve))
    list_of_cells1 <- salve %>%
      filter(if_any(all_of(available_cols), ~ . != 0)) %>%
      select(cellID) %>%
      unlist()
    list_of_cells2 <- salve %>% filter(total != 0) %>%
      select(cellID) %>% filter(!(cellID %in% list_of_cells1)) %>% unlist()

    result <- analyze_cell_subset(
      seurat_obj,
      list_of_cells1,
      list_of_cells2,
      group_names = c("SIV_spliced", "SIV_other")
    )
    save_subset_analysis(result, paste0(output.dir, "combined_splicedvnot/", sample_name, "/"))
    
    
    cat("Processed sample:", sample_name, "\n")
  }, error = function(e) {
    cat("Error processing sample", sample_name, ":", conditionMessage(e), "\n\n")
  })
}
```



### cNMF

Pattern annotation
```{r}
# =============================================================================
# Configuration
# =============================================================================
load_dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS024/singleCell/cNMF/"
rds_dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS024/singleCell/Seurat/RDS/"
output_dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS024/joint/plots/cNMF/"
run_name <- "GFP_invitro"
selected_k <- 6

if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}

# =============================================================================
# 1. Load cNMF Outputs
# =============================================================================
usage_file <- file.path(load_dir, paste0(run_name, "_usage_K", selected_k, ".csv"))
spectra_file <- file.path(load_dir, paste0(run_name, "_spectra_K", selected_k, ".csv"))
spectra_tpm_file <- file.path(load_dir, paste0(run_name, "_spectra_tpm_K", selected_k, ".csv"))
top_genes_file <- file.path(load_dir, paste0(run_name, "_top_genes_K", selected_k, ".csv"))

organism <- "Macaca mulatta"
n_top_genes_ora <- 50
gsea_pval_cutoff <- 0.05
ora_pval_cutoff <- 0.05

usage <- read.csv(usage_file, row.names = 1, check.names = FALSE)
spectra <- read.csv(spectra_file, row.names = 1, check.names = FALSE)
spectra_tpm <- read.csv(spectra_tpm_file, row.names = 1, check.names = FALSE)
top_genes <- read.csv(top_genes_file, row.names = 1, check.names = FALSE)

#cat("Loaded cNMF outputs:\n")
#cat("  Usage matrix:", nrow(usage), "cells x", ncol(usage), "patterns\n")
#cat("  Spectra (Z-score) matrix:", nrow(spectra), "genes x", ncol(spectra), "patterns\n")
#cat("  Spectra (TPM) matrix:", nrow(spectra_tpm), "genes x", ncol(spectra_tpm), "patterns\n")

cat("\nTop 20 genes per pattern (by Z-score loading):\n")
for (pat in colnames(spectra)) {
  pat_scores <- spectra[[pat]]
  names(pat_scores) <- rownames(spectra)
  top_20 <- names(sort(pat_scores, decreasing = TRUE))[1:20]
  cat(paste0("  ", pat, ": ", paste(top_20, collapse = ", "), "\n"))
}

# =============================================================================
# 2. Build Gene Set Collections from MSigDB
# =============================================================================
hallmark_sets <- msigdbr(species = organism, collection = "H") %>%
  split(x = .$gene_symbol, f = .$gs_name)

reactome_sets <- msigdbr(species = organism, collection = "C2", subcollection = "CP:REACTOME") %>%
  split(x = .$gene_symbol, f = .$gs_name)

gobp_sets <- msigdbr(species = organism, collection = "C5", subcollection = "GO:BP") %>%
  split(x = .$gene_symbol, f = .$gs_name)

#cat("\nGene set collections loaded:\n")
#cat("  Hallmark:", length(hallmark_sets), "sets\n")
#cat("  Reactome:", length(reactome_sets), "sets\n")
#cat("  GO:BP:", length(gobp_sets), "sets\n")

# =============================================================================
# 3. Preranked GSEA per Pattern
# =============================================================================
# gene_spectra_score Z-scores reflect expected change in gene expression per
# unit increase in pattern usage. fgsea uses the full ranked vector per pattern
# rather than a fixed top-N cutoff (Kotliar et al., eLife 2019, Methods).s

cat("\nRunning preranked GSEA (Hallmark)...\n")
suppressWarnings(
  gsea_hallmark <- run_preranked_gsea(spectra, hallmark_sets, gsea_pval_cutoff)
)

cat("Running preranked GSEA (Reactome)...\n")
suppressWarnings(
  gsea_reactome <- run_preranked_gsea(spectra, reactome_sets, gsea_pval_cutoff)
)

cat("Running preranked GSEA (GO:BP)...\n")
suppressWarnings(
  gsea_gobp <- run_preranked_gsea(spectra, gobp_sets, gsea_pval_cutoff)
)

cat("\nTop 5 Hallmark terms per pattern (by NES):\n")
for (pat in colnames(spectra)) {
  pat_res <- gsea_hallmark[gsea_hallmark$pattern == pat, ]
  if (nrow(pat_res) == 0) {
    cat(paste0("  ", pat, ": no significant terms\n"))
    next
  }
  pat_res <- pat_res[order(abs(pat_res$NES), decreasing = TRUE), ]
  top_terms <- head(pat_res, 5)
  for (j in seq_len(nrow(top_terms))) {
    direction <- ifelse(top_terms$NES[j] > 0, "+", "-")
    term <- gsub("^HALLMARK_", "", top_terms$pathway[j])
    cat(sprintf("  %s: [%s] %s (NES=%.2f, padj=%.1e)\n",
                pat, direction, term, top_terms$NES[j], top_terms$padj[j]))
  }
}

# =============================================================================
# 4. ORA per Pattern (Top 50 Genes) -- Hypergeometric Test
# =============================================================================

top50_per_pattern <- get_top_n_genes(spectra, n = n_top_genes_ora)
background <- rownames(spectra)

cat("\nRunning ORA on top", n_top_genes_ora, "genes per pattern (GO:BP)...\n")
ora_results <- list()
for (pat in names(top50_per_pattern)) {
  ora_results[[pat]] <- run_ora_hypergeometric(
    query_genes = top50_per_pattern[[pat]],
    gene_sets = gobp_sets,
    background_genes = background,
    pval_cutoff = ora_pval_cutoff
  )
}

cat("\nTop 5 ORA GO:BP terms per pattern:\n")
for (pat in names(ora_results)) {
  res <- ora_results[[pat]]
  if (nrow(res) == 0) {
    cat(paste0("  ", pat, ": no significant terms\n"))
    next
  }
  top_terms <- head(res, 5)
  for (j in seq_len(nrow(top_terms))) {
    cat(sprintf("  %s: %s (padj=%.1e, %d/%d genes)\n",
                pat, gsub("^GOBP_", "", top_terms$pathway[j]),
                top_terms$padj[j], top_terms$overlap[j], n_top_genes_ora))
  }
}

# =============================================================================
# 5. Visualization
# =============================================================================

# --- 5a. GSEA Hallmark NES heatmap ---
if (nrow(gsea_hallmark) > 0) {
  nes_matrix <- gsea_hallmark %>%
    select(pattern, pathway, NES) %>%
    mutate(pathway = gsub("^HALLMARK_", "", pathway)) %>%
    pivot_wider(names_from = pattern, values_from = NES, values_fill = 0) %>%
    tibble::column_to_rownames("pathway")

  nes_mat <- as.matrix(nes_matrix)
  max_abs <- max(abs(nes_mat), na.rm = TRUE)

  p_hallmark <- pheatmap::pheatmap(
    nes_mat,
    color = colorRampPalette(c("blue", "white", "red"))(100),
    breaks = seq(-max_abs, max_abs, length.out = 101),
    cluster_cols = TRUE,
    cluster_rows = TRUE,
    fontsize_row = 7,
    fontsize_col = 10,
    main = "GSEA Hallmark NES per cNMF Pattern",
    silent = TRUE
  )

  ggsave(
    file.path(output_dir, "cNMF_gsea_hallmark_heatmap.pdf"),
    plot = p_hallmark,
    width = 10,
    height = max(6, nrow(nes_mat) * 0.25)
  )
  cat("\nSaved Hallmark GSEA heatmap.\n")
}

# --- 5b. Top genes dot plot per pattern ---
n_show <- 20
plot_data <- list()
for (pat in colnames(spectra)) {
  scores <- spectra[[pat]]
  names(scores) <- rownames(spectra)
  top_g <- names(sort(scores, decreasing = TRUE))[1:n_show]

  plot_data[[pat]] <- data.frame(
    gene = factor(top_g, levels = rev(top_g)),
    pattern = pat,
    z_score = spectra[top_g, pat],
    tpm = spectra_tpm[top_g, pat],
    stringsAsFactors = FALSE
  )
}
plot_df <- do.call(rbind, plot_data)

p_dotplot <- ggplot(plot_df, aes(x = pattern, y = gene, color = z_score, size = tpm)) +
  geom_point() +
  scale_color_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0,
                        name = "Z-score") +
  scale_size_continuous(name = "TPM loading", range = c(0.5, 4)) +
  facet_wrap(~pattern, scales = "free_y", nrow = 1) +
  theme_minimal(base_size = 9) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    strip.text = element_text(size = 10, face = "bold"),
    panel.grid.major.x = element_blank()
  ) +
  labs(x = NULL, y = NULL, title = "Top 20 genes per cNMF pattern")

ggsave(
  file.path(output_dir, "cNMF_top_genes_dotplot.pdf"),
  plot = p_dotplot,
  width = 3 * selected_k,
  height = 8
)
cat("Saved top genes dot plot.\n")

# -- plot GSEA --

n_top_gsea_per_collection <- 5
for (pat in colnames(spectra)) {
  
  combined_list <- list()
  
  for (collection_name in c("Hallmark", "Reactome", "GO:BP")) {
    
    gsea_df <- switch(collection_name,
      "Hallmark" = gsea_hallmark,
      "Reactome" = gsea_reactome,
      "GO:BP"    = gsea_gobp
    )
    
    if (is.null(gsea_df) || nrow(gsea_df) == 0) next
    
    pat_df <- gsea_df[gsea_df$pattern == pat, ]
    if (nrow(pat_df) == 0) next
    
    prefix_pattern <- switch(collection_name,
      "Hallmark" = "^HALLMARK_",
      "Reactome" = "^REACTOME_",
      "GO:BP"    = "^GOBP_"
    )
    
    pat_df$pathway_clean <- gsub(prefix_pattern, "", pat_df$pathway)
    pat_df$pathway_clean <- gsub("_", " ", pat_df$pathway_clean)
    pat_df$pathway_clean <- tolower(pat_df$pathway_clean)
    pat_df$pathway_clean <- paste0(toupper(substring(pat_df$pathway_clean, 1, 1)),
                                    substring(pat_df$pathway_clean, 2))
    pat_df$pathway_clean <- ifelse(nchar(pat_df$pathway_clean) > 50,
                                    paste0(substr(pat_df$pathway_clean, 1, 47), "..."),
                                    pat_df$pathway_clean)
    
    # Top N by |NES| per collection
    pat_df <- pat_df[order(abs(pat_df$NES), decreasing = TRUE), ]
    pat_df <- head(pat_df, n_top_gsea_per_collection)
    pat_df$collection <- collection_name
    
    combined_list[[collection_name]] <- pat_df
  }
  
  if (length(combined_list) == 0) next
  combined_df <- do.call(rbind, combined_list)
  
  combined_df$neg_log10_padj <- -log10(combined_df$padj)
  
  # Set collection as ordered factor for grouping
  combined_df$collection <- factor(combined_df$collection,
                                    levels = c("Hallmark", "Reactome", "GO:BP"))
  
  # Within each collection, order by NES
  combined_df <- combined_df[order(combined_df$collection, combined_df$NES), ]
  
  # Create unique y labels (in case of duplicate pathway names across collections)
  combined_df$y_label <- make.unique(combined_df$pathway_clean, sep = " ")
  combined_df$y_label <- factor(combined_df$y_label, levels = combined_df$y_label)
  
  # Symmetric NES axis
  max_abs_nes <- max(abs(combined_df$NES)) * 1.05
  
  p <- ggplot(combined_df, aes(x = NES, y = y_label, fill = NES, size = neg_log10_padj)) +
    geom_point(shape = 21, stroke = 0.3) +
    scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B", midpoint = 0,
                         name = "NES") +
    scale_size_continuous(name = expression(-log[10]~p[adj]), range = c(2, 7)) +
    scale_x_continuous(limits = c(-max_abs_nes, max_abs_nes)) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey40", linewidth = 0.3) +
    facet_grid(collection ~ ., scales = "free_y", space = "free_y", switch = "y") +
    theme_minimal(base_size = 10) +
    theme(
      panel.grid.major.y = element_blank(),
      panel.grid.minor = element_blank(),
      axis.text.y = element_text(size = 9),
      strip.text.y.left = element_text(size = 10, face = "bold", angle = 0),
      strip.placement = "outside",
      plot.title = element_text(size = 11, face = "bold"),
      legend.position = "right"
    ) +
    labs(
      x = "Normalized Enrichment Score (NES)",
      y = NULL,
      title = paste0(pat, " \u2014 GSEA")
    )
  
  n_terms <- nrow(combined_df)
  ggsave(
    file.path(output_dir, paste0("cNMF_gsea_combined_", pat, ".pdf")),
    plot = p,
    width = 9,
    height = max(4, n_terms * 0.32 + 2)
  )
}
cat("Saved combined GSEA dot plots per pattern.\n")

# Plot ORA
n_top_ora <- 10

for (pat in names(ora_results)) {
  res <- ora_results[[pat]]
  if (is.null(res) || nrow(res) == 0) next
  
  res$pathway_clean <- gsub("^GOBP_", "", res$pathway)
  res$pathway_clean <- gsub("_", " ", res$pathway_clean)
  res$pathway_clean <- tolower(res$pathway_clean)
  res$pathway_clean <- paste0(toupper(substring(res$pathway_clean, 1, 1)),
                               substring(res$pathway_clean, 2))
  res$pathway_clean <- ifelse(nchar(res$pathway_clean) > 50,
                               paste0(substr(res$pathway_clean, 1, 47), "..."),
                               res$pathway_clean)
  
  res$gene_ratio <- res$overlap / res$query_size
  res$neg_log10_padj <- -log10(res$padj)
  
  # Top N by padj
  res <- head(res[order(res$padj), ], n_top_ora)
  
  # Order by gene ratio for plotting
  res <- res[order(res$gene_ratio), ]
  res$pathway_clean <- make.unique(res$pathway_clean, sep = " ")
  res <- res[order(res$gene_ratio), ]
  res$pathway_clean <- factor(res$pathway_clean, levels = res$pathway_clean)
  
  p <- ggplot(res, aes(x = gene_ratio, y = pathway_clean, size = overlap, color = neg_log10_padj)) +
    geom_point() +
    scale_color_viridis_c(option = "inferno", direction = -1, end = 0.85,
                          name = expression(-log[10]~p[adj])) +
    scale_size_continuous(name = "Overlap", range = c(2, 7)) +
    theme_minimal(base_size = 10) +
    theme(
      panel.grid.major.y = element_blank(),
      panel.grid.minor = element_blank(),
      axis.text.y = element_text(size = 9),
      plot.title = element_text(size = 11, face = "bold"),
      legend.position = "right"
    ) +
    labs(
      x = "Gene ratio (overlap / query size)",
      y = NULL,
      title = paste0(pat, " \u2014 ORA GO:BP (top ", n_top_genes_ora, " genes)")
    )
  
  ggsave(
    file.path(output_dir, paste0("cNMF_ora_gobp_", pat, ".pdf")),
    plot = p,
    width = 8,
    height = max(3, nrow(res) * 0.35 + 1)
  )
}
cat("Saved ORA GO:BP dot plots per pattern.\n")

# --- 5c. Save results ---
collapse_leading_edge <- function(df) {
  if ("leadingEdge" %in% colnames(df) && is.list(df$leadingEdge)) {
    df$leadingEdge <- vapply(df$leadingEdge, paste, collapse = ";", FUN.VALUE = character(1))
  }
  df
}

write.csv(collapse_leading_edge(gsea_hallmark), file.path(output_dir, "gsea_hallmark_results.csv"), 
          row.names = FALSE)
write.csv(collapse_leading_edge(gsea_reactome), file.path(output_dir, "gsea_reactome_results.csv"), 
          row.names = FALSE)
write.csv(collapse_leading_edge(gsea_gobp), file.path(output_dir, "gsea_gobp_results.csv"), 
          row.names = FALSE)

ora_combined <- do.call(rbind, lapply(names(ora_results), function(pat) {
  df <- ora_results[[pat]]
  if (nrow(df) > 0) df$pattern <- pat
  df
}))
if (!is.null(ora_combined) && nrow(ora_combined) > 0) {
  write.csv(ora_combined, file.path(output_dir, "ora_gobp_results.csv"), row.names = FALSE)
}

top50_df <- data.frame(
  lapply(top50_per_pattern, function(x) { length(x) <- n_top_genes_ora; x }),
  check.names = FALSE
)
write.csv(top50_df, file.path(output_dir, "top50_genes_per_pattern.csv"), row.names = FALSE)

cat("All annotation outputs saved to:", output_dir, "\n")

# Plot cNMF pattern UMAP
seurat_obj <- readRDS(paste0(rds_dir, run_name, ".rds"))
seurat_obj <- add_cnmf_to_seurat(seurat_obj, usage)
fplot <- FeaturePlot(seurat_obj, features = paste0("Pattern_", 1:selected_k), ncol = 4) &
scale_color_viridis_c()
ggsave(fplot, file = paste0(output_dir, run_name, "_umap_pattern_weight.pdf"),
       width = 18, height = 8)



# for (i in 1:length(samples)) {
#   run_name <- samples[i]
#   selected_k <- selected_ks[i]
#   cat("Loading: ",run_name, "\n")
# 
#   usage_file <- file.path(load_dir, paste0(run_name, "_usage_K", selected_k, ".csv"))
#   usage <- read.csv(usage_file, row.names = 1, check.names = FALSE)
#   
#   # Plot cNMF pattern UMAP
#   rds_file <- list.files(rds_dir, pattern = paste0("^", run_name, "\\.rds$"), 
#                          ignore.case = TRUE, full.names = TRUE)
#   seurat_obj <- readRDS(rds_file[1])
#   seurat_obj <- add_cnmf_to_seurat(seurat_obj, usage)
#   
#   pattern_cols <- paste0("Pattern_", 1:selected_k)
#   keep_patterns <- 1:4
#   exclude_patterns <- setdiff(1:selected_k, keep_patterns)
# 
#   dominant_pattern <- apply(seurat_obj@meta.data[, pattern_cols], 1, which.max)
#   seurat_sub <- subset(seurat_obj, cells = colnames(seurat_obj)[dominant_pattern %in% keep_patterns])
# 
#   cat("Cells before:", ncol(seurat_obj), "\n")
#   cat("Cells after:", ncol(seurat_sub), "\n")
# 
#   saveRDS(seurat_sub, file = paste0(rds_dir, run_name, "_pattern1.4.rds"))
#   sce <- SingleCellExperiment(
#     assays = list(counts = GetAssayData(seurat_sub, layer = "counts")),
#     colData = seurat_sub@meta.data
#   )
#   suppressMessages(
#     writeH5AD(sce, file = paste0(rds_dir, run_name, "_pattern1.4.h5ad"))
#   )
# }


```

Pattern markers for drug targets
```{r}
# =============================================================================
# Pattern Marker Identification
# =============================================================================
# Step 1: Pattern markers weighted for druggability
result <- find_pattern_markers(
  spectra_tpm = spectra_tpm,
  spectra_score = spectra,
  n_markers = 100,
  method = "rank_product",
  min_score = 5,             # require meaningful expression
  exclude_shared = TRUE,     # remove multi-pattern genes
  shared_threshold = 2.0     # require 2x enrichment over next-best
)

# Step 2: Filter against a druggability annotation
# (e.g., DGIdb categories, surface proteins, kinases)
# druggable_genes <- read.csv("dgidb_druggable_genome.csv")$gene
# for (pat in names(result$markers)) {
#   targetable <- intersect(result$markers[[pat]], druggable_genes)
#   cat(sprintf("  %s: %d druggable markers out of %d\n",
#               pat, length(targetable), length(result$markers[[pat]])))
# }

# Print top 20 markers per pattern:
markers <- result$markers
for (pat in names(markers)) {
  cat(sprintf("  %s: %s\n", pat, paste(head(markers[[pat]], 20), collapse = ", ")))
}
```


SIV+ vs SIV-
```{r}
# whole sample
load_dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS016/singleCell/cNMF/"
joint_dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/joint/dataframes/"
samples <- c("invitro", "D195", "Pacute")
dfs <- c("Invitro", "D195", "Pacute")
selected_ks <- c(7, 8, 8)
# subpatterns
load_dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS016/singleCell/cNMF/subpattern/"
joint_dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/joint/dataframes/"
samples <- c("invitro_pattern1.4", "D195_pattern1.4", "Pacute_pattern1.4")
dfs <- c("Invitro", "D195", "Pacute")
selected_ks <- c(5, 6, 6)

for (i in 1:length(samples)) {
  run_name <- samples[i]
  df_name <- dfs[i]
  selected_k <- selected_ks[i]
  output_dir <- paste0("/projects/b1042/GoyalLab/egrody/extractedData/EGS016/joint/plots/cNMF/", 
                       run_name, "/")
  if (!dir.exists(output_dir)) {
    dir.create(output_dir, recursive = TRUE)
  }
  cat("Loading: ", run_name, "\n")
  
  usage_file <- file.path(load_dir, paste0(run_name, "_usage_K", selected_k, ".csv"))
  usage <- read.csv(usage_file, row.names = 1, check.names = FALSE)
  
  both <- read.csv(paste0(joint_dir, df_name, "_both.csv"))
  GEXonly <- read.csv(paste0(joint_dir, df_name, "_onlySingleCell.csv"))
  SALVEonly <- read.csv(paste0(joint_dir, df_name, "_onlySALVE.csv"))
  all <- read.csv(paste0(joint_dir, df_name, "_all.csv"))
  
  infected <- rbind(SALVEonly, both)
  infected_cellids <- infected$cellID
  comparison <- compare_pattern_usage(usage, infected_cellids)
  
  plot_pattern_violins(comparison, file.path(output_dir, "cnmf_infected_violins.pdf"))
  plot_log2fc(comparison, file.path(output_dir, "cnmf_infected_log2fc.pdf"))
  #plot_marker_heatmap(spectra, markers, n_genes = 20, file.path(output_dir, "cnmf_marker_heatmap.pdf"))
}


```

splice+ vs splice-
```{r}
# whole sample
load_dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS016/singleCell/cNMF/"
joint_dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/joint/dataframes/"
samples <- c("invitro", "D195", "Pacute")
dfs <- c("Invitro", "D195", "Pacute")
selected_ks <- c(7, 8, 8)
# subpatterns
load_dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS016/singleCell/cNMF/subpattern/"
joint_dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/joint/dataframes/"
samples <- c("invitro_pattern1.4", "D195_pattern1.4", "Pacute_pattern1.4")
dfs <- c("Invitro", "D195", "Pacute")
selected_ks <- c(5, 6, 6)

for (i in 1:length(samples)) {
  run_name <- samples[i]
  df_name <- dfs[i]
  selected_k <- selected_ks[i]
  output_dir <- paste0("/projects/b1042/GoyalLab/egrody/extractedData/EGS016/joint/plots/cNMF/", 
                       run_name, "/")
  if (!dir.exists(output_dir)) {
    dir.create(output_dir, recursive = TRUE)
  }
  cat("Loading: ", run_name, "\n")
  
  usage_file <- file.path(load_dir, paste0(run_name, "_usage_K", selected_k, ".csv"))
  usage <- read.csv(usage_file, row.names = 1, check.names = FALSE)
  
  both <- read.csv(paste0(joint_dir, df_name, "_both.csv"))
  GEXonly <- read.csv(paste0(joint_dir, df_name, "_onlySingleCell.csv"))
  SALVEonly <- read.csv(paste0(joint_dir, df_name, "_onlySALVE.csv"))
  all <- read.csv(paste0(joint_dir, df_name, "_all.csv"))
  
  infected <- rbind(SALVEonly, both)
  infected_cellids <- infected$cellID
  splicers <- infected %>%
    filter(SS > 0 | MS > 0)
  splicers_cellids <- splicers$cellID
  nonsplicers_cellids <- setdiff(infected_cellids, splicers_cellids)
  
  if (length(splicers_cellids) == 0) {
    cat("Skipping", run_name, "-- no cells SIV+splice+\n")
    next
  }
  
  comparison <- compare_pattern_usage(usage, splicers_cellids, nonsplicers_cellids, 
                                      group_labels = c("splicers", "nonsplicers"))
  
  plot_pattern_violins(comparison, file.path(output_dir, "cnmf_splicing_violins.pdf"), 
                       group_names = c("splicers", "nonsplicers"))
  plot_log2fc(comparison, file.path(output_dir, "cnmf_splicing_log2fc.pdf"),
              group_names = c("splicers", "nonsplicers"))
  #plot_marker_heatmap(spectra, markers, n_genes = 20, file.path(output_dir, "cnmf_marker_heatmap.pdf"))
}


```


