---
title: "EGS009SALVEseq"
author: "Emanuelle Grody"
date: "2024-Winter"
output: html_document
---


```{r, echo = FALSE}
source("~/SALVEseq/packages.R")
source("~/SALVEseq/functions.R")
```

# SALVE Analysis

### Loading in SALVEseq 

```{r}
#output_dir = "/projects/b1042/GoyalLab/egrody/20241127_EGS009/analysis/Seurat/GEX_Mmulmac/"
input_root = "/projects/b1042/GoyalLab/egrody/20241127_EGS009/analysis/counts/"


objects <- list()
expression_data_list <- list()
tissues <- c("P","L","G")
conditions <- c("acute", "ART")
viral_genes <- c("pol", "SSenv", "nef")

for (i in tissues) {
  for (j in conditions) {
    for (k in viral_genes) {
      sample_name <- paste(i, j, k, sep = "_")
      input_dir <- paste0(input_root, "run_count_", sample_name, "/outs/filtered_feature_bc_matrix/")
      
      result <- try({
        objects[[sample_name]] <- Read10X(data.dir = input_dir) %>% CreateSeuratObject()
        
        viral_gene <- if(k == "SSenv") "env" else k
        
        # Get counts matrix and sum all counts for each cell
        counts_matrix <- GetAssayData(objects[[sample_name]], slot = "counts")
        total_counts <- Matrix::colSums(counts_matrix)
        
        temp_data <- data.frame(
          cell_id = names(total_counts),
          count = as.numeric(total_counts),
          gene = viral_gene,
          sample = paste(i, j, sep = "_")
        )
        
        expression_data_list[[sample_name]] <- temp_data
      }, silent = TRUE)
    }
  }
}


comparecells_SALVE <- bind_rows(expression_data_list) %>%
  pivot_wider(
    id_cols = c(cell_id, sample),
    names_from = gene,
    values_from = count,
    values_fill = 0
  ) %>%
  mutate(across(c(pol, env, nef), ~ifelse(. <= 1, 0, .))) %>%
  filter(!(pol == 0 & env == 0 & nef == 0))

output_dir <- "/projects/b1042/GoyalLab/egrody/20241127_EGS009/analysis/joint/"
file_path <- paste0(output_dir, "comparecells_SALVE.csv")
write.csv(comparecells_SALVE, file_path)
```

### W2 and D13

```{r}
input_root = "/projects/b1042/GoyalLab/egrody/20241127_EGS009/analysis/counts/"
output_dir <- "/projects/b1042/GoyalLab/egrody/20241127_EGS009/analysis/joint/"

conditions <- c("W2_acute", "D13_acute")
viral_genes <- c("pol", "SSenv", "nef")

for (j in conditions) {
  # Reset list for each condition
  objects <- list()
  expression_data_list <- list()
  
  # Get the short name (W2 or D13) for file naming
  condition_short <- strsplit(j, "_")[[1]][1]
  
  for (k in viral_genes) {
    sample_name <- paste(j, k, sep = "_")
    input_dir <- paste0(input_root, "run_count_", sample_name, "/outs/filtered_feature_bc_matrix/")
    
    result <- try({
      objects[[sample_name]] <- Read10X(data.dir = input_dir) %>% CreateSeuratObject()
      
      viral_gene <- if(k == "SSenv") "env" else k
      
      counts_matrix <- GetAssayData(objects[[sample_name]], slot = "counts")
      total_counts <- Matrix::colSums(counts_matrix)
      
      temp_data <- data.frame(
        cell_id = names(total_counts),
        count = as.numeric(total_counts),
        gene = viral_gene,
        sample = condition_short 
      )
      
      expression_data_list[[sample_name]] <- temp_data
    }, silent = TRUE)
  }
  
  comparecells_SALVE <- bind_rows(expression_data_list) %>%
    pivot_wider(
      id_cols = c(cell_id, sample),
      names_from = gene,
      values_from = count,
      values_fill = 0
    ) %>%
    mutate(across(c(pol, env, nef), ~ifelse(. <= 1, 0, .))) %>%
    filter(!(pol == 0 & env == 0 & nef == 0))
  
  file_path <- paste0(output_dir, "comparecells_", condition_short, ".csv")
  write.csv(comparecells_SALVE, file = file_path, row.names = FALSE)
}

```

## Deconvolve UMIs

Ran the deconvolve_umis.py script for each sample from P_acute using the list of barcodes from above.

### Reading back in
```{r}
input_dir <- "/projects/b1042/GoyalLab/egrody/20241127_EGS009/analysis/deconvolve/D13/"
genes_umis <- list()
for (k in viral_genes) {
  genes_umis[[k]] <- read.csv(paste0(input_dir, "UMIs_", k, ".csv"))
}
```

Are there UMIs in common to deconvolve?
```{r}
common_barcodes <- intersect(intersect(genes_umis[[1]]$barcode, genes_umis[[2]]$barcode), genes_umis[[3]]$barcode)

# For each common barcode, check for duplicate UMIs
for(barcode in common_barcodes) {
  # Get UMIs for this barcode from each file
  umis1 <- genes_umis[[1]]$umi[genes_umis[[1]]$barcode == barcode]
  umis2 <- genes_umis[[2]]$umi[genes_umis[[2]]$barcode == barcode]
  umis3 <- genes_umis[[3]]$umi[genes_umis[[3]]$barcode == barcode]
  
  # Find UMIs that appear in multiple files
  duplicate_umis <- intersect(intersect(umis1, umis2), umis3)
  
  if(length(duplicate_umis) > 0) {
    cat("Barcode:", barcode, "\n")
    cat("Duplicate UMIs:", paste(duplicate_umis, collapse=", "), "\n\n")
  }
}
```
No there aren't.

# Joint Analysis
## Current Analysis
### Loading data

```{r}
input_dir <- "/projects/b1042/GoyalLab/egrody/20241127_EGS009/analysis/joint/"
file_path <- paste0(output_dir, "comparecells_GEX_Mmulmac.csv")
comparecells_GEX <- read.csv(file_path)
file_path <- paste0(output_dir, "comparecells_SALVE.csv")
comparecells_SALVE <- read.csv(file_path)
```

### Joining data

```{r}
comparecells_SALVE <- comparecells_SALVE %>% filter(sample == "P_acute") %>% #they all are
  select(-sample)
joint <- inner_join(comparecells_SALVE, comparecells_GEX, by = "cell_id")
output_dir <- "/projects/b1042/GoyalLab/egrody/20241127_EGS009/analysis/joint/"
file_path <- paste0(output_dir, "comparecells_joint.csv")
write.csv(joint, file_path)
```



# Future
### Linking to 10X
To see what are in common, we use left_join to look at only the cells with 10X anchors:
```{r}
env_paint_umap = left_join(SingleCell_d195_umap_env, SALVE_d195, by = "cellID")
env_paint_umap[is.na(env_paint_umap)] <- 0

left_both <- sum((env_paint_umap$log1pSingleCell != 0 & env_paint_umap$log1pSALVE != 0), na.rm = TRUE)
left_10Xonly <- sum(env_paint_umap$log1pSingleCell != 0, na.rm = TRUE) - left_both
left_SALVEonly <- sum(env_paint_umap$log1pSALVE != 0, na.rm = TRUE) - left_both
cat("\n\nenv\nTotal cells: \t\t", nrow(env_paint_umap),
  "\nCells in both: \t\t", left_both,
  "\nCells in 10X only: \t", left_10Xonly,
    "\nCells in SALVE only: \t", left_SALVEonly)

```

```{r}
input.dir <- "/projects/b1042/GoyalLab/egrody/20240116_VISER_SALVEseq/EGS004/scRNAseq/counts/cellranger_count_D195/outs/raw_feature_bc_matrix/"
d195_raw <- Read10X(input.dir)

# Calculate UMI counts per barcode
umi_counts <- colSums(d195_raw)

# Create a data frame with barcode and UMI count
plot_data <- data.frame(barcode = names(umi_counts), 
                        umi_count = umi_counts)

# Sort the data frame by UMI count in descending order
plot_data <- plot_data %>% 
  arrange(desc(umi_count)) %>%
  mutate(barcode = substr(barcode, 1, nchar(barcode) - 2))

#isolating out only the SALVE+ cellIDs
SALVE_cellids <- SALVE_d195 %>% select(cellID)
cat(sum(SALVE_cellids[[1]] %in% plot_data$barcode), "out of", nrow(SALVE_cellids), 
    "SALVEseq cellIDs are found in raw cellID list")

# Add a column to indicate whether each barcode should be highlighted
plot_data$highlight <- plot_data$barcode %in% SALVE_cellids[[1]]

# Add rank and handle zero/negative values
plot_data <- plot_data %>%
  arrange(desc(umi_count)) %>%
  mutate(
    rank = row_number(),
    umi_count_adj = ifelse(umi_count <= 0, 0.1, umi_count)  # Replace 0 or negative with 0.1
  )

# Create the plot
umi_rank_plot <- ggplot(plot_data %>% arrange(highlight), aes(x = rank, y = umi_count_adj)) +
  geom_point(aes(color = highlight), alpha = 0.6) +
  scale_x_log10(
    labels = scales::label_number(),
    breaks = scales::breaks_log(n = 6)
  ) +
  scale_y_log10(
    labels = scales::label_number(),
    breaks = scales::breaks_log(n = 6)
  ) +
  scale_color_manual(values = c("FALSE" = "grey", "TRUE" = "red")) +
  labs(x = "Rank of Barcode", 
       y = "UMI Count (Adjusted)",
       title = "UMI Count vs. Barcode Rank (Log-Scaled Axes)") +
  theme_minimal() +
  theme(legend.position = "none")

umi_rank_plot

# Save the plot
#ggsave("/projects/b1042/GoyalLab/egrody/20240116_VISER_SALVEseq/EGS003/joint/barcodeRankPlot_total_env.png", umi_rank_plot, width = 10, height = 8)
```


### Correlation plots
```{r}

#  now let's repeat the correlation analysis above
jointTableFilter = inner_join(scViralQuant_env, PL_env_clean, by = "cellID") 
jointTableTat = inner_join(scViralQuant_tat, PL_tat1_clean, by = "cellID") 

#  mutate(log1pViserCount = log1p(VISERcount), log2ViserCount = log2(VISERcount), lnViserCount = log(VISERcount), normalizedViserCount = 1000*VISERcount/sum(VISERcount), logNormViser = log1p(normalizedViserCount), log1pSingleCell = log1p(SingleCellcount))


output.dir <- "/projects/b1042/GoyalLab/egrody/20240116_VISER_SALVEseq/EGS003/Seurat/"
cat("Correlation of env between 10X data and log1p SALVE data:", cor(jointTableFilter$log1pSingleCell, jointTableFilter$log1pSALVE))
corrplot <- ggplot(data = jointTableFilter, aes(x=log1pSALVE, y=log1pSingleCell)) +
  geom_point() + 
  xlim(0,7) + 
  ylim(0,7)
corrplot
ggsave(corrplot, file = paste0(output.dir, "PLenv_correlationplot.svg"))

cat("Correlation of tat between 10X data and log1p SALVE data:", cor(jointTableTat$log1pSingleCell, jointTableTat$log1pSALVE))
corrplot <- ggplot(data = jointTableTat, aes(x=log1pSALVE, y=log1pSingleCell)) +
  geom_point() +
  xlim(0,7) +
  ylim(0,7)
ggsave(corrplot, file = paste0(output.dir, "PLtat1_correlationplot.svg"))
```


### Paint UMAPs
```{r}

output.dir <- "/projects/b1042/GoyalLab/egrody/20240116_VISER_SALVEseq/EGS003/joint/"
umap_env_salve <- create_umap_plot(
  data = env_paint_umap,
  method = "SALVE",
  title = "env from SALVE",
  output_file_name = "umap_env_SALVE.svg",
  output_dir = output.dir
)
create_umap_plot(
  data = tat_paint_umap,
  method = "SALVE",
  title = "tat from SALVE",
  output_file_name = "umap_tat_SALVE.svg",
  output_dir = output.dir
)
create_umap_plot(
  data = env_paint_umap,
  method = "10X",
  title = "env from 10X",
  output_file_name = "umap_env_SingleCell.svg",
  output_dir = output.dir
)
create_umap_plot(
  data = tat_paint_umap,
  method = "10X",
  title = "env from 10X",
  output_file_name = "umap_env_SingleCell.svg",
  output_dir = output.dir
)
```

```{r betweenMethods UMAP}
create_identity_umap(
   data = env_paint_umap,
   title = "env between methods",
   output_file_name = "umap_env_comparison.svg",
   output_dir = output.dir
)

create_identity_umap(
   data = tat_paint_umap,
   title = "tat between methods",
   output_file_name = "umap_tat_comparison.svg",
   output_dir = output.dir
)
```

### Unanchored cells
Do the unanchored cells exist in the raw cellIDs from 10X?
```{r}
input.dir <- "/projects/b1042/GoyalLab/egrody/20231017_VISER/analysis/counts/Mmul_10_only/run_count_invitro/outs/raw_feature_bc_matrix_bbmap/"
invitro_raw <- Read10X(input.dir)

# Calculate UMI counts per barcode
umi_counts <- colSums(invitro_raw)

# Create a data frame with barcode and UMI count
plot_data <- data.frame(barcode = names(umi_counts), 
                        umi_count = umi_counts)

# Sort the data frame by UMI count in descending order
plot_data <- plot_data %>% 
  arrange(desc(umi_count)) %>%
  mutate(barcode = substr(barcode, 1, nchar(barcode) - 2))

#isolating out only the SALVE+ cellIDs
SALVE_cellids <- invitro_paint_umap_full %>% filter(log1pSALVE > 0) %>% select(cellID)
cat(sum(SALVE_cellids[[1]] %in% plot_data$barcode), "out of", nrow(SALVE_cellids), 
    "SALVEseq cellIDs are found in raw cellID list")

# Add a column to indicate whether each barcode should be highlighted
plot_data$highlight <- plot_data$barcode %in% SALVE_cellids[[1]]

# Add rank and handle zero/negative values
plot_data <- plot_data %>%
  arrange(desc(umi_count)) %>%
  mutate(
    rank = row_number(),
    umi_count_adj = ifelse(umi_count <= 0, 0.1, umi_count)  # Replace 0 or negative with 0.1
  )

# Create the plot
umi_rank_plot <- ggplot(plot_data %>% arrange(highlight), aes(x = rank, y = umi_count_adj)) +
  geom_point(aes(color = highlight), alpha = 0.6) +
  scale_x_log10(
    labels = scales::label_number(),
    breaks = scales::breaks_log(n = 6)
  ) +
  scale_y_log10(
    labels = scales::label_number(),
    breaks = scales::breaks_log(n = 6)
  ) +
  scale_color_manual(values = c("FALSE" = "grey", "TRUE" = "red")) +
  labs(x = "Rank of Barcode", 
       y = "UMI Count (Adjusted)",
       title = "UMI Count vs. Barcode Rank (Log-Scaled Axes)") +
  theme_minimal() +
  theme(legend.position = "none")

umi_rank_plot

# Save the plot
ggsave("/projects/b1042/GoyalLab/egrody/20240116_VISER_SALVEseq/EGS003/joint/barcodeRankPlot_total_env.svg", umi_rank_plot, width = 10, height = 8, device = "svg")
```