---
title: "EGS004scRNAseq"
author: "Emanuelle Grody"
date: "2024-01-30"
output: html_document
---

The goals of this scRNAseq analysis was to recover viral reads from the 10X data. I tried a couple of different strategies to accomplish this:
* Augmenting the Macaca mulatta reference genome with the SIVmac239 genomes
** mac239 genome as one gene "no individual genes"
** mac239 genome with each gene location annotated "all genes annotated"
* Recover viral reads from unaligned reads using scViralQuant package

The first strategy was used in the Previous Analysis section and will not be pursued moving forward. I found that the implementation of scViralQuant was easier and faster.

```{r, echo = FALSE}
source("~/SALVEseq/packages.R")
source("~/SALVEseq/functions.R")
```


# Single Cell Data

We take as an input the count matrix(es). To get from base calls to count matrices, run cellranger with custom reference Mmu.

## Previous: SPQ
### scViralQuant

For this analysis, the reference genome is regular macaque reference. Running the scPathoQuant package \href{https://github.com/galelab/scPathoQuant} creates a new supplemented feature barcode matrix by finding viral reads that were discarded during cellranger count alignment and adding them back to the feature matrix. We read in this updated feature matrix and perform all Seurat analysis steps without alteration.

```{r Seurat}

output_dir = "/projects/b1042/GoyalLab/egrody/20240116_VISER_SALVEseq/EGS004/scRNAseq/Seurat/filtered_scPathoQuant/"
d13 <- SeuratPipeline("/projects/b1042/GoyalLab/egrody/20240116_VISER_SALVEseq/EGS004/scRNAseq/counts/cellranger_count_D13/outs/filtered_feature_bc_matrix_scPathoQuant_bbmap/", "d13", output_dir = output_dir, plots = "FALSE")
d83 <- SeuratPipeline("/projects/b1042/GoyalLab/egrody/20240116_VISER_SALVEseq/EGS004/scRNAseq/counts/cellranger_count_D83/outs/filtered_feature_bc_matrix_scPathoQuant_bbmap/", "d83", output_dir = output_dir, plots = "FALSE")
d195 <- SeuratPipeline("/projects/b1042/GoyalLab/egrody/20240116_VISER_SALVEseq/EGS004/scRNAseq/counts/cellranger_count_D195/outs/filtered_feature_bc_matrix_scPathoQuant_bbmap/", "d195", output_dir = output_dir, plots = "FALSE")

output_dir = "/projects/b1042/GoyalLab/egrody/20240116_EGS003_004/EGS004/scRNAseq/Seurat/filtered/"
d0 <- SeuratPipeline("/projects/b1042/GoyalLab/egrody/20240116_EGS003_004/EGS004/scRNAseq/counts/Mmul_10/cellranger_count_D0/outs/filtered_feature_bc_matrix/", "d0", output_dir = output_dir, plots = "TRUE")
d13 <- SeuratPipeline("//projects/b1042/GoyalLab/egrody/20240116_EGS003_004/EGS004/scRNAseq/counts/Mmul_10/cellranger_count_D13/outs/filtered_feature_bc_matrix/", "d13", output_dir = output_dir, plots = "TRUE")
d83 <- SeuratPipeline("/projects/b1042/GoyalLab/egrody/20240116_EGS003_004/EGS004/scRNAseq/counts/Mmul_10/cellranger_count_D83/outs/filtered_feature_bc_matrix/", "d83", output_dir = output_dir, plots = "TRUE")
d195 <- SeuratPipeline("/projects/b1042/GoyalLab/egrody/20240116_EGS003_004/EGS004/scRNAseq/counts/Mmul_10/cellranger_count_D195/outs/filtered_feature_bc_matrix/", "d195", output_dir = output_dir, plots = "TRUE")
```

```{r}
FeaturePlot(d0, "KLRB1")
FeaturePlot(d13, "KLRB1")

KLRB1_d0 <- targetExpressionDF(d0, "KLRB1")
KLRB1_d13 <- targetExpressionDF(d13, "KLRB1")

cat(
  "D0 # cells KLRB1+:\t\t", sum(KLRB1_d0$KLRB1 > 0), "/", nrow(KLRB1_d0),
  "\nD0 sum total KLRB1 expression:\t", sum(KLRB1_d0$KLRB1),
  "\nD0 average KLRB1 expression:\t", sum(KLRB1_d0$KLRB1)/nrow(KLRB1_d0),
  "\n\nD13 # cells KLRB1+:\t\t", sum(KLRB1_d13$KLRB1 > 0), "/", nrow(KLRB1_d13),
  "\nD13 sum total KLRB1 expression:\t", sum(KLRB1_d13$KLRB1),
  "\nD13 average KLRB1 expression:\t", sum(KLRB1_d13$KLRB1)/nrow(KLRB1_d13)
)
```


### Without virus UMAP coords

```{r}
# Function to extract and export cell metadata from Seurat object
export_cell_metadata <- function(seurat_obj, filename = "cell_metadata.csv") {
  # Get cell embeddings (UMAP coordinates)
  umap_coords <- as.data.frame(Embeddings(seurat_obj, reduction = "umap"))
  colnames(umap_coords) <- c("UMAP1", "UMAP2")
  
  # Get cluster identities
  clusters <- as.data.frame(Idents(seurat_obj))
  colnames(clusters) <- "Cluster"
  
  # Combine data
  metadata <- data.frame(
    cellID = rownames(umap_coords),
    UMAP1 = umap_coords$UMAP1,
    UMAP2 = umap_coords$UMAP2,
    Cluster = clusters$Cluster
  )
  
  # Export to CSV
  write.csv(metadata, file = filename, row.names = FALSE)
  
  return(metadata)
}

cell_data <- export_cell_metadata(d13,                            "/projects/b1042/GoyalLab/egrody/20240116_VISER_SALVEseq/EGS004/scRNAseq/d13_umapclusters.csv")
```


### Alternative
If we are loading in the filtered data without viral gene annotation, we join the scPathoQuant output (not normalized?) to the cellIDs. 

```{r}
virusfreeUMAPscPathoQuant <- function(sample, input.directory, seurat_object, gene_list = NULL) {
  expression_file <- paste0(input.directory, sample, "/pathogen_al_gene_counts_mac239.csv")
  if (!file.exists(expression_file)) {
    stop("scPathoQuant output file does not exist: ", expression_file)
  }
  
  expression <- tryCatch({
    read.csv(expression_file)
  }, error = function(e) {
    stop("Error reading CSV file: ", e$message)
  })
  
  if (!is.data.frame(expression)) {
    stop("Expression data is not a dataframe")
  }
  
  if (ncol(expression) < 3) {
    stop("Expression data should have at least 3 columns: cell ID, gene, and count")
  }
  
  colnames(expression)[1:3] <- c("cellID", "Gene", "Count")
  if (!is.null(gene_list)) {
    expression <- expression[expression$Gene %in% gene_list, ]
    if (nrow(expression) == 0) {
      stop("No genes from the provided gene_list found in the expression data")
    }
  } else {
    gene_list <- unique(expression$Gene)
  }
  if (!requireNamespace("Seurat", quietly = TRUE)) {
    stop("Seurat package is required but not installed")
  }
  
  if (!"umap" %in% names(seurat_object@reductions)) {
    stop("UMAP reduction not found in Seurat object")
  }
  
  cell_metadata <- data.frame(
    cellID = colnames(seurat_object),
    UMAP1 = seurat_object@reductions$umap@cell.embeddings[,1],
    UMAP2 = seurat_object@reductions$umap@cell.embeddings[,2]
  )
  
  expression_wide <- reshape2::dcast(expression, cellID ~ Gene, value.var = "Count", fill = 0)
  combined_data <- merge(cell_metadata, expression_wide, by = "cellID", all.x = TRUE)
  
  combined_data[is.na(combined_data)] <- 0
  missing_genes <- setdiff(gene_list, colnames(combined_data))
  if (length(missing_genes) > 0) {
    combined_data[, missing_genes] <- 0
  }
  
  gene_cols <- setdiff(colnames(combined_data), c("cellID", "UMAP1", "UMAP2"))
  combined_data <- combined_data[, c("cellID", "UMAP1", "UMAP2", gene_cols)]
  
  return(combined_data)
}

output.dir <- "/projects/b1042/GoyalLab/egrody/20240116_VISER_SALVEseq/EGS004/scRNAseq/"
input.dir <- "/projects/b1042/GoyalLab/egrody/20240116_VISER_SALVEseq/EGS004/scRNAseq/scPathoQuant/"
gene_list <- c("5pU3", "5pR", "5pU5", "5pUTR",
               "gag", "pol", "vif", "vpx",
               "vpr", "tat", "rev", "env",
               "nef", "3pU3", "3pR", "3pU5")
combined_d13 <- virusfreeUMAPscPathoQuant("D13", input.dir, d13, gene_list)
write.csv(combined_d13, paste0(output.dir, "d13_virusfreeUMAPscPathoQuant.csv"))
combined_d83 <- virusfreeUMAPscPathoQuant("D83", input.dir, d83, gene_list)
write.csv(combined_d83, paste0(output.dir, "d83_virusfreeUMAPscPathoQuant.csv"))
combined_d195 <- virusfreeUMAPscPathoQuant("D195", input.dir, d195, gene_list)
write.csv(combined_d195, paste0(output.dir, "d195_virusfreeUMAPscPathoQuant.csv"))
```

### scPathoQuant UMAP coords
If we are loading in the scPathoQuant data, we next isolate the viral counts from the Seurat objects.
```{r output viral expression}

objects <- list(d13 = d13, d83 = d83, d195 = d195)
genes <- c("env", "tat", "mac239")
output_dir <- "/projects/b1042/GoyalLab/egrody/20240116_VISER_SALVEseq/EGS004/scRNAseq/"
save_scPQ_expression(
  objects, 
  genes, 
  output_dir = output_dir)

```

### Mmul10 UMAP painted with scViralQuant (skip)

I use the scViralQuant alignment information from the previous section, run Seurat on the non-scViralQuant original matrices, and join the two datasets by cellID.

```{r}
w2_painted <- SeuratPipeline("/projects/b1042/GoyalLab/egrody/20231017_VISER/counts/Mmul_10_only/run_count_W2/outs/raw_feature_bc_matrix/", "w2")

SingleCell_w2_painted <- sctargetExpressionDF(w2_painted, "env")
SingleCell_w2_painted_umap <- scUMAPViralCellID(w2_painted, SingleCell_w2)
```

Plotting.
```{r}
plotUMAP(data = SingleCell_w2_painted_umap,
         colorby = log1pSingleCell,
         title = "scViralQuant onto preclustered 10X, log counts",
         output_dir = output_dir,
         saveas = "scviralquant_preclustered10X_w2")
```


## Current Analysis
### cellranger Mmulmac

Current approach to getting viral reads from 10X data is to align to concatenated macaque + SIVmac239 genomes with full mac239 annotation ("Mmul_10_mac239annot" or "Mmul_10_mac239full").


```{r Seurat}

output_dir = "/projects/b1042/GoyalLab/egrody/20241209_EGS004/analysis/Seurat/plots/"
d13 <- SeuratPipeline("/projects/b1042/GoyalLab/egrody/20241209_EGS004/analysis/counts/run_count_EG_JK85_D13/outs/filtered_feature_bc_matrix/", "d13", output_dir = output_dir, plots = TRUE)

```

Next we isolate the viral counts from our Seurat object.
```{r targetExpressionDF}

output.dir <- "/projects/b1042/GoyalLab/egrody/20241209_EGS004/analysis/Seurat/"
SingleCell_USSSMS <- targetExpressionDF(d13, c("pol", "env", "nef"))
write.csv(SingleCell_env, paste0(output.dir, "d13_Mmulmac_env.csv"))
SingleCell_tat <- targetExpressionDF(d13, "tat")
write.csv(SingleCell_tat, paste0(output.dir, "d13_Mmulmac_tat.csv"))

cat(
  "# cells env+:", sum(SingleCell_env$Count > 0),
  "\n# cells tat+:", sum(SingleCell_tat$Count > 0)
)
```


