---
title: "EGS004VISER"
author: "Emanuelle Grody"
date: "2024-06-23"
output: html_document
---

This analysis is to be run after EGS004scRNAseq.Rmd.

```{r, echo = FALSE}
source("~/SALVEseq/packages.R")
source("~/SALVEseq/functions.R")
```


# invitro EGS002 v 004
```{r}
input.dir = "/projects/b1042/GoyalLab/egrody/20240116_VISER_SALVEseq/EGS004/seqIO/"
samples = c("invitro")
reads = data.frame()
for (sample in samples) {
  file_name <- paste0(input.dir, sample, "/", sample, "_shavedReads.txt")
   unique_data <- read.csv(file_name, header = TRUE) %>% 
    unique()
  unique_data$sample <- sample
  
  cat("Sample: ", sample, "\nNo. of unique targets: ", nrow(unique_data), "\n")
  reads <- bind_rows(reads,unique_data)
}

EGS004_invitro_full <- reads %>% filter(sample == "invitro") %>% select(-sample)

```

Looking at target with pairwise LV:
```{r}
set.seed(2059)
subsample1 = sample(EGS004_invitro_full$target,10000) #can lower to 5k to reduce run time
subsample2 = sample(EGS004_invitro_full$target,10000)
subsample3 = sample(EGS004_invitro_full$target,10000)
BarcodesLv1 = as.integer(stringdistmatrix(subsample1, method = "lv"))
BarcodesLv2 = as.integer(stringdistmatrix(subsample2, method = "lv"))
BarcodesLv3 = as.integer(stringdistmatrix(subsample3, method = "lv"))
lBarcodesLv = length(BarcodesLv1)

BarcodesLv = tibble(
  lvdist = c(BarcodesLv1, BarcodesLv2, BarcodesLv3),
  subsamNum = c(rep("subsamping1", lBarcodesLv), rep("subsamping2", lBarcodesLv), rep("subsamping3", lBarcodesLv)))

BarcodesLvHist <- BarcodesLv %>% group_by(subsamNum, lvdist) %>% summarise(length(lvdist)) %>%
  group_by(subsamNum) %>% mutate(totalNum = sum(`length(lvdist)`), fracLvDist = `length(lvdist)`/totalNum)

BarcodesLvHistPlot <- ggplot(BarcodesLvHist, aes(lvdist, fracLvDist)) +
  geom_bar(width = 0.5, stat = 'identity') +
  facet_wrap(facets = vars(subsamNum)) +
  theme_classic()
BarcodesLvHistPlot
```

Cheating and using the starcode consensus from EGS002 invitro (trimmed to 26bp): ACACTGCACTGCGAACCAGAGAAGGC
```{r}
# Add LV dist column
starcodeTarget <- "ACACTGCACTGCGAACCAGAGAAGGC"

EGS004_invitro_full$dist <- adist(EGS004_invitro_full$target, starcodeTarget)

# Plotting
plotme <- EGS004_invitro_full$dist
hist(plotme, breaks = 50, xlab = "Levenshtein Distance: recovered to consensus", main = "EGS004 invitro")

# Trimming at 6
EGS004_invitro_trim <- EGS004_invitro_full %>% filter(dist < 6) %>% select(-dist, -target) %>% unique() %>% count(cellID)

#whitelist
whitelist_location <- "/home/egy2296/packages/cellranger-7.2.0/lib/python/cellranger/barcodes/3M-february-2018.txt.gz"
whitelist <- as.list(readLines(whitelist_location))

EGS004_invitro_whitelist <- EGS004_invitro_trim %>% filter(cellID %in% whitelist) %>% filter(n > 1)

output.dir <- "/projects/b1042/GoyalLab/egrody/20240116_VISER_SALVEseq/EGS004/seqIO/cleaned/"
write.csv(EGS004_invitro_whitelist, paste0(output.dir, "invitro_EGS004_whitelist_uniqued.csv"))
```


Comparing to EGS002 scRNAseq list:
```{r}
input.dir = "/projects/b1042/GoyalLab/egrody/20231017_VISER/analysis/scViralQuant/outputs/"
scRNAinvitro <- read.csv(paste0(input.dir, "invitro_scViralQuantexpression.csv"))

invitro_scRNAEGS002_left = left_join(scRNAinvitro, VISER_invitro_whitelist, by = "cellID") %>% rename(VISERcount = n) %>% 
                    mutate(log2ViserCount = log2(VISERcount))
invitro_scRNAEGS002_left[is.na(invitro_scRNAEGS002_left)] <- 0

# how many cells from each method?
joint_scRNAEGS002 <- invitro_scRNAEGS002_left %>% filter(log2ViserCount != 0)
cat("env+ cells from EGS002 (anchored):", nrow(joint_scRNAEGS002))

invitro_scRNAEGS004_left = left_join(scRNAinvitro, EGS004_invitro_whitelist, by = "cellID") %>% rename(VISERcount = n) %>% 
                    mutate(log2ViserCount = log2(VISERcount))
invitro_scRNAEGS004_left[is.na(invitro_scRNAEGS004_left)] <- 0

# how many cells from each method?
joint_scRNAEGS004 <- invitro_scRNAEGS004_left %>% filter(log2ViserCount != 0)
cat("\nenv+ cells from EGS004 (anchored):", nrow(joint_scRNAEGS004))

full_join <- full_join(joint_scRNAEGS002, joint_scRNAEGS004, by = "cellID")
full_join[is.na(full_join)] <- 0
only_EGS002 <- full_join %>% filter(log2ViserCount.y == 0)
only_EGS004 <- full_join %>% filter(log2ViserCount.x == 0)
joint <- full_join %>% filter(log2ViserCount.x != 0 & log2ViserCount.y != 0)

#correlations between all three
cat("Correlation between seqIO and scPathoQuant env counts:", cor(scPQ_v_seqIO$counts, scPQ_v_seqIO$umi))
corrplot <- ggplot(data = scPQ_v_seqIO, aes(x=counts, y=umi)) +
  geom_point() + 
  labs(x = "seqIO counts", y = "scPathoQuant counts", title = "VISER D13 env+ cells") +
  theme_classic()
corrplot
```


# env Analysis
## Previous Analysis (skip)
```{r, eval = FALSE}
input.dir = "/projects/b1042/GoyalLab/egrody/20240116_VISER_SALVEseq/EGS004/seqIO/"
samples = c("VISER_D13", "VISER_D83", "VISER_D195")
reads = data.frame()
for (sample in samples) {
  file_name <- paste0(input.dir, sample, "/", sample, "_shavedReads.txt")
   unique_data <- read.csv(file_name, header = TRUE) %>% 
    unique()
  unique_data$sample <- sample
  
  cat("Sample: ", sample, "\nNo. of unique targets: ", nrow(unique_data), "\n")
  reads <- bind_rows(reads,unique_data)
}

VISER_D13 <- reads %>% filter(sample == "VISER_D13")
VISER_D83 <- reads %>% filter(sample == "VISER_D83")
VISER_D195 <- reads %>% filter(sample == "VISER_D195")

```
### Pairwise LV Dist Histogram
The first step is to see what the pairwise LV histogram looks like to see if we have more similar or dissimilar sequences.
```{r}
set.seed(2059)
#subsample1 = sample(VISER_D13$target,10000) #can lower to 5k to reduce run time
#subsample2 = sample(temp_data$target,10000)
#subsample3 = sample(temp_data$target,10000)
BarcodesLv1 = as.integer(stringdistmatrix(VISER_D195$target, method = "lv"))
lBarcodesLv = length(BarcodesLv1)

BarcodesLv = tibble(
  lvdist = BarcodesLv1,
  subsamNum = rep("D195", lBarcodesLv))


BarcodesLvHistPlot <- ggplot(BarcodesLv, aes(x = lvdist)) + 
  geom_histogram(aes(y = ..count../sum(..count..)), bins = 30) +
  scale_y_continuous(labels = scales::percent) +
  labs(y="Fractional Abundance") +
  theme_classic()
BarcodesLvHistPlot
#ggsave(plot = BarcodesLvHistPlot, file = paste0(readsdirectory, 'invitro_LVhist.svg'))
```


### starcode
```{r}
preprocess_starcode <- function(sample, length) {
  readsdirectory <- paste0(input.dir, sample, "/")
  
  # Get list of CSV files in the directory
  file_name <- list.files(readsdirectory, pattern = "*_shavedReads.txt", full.names = TRUE)
  
  # Ensure file_name is not empty
  if (length(file_name) > 0) {
    # output for starcode
    VISER_out <- read.csv(file_name, header = TRUE) %>% select(target)
    VISER_out$target <- substr(VISER_out$target, 1, length)
    VISER_out <- VISER_out %>% unique()
    write.table(VISER_out$target, paste0(output.dir, sample, "shavedReadsList_", length, ".txt"), sep = ",", quote = FALSE, row.names = FALSE, col.names = FALSE)
  }
}


input.dir = "/projects/b1042/GoyalLab/egrody/20240116_VISER_SALVEseq/EGS004/seqIO/"
output.dir = "/projects/b1042/GoyalLab/egrody/20240116_VISER_SALVEseq/EGS004/starcode/"

samples = c("VISER_D13", "VISER_D83", "VISER_D195")
starcode_lengths <- c(18, 24, 30)

for (sample in samples) {
  for (i in starcode_lengths) {
    preprocess_starcode(sample, i)
  }
}

```
I saved three different lengths of target sequences: 18bp, 24bp, and 30bp ("full length"), because starcode has a maximum LV dist of 8. I ran starcode in Quest using the starcodeRun.py script over these files. From that analysis, I got a list of the most popular consensus sequences for each sample for each target length. I used those consensus sequences to identify a cutoff for target reads.


### Target cutoff
```{r, eval = FALSE}
postprocess_starcode <- function(sample, length, readsdirectory) {
  # Get list of CSV files in the directory
  all_files <- list.files(readsdirectory, pattern = "*_d8.txt", full.names = TRUE)
  file_name <- all_files[grepl(sample, all_files) & grepl(length, all_files)]
  
  # Ensure file_name is not empty
  if (length(file_name) > 0) {
    # output for starcode
    VISER_out <- read.csv(file_name, header = FALSE, sep = "\t") %>% slice(1:5)
    colnames(VISER_out) <- c("consensus", "count")
    starcode_consensus[sample, length] <<- VISER_out$consensus[1]
  }
  return(starcode_consensus)
}


input.dir = "/projects/b1042/GoyalLab/egrody/20240116_VISER_SALVEseq/EGS004/starcode/"

samples = c("VISER_D13", "VISER_D83", "VISER_D195")
starcode_lengths <- c("18_", "24_", "30_")
starcode_consensus <- data.frame(matrix(NA, nrow = length(samples), ncol = length(starcode_lengths)))
rownames(starcode_consensus) <- samples
colnames(starcode_consensus) <- starcode_lengths

for (sample in samples) {
  for (i in starcode_lengths) {
    postprocess_starcode(sample, i, input.dir)
  }
}


```

Cutoff by LV
```{r, eval = FALSE}
for (sample in samples) {
  df <- get(sample)
  df$dist <- NA
  for (i in 1:nrow(df)) {
    # Calculate the Levenshtein distance between the current string and the constant
    dist <- adist(df$target[i], starcode_consensus[sample, "30_"])
    # Assign the distance to the 'dist' column
    df$dist[i] <- dist
    assign(sample, df)
  }
}

hist(VISER_D195$dist, breaks = 50, xlab = "Levenshtein Distance: recovered to expected")
```
This histogram has two peaks: one centered at 3 and one centered at 16. Cutoff is 8.

```{r}
VISER_D13_starcodecut <- VISER_D13 %>% filter(dist < 8) %>% select(cellID, UMI) %>% unique() %>% count(cellID) %>% rename(counts = n)
hist(VISER_D13_starcodecut$counts, ylim = c(0,70), breaks = 50, main = "VISER_D13 from seqIO", xlab = "env UMIs")
```


### scPathoQuant test analysis
Instead of using seqIO pipeline, I tested using cellranger count --> scPathoQuant to align reads

```{r}
input.dir <- "/projects/b1042/GoyalLab/egrody/20240116_VISER_SALVEseq/EGS004/scPathoQuant/VISER_D13/"
file <- "pathogen_al_gene_counts_mac239.csv"
VISER_scPQ <- read.csv(paste0(input.dir, file), header = TRUE) %>% rename(cellID = cell_barcode) %>% 
  mutate(cellID = str_sub(cellID, end = -3)) %>% unique()

#compare to seqIO analysis
hist(VISER_scPQ$umi, breaks = 50, main = "VISER_D13 from scPathoQuant", xlab = "env UMIs")
```

Overlap?
```{r}
scPQ_v_seqIO <- left_join(VISER_D13_starcodecut, VISER_scPQ, by = "cellID") %>% filter(!is.na(umi))
cat("Correlation between seqIO and scPathoQuant env counts:", cor(scPQ_v_seqIO$counts, scPQ_v_seqIO$umi))
corrplot <- ggplot(data = scPQ_v_seqIO, aes(x=counts, y=umi)) +
  geom_point() + 
  labs(x = "seqIO counts", y = "scPathoQuant counts", title = "VISER D13 env+ cells") +
  theme_classic()
corrplot

VISER_D13_prestarcode <- VISER_D13 %>% select(cellID, UMI) %>% unique() %>% count(cellID) %>% rename(counts = n)
scPQ_v_seqIOfull <- left_join(VISER_D13_prestarcode, VISER_scPQ, by = "cellID") %>% filter(!is.na(umi))
cat("Correlation between seqIO prestarcode and scPathoQuant env counts:", cor(scPQ_v_seqIO$counts, scPQ_v_seqIO$umi))
corrplot <- ggplot(data = scPQ_v_seqIO, aes(x=counts, y=umi)) +
  geom_point() + 
  labs(x = "seqIO counts", y = "scPathoQuant counts", title = "VISER D13 env+ cells") +
  theme_classic()
corrplot
```




## Necessary pre-analysis

### Load seqIO output (skip)
```{r}
input.dir = "/projects/b1042/GoyalLab/egrody/20240116_VISER_SALVEseq/EGS004/seqIO/"
samples = c("VISER_D13", "VISER_D83", "VISER_D195")
reads = data.frame()
for (sample in samples) {
  file_name <- paste0(input.dir, sample, "/", sample, "_shavedReads.txt")
   unique_data <- read.csv(file_name, header = TRUE) %>% 
    unique()
  unique_data$sample <- sample
  
  #cat("Sample: ", sample, "\nNo. of unique targets: ", nrow(unique_data), "\n")
  reads <- bind_rows(reads,unique_data)
}

VISER_D13 <- reads %>% filter(sample == "VISER_D13") %>% 
  select(cellID, UMI) %>%
  group_by(cellID, UMI) %>%
  summarise(count = n(), .groups = "drop") %>%
  ungroup()
VISER_D83 <- reads %>% filter(sample == "VISER_D83") %>% 
  select(cellID, UMI) %>%
  group_by(cellID, UMI) %>%
  summarise(count = n(), .groups = "drop") %>%
  ungroup()
VISER_D195 <- reads %>% filter(sample == "VISER_D195") %>% 
  select(cellID, UMI) %>%
  group_by(cellID, UMI) %>%
  summarise(count = n(), .groups = "drop") %>%
  ungroup()
```


### UMI correction (skip)
Error correction for UMI sequences:
```{r error correct UMI}
umi_error_correction <- function(df) {
  # Function to calculate Hamming distance
  hamming_distance <- function(a, b) {
    stringdist(a, b, method = "hamming")
  }
  
  # Group by cellID and perform error correction within each group
  df_corrected <- df %>%
    group_by(cellID) %>%
    group_modify(~ {
      umis <- .x$UMI
      counts <- .x$count
      n <- length(umis)
      
      if (n > 1) {  # Only proceed if there's more than one UMI
        # Calculate all pairwise Hamming distances
        dist_matrix <- outer(umis, umis, Vectorize(hamming_distance))
        
        # Identify UMIs to be merged (Hamming distance = 1)
        to_merge <- which(dist_matrix == 1, arr.ind = TRUE)
        
        # Perform merging if there are UMIs to merge
        if (nrow(to_merge) > 0) {
          for (i in 1:nrow(to_merge)) {
            idx1 <- to_merge[i, 1]
            idx2 <- to_merge[i, 2]
            
            if (idx1 < idx2 && !is.na(umis[idx1]) && !is.na(umis[idx2])) {
              if (counts[idx1] >= counts[idx2]) {
                counts[idx1] <- counts[idx1] + counts[idx2]
                umis[idx2] <- NA
                counts[idx2] <- NA
              } else {
                counts[idx2] <- counts[idx1] + counts[idx2]
                umis[idx1] <- NA
                counts[idx1] <- NA
              }
            }
          }
          
          # Remove merged (NA) entries
          valid_indices <- which(!is.na(umis))
          umis <- umis[valid_indices]
          counts <- counts[valid_indices]
        }
      }
      
      data.frame(UMI = umis, count = counts)
    }) %>%
    ungroup()
  
  return(df_corrected)
}

VISER_D13_UMIcorrect <- umi_error_correction(VISER_D13)
VISER_D83_UMIcorrect <- umi_error_correction(VISER_D83)
VISER_D195_UMIcorrect <- umi_error_correction(VISER_D195)
```

### cellID correction (skip)
Error correction for cellIDs with whitelist reference:
```{r error correct cellID}
whitelist_location <- "/home/egy2296/packages/cellranger-7.2.0/lib/python/cellranger/barcodes/3M-february-2018.txt.gz"
whitelist <- as.list(fread(whitelist_location, sep = "\n", header = FALSE, data.table = FALSE)$V1)

efficient_cellid_correction <- function(data, whitelist, cellid_column = "cellID") {
  # Ensure whitelist is a character vector
  whitelist <- as.character(unlist(whitelist))
  
  # Step 1: Count the observed frequency of whitelist barcodes
  whitelist_counts <- data %>%
    filter(!!sym(cellid_column) %in% whitelist) %>%
    group_by(!!sym(cellid_column)) %>%
    summarise(count = n(), .groups = "drop") %>%
    rename(barcode = !!sym(cellid_column))
  
  # Step 2: Identify barcodes not in the whitelist
  non_whitelist_barcodes <- setdiff(unique(data[[cellid_column]]), whitelist)
  
  # Function to find the closest whitelist barcode
  find_closest_barcode <- function(barcode, whitelist) {
    distances <- stringdist(barcode, whitelist, method = "hamming")
    closest <- which(distances == 1)
    if (length(closest) > 0) {
      return(whitelist[closest[1]])  # Return the first match if multiple exist
    } else {
      return(NA)
    }
  }
  
  # Create a correction map for non-whitelist barcodes
  correction_map <- sapply(non_whitelist_barcodes, find_closest_barcode, whitelist = whitelist)
  correction_map <- data.frame(
    original = names(correction_map),
    corrected = unname(correction_map)
  )
  
  # Apply corrections to the dataset
  data_corrected <- data %>%
    left_join(correction_map, by = setNames("original", cellid_column)) %>%
    mutate(corrected_cellID = coalesce(corrected, !!sym(cellid_column))) %>%
    select(-corrected)
  
  return(list(
    data_corrected = data_corrected,
    whitelist_counts = whitelist_counts,
    correction_map = correction_map
  ))
}

result <- efficient_cellid_correction(VISER_D13_UMIcorrect, whitelist)
corrected_data <- result$data_corrected
VISER_D13_correct <- corrected_data %>%
   group_by(corrected_cellID, UMI) %>%
   summarise(count = sum(count), .groups = 'drop')

result <- efficient_cellid_correction(VISER_D83_UMIcorrect, whitelist)
corrected_data <- result$data_corrected
VISER_D83_correct <- corrected_data %>%
   group_by(corrected_cellID, UMI) %>%
   summarise(count = sum(count), .groups = 'drop')

result <- efficient_cellid_correction(VISER_D195_UMIcorrect, whitelist)
corrected_data <- result$data_corrected
VISER_D195_correct <- corrected_data %>%
   group_by(corrected_cellID, UMI) %>%
   summarise(count = sum(count), .groups = 'drop')

output.dir <- "/projects/b1042/GoyalLab/egrody/20240116_VISER_SALVEseq/EGS004/VISER/"
write.csv(VISER_D83_correct, paste0(output.dir, "VISER_D83_correct_full.csv"))
write.csv(VISER_D195_correct, paste0(output.dir, "VISER_D195_correct_full.csv"))
```

D13 had many cellIDs and UMIs so I made a more efficient cellID error correction function. It will take ~2hrs to run.
```{r alternative cellID correction}
efficient_cellid_correction <- function(data, whitelist_location, cellid_column = "cellID", chunks = 10) {
  # Read whitelist directly as a character vector
  whitelist <- scan(whitelist_location, what = "character", quiet = TRUE)
  
  # Function to find the closest whitelist barcode
  find_closest_barcode <- function(barcode, whitelist) {
    distances <- stringdist(barcode, whitelist, method = "hamming")
    closest <- which(distances == 1)
    if (length(closest) > 0) {
      return(whitelist[closest[1]])  # Return the first match if multiple exist
    } else {
      return(NA_character_)
    }
  }
  
  # Function to process a chunk of data
  process_chunk <- function(chunk) {
    # Count the observed frequency of whitelist barcodes in this chunk
    chunk_whitelist_counts <- chunk %>%
      filter(!!sym(cellid_column) %in% whitelist) %>%
      count(!!sym(cellid_column), name = "count") %>%
      rename(barcode = !!sym(cellid_column))
    
    # Identify unique non-whitelist barcodes in this chunk
    chunk_non_whitelist_barcodes <- setdiff(unique(chunk[[cellid_column]]), whitelist)
    
    # Create a correction map for non-whitelist barcodes in this chunk
    chunk_correction_map <- data.frame(
      original = chunk_non_whitelist_barcodes,
      corrected = vapply(chunk_non_whitelist_barcodes, find_closest_barcode, character(1), whitelist = whitelist)
    )
    
    # Apply corrections to the chunk
    chunk_corrected <- chunk %>%
      left_join(chunk_correction_map, by = setNames("original", cellid_column)) %>%
      mutate(corrected_cellID = coalesce(corrected, !!sym(cellid_column))) %>%
      select(-corrected)
    
    list(
      data_corrected = chunk_corrected,
      whitelist_counts = chunk_whitelist_counts,
      correction_map = chunk_correction_map
    )
  }
  
  # Split data into chunks
  chunk_size <- ceiling(nrow(data) / chunks)
  data_chunks <- split(data, ceiling(seq_len(nrow(data)) / chunk_size))
  
  # Process each chunk
  results <- lapply(data_chunks, process_chunk)
  
  # Combine results
  data_corrected <- do.call(rbind, lapply(results, function(x) x$data_corrected))
  whitelist_counts <- do.call(rbind, lapply(results, function(x) x$whitelist_counts)) %>%
    group_by(barcode) %>%
    summarise(count = sum(count), .groups = 'drop')
  correction_map <- do.call(rbind, lapply(results, function(x) x$correction_map)) %>% distinct()
  
  # Clean up
  rm(whitelist, data_chunks, results)
  gc()
  
  return(list(
    data_corrected = data_corrected,
    whitelist_counts = whitelist_counts,
    correction_map = correction_map
  ))
}

# Usage
whitelist_location <- "/home/egy2296/packages/cellranger-7.2.0/lib/python/cellranger/barcodes/3M-february-2018.txt.gz"

result <- efficient_cellid_correction(VISER_D13_UMIcorrect, whitelist_location, chunks = 10)
corrected_data <- result$data_corrected
VISER_D13_correct <- corrected_data %>%
  group_by(corrected_cellID, UMI) %>%
  summarise(count = sum(count), .groups = 'drop')
output.dir <- "/projects/b1042/GoyalLab/egrody/20240116_VISER_SALVEseq/EGS004/VISER/"
write.csv(VISER_D13_correct, paste0(output.dir, "VISER_D13_correct_full.csv"))
```



## Current Analysis

### Read in data
Read back in 10X data and VISER sequences.
```{r}
#SingleCell
input.dir <- "/projects/b1042/GoyalLab/egrody/20240116_VISER_SALVEseq/EGS004/scRNAseq/"
SingleCell_d13_umap <- read.csv(paste0(input.dir, "d13_virusfreeUMAPscPathoQuant.csv")) 
SingleCell_d13_umap_env <- SingleCell_d13_umap %>%
  mutate(env = ifelse(env == 1, 0, env)) %>%
  select(cellID, UMAP1, UMAP2, env) %>%
  mutate(log1pSingleCell = log1p(env)) %>%
  mutate(cellID = str_sub(cellID, end = -3))
SingleCell_d83_umap <- read.csv(paste0(input.dir, "d83_virusfreeUMAPscPathoQuant.csv")) 
SingleCell_d83_umap_env <- SingleCell_d83_umap %>%
  select(cellID, UMAP1, UMAP2, env) %>%
  mutate(env = ifelse(env == 1, 0, env)) %>%
  mutate(log1pSingleCell = log1p(env)) %>%
  mutate(cellID = str_sub(cellID, end = -3))
SingleCell_d195_umap <- read.csv(paste0(input.dir, "d195_virusfreeUMAPscPathoQuant.csv")) 
SingleCell_d195_umap_env <- SingleCell_d195_umap %>%
  mutate(env = ifelse(env == 1, 0, env)) %>%
  select(cellID, UMAP1, UMAP2, env) %>%
  mutate(log1pSingleCell = log1p(env)) %>%
  mutate(cellID = str_sub(cellID, end = -3))


#VISER
input.dir <- "/projects/b1042/GoyalLab/egrody/20240116_VISER_SALVEseq/EGS004/VISER/"
VISER_D13_correct <- read.csv(paste0(input.dir, "VISER_D13_correct_full.csv"))
VISER_D83_correct <- read.csv(paste0(input.dir, "VISER_D83_correct_full.csv"))
VISER_D195_correct <- read.csv(paste0(input.dir, "VISER_D195_correct_full.csv"))

# Filter by raw barcode list and whitelist
d13_raw_data <- Read10X("/projects/b1042/GoyalLab/egrody/20240116_VISER_SALVEseq/EGS004/scRNAseq/counts/cellranger_count_D13/outs/raw_feature_bc_matrix/")
d13_raw_barcodes <- substr(colnames(d13_raw_data), 1, nchar(colnames(d13_raw_data)) - 2)
d83_raw_data <- Read10X("/projects/b1042/GoyalLab/egrody/20240116_VISER_SALVEseq/EGS004/scRNAseq/counts/cellranger_count_D83/outs/raw_feature_bc_matrix/")
d83_raw_barcodes <- substr(colnames(d83_raw_data), 1, nchar(colnames(d83_raw_data)) - 2)
d195_raw_data <- Read10X("/projects/b1042/GoyalLab/egrody/20240116_VISER_SALVEseq/EGS004/scRNAseq/counts/cellranger_count_D195/outs/raw_feature_bc_matrix/")
d195_raw_barcodes <- substr(colnames(d195_raw_data), 1, nchar(colnames(d195_raw_data)) - 2)

VISER_D13_correct <- VISER_D13_correct %>% 
  filter(count > 1) %>%
  rename(cellID = corrected_cellID) %>%
  filter(cellID %in% d13_raw_barcodes) %>%
  count(cellID) %>%
  mutate(log1pVISER = log1p(n))

VISER_D83_correct <- VISER_D83_correct %>% 
  filter(count > 1) %>%
  rename(cellID = corrected_cellID) %>%
  filter(cellID %in% d83_raw_barcodes) %>%
  count(cellID) %>%
  mutate(log1pVISER = log1p(n))

VISER_D195_correct <- VISER_D195_correct %>% 
  filter(count > 1) %>%
  rename(cellID = corrected_cellID) %>%
  filter(cellID %in% d195_raw_barcodes) %>%
  count(cellID) %>%
  mutate(log1pVISER = log1p(n))
```

Alternatively, filter by whitelist (skip)
```{r}
whitelist_location <- "/home/egy2296/packages/cellranger-7.2.0/lib/python/cellranger/barcodes/3M-february-2018.txt.gz"
whitelist <- readLines(whitelist_location, warn = FALSE)

VISER_D13_correct <- VISER_D13_correct %>% 
  filter(count > 1) %>%
  rename(cellID = corrected_cellID) %>%
  filter(cellID %in% whitelist) %>%
  count(cellID) %>%
  mutate(log1pVISER = log1p(n))

VISER_D83_correct <- VISER_D83_correct %>% 
  filter(count > 1) %>%
  rename(cellID = corrected_cellID) %>%
  filter(cellID %in% whitelist) %>%
  count(cellID) %>%
  mutate(log1pVISER = log1p(n))

VISER_D195_correct <- VISER_D195_correct %>% 
  filter(count > 1) %>%
  rename(cellID = corrected_cellID) %>%
  filter(cellID %in% whitelist) %>%
  count(cellID) %>%
  mutate(log1pVISER = log1p(n))

```


### joint tables
To see what are in common, we use left_join to look at only the cells with 10X anchors:
```{r}
D13_paint_umap = left_join(SingleCell_d13_umap_env, VISER_D13_correct, by = "cellID")
D13_paint_umap[is.na(D13_paint_umap)] <- 0

left_both <- sum((D13_paint_umap$log1pSingleCell != 0 & D13_paint_umap$log1pVISER != 0), na.rm = TRUE)
left_10Xonly <- sum(D13_paint_umap$log1pSingleCell != 0, na.rm = TRUE) - left_both
left_VISERonly <- sum(D13_paint_umap$log1pVISER != 0, na.rm = TRUE) - left_both
cat("D13\nTotal cells: \t\t", nrow(D13_paint_umap),
  "\nCells in both: \t\t", left_both,
  "\nCells in 10X only: \t", left_10Xonly,
    "\nCells in VISER only: \t", left_VISERonly)

D83_paint_umap = left_join(SingleCell_d83_umap_env, VISER_D83_correct, by = "cellID")
D83_paint_umap[is.na(D83_paint_umap)] <- 0

left_both <- sum((D83_paint_umap$log1pSingleCell != 0 & D83_paint_umap$log1pVISER != 0), na.rm = TRUE)
left_10Xonly <- sum(D83_paint_umap$log1pSingleCell != 0, na.rm = TRUE) - left_both
left_VISERonly <- sum(D83_paint_umap$log1pVISER != 0, na.rm = TRUE) - left_both
cat("\n\nD83\nTotal cells: \t\t", nrow(D83_paint_umap),
  "\nCells in both: \t\t", left_both,
  "\nCells in 10X only: \t", left_10Xonly,
    "\nCells in VISER only: \t", left_VISERonly)

D195_paint_umap = left_join(SingleCell_d195_umap_env, VISER_D195_correct, by = "cellID")
D195_paint_umap[is.na(D195_paint_umap)] <- 0

left_both <- sum((D195_paint_umap$log1pSingleCell != 0 & D195_paint_umap$log1pVISER != 0), na.rm = TRUE)
left_10Xonly <- sum(D195_paint_umap$log1pSingleCell != 0, na.rm = TRUE) - left_both
left_VISERonly <- sum(D195_paint_umap$log1pVISER != 0, na.rm = TRUE) - left_both
cat("\n\nD195\nTotal cells: \t\t", nrow(D195_paint_umap),
  "\nCells in both: \t\t", left_both,
  "\nCells in 10X only: \t", left_10Xonly,
    "\nCells in VISER only: \t", left_VISERonly)

```

Now looking at all the VISER data, even without anchors, using full_join:
```{r}
D13_paint_umap_full = full_join(SingleCell_d13_umap_env, VISER_D13_correct, by = "cellID")
D13_paint_umap_full[is.na(D13_paint_umap_full)] <- 0

full_both <- sum((D13_paint_umap_full$log1pSingleCell != 0 & D13_paint_umap_full$log1pVISER != 0), na.rm = TRUE)
full_10Xonly <- sum(D13_paint_umap_full$log1pSingleCell != 0, na.rm = TRUE) - full_both
full_VISERonly <- sum(D13_paint_umap_full$log1pVISER != 0, na.rm = TRUE) - full_both
unanchored <- full_VISERonly - left_VISERonly
cat("D13\nTotal cells: \t\t", nrow(D13_paint_umap_full),
  "\nCells in both: \t\t", full_both,
  "\nCells in 10X only: \t", full_10Xonly,
    "\nCells in VISER only: \t", full_VISERonly,
  "\nUnanchored VISER: \t", unanchored)

D83_paint_umap_full = full_join(SingleCell_d83_umap_env, VISER_D83_correct, by = "cellID")
D83_paint_umap_full[is.na(D83_paint_umap_full)] <- 0

full_both <- sum((D83_paint_umap_full$log1pSingleCell != 0 & D83_paint_umap_full$log1pVISER != 0), na.rm = TRUE)
full_10Xonly <- sum(D83_paint_umap_full$log1pSingleCell != 0, na.rm = TRUE) - full_both
full_VISERonly <- sum(D83_paint_umap_full$log1pVISER != 0, na.rm = TRUE) - full_both
unanchored <- full_VISERonly - left_VISERonly
cat("D83\nTotal cells: \t\t", nrow(D83_paint_umap_full),
  "\nCells in both: \t\t", full_both,
  "\nCells in 10X only: \t", full_10Xonly,
    "\nCells in VISER only: \t", full_VISERonly,
  "\nUnanchored VISER: \t", unanchored)

D195_paint_umap_full = full_join(SingleCell_d195_umap_env, VISER_D195_correct, by = "cellID")
D195_paint_umap_full[is.na(D195_paint_umap_full)] <- 0

full_both <- sum((D195_paint_umap_full$log1pSingleCell != 0 & D195_paint_umap_full$log1pVISER != 0), na.rm = TRUE)
full_10Xonly <- sum(D195_paint_umap_full$log1pSingleCell != 0, na.rm = TRUE) - full_both
full_VISERonly <- sum(D195_paint_umap_full$log1pVISER != 0, na.rm = TRUE) - full_both
unanchored <- full_VISERonly - left_VISERonly
cat("\n\nD195\nTotal cells: \t\t", nrow(D195_paint_umap_full),
  "\nCells in both: \t\t", full_both,
  "\nCells in 10X only: \t", full_10Xonly,
    "\nCells in VISER only: \t", full_VISERonly,
  "\nUnanchored VISER: \t", unanchored)
```

### Correlation plots (skip)

Because the yield from the ART samples (D83 and D195) is so low, this analysis does not work, i.e. get a warning "Warning: the standard deviation is zero":
```{r}
jointTableD83 = inner_join(SingleCell_d83_umap_env, VISER_D83_correct, by = "cellID") 
jointTableD195 = inner_join(SingleCell_d195_umap_env, VISER_D195_correct, by = "cellID") 

#  mutate(log1pViserCount = log1p(VISERcount), log2ViserCount = log2(VISERcount), lnViserCount = log(VISERcount), normalizedViserCount = 1000*VISERcount/sum(VISERcount), logNormViser = log1p(normalizedViserCount), log1pSingleCell = log1p(SingleCellcount))


output.dir <- "/projects/b1042/GoyalLab/egrody/20240116_VISER_SALVEseq/EGS004/joint/"
cat("Correlation of D83 between 10X data and log1p VISER data:", cor(jointTableD83$log1pSingleCell, jointTableD83$log1pVISER))
corrplot <- ggplot(data = jointTableD83, aes(x=log1pVISER, y=log1pSingleCell)) +
  geom_point() + 
  xlim(0,7) + 
  ylim(0,7)
corrplot
ggsave(corrplot, file = paste0(output.dir, "D83_correlationplot.svg"))

cat("\nCorrelation of D195 between 10X data and log1p VISER data:", cor(jointTableD195$log1pSingleCell, jointTableD195$log1pVISER))
corrplot <- ggplot(data = jointTableD195, aes(x=log1pVISER, y=log1pSingleCell)) +
  geom_point() +
  xlim(0,7) +
  ylim(0,7)
ggsave(corrplot, file = paste0(output.dir, "D195_correlationplot.svg"))
```


### Paint UMAPs
```{r}
create_umap_plot <- function(data, method, title, output_file_name, output_dir = "") {
  color_column <- if (method == "VISER") {
    "log1pVISER"
  } else if (method == "10X") {
    "log1pSingleCell"
  } else {
    stop("Invalid method. Please use 'VISER' or '10X'.")
  }
  
  if (!color_column %in% names(data)) {
    stop(paste("Column", color_column, "not found in the data."))
  }
  
  global_min <- min(min(data$log1pVISER, na.rm = TRUE), 
                   min(data$log1pSingleCell, na.rm = TRUE))
  global_max <- max(max(data$log1pVISER, na.rm = TRUE), 
                   max(data$log1pSingleCell, na.rm = TRUE))
  data$highlight <- data[[color_column]] > 0
  
  paintplot <- ggplot(data %>% arrange(highlight), aes(x = UMAP1, y = UMAP2)) +
    geom_point(aes(color = .data[[color_column]]), size = 1, shape = 16) +
    scale_color_gradient(low = "gray93", high = "darkblue",
                        limits = c(global_min, global_max)) +
    theme_classic() +
    theme(legend.position = "bottom",
          legend.title = element_text(size = rel(0.6)),
          legend.text = element_text(size = rel(0.6), angle = 30),
          axis.text = element_blank(),
          axis.ticks = element_blank()) +
    labs(title = title, color = method)

  ggsave(paintplot, file = file.path(output_dir, output_file_name), width = 7, height = 7)
  return(paintplot)
}

output.dir <- "/projects/b1042/GoyalLab/egrody/20240116_VISER_SALVEseq/EGS004/joint/"
create_umap_plot(
  data = D13_paint_umap,
  method = "VISER",
  title = "D13 SALVEseq v1",
  output_file_name = "umap_D13_VISER.svg",
  output_dir = output.dir
)
create_umap_plot(
  data = D83_paint_umap,
  method = "VISER",
  title = "D83 SALVEseq v1",
  output_file_name = "umap_D83_VISER.svg",
  output_dir = output.dir
)
create_umap_plot(
  data = D195_paint_umap,
  method = "VISER",
  title = "D195 SALVEseq v1",
  output_file_name = "umap_D195_VISER.svg",
  output_dir = output.dir
)
create_umap_plot(
  data = D13_paint_umap,
  method = "10X",
  title = "D13 10X",
  output_file_name = "umap_D13_SingleCell.svg",
  output_dir = output.dir
)
create_umap_plot(
  data = D83_paint_umap,
  method = "10X",
  title = "D83 10X",
  output_file_name = "umap_D83_SingleCell.svg",
  output_dir = output.dir
)
create_umap_plot(
  data = D195_paint_umap,
  method = "10X",
  title = "D195 10X",
  output_file_name = "umap_D195_SingleCell.svg",
  output_dir = output.dir
)
```

### UMAP with 10X-scPQ clustering
Now creating the UMAPs from Seurat object clustered with standard 10X virus expression (for mac239 only)
```{r}
input.dir <- "/projects/b1042/GoyalLab/egrody/20240116_VISER_SALVEseq/EGS004/scRNAseq/"
D13_10XscPQcluster <- read.csv(paste0(input.dir, "d13_scPQexpression_mac239.csv"))
D13_10XscPQcluster <- D13_10XscPQcluster %>%
  mutate(Count = ifelse(Count == 1, 0, Count)) %>%
  select(cellID, UMAP1, UMAP2, Count) %>%
  mutate(log1pSingleCell = log1p(Count)) %>%
  mutate(cellID = str_sub(cellID, end = -3))
D83_10XscPQcluster <- read.csv(paste0(input.dir, "d83_scPQexpression_mac239.csv"))
D195_10XscPQcluster <- read.csv(paste0(input.dir, "d195_scPQexpression_mac239.csv"))

D13_paint_umap = left_join(D13_10XscPQcluster, VISER_D13_correct, by = "cellID")
D13_paint_umap[is.na(D13_paint_umap)] <- 0
D13_paint_umap <- D13_paint_umap %>% 
  rename(SingleCell = Count, VISER = n) %>%
  mutate(log1pSingleCell = log1p(SingleCell))

D83_paint_umap = left_join(D83_10XscPQcluster, VISER_D83_correct, by = "cellID")
D83_paint_umap[is.na(D83_paint_umap)] <- 0
D83_paint_umap <- D83_paint_umap %>% 
  rename(SingleCell = Count, VISER = n) %>%
  mutate(log1pSingleCell = log1p(SingleCell))
  
D195_paint_umap = left_join(D195_10XscPQcluster, VISER_D195_correct, by = "cellID")
D195_paint_umap[is.na(D195_paint_umap)] <- 0
D195_paint_umap <- D195_paint_umap %>% 
  rename(SingleCell = Count, VISER = n) %>%
  mutate(log1pSingleCell = log1p(SingleCell))

output.dir <- "/projects/b1042/GoyalLab/egrody/20240116_VISER_SALVEseq/EGS004/joint/"
create_umap_plot(
  data = D13_paint_umap,
  method = "VISER",
  title = "D13 SALVEseq v1",
  output_file_name = "umap_10XscPQcluster_D13_VISER.svg",
  output_dir = output.dir
)
create_umap_plot(
  data = D83_paint_umap,
  method = "VISER",
  title = "D83 SALVEseq v1",
  output_file_name = "umap_10XscPQcluster_D83_VISER.svg",
  output_dir = output.dir
)
create_umap_plot(
  data = D195_paint_umap,
  method = "VISER",
  title = "D195 SALVEseq v1",
  output_file_name = "umap_10XscPQcluster_D195_VISER.svg",
  output_dir = output.dir
)
create_umap_plot(
  data = D13_paint_umap,
  method = "10X",
  title = "D13 10X",
  output_file_name = "umap_10XscPQcluster_D13_SingleCell.svg",
  output_dir = output.dir
)
create_umap_plot(
  data = D83_paint_umap,
  method = "10X",
  title = "D83 10X",
  output_file_name = "umap_10XscPQcluster_D83_SingleCell.svg",
  output_dir = output.dir
)
create_umap_plot(
  data = D195_paint_umap,
  method = "10X",
  title = "D195 10X",
  output_file_name = "umap_10XscPQcluster_D195_SingleCell.svg",
  output_dir = output.dir
)
```




### Unanchored cells
Do the unanchored cells exist in the raw cellIDs from 10X?

```{r}
create_barcode_rank_plot <- function(sample, input_dir, joint_umap_df) {
  # Construct the full input directory path
  full_input_dir <- file.path(input_dir, paste0("cellranger_count_", sample), "outs", "raw_feature_bc_matrix")
  
  # Read the 10X data
  raw_data <- Read10X(full_input_dir)
  
  # Calculate UMI counts per barcode
  umi_counts <- colSums(raw_data)
  
  # Create a data frame with barcode and UMI count
  plot_data <- data.frame(barcode = names(umi_counts), 
                          umi_count = umi_counts)
  
  # Sort the data frame by UMI count in descending order and trim barcode
  plot_data <- plot_data %>% 
    arrange(desc(umi_count)) %>%
    mutate(barcode = substr(barcode, 1, nchar(barcode) - 2))
  
  # Isolate VISER+ cellIDs
  VISER_cellids <- joint_umap_df %>% 
    filter(log1pVISER > 0) %>% 
    select(cellID)
  
  # Print the number of matching cellIDs
  cat(sum(VISER_cellids[[1]] %in% plot_data$barcode), "out of", nrow(VISER_cellids), 
      "VISER cellIDs are found in raw cellID list\n")
  
  # Add a column to indicate whether each barcode should be highlighted
  plot_data$highlight <- plot_data$barcode %in% VISER_cellids[[1]]
  
  # Add rank and handle zero/negative values
  plot_data <- plot_data %>%
    arrange(desc(umi_count)) %>%
    mutate(
      rank = row_number(),
      umi_count_adj = ifelse(umi_count <= 0, 0.1, umi_count)  # Replace 0 or negative with 0.1
    )
  
  # Create the plot
  barcode_rank_plot <- ggplot(plot_data %>% arrange(highlight), aes(x = rank, y = umi_count_adj)) +
    geom_point(aes(color = highlight), alpha = 0.6) +
    scale_x_log10(
      labels = scales::label_number(),
      breaks = scales::breaks_log(n = 6)
    ) +
    scale_y_log10(
      labels = scales::label_number(),
      breaks = scales::breaks_log(n = 6)
    ) +
    scale_color_manual(values = c("FALSE" = "grey", "TRUE" = "red")) +
    labs(x = "Rank of Barcode", 
         y = "UMI Count (Adjusted)",
         title = paste("UMI Count vs. Barcode Rank (Log-Scaled Axes) -", sample)) +
    theme_minimal() +
    theme(legend.position = "none")
  
  return(barcode_rank_plot)
}


input_dir <- "/projects/b1042/GoyalLab/egrody/20240116_VISER_SALVEseq/EGS004/scRNAseq/counts"
D13_rank_plot <- create_barcode_rank_plot("D13", input_dir, D13_paint_umap_full)
D13_rank_plot
ggsave("/projects/b1042/GoyalLab/egrody/20240116_VISER_SALVEseq/EGS004/joint/barcodeRankPlot_D13.png", D13_rank_plot, width = 10, height = 8, device = "png")

D83_rank_plot <- create_barcode_rank_plot("D83", input_dir, D83_paint_umap_full)
D83_rank_plot
ggsave("/projects/b1042/GoyalLab/egrody/20240116_VISER_SALVEseq/EGS004/joint/barcodeRankPlot_D83.png", D83_rank_plot, width = 10, height = 8, device = "png")

D195_rank_plot <- create_barcode_rank_plot("D195", input_dir, D195_paint_umap_full)
D195_rank_plot
ggsave("/projects/b1042/GoyalLab/egrody/20240116_VISER_SALVEseq/EGS004/joint/barcodeRankPlot_D195.png", D195_rank_plot, width = 10, height = 8, device = "png")
```

### UMAPs colored by UMI count

```{r}
umap_by_umi_count <- function(seurat_obj, 
                                   assay = "RNA", 
                                   color_scale = scale_color_viridis_c()) {
  
  # Calculate total UMI count for each cell
  umi_counts <- colSums(seurat_obj[[assay]]@counts)
  
  # Log-transform the UMI counts
  log_umi_counts <- log1p(umi_counts)
  
  # Extract UMAP coordinates
  umap_coords <- Embeddings(seurat_obj, reduction = "umap")
  
  # Create a data frame for plotting
  plot_data <- data.frame(
    UMAP_1 = umap_coords[,1],
    UMAP_2 = umap_coords[,2],
    log_umi_counts = log_umi_counts
  )
  
  # Create the UMAP plot using ggplot2
  plot <- ggplot(plot_data, aes(x = UMAP_1, y = UMAP_2, color = log_umi_counts)) +
    geom_point(size = 1) +
    color_scale +
    theme_minimal() +
    labs(title = "UMAP colored by log(UMI count)",
         x = "UMAP_1",
         y = "UMAP_2",
         color = "log(UMI count)") +
    theme(legend.position = "right")
  
  return(plot)
}

# umap_plot <- umap_by_umi_count(seurat_obj)
# print(umap_plot)
```

