---
title: "EGS016SALVEseq"
author: "Emanuelle Grody"
date: "2025-07-21"
output: html_document
---

```{r, message=FALSE, warning=FALSE}
source("~/SALVEseq/packages.R")
source("~/SALVEseq/functions.R")
```

# Individual Datasets
## KLRB1

GEX: bamsort
```{r GEX bamsort}
input.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS016/singleCell/bamsort/alignment/"
output.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS016/singleCell/bamsort/reads/KLRB1"

raw_cellIDs <- read.csv("/projects/b1042/GoyalLab/egrody/extractedData/EGS014/singleCell/raw_cellIDs/Uninfected_raw_cellIDs.csv")
#raw_cellIDs <- split(raw_cellIDs_df$cellID, raw_cellIDs_df$sample)

# processing
#process_bamsort("SALVE", samples$samples, input.dir, output.dir, raw_cellIDs)


filename <- "W0_GEX_bamsort_alignment_KLRB1.csv"
sample_files <- paste0(input.dir, filename)

extracted_data <- NULL  # Initialize as NULL for each file
all_data <- NULL

tryCatch({
  # Read the CSV file, expecting standard column names
  file_data <- fread(sample_files, data.table = FALSE)
  
  # Check if file has content
  if(nrow(file_data) == 0) {
    cat("Warning: File is empty:", filenames, "\n")
    next
  }
  
  # Simplify column mapping - expect standard column names
  required_cols <- c("cellID", "UMI", "count")
  
  # Check if all required columns exist
  missing_cols <- setdiff(required_cols, colnames(file_data))
  
  if (length(missing_cols) > 0) {
    cat("Warning: Missing required columns:", paste(missing_cols, collapse=", "), "\n")
    cat("Available columns:", paste(colnames(file_data), collapse=", "), "\n")
    next
  }
  
  # Extract data with the required columns and add category
  extracted_data <- file_data %>%
    select(cellID, UMI, count) %>%
    rename(read = count) %>%  # Rename count to read for consistency with later code
    mutate(category = "KLRB1")
  
}, error = function(e) {
  cat("Error reading file:", filenames[j], "\nLikely bad data file\n")
  cat("Error message:", conditionMessage(e), "\n")
})

# Add to the combined data frame only if we have data
if (!is.null(extracted_data) && nrow(extracted_data) > 0) {
  all_data <- rbind(all_data, extracted_data)
}
# Remove any rows with NA values and unique
all_data <- all_data %>% 
  filter(!is.na(cellID) & !is.na(UMI) & !is.na(read)) %>%
  unique()

# Count reads per UMI
tryCatch({
  umi_read_counts <- all_data %>%
    group_by(cellID, UMI, category) %>%
    summarize(
      read_count = sum(as.numeric(read)),  # Sum the count values
      .groups = 'drop'
    )
  
  cat("Gathered read counts for", nrow(umi_read_counts), "UMIs from", 
      n_distinct(umi_read_counts$cellID), "cells\n")
  
  # Create a wide format with UMI counts per category
  umi_by_category <- umi_read_counts %>%
    group_by(cellID, category) %>%
    summarize(
      category_UMIs = n_distinct(UMI),
      category_reads = sum(read_count),
      .groups = 'drop'
    )
  
  # Use pivot_wider to create a wide format
  umi_by_category_wide <- tidyr::pivot_wider(
    umi_by_category,
    id_cols = cellID,
    names_from = category,
    names_sep = "_",
    values_from = c(category_UMIs, category_reads),
    values_fill = 0)
  
  # Apply raw_cellIDs filtering
    valid_cells <- raw_cellIDs$cellID
    cat("Keeping only cells in", 
        length(valid_cells), "valid cells from 10X data\n")
    
    # Filter the UMI-level data
    filtered_umi_read_counts <- umi_read_counts %>%
      filter(cellID %in% valid_cells)
  
  
  # Save the detailed UMI-level data (both filtered and unfiltered)
  umi_level_file <- file.path(output.dir, "W0_UMI_read_counts_raw.csv")
  write.csv(umi_read_counts, file = umi_level_file, row.names = FALSE)
  
  filtered_umi_level_file <- file.path(output.dir, "W0_UMI_read_counts_full.csv")
  write.csv(filtered_umi_read_counts, file = filtered_umi_level_file, row.names = FALSE)
  
  cat("Successfully processed\n")
  rm(umi_by_category, umi_by_category_wide, umi_read_counts)
}, error = function(e) {
  cat("Error processing data for sample", sample, ":", conditionMessage(e), "\n")
  
  # Try to save the raw data at least
  raw_file <- file.path(output.dir, paste0(sample, "_raw_data.csv"))
  write.csv(all_data, file = raw_file, row.names = FALSE)
  cat("Saved raw data to:", raw_file, "\n")
})

# minimums
input <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS016/singleCell/bamsort/reads/KLRB1/W0_UMI_read_counts_full.csv"
output_dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS016/singleCell/bamsort/reads/KLRB1/minimum/" 

W0_min <- set_minimums("KLRB1", input, min_reads = 1, min_umi = 1)
write.csv(W0_min, paste0(output_dir, "W0_minread1_minumi1.csv"))
GEX_KLRB1_W0 <- W0_min %>%
  select(-total) %>%
  mutate(KLRB1 = log1p(KLRB1))

GEX_KLRB1_W0_expressionDF <- read.csv("/projects/b1042/GoyalLab/egrody/extractedData/EGS002/singleCell/counts/Mmul_10_only/expressionDF/W0_KLRB1_raw.csv", row.names = "X")
cat("# cells KLRB1+ from expressionDF: \t",
  sum(GEX_KLRB1_W0_expressionDF$KLRB1 > 0),
  "\n# cells KLRB1+ from bamsort: \t\t",
  sum(W0_min$KLRB1 > 0))
GEX_KLRB1_W0_expressionDF <- GEX_KLRB1_W0_expressionDF %>% select(-KLRB1)
GEX_KLRB1_W0 <- left_join(GEX_KLRB1_W0_expressionDF, GEX_KLRB1_W0, by = "cellID")
GEX_KLRB1_W0[is.na(GEX_KLRB1_W0)] <- 0

max(GEX_KLRB1_W0$KLRB1)
```

SALVE: bamsort
```{r SALVE}
input.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS016/SALVE/bamsort/alignment/KLRB1/"
output.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS016/SALVE/bamsort/reads/KLRB1/"

raw_cellIDs <- read.csv("/projects/b1042/GoyalLab/egrody/extractedData/EGS014/singleCell/raw_cellIDs/Uninfected_raw_cellIDs.csv")

filenames <- "W0_KLRB1_PL_bamsort_alignment_KLRB1.csv"
sample_files <- paste0(input.dir, filenames)

all_data <- data.frame()
extracted_data <- NULL

tryCatch({
  # Read the CSV file
  file_data <- fread(sample_files, data.table = FALSE)
  
  # Check if file has content
  if(nrow(file_data) == 0) {
    cat("Warning: File is empty:", filenames, "\n")
    stop("Empty file")  # Use stop() instead of next
  }
  
  # Check required columns
  required_cols <- c("cellID", "UMI", "count")
  missing_cols <- setdiff(required_cols, colnames(file_data))
  
  if (length(missing_cols) > 0) {
    cat("Warning: Missing required columns:", paste(missing_cols, collapse=", "), "\n")
    cat("Available columns:", paste(colnames(file_data), collapse=", "), "\n")
    stop("Missing columns")  # Use stop() instead of next
  }
  
  # Extract data
  extracted_data <- file_data %>%
    select(cellID, UMI, count) %>%
    rename(read = count) %>%
    mutate(category = "KLRB1")
  
}, error = function(e) {
  cat("Error reading file:", filenames, "\nLikely bad data file\n")
  cat("Error message:", conditionMessage(e), "\n")
})

# Add to combined data frame
if (!is.null(extracted_data) && nrow(extracted_data) > 0) {
  all_data <- rbind(all_data, extracted_data)
}

# Remove rows with NA values and get unique records
all_data <- all_data %>% 
  filter(!is.na(cellID) & !is.na(UMI) & !is.na(read)) %>%
  unique()

# Count reads per UMI
tryCatch({
  umi_read_counts <- all_data %>%
    group_by(cellID, UMI, category) %>%
    summarize(
      read_count = sum(as.numeric(read)),
      .groups = 'drop'
    )
  
  cat("Gathered read counts for", nrow(umi_read_counts), "UMIs from", 
      n_distinct(umi_read_counts$cellID), "cells\n")
  
  # Create wide format with UMI counts per category
  umi_by_category <- umi_read_counts %>%
    group_by(cellID, category) %>%
    summarize(
      category_UMIs = n_distinct(UMI),
      category_reads = sum(read_count),
      .groups = 'drop'
    )
  
  umi_by_category_wide <- tidyr::pivot_wider(
    umi_by_category,
    id_cols = cellID,
    names_from = category,
    names_sep = "_",
    values_from = c(category_UMIs, category_reads),
    values_fill = 0)
  
  # Apply raw_cellIDs filtering
  valid_cells <- raw_cellIDs$cellID
  cat("Keeping only cells in", 
      length(valid_cells), "valid cells from 10X data\n")
  
  filtered_umi_read_counts <- umi_read_counts %>%
    filter(cellID %in% valid_cells)
  
  # Save outputs
  umi_level_file <- file.path(output.dir, "W0_UMI_read_counts_raw.csv")
  write.csv(umi_read_counts, file = umi_level_file, row.names = FALSE)
  
  filtered_umi_level_file <- file.path(output.dir, "W0_UMI_read_counts_full.csv")
  write.csv(filtered_umi_read_counts, file = filtered_umi_level_file, row.names = FALSE)
  
  cat("Successfully processed\n")
  rm(umi_by_category, umi_by_category_wide, umi_read_counts)
  
}, error = function(e) {
  cat("Error processing data:", conditionMessage(e), "\n")
  
  # Save raw data as fallback
  raw_file <- file.path(output.dir, "raw_data.csv")
  write.csv(all_data, file = raw_file, row.names = FALSE)
  cat("Saved raw data to:", raw_file, "\n")
})


# minimums
input <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS016/SALVE/bamsort/reads/KLRB1/W0_UMI_read_counts_full.csv"
output_dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS016/SALVE/bamsort/reads/KLRB1/minimum/" 
W0_min <- set_minimums("KLRB1", input, min_reads = 5, min_umi = 1)
write.csv(W0_min, paste0(output_dir, "SALVE_KLRB1_W0_SALVE_minreads5.csv"))
SALVE_KLRB1_W0 <- W0_min %>%
  select(-total) %>%
  mutate(KLRB1 = log1p(KLRB1))

#max(SALVE_KLRB1_W0$KLRB1)
```

### Saturation
Compare recovery: UMIs and cells
```{r}
input <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS016/singleCell/bamsort/reads/KLRB1/W0_UMI_read_counts_full.csv"
GEX_nomin <- read.csv(input)
input <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS016/SALVE/bamsort/reads/KLRB1/W0_UMI_read_counts_full.csv"
SALVE_nomin <- read.csv(input)
SALVE_nomin <- SALVE_nomin %>%
  filter(read_count >= 5)
SALVE_nomin <- SALVE_nomin %>%
  filter(cellID %in% raw_cellIDs$cellID)

# Compare UMIs
## Create combined identifiers
SALVE_pairs <- paste(SALVE_nomin$cellID, SALVE_nomin$UMI, sep = "_")
GEX_pairs <- paste(GEX_nomin$cellID, GEX_nomin$UMI, sep = "_")
## Find which pairs from SALVE are in GEX
SALVE_in_GEX <- SALVE_pairs %in% GEX_pairs
GEX_in_SALVE <- GEX_pairs %in% SALVE_pairs
## Calculate proportions
prop_SALVE_in_GEX <- sum(SALVE_in_GEX) / length(SALVE_in_GEX)
prop_GEX_in_SALVE <- sum(GEX_in_SALVE) / length(GEX_in_SALVE)

cat("Fraction cellID-UMI pairs from...\nGEX in SALVE: ",
    prop_GEX_in_SALVE,
    "\nSALVE in GEX: ",
    prop_SALVE_in_GEX)

# Compare cells
## Create combined identifiers
SALVE_cells <- unique(SALVE_nomin$cellID)
GEX_cells <- unique(GEX_nomin$cellID)
## Find which cells from SALVE are in GEX
SALVEc_in_GEXc <- SALVE_cells %in% GEX_cells
GEXc_in_SALVEc <- GEX_cells %in% SALVE_cells
## Calculate proportions
prop_cells_SALVE_in_GEX <- sum(SALVEc_in_GEXc) / length(SALVEc_in_GEXc)
prop_cells_GEX_in_SALVE <- sum(GEXc_in_SALVEc) / length(GEXc_in_SALVEc)

cat("Fraction cells from...\nGEX in SALVE: ",
    prop_cells_GEX_in_SALVE,
    "\nSALVE in GEX: ",
    prop_cells_SALVE_in_GEX)

cat("\nNumbers for text\n\nTotal GEX UMI: ",
    length(GEX_pairs),
    "\nNumber UMI GEX only: \t\t",
    length(GEX_pairs) - sum(GEX_in_SALVE),
    "\nFraction UMI GEX only: \t\t",
    1-prop_GEX_in_SALVE,
    "\nNumber cells GEX only: \t\t",
    length(GEX_cells) - sum(GEXc_in_SALVEc),
    "\nFraction cells GEX only: \t",
    1-prop_cells_GEX_in_SALVE)
```

Subsample read-weighted UMI
```{r}
output.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS016/SALVE/saturation/sample_UMI/"
input <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS016/SALVE/bamsort/reads/KLRB1/W0_UMI_read_counts_full.csv"
SALVE_nomin <- read.csv(input) %>%
  rename(reads = read_count)

  
results <- sample_UMI_weighted(SALVE_nomin)
summary <- analyze_UMI_sampling(results, title = "Weighted UMI Saturation: KLRB1 W0")

results_file <- paste0(output.dir, "KLRB1_W0_sample_UMI_results.csv")
summary_file <- paste0(output.dir, "KLRB1_W0_sample_UMI_summary.csv")
write.csv(results, results_file, row.names = FALSE)
write.csv(summary, summary_file, row.names = FALSE)

```

```{r}
model_results <- fit_models(results, target_coverage = 95, coverage_column = "pair_coverage")
model_results <- fit_models(results, target_coverage = 98.9, coverage_column = "pair_coverage")
model_results <- fit_models(results, target_coverage = 100, coverage_column = "pair_coverage")
model_results <- fit_models(results, target_coverage = 120)
```

To sequence more or not to sequence more, also not sure if it's correct
```{r}
decision <- make_sequencing_decision(SALVE_nomin, target_coverage = 80, coverage_column = "pair_coverage")

# Function to extract saturation data from model results
extract_saturation_data <- function(model_results, reads_GEX) {
  
  # Get prediction data
  pred_data <- model_results$predictions
  model_columns <- setdiff(colnames(pred_data), "percentage")
  
  # Calculate total reads at different sampling percentages
  total_reads <- sum(reads_GEX$reads)
  
  # Initialize results dataframe
  saturation_data <- data.frame(
    sampling_percentage = pred_data$percentage,
    reads_millions = (pred_data$percentage / 100) * total_reads / 1000000
  )
  
  # Get coverage data from best model
  valid_coverage <- c()
  for (col in model_columns) {
    if (col %in% colnames(pred_data)) {
      model_data <- pred_data[[col]]
      # Remove infinite, NA, and unrealistic values
      model_data <- model_data[is.finite(model_data) & model_data >= 0 & model_data <= 200]
      
      if (length(model_data) > 0) {
        valid_coverage <- pred_data[[col]]
        break  # Use first valid model
      }
    }
  }
  
  saturation_data$coverage <- valid_coverage
  
  # Calculate sequencing saturation (% of maximum possible coverage achieved)
  max_coverage <- max(saturation_data$coverage, na.rm = TRUE)
  saturation_data$sequencing_saturation <- (saturation_data$coverage / max_coverage) * 100
  
  # Information saturation is essentially the coverage itself
  # (what % of the information/genes/pairs you're capturing)
  saturation_data$information_saturation <- saturation_data$coverage
  
  return(list(
    data = saturation_data,
    max_coverage = max_coverage,
    total_reads = total_reads
  ))
}

# Function to plot sequencing saturation
plot_sequencing_saturation <- function(saturation_data, saturation_target = 90, 
                                      title = "Sequencing Saturation Curve") {
  
  data <- saturation_data$data
  total_reads <- saturation_data$total_reads
  
  # Calculate actual reads from sampling percentage
  data$actual_reads <- (data$sampling_percentage / 100) * total_reads
  
  p1 <- ggplot(data, aes(x = actual_reads, y = sequencing_saturation)) +
    geom_line(color = "blue", size = 1.2) +
    geom_point(color = "blue", size = 2) +
    labs(
      title = title,
      subtitle = "How efficiently you're capturing your library's potential",
      x = "Total Reads Sequenced",
      y = "Sequencing Saturation (%)",
      caption = "100% = you've captured all complexity this library can provide"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(size = 14, face = "bold"),
      plot.subtitle = element_text(size = 12, color = "gray60"),
      axis.title = element_text(size = 12),
      panel.grid.minor = element_blank()
    ) +
    scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 20)) +
    scale_x_continuous(labels = scales::comma_format()) +
    geom_hline(yintercept = saturation_target, linetype = "dashed", color = "red", alpha = 0.7) +
    annotate("text", x = max(data$actual_reads) * 0.7, y = saturation_target + 2, 
             label = paste0(saturation_target, "% efficiency"), color = "red", size = 3)
  
  return(p1)
}

# Function to plot information saturation
plot_information_saturation <- function(saturation_data, target_coverage = 90, 
                                       title = "Information Saturation Curve") {
  
  data <- saturation_data$data
  
  p2 <- ggplot(data, aes(x = reads_millions, y = information_saturation)) +
    geom_line(color = "darkgreen", size = 1.2) +
    geom_point(color = "darkgreen", size = 2) +
    labs(
      title = title,
      subtitle = "How much biological information you're capturing",
      x = "Sequencing Depth (Million Reads)",
      y = "Information Saturation (%)",
      caption = "% of genes/pairs/features detected in your experiment"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(size = 14, face = "bold"),
      plot.subtitle = element_text(size = 12, color = "gray60"),
      axis.title = element_text(size = 12),
      panel.grid.minor = element_blank()
    ) +
    scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 20)) +
    geom_hline(yintercept = target_coverage, linetype = "dashed", color = "red", alpha = 0.7) +
    annotate("text", x = max(data$reads_millions) * 0.7, y = target_coverage + 2, 
             label = paste0(target_coverage, "% target"), color = "red", size = 3)
  
  return(p2)
}


sat_data <- extract_saturation_data(decision$model_results, SALVE_nomin)
# Plot sequencing saturation
p1 <- plot_sequencing_saturation(sat_data, saturation_target = 80)
print(p1)

# Plot information saturation  
p2 <- plot_information_saturation(sat_data, target_coverage = 80)
print(p2)


target_saturation <- 90
data <- sat_data$data
total_reads <- sat_data$total_reads
data$actual_reads <- (data$sampling_percentage / 100) * total_reads
closest_idx <- which.min(abs(data$sequencing_saturation - target_saturation))
reads_needed <- data$actual_reads[closest_idx]

cat(sprintf("For %.0f%% sequencing saturation: %s reads\n", 
            target_saturation, scales::comma(round(reads_needed))))
```

Saturation analysis, not sure if it's correct:
```{r}
# Function to analyze sequencing and information saturation
analyze_saturation <- function(df) {
  # Assumes df has columns: cellID, UMI, reads
  
  cat("=== BASIC STATISTICS ===\n")
  cat("Total cells:", length(unique(df$cellID)), "\n")
  cat("Total UMIs:", length(unique(df$UMI)), "\n")
  cat("Total reads:", sum(df$reads), "\n")
  cat("Mean reads per UMI:", round(mean(df$reads), 2), "\n\n")
  
  # 1. SEQUENCING SATURATION (corrected)
  cat("=== SEQUENCING SATURATION ===\n")
  
  cell_stats <- df %>%
    group_by(cellID) %>%
    summarise(
      total_reads = sum(reads),
      unique_umis = n(),
      .groups = 'drop'
    ) %>%
    mutate(
      # CORRECTED: Sequencing saturation = (total_reads - unique_umis) / total_reads
      # This represents the fraction of reads that are duplicates
      seq_saturation = ifelse(total_reads > 0, 
                             (total_reads - unique_umis) / total_reads, 
                             0)
    )
  
  cat("Mean sequencing saturation:", round(mean(cell_stats$seq_saturation), 3), "\n")
  cat("Median sequencing saturation:", round(median(cell_stats$seq_saturation), 3), "\n")
  cat("Range:", round(min(cell_stats$seq_saturation), 3), "-", round(max(cell_stats$seq_saturation), 3), "\n\n")
  
  # 2. INFORMATION SATURATION (corrected approach)
  cat("=== INFORMATION SATURATION ===\n")
  
  # Better approach: Calculate saturation curves per cell, then average
  fractions <- seq(0.1, 1.0, by = 0.1)
  
  # Function to subsample reads for a single cell
  subsample_cell <- function(cell_data, fraction) {
    total_reads <- sum(cell_data$reads)
    n_reads_to_sample <- round(total_reads * fraction)
    
    if (n_reads_to_sample == 0) return(0)
    if (n_reads_to_sample >= total_reads) return(nrow(cell_data))
    
    # Create probability weights for sampling UMIs based on read counts
    probs <- cell_data$reads / total_reads
    
    # Sample UMIs with replacement, weighted by read counts
    sampled_reads <- 0
    detected_umis <- character(0)
    
    while (sampled_reads < n_reads_to_sample) {
      # Sample one UMI based on read probabilities
      sampled_umi_idx <- sample(1:nrow(cell_data), 1, prob = probs)
      sampled_umi <- cell_data$UMI[sampled_umi_idx]
      
      detected_umis <- c(detected_umis, sampled_umi)
      sampled_reads <- sampled_reads + 1
    }
    
    return(length(unique(detected_umis)))
  }
  
  # Perform subsampling for cells with enough reads
  min_reads_threshold <- 50  # Only analyze cells with sufficient reads
  cells_to_analyze <- cell_stats %>% 
    filter(total_reads >= min_reads_threshold) %>% 
    pull(cellID)
  
  if (length(cells_to_analyze) == 0) {
    cat("Warning: No cells have sufficient reads for saturation analysis\n")
    cells_to_analyze <- unique(df$cellID)[1:min(10, length(unique(df$cellID)))]
  }
  
  saturation_curves <- data.frame()
  
  for (cell in cells_to_analyze) {
    cell_data <- df[df$cellID == cell, ]
    
    for (frac in fractions) {
      umis_detected <- subsample_cell(cell_data, frac)
      
      saturation_curves <- rbind(saturation_curves, data.frame(
        cellID = cell,
        fraction = frac,
        umis_detected = umis_detected,
        total_reads_sampled = round(sum(cell_data$reads) * frac)
      ))
    }
  }
  
  # Calculate average saturation curve
  avg_saturation <- saturation_curves %>%
    group_by(fraction) %>%
    summarise(
      mean_umis = mean(umis_detected),
      median_umis = median(umis_detected),
      se_umis = sd(umis_detected) / sqrt(n()),
      .groups = 'drop'
    )
  
  # Calculate information saturation metrics
  max_umis <- avg_saturation$mean_umis[avg_saturation$fraction == 1.0]
  umis_at_90pct <- avg_saturation$mean_umis[avg_saturation$fraction == 0.9]
  umis_at_80pct <- avg_saturation$mean_umis[avg_saturation$fraction == 0.8]
  
  info_saturation_90 <- umis_at_90pct / max_umis
  info_saturation_80 <- umis_at_80pct / max_umis
  
  cat("Information saturation (90% vs 100% reads):", round(info_saturation_90, 3), "\n")
  cat("Information saturation (80% vs 100% reads):", round(info_saturation_80, 3), "\n")
  
  # Calculate slope in the last 20% to assess if plateau is reached
  slope_data <- avg_saturation %>% filter(fraction >= 0.8)
  if (nrow(slope_data) >= 2) {
    slope <- (max(slope_data$mean_umis) - min(slope_data$mean_umis)) / 
             (max(slope_data$fraction) - min(slope_data$fraction))
    cat("UMI detection rate in final 20% of reads:", round(slope, 2), "UMIs per 10% read increase\n")
  }
  cat("\n")
  
  # 3. VISUALIZATION
  
  # Plot 1: Sequencing saturation distribution
  p1 <- ggplot(cell_stats, aes(x = seq_saturation)) +
    geom_histogram(bins = 30, fill = "lightblue", alpha = 0.7, color = "black") +
    labs(title = "Distribution of Sequencing Saturation",
         subtitle = paste("Mean:", round(mean(cell_stats$seq_saturation), 3)),
         x = "Sequencing Saturation (Fraction of Duplicate Reads)", 
         y = "Number of Cells") +
    theme_minimal()
  
  # Plot 2: Information saturation curve with error bars
  p2 <- ggplot(avg_saturation, aes(x = fraction * 100, y = mean_umis)) +
    geom_line(color = "blue", size = 1) +
    geom_point(color = "blue", size = 2) +
    geom_errorbar(aes(ymin = mean_umis - se_umis, ymax = mean_umis + se_umis), 
                  width = 2, color = "blue", alpha = 0.7) +
    labs(title = "Information Saturation Curve",
         subtitle = paste("Based on", length(cells_to_analyze), "cells with ≥", min_reads_threshold, "reads"),
         x = "Percentage of Total Reads (%)",
         y = "Mean UMIs Detected") +
    scale_x_continuous(breaks = seq(10, 100, 10)) +
    theme_minimal()
  
  # Plot 3: Reads vs UMIs per cell (log scale for better visualization)
  p3 <- ggplot(cell_stats, aes(x = total_reads, y = unique_umis)) +
    geom_point(alpha = 0.6, color = "darkgreen") +
    geom_smooth(method = "loess", color = "red", se = TRUE) +
    labs(title = "UMIs vs Total Reads per Cell",
         x = "Total Reads (log scale)",
         y = "Unique UMIs (log scale)") +
    scale_x_log10() +
    scale_y_log10() +
    theme_minimal()
  
  print(p1)
  print(p2)
  print(p3)
  
  # Return summary statistics
  return(list(
    cell_stats = cell_stats,
    saturation_curves = saturation_curves,
    avg_saturation = avg_saturation,
    overall_seq_saturation = mean(cell_stats$seq_saturation),
    info_saturation_90vs100 = info_saturation_90,
    info_saturation_80vs100 = info_saturation_80
  ))
}


results <- analyze_saturation(SALVE_nomin)
```


## SALVE 

Using both EGS016 and EGS018 data
### bamsort
Run this once first to get raw cellIDs:
```{r raw_cellIDs}
# Load and save raw cell IDs from 10X data
raw_cellIDs <- list()
output_dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS016/singleCell/raw_cellIDs" 

# Create output directory if it doesn't exist
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}


samples <- data.frame(
  datasets = c(
    "LP29_D0",
    "LP29_D195",
    "LP29_D83",
    "invitro",
    "P_acute_GEX",
    "W0",
    "W2"
  ),
  samples = c(
    "D0",
    "D195",
    "D83",
    "Invitro",
    "Pacute",
    "W0",
    "W2"
  ),
  folders = c(
    "/projects/b1042/GoyalLab/egrody/extractedData/EGS004/singleCell/counts/Mmul_10_mac239/",
    "/projects/b1042/GoyalLab/egrody/extractedData/EGS004/singleCell/counts/Mmul_10_mac239/",
    "/projects/b1042/GoyalLab/egrody/extractedData/EGS004/singleCell/counts/Mmul_10_mac239/",
    "/projects/b1042/GoyalLab/egrody/extractedData/EGS002/singleCell/counts/Mmul_10_mac239/",
    "/projects/b1042/GoyalLab/egrody/extractedData/EGS009/singleCell/counts/Mmul_10_mac239/",
    "/projects/b1042/GoyalLab/egrody/extractedData/EGS002/singleCell/counts/Mmul_10_mac239/",
    "/projects/b1042/GoyalLab/egrody/extractedData/EGS002/singleCell/counts/Mmul_10_mac239/"
  )
)

all_cellIDs <- data.frame(cellID = character(), sample = character(), stringsAsFactors = FALSE)

for (i in 1:nrow(samples)) {
  dataset_name <- samples$datasets[i]
  sample_name <- samples$samples[i]
  rawDataFolder <- paste0(samples$folders[i], "Mmul_10_mac239_", samples$datasets[i], "/outs/raw_feature_bc_matrix/")
  
  tryCatch({
    rawdata <- Read10X(rawDataFolder)
    umi_counts <- colSums(rawdata)
    cell_ids <- names(umi_counts[umi_counts > 100]) # keep only cells with transcriptomes
    raw_cellIDs[[sample_name]] <- cell_ids
    cat("Loaded", length(cell_ids), "raw cellIDs for sample", sample_name, "\n")
    
    # Add to the combined dataframe
    sample_df <- data.frame(
      cellID = cell_ids,
      sample = rep(sample_name, length(cell_ids)),
      stringsAsFactors = FALSE
    )
    all_cellIDs <- rbind(all_cellIDs, sample_df)
    
    # Save individual sample's cell IDs to CSV
    sample_file <- file.path(output_dir, paste0(sample_name, "_raw_cellIDs.csv"))
    write.csv(data.frame(cellID = cell_ids), sample_file, row.names = FALSE)
    cat("Saved", length(cell_ids), "cell IDs to", sample_file, "\n")
    
  }, error = function(e) {
    cat("Error processing", dataset_name, ":", conditionMessage(e), "\n")
  })
  
  if (exists("rawdata")) {
    rm(rawdata)
    gc()
  }
}

# Save the combined cell IDs to a single CSV
combined_file <- file.path(output_dir, "all_raw_cellIDs.csv")
write.csv(all_cellIDs, combined_file, row.names = FALSE)
cat("Saved combined cell IDs to", combined_file, "\n")

# To easily load this data in the future:
# raw_cellIDs_df <- read.csv("path/to/output/directory/all_raw_cellIDs.csv")
# raw_cellIDs <- split(raw_cellIDs_df$cellID, raw_cellIDs_df$sample)
```

Run bamsort_combined (alignment_reads and splice) first.
Making full and minimum filtered lists:
```{r}
samples <- c(
    "D0",
    "D195",
    "Invitro",
    "Pacute",
    "W0",
    "W2"
)


input.dirs <- c("/projects/b1042/GoyalLab/egrody/extractedData/EGS016/SALVE/bamsort/alignment/v5coordinates/", 
          "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/SALVE/bamsort/alignment/v5coordinates/")
output.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/SALVE/bamsort/reads/v6/"


raw_cellIDs_df <- read.csv("/projects/b1042/GoyalLab/egrody/extractedData/EGS016/singleCell/raw_cellIDs/all_raw_cellIDs.csv")
raw_cellIDs <- split(raw_cellIDs_df$cellID, raw_cellIDs_df$sample)

process_bamsortv6(samples, input.dirs, output.dir, raw_cellIDs,
                  correction_dir = "/projects/b1042/GoyalLab/egrody/extractedData/EGS016/SALVE/bamsort/splice/combined/")

# minimums
input_dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/SALVE/bamsort/reads/v6/"
output_dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/SALVE/bamsort/reads/v6/minimum/" 

viremia <- c(
    "Invitro",
    "Pacute",
    "W2"
  )
ART <- c(
    "D0",
    "D195",
    "W0"
  )
process_all_set_minimumsv6("SALVE", viremia, input_dir, TRUE, output_dir, min_reads = 5, min_umi = 2,
                           aggregate_valid_umi = FALSE)
process_all_set_minimumsv6("SALVE", ART, input_dir, TRUE, output_dir, min_reads = 2, min_umi = 2,
                           aggregate_valid_umi = FALSE)

```



## GEX
### bamsort
```{r}
samples <- data.frame(
  datasets = c(
    "LP29_D0",
    "LP29_D195",
    "LP29_D83",
    "invitro",
    "P_acute_GEX",
    "W0",
    "W2"
  ),
  samples = c(
    "D0",
    "D195",
    "D83",
    "Invitro",
    "Pacute",
    "W0",
    "W2"
  )
)

# Process
input.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS016/singleCell/bamsort/alignment/v5/"
output.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS016/singleCell/bamsort/reads/v6/"

raw_cellIDs_df <- read.csv("/projects/b1042/GoyalLab/egrody/extractedData/EGS016/singleCell/raw_cellIDs/all_raw_cellIDs.csv")
raw_cellIDs <- split(raw_cellIDs_df$cellID, raw_cellIDs_df$sample)

process_bamsortv6(samples$samples, input.dir, output.dir, raw_cellIDs)

# Minimums
input_dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS016/singleCell/bamsort/reads/v6/"
output_dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS016/singleCell/bamsort/reads/v6/minimum/"

process_all_set_minimumsv6("GEX", samples$samples, input_dir, TRUE, output_dir, min_reads_cell = 2)

```

### UMAP coords and cell size

```{r}
samples <- data.frame(
  datasets = c(
    "LP29_D0",
    #"JK85_D13",
    "LP29_D195",
    "LP29_D83",
    "invitro",
    "P_acute_combined",
    "W0",
    "W2"
  ),
  samples = c(
    "D0",
    #"D13",
    "D195",
    "D83",
    "Invitro",
    "Pacute",
    "W0",
    "W2"
  ),
  folders = c(
    "/projects/b1042/GoyalLab/egrody/extractedData/EGS004/singleCell/counts/Mmul_10_mac239/",
    #"/projects/b1042/GoyalLab/egrody/extractedData/EGS004/singleCell/counts/Mmul_10_mac239/",
    "/projects/b1042/GoyalLab/egrody/extractedData/EGS004/singleCell/counts/Mmul_10_mac239/",
    "/projects/b1042/GoyalLab/egrody/extractedData/EGS004/singleCell/counts/Mmul_10_mac239/",
    "/projects/b1042/GoyalLab/egrody/extractedData/EGS002/singleCell/counts/Mmul_10_mac239/",
    "/projects/b1042/GoyalLab/egrody/extractedData/EGS009/singleCell/counts/Mmul_10_mac239/",
    "/projects/b1042/GoyalLab/egrody/extractedData/EGS002/singleCell/counts/Mmul_10_mac239/",
    "/projects/b1042/GoyalLab/egrody/extractedData/EGS002/singleCell/counts/Mmul_10_mac239/"
  )
)

output.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS016/singleCell/Seurat/UMAPcoords/"
if (!dir.exists(output.dir)) {
  dir.create(output.dir, recursive = TRUE)
}

for (i in 1:nrow(samples)) {
  sample_name <- samples$samples[i]
  dataset <- samples$datasets[i]
  folder <- samples$folders[i]
  input.dir <- paste0(folder, "Mmul_10_mac239_", dataset, "/outs/filtered_feature_bc_matrix/")
  
  tryCatch({
    seurat_obj <- SeuratPipeline(input.dir, sample_name, 
                                 output_dir = "/projects/b1042/GoyalLab/egrody/extractedData/EGS016/singleCell/Seurat/plots/", 
                                 plots = TRUE, rds = TRUE)
    
    # Generate expression data frame without gene expression
    sample_df <- targetExpressionDF(seurat_obj, genes = "")

    # Add sample and target columns to the dataframe
    sample_df$size <- colSums(GetAssayData(seurat_obj, assay = "RNA", slot = "counts"))
    sample_df$sample <- sample_name

    write.csv(sample_df, paste0(output.dir, sample_name, "UMAP_coords.csv"))
    
    cat("Processed sample:", sample_name, "\n")
  }, error = function(e) {
    cat("Error processing sample", sample_name, ":", conditionMessage(e), "\n\n")
  })
}
```


### Seurat deep dive

```{r}
calculate_object_diversity <- function(seurat_obj, celltype_col = "seurat_clusters") {
  
  # Check if celltype column exists
  if (!celltype_col %in% colnames(seurat_obj@meta.data)) {
    stop(paste("Column", celltype_col, "not found in metadata. Available columns:", 
               paste(colnames(seurat_obj@meta.data), collapse = ", ")))
  }
  
  # 1. Cell type diversity (Shannon)
  cell_counts <- table(seurat_obj@meta.data[[celltype_col]])
  proportions <- cell_counts / sum(cell_counts)
  
  shannon <- -sum(proportions * log(proportions))
  simpson <- 1 - sum(proportions^2)
  n_celltypes <- length(cell_counts)
  
  # 2. Gene expression variance
  expr_data <- GetAssayData(seurat_obj, slot = "data")
  
  # Calculate coefficient of variation for each gene
  gene_cv <- apply(expr_data, 1, function(x) {
    if(mean(x) == 0) return(0)
    sd(x) / mean(x)
  })
  
  mean_gene_cv <- mean(gene_cv, na.rm = TRUE)
  median_gene_cv <- median(gene_cv, na.rm = TRUE)
  
  # Return results
  return(list(
    shannon_diversity = shannon,
    simpson_diversity = simpson,
    n_celltypes = n_celltypes,
    mean_gene_cv = mean_gene_cv,
    median_gene_cv = median_gene_cv
  ))
}

# Now calculate diversity for each sample
diversity_invitro <- calculate_object_diversity(seurat_df$Invitro)
diversity_pacute <- calculate_object_diversity(seurat_df$Pacute)
print(diversity_invitro)
print(diversity_pacute)




output.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS016/singleCell/Seurat/plots/"

features <- list(
  CD4 = "CD4",
  CD3 = c("CD3D", "CD3E", "CD3G"),
  CD8 = c("CD8A", "CD8B"),
  Treg = c("FOXP3", "IL2RA"),
  Bcell = c("CD19", "CD20", "CD79A", "CD79B"),
  NK = c("NCAM1", "FCGR3A"),
  Mono = c("CSF1R", "LYZ"),
  DC = c("CD1C", "CLEC9A", "LILRA4", "FCER1A")
)

check_available_genes <- function(seurat_obj, genes) {
  available_genes <- rownames(seurat_obj)
  present_genes <- intersect(genes, available_genes)
  missing_genes <- setdiff(genes, available_genes)
  
  if (length(missing_genes) > 0) {
    cat("Missing genes:", paste(missing_genes, collapse = ", "), "\n")
  }
  
  return(present_genes)
}

# Loop through each sample
for (sample_name in names(seurat_df)) {
  current_obj <- seurat_df[[sample_name]]
  
  cat("Creating plots for sample:", sample_name, "\n")
  
  # Loop through each celltype for this sample
  for (celltype_name in names(features)) {
    genes <- features[[celltype_name]]
    
    cat("  Processing celltype:", celltype_name, "\n")
    
    # Check which genes are available
    available_genes <- check_available_genes(current_obj, genes)
    
    # Skip if no genes are available
    if (length(available_genes) == 0) {
      cat("    Skipping", celltype_name, "for", sample_name, "- no genes found\n")
      next
    }
    
    # Create individual FeaturePlots for each available gene
    gene_plots <- list()
    
    for (gene in available_genes) {
      plot <- FeaturePlot(current_obj, features = gene, pt.size = 0.5) +
        ggtitle(gene) +
        theme(plot.title = element_text(size = 14, hjust = 0.5, face = "bold"),
              axis.text = element_blank(),
              axis.ticks = element_blank(),
              legend.position = "right")
      
      gene_plots[[gene]] <- plot
    }
    
    # Arrange in grid based on number of available genes
    n_genes <- length(gene_plots)
    
    if (n_genes == 1) {
      # Single gene - just use the plot directly
      combined_plot <- gene_plots[[1]]
      width <- 8
      height <- 6
    } else {
      # Multiple genes - arrange in grid
      if (n_genes <= 4) {
        ncol <- 2
        nrow <- 2
      } else if (n_genes <= 6) {
        ncol <- 3
        nrow <- 2
      } else {
        ncol <- 3
        nrow <- ceiling(n_genes / 3)
      }
      
      combined_plot <- plot_grid(plotlist = gene_plots, 
                                ncol = ncol, 
                                nrow = nrow,
                                align = "hv")
      
      width <- ncol * 4
      height <- nrow * 4
    }
    
    # Add overall title
    title_plot <- ggdraw() + 
      draw_label(paste(sample_name, "-", celltype_name, "Markers"), 
                size = 16, fontface = "bold")
    
    final_plot <- plot_grid(title_plot, combined_plot, 
                           ncol = 1, 
                           rel_heights = c(0.1, 0.9))
    
    # Save as SVG - ONE FILE PER CELLTYPE
    filename <- paste0(output.dir, sample_name, "_FeaturePlot_", celltype_name, ".svg")
    
    ggsave(filename, final_plot, 
           width = width, height = height + 1, 
           device = "svg", dpi = 300)
    
    cat("    Saved:", filename, "\n")
  }
}
```

Coexpression of CD4 and CD3s
```{r}
samples <- c(
    "D0",
    "D195",
    "D83",
    "Invitro",
    "Pacute",
    "W0",
    "W2"
)

genes <- c(
  "CD4",
  "CD3D", "CD3E", "CD3G",
  "CD8A", "CD8B"
)

input.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS016/singleCell/Seurat/RDS/"
output.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS016/singleCell/Seurat/expressionDF/"
if (!dir.exists(output.dir)) {
  dir.create(output.dir, recursive = TRUE)
}
for (i in 1:length(samples)) {
  sample_name <- samples[i]
  input_file <- paste0(input.dir, sample_name, ".rds")
  
  tryCatch({
    seurat_obj <- readRDS(input_file)

    # Generate expression data frame
    sample_df <- targetExpressionDF(seurat_obj, genes = genes)
    sample_df$sample <- sample_name

    write.csv(sample_df, paste0(output.dir, sample_name, "_CD4T.csv"))
    
    cat("Processed sample:", sample_name, "\n")
  }, error = function(e) {
    cat("Error processing sample", sample_name, ":", conditionMessage(e), "\n\n")
  })
}

invitro_coex <- targetExpressionDF(seurat_objects$Invitro, genes = genes)
invitro_coexp <- invitro_coex %>%
  mutate(CD3 = ifelse(CD3D > 0 | CD3E > 0 | CD3G > 0, 1, 0)) %>%
  mutate(CD4 = ifelse(CD4 > 0 , 1, 0)) %>%
  mutate(CD8 = ifelse(CD8A > 0 , 1, 0)) %>%
  select(cellID, CD4, CD3, CD8)

current_df <- joint_data$Invitro
current_df <- left_join(current_df, invitro_coexp)
celltype_df <- current_df %>%
  mutate(GEX = ifelse(CD4 > 0 & CD3 > 0, GEX, 0)) %>% #keep only CD4s
  mutate(SALVE = ifelse(CD4 > 0 & CD3 > 0, SALVE, 0)) %>%
  filter(CD4 == 0 | CD8 == 0)

both <- celltype_df %>% filter(GEX != 0 & SALVE != 0)
onlySingleCell <- celltype_df %>% 
  filter(GEX != 0) %>%
  filter(!(cellID %in% both$cellID))
onlySALVE <- celltype_df %>% 
  filter(SALVE != 0) %>%
  filter(!(cellID %in% both$cellID))
cd4 <- celltype_df %>%
  filter(CD4 != 0 & CD3 != 0)
cd4.8 <- celltype_df %>%
  filter(CD4 != 0 & CD8 != 0)

cat("Sample\tCellType\tIn_type\tTotal_cells\tboth\t10X_only\tSALVE_only\n",
    "Invitro\t",
    "CD4 Ts\t",
    nrow(cd4), "\t",
    nrow(celltype_df), "\t",
    nrow(both), "\t",
    nrow(onlySingleCell), "\t",
    nrow(onlySALVE), "\n")
print(nrow(cd4.8))

```


### Supervised clustering
```{r}
# Loading Seurat object
samples <- data.frame(
  datasets = c(
    "invitro"
  ),
  samples = c(
    "Invitro"
  ),
  folders = c(
    "/projects/b1042/GoyalLab/egrody/extractedData/EGS002/singleCell/counts/Mmul_10_mac239/"
  )
)

sample_name <- samples$samples
dataset <- samples$datasets
folder <- samples$folders
input.dir <- paste0(folder, "Mmul_10_mac239_", dataset, "/outs/filtered_feature_bc_matrix/")

# Run everything from SeuratPipeline until clustering
data <- Read10X(data.dir = input.dir)
sample <- CreateSeuratObject(counts = data, project = sample_name, min.cells = 3, min.features = 200)
sample <- Add_Mito_Ribo(sample, species = "macaque")
sample <- subset(sample, subset = nFeature_RNA > 200 & nFeature_RNA < 3500 & nCount_RNA < 20000 & percent_mito < 5)
if ("CD4" %in% rownames(sample) && ("CD8A" %in% rownames(sample) || "CD8B" %in% rownames(sample))) {
  cd4_expr <- GetAssayData(sample, layer = "counts")["CD4", ]
  cd8a_expr <- if ("CD8A" %in% rownames(sample)) GetAssayData(sample, layer = "counts")["CD8A", ] else rep(0, ncol(sample))
  cd8b_expr <- if ("CD8B" %in% rownames(sample)) GetAssayData(sample, layer = "counts")["CD8B", ] else rep(0, ncol(sample))
  
  cd4_pos <- cd4_expr > 0
  cd8_pos <- (cd8a_expr > 0) | (cd8b_expr > 0)
  doublets <- cd4_pos & cd8_pos
  
  sample <- subset(sample, cells = colnames(sample)[!doublets])
}
if ("mac239" %in% rownames(sample)) {
  sample <- sample[!rownames(sample) %in% "mac239", ]
}
sample <- NormalizeData(sample, verbose = FALSE)
sample <- FindVariableFeatures(sample, selection.method = "vst", nfeatures = 2000, verbose = FALSE)
genes <- rownames(sample)
sample <- ScaleData(sample, features = genes, verbose = FALSE)
sample <- RunPCA(sample, features = VariableFeatures(object = sample), verbose = FALSE)
sample <- FindNeighbors(sample, dims = 1:30, verbose = FALSE)
sample <- FindClusters(sample, resolution = 0.3, verbose = FALSE)
sample <- RunUMAP(sample, dims = 1:30, verbose = FALSE)
DimPlot(sample, label = TRUE, pt.size = 0.5) +
  ggtitle("Unsupervised Classification")


# Define marker gene sets for each T cell subset
effector_markers <- c("GZMA", "GZMB", "GZMK", "PRF1", "NKG7", "IFNG", "TNF", "CCL3", "CCL4")
memory_markers <- c("IL7R", "CCR7", "SELL", "TCF7", "LEF1", "CD27", "CD28")
treg_markers <- c("FOXP3", "IL2RA", "CTLA4", "IKZF2", "TIGIT", "TNFRSF18", "TNFRSF4")

# Add module scores to your Seurat object
seurat_obj <- AddModuleScore(
  object = sample,
  features = list(effector_markers),
  name = "Effector_Score"
)

seurat_obj <- AddModuleScore(
  object = seurat_obj,
  features = list(memory_markers),
  name = "Memory_Score"
)

seurat_obj <- AddModuleScore(
  object = seurat_obj,
  features = list(treg_markers),
  name = "Treg_Score"
)

# Classify cells with hierarchical priority: CD8 > Treg > Effector > Memory > Other
seurat_obj$supervised_cluster <- "Other"

# First identify CD8+ T cells (highest priority - distinct lineage)
# Use normalized expression data
cd8_expression <- GetAssayData(seurat_obj, slot = "data")["CD8A", ]
cd8_threshold <- quantile(cd8_expression[cd8_expression > 0], 0.3, na.rm = TRUE)

# If CD8A is detected (non-zero), use expression threshold
# Otherwise, all cells are considered CD8-negative
if(sum(cd8_expression > 0) > 0) {
  seurat_obj$supervised_cluster[cd8_expression > cd8_threshold] <- "CD8"
}

# Then identify Tregs from CD8-negative cells
non_cd8 <- seurat_obj$supervised_cluster != "CD8"
treg_threshold <- quantile(seurat_obj$Treg_Score1[non_cd8], 0.75)
seurat_obj$supervised_cluster[non_cd8 & 
                               seurat_obj$Treg_Score1 > treg_threshold] <- "Treg"

# Then identify effector cells from remaining CD8-negative, non-Treg cells
non_treg_non_cd8 <- seurat_obj$supervised_cluster %in% c("Other")
effector_threshold <- quantile(seurat_obj$Effector_Score1[non_treg_non_cd8], 0.7)
seurat_obj$supervised_cluster[non_treg_non_cd8 & 
                               seurat_obj$Effector_Score1 > effector_threshold] <- "Effector"

# Then identify memory cells from remaining cells
remaining <- seurat_obj$supervised_cluster == "Other"
memory_threshold <- quantile(seurat_obj$Memory_Score1[remaining], 0.7)
seurat_obj$supervised_cluster[remaining & 
                              seurat_obj$Memory_Score1 > memory_threshold] <- "Memory"

# Convert to factor with desired order
seurat_obj$supervised_cluster <- factor(seurat_obj$supervised_cluster,
                                        levels = c("CD8", "Treg", "Effector", "Memory", "Other"))

# Visualize results
FeaturePlot(seurat_obj, features = c("CD8A", "Effector_Score1", "Memory_Score1", "Treg_Score1"))




# Get CD8A expression from normalized data
cd8a_expression <- GetAssayData(seurat_obj, slot = "data")["CD8A", ]

# Initialize classification
seurat_obj$supervised_cluster <- "Other"

# 1. Identify CD8+ T cells (use stricter threshold given clear separation)
# Only cells with CD8A > 2 appear to be true CD8+ cells based on your plot
cd8_threshold <- 2.0
seurat_obj$supervised_cluster[cd8a_expression > cd8_threshold] <- "CD8"

# 2. Identify Tregs from CD8-negative cells
# Use a more stringent threshold since Treg_Score1 is relatively low overall
non_cd8 <- seurat_obj$supervised_cluster != "CD8"
treg_threshold <- quantile(seurat_obj$Treg_Score1[non_cd8], 0.80)  # Top 20%
seurat_obj$supervised_cluster[non_cd8 & 
                               seurat_obj$Treg_Score1 > treg_threshold] <- "Treg"

# 3. Identify effector cells from remaining cells
# Effector score shows clearer signal
remaining <- seurat_obj$supervised_cluster == "Other"
effector_threshold <- quantile(seurat_obj$Effector_Score1[remaining], 0.65)  # Top 35%
seurat_obj$supervised_cluster[remaining & 
                               seurat_obj$Effector_Score1 > effector_threshold] <- "Effector"

# 4. Identify memory cells from remaining cells
remaining <- seurat_obj$supervised_cluster == "Other"
memory_threshold <- quantile(seurat_obj$Memory_Score1[remaining], 0.65)  # Top 35%
seurat_obj$supervised_cluster[remaining & 
                              seurat_obj$Memory_Score1 > memory_threshold] <- "Memory"

# Convert to factor
seurat_obj$supervised_cluster <- factor(seurat_obj$supervised_cluster,
                                        levels = c("CD8", "Treg", "Effector", "Memory", "Other"))

# Visualize the classification
DimPlot(seurat_obj, group.by = "supervised_cluster", label = TRUE, pt.size = 0.5) +
  ggtitle("Supervised T Cell Classification")

# Check cluster sizes
table(seurat_obj$supervised_cluster)

```


# Joint Dataset
## Joint

### KLRB1
```{r}
# Loading
input.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS016/SALVE/bamsort/reads/KLRB1/minimum/"
W0_min <- read.csv(paste0(input.dir, "SALVE_KLRB1_W0_SALVE_minreads5.csv"), row.names = "X")
SALVE_KLRB1_W0 <- W0_min %>%
  select(-total) %>%
  mutate(KLRB1 = log1p(KLRB1))

input.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS002/singleCell/bamsort/reads/KLRB1/minimum/" 
GEX_KLRB1_W0 <- read.csv(paste0(input.dir, "W0_minread1_minumi1.csv"), row.names = "X")
GEX_KLRB1_W0 <- GEX_KLRB1_W0 %>%
  select(-total) %>%
  mutate(KLRB1 = log1p(KLRB1))
GEX_KLRB1_W0_expressionDF <- read.csv("/projects/b1042/GoyalLab/egrody/extractedData/EGS002/singleCell/counts/Mmul_10_only/expressionDF/W0_CD4T.csv", row.names = "X")
GEX_KLRB1_W0_expressionDF <- GEX_KLRB1_W0_expressionDF %>% select(cellID, UMAP1, UMAP2, cluster)
GEX_KLRB1_W0 <- left_join(GEX_KLRB1_W0_expressionDF, GEX_KLRB1_W0, by = "cellID")


# Joining
W0_joint <- left_join(GEX_KLRB1_W0, SALVE_KLRB1_W0, by = "cellID") %>%
  rename(GEX = KLRB1.x, SALVE = KLRB1.y)
W0_joint[is.na(W0_joint)] <- 0


# Numbers
name <- "W0_joint"
current_df <- get(name)
both <- current_df %>% filter(GEX != 0 & SALVE != 0)
onlySingleCell <- current_df %>% 
  filter(GEX != 0) %>%
  filter(!(cellID %in% both$cellID))
onlySALVE <- current_df %>% 
  filter(SALVE != 0) %>%
  filter(!(cellID %in% both$cellID))

cat("\nSample\tTotal_cells\tboth\t10X_only\tSALVE_only\n",
    name, "\t",
    nrow(current_df), "\t",
    nrow(both), "\t",
    nrow(onlySingleCell), "\t",
    nrow(onlySALVE), "\n")

  
# for in-text comparison of # UMI and # cells between methods
# run the following on non-log1p transformed KLRB1 values
cat("\n# UMI in GEX total:\t", sum(current_df$GEX),
    "\n# UMI in GEX only:\t", sum(onlySingleCell$GEX),
    "\n% UMI in GEX only:\t", round(sum(onlySingleCell$GEX) / sum(current_df$GEX) * 100, 2), "%",
    "\n# Cells in GEX total:\t", sum(current_df$GEX > 0),
    "\n# Cells in GEX only:\t", sum(onlySingleCell$GEX > 0),
    "\n% Cells in GEX only:\t", round(sum(onlySingleCell$GEX > 0) / sum(current_df$GEX > 0) * 100, 2), "%")

# plotting
output.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS016/joint/plots/KLRB1/"

plotUMAP(W0_joint, colorby = GEX,
         "W0 joint KLRB1: GEX (bamsort)",
         output.dir,
         "KLRB1_W0_GEX_bamsort.pdf",
         comparison = FALSE, color = "gray25")
plotUMAP(W0_joint, colorby = SALVE,
         "W0 joint KLRB1: SALVE (bamsort)",
         output.dir,
         "KLRB1_W0_SALVE_bamsort_minreads5.pdf",
         comparison = FALSE, color = "gray25")

output_dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS016/joint/dataframes/"
write.csv(both, paste0(output_dir, "KLRB1_both.csv"), row.names = FALSE)
write.csv(onlySingleCell, paste0(output_dir, "KLRB1_onlySingleCell.csv"), row.names = FALSE)
write.csv(onlySALVE, paste0(output_dir, "KLRB1_onlySALVE.csv"), row.names = FALSE)
write.csv(W0_joint, paste0(output_dir, "KLRB1_all.csv"), row.names = FALSE)
```
Correlations
```{r}
either <- W0_joint %>% filter(GEX != 0 | SALVE != 0)
both <- W0_joint %>% filter(GEX != 0 & SALVE != 0)

p <- ggplot(data = W0_joint, aes(x = SALVE, y = GEX)) +
  #geom_smooth(method = "lm", se = TRUE, color = "red", alpha = 0.7) +
  geom_jitter(alpha = 0.6, size = 1.5, width = 0.1, height = 0.1) +
  #geom_abline(intercept = 0, slope = 1, color = "red") +
  labs(
    x = "SALVE",
    y = "GEX",
    title = paste("W0 KLRB1 Correlation: SALVE vs GEX")
  ) +
  theme_minimal() +
  coord_fixed(ratio = 1) +
  xlim(0, 3) +
  ylim(0, 3)
print(p)
p_exclude_zero <- ggplot(data = subset(W0_joint, SALVE > 0 | GEX > 0), aes(x = SALVE, y = GEX)) +
  #geom_abline(intercept = 0, slope = 1, color = "red", linewidth = 1) +
  geom_smooth(method = "lm", se = TRUE, color = "white", alpha = 0.7) +
  geom_bin2d(bins = 18, alpha = 0.8) +
  labs(
    x = "SALVE",
    y = "GEX",
    title = "2D Binning (Excluding 0,0 Point)"
  ) +
  theme_minimal() +
  coord_cartesian(xlim = c(0, 3), ylim = c(0, 3)) +
  scale_fill_viridis_c(name = "Count") +
  theme(aspect.ratio = 1) +
  guides(fill = guide_colorbar(raster = TRUE))
print(p_exclude_zero)
ggsave("/projects/b1042/GoyalLab/egrody/extractedData/EGS016/joint/plots/KLRB1/KLRB1_correlation_jitter.svg", 
       plot = p)
ggsave("/projects/b1042/GoyalLab/egrody/extractedData/EGS016/joint/plots/KLRB1/KLRB1_correlation_2Dbinning.svg", 
       plot = p_exclude_zero)

max(W0_joint$SALVE)

cat("Pearson\t\tSpearman\t\tJaccard\n",
    cor(W0_joint$SALVE, W0_joint$GEX, method = "pearson"), "\t",
    cor(W0_joint$SALVE, W0_joint$GEX, method = "spearman"), "\t",
    sum((W0_joint$SALVE > 0) & (W0_joint$GEX > 0)) / sum((W0_joint$SALVE > 0) | (W0_joint$GEX > 0)), "\n",
    cor(either$SALVE, either$GEX, method = "pearson"), "\t",
    cor(either$SALVE, either$GEX, method = "spearman"), "\t",
    sum((either$SALVE > 0) & (either$GEX > 0)) / sum((either$SALVE > 0) | (either$GEX > 0)), "\n",
    cor(both$SALVE, both$GEX, method = "pearson"), "\t",
    cor(both$SALVE, both$GEX, method = "spearman"), "\t",
    sum((both$SALVE > 0) & (both$GEX > 0)) / sum((both$SALVE > 0) | (both$GEX > 0))
)

```


### bamsort
```{r}
# Loading
samples <-  c(
    "D0",
    "D195",
    "Invitro",
    "Pacute",
    "W0",
    "W2"
)

SALVE.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/SALVE/bamsort/reads/v6/minimum/"
GEX.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS016/singleCell/bamsort/reads/v6/minimum/"

joint_data <- list()

for (i in 1:length(samples)) {
  sample_name <- samples[i]
  
  # Read all data files
  salve_file <- paste0(SALVE.dir, sample_name, "_SALVE_filtered.csv")
  gex_file <- paste0(GEX.dir, sample_name, "_GEX_filtered.csv")
  umap_file <- paste0("/projects/b1042/GoyalLab/egrody/extractedData/EGS016/singleCell/Seurat/UMAPcoords/", 
                      sample_name, "UMAP_coords.csv")
  cd4_file <- paste0("/projects/b1042/GoyalLab/egrody/extractedData/EGS016/singleCell/Seurat/expressionDF/", 
                     sample_name, "_CD4T.csv")
  
  salve_data <- if(file.exists(salve_file)) read.csv(salve_file) else NULL
  gex_data <- if(file.exists(gex_file)) read.csv(gex_file) else NULL
  umap_data <- if(file.exists(umap_file)) read.csv(umap_file, row.names = "X") else NULL
  cd4_data <- if(file.exists(cd4_file)) read.csv(cd4_file, row.names = "X") else NULL
  
  # Skip if no data at all
  if (is.null(salve_data) && is.null(gex_data) && is.null(umap_data)) {
    cat("Skipping", sample_name, ": no data\n")
    next
  }
  
  # Get cellIDs from Mmul_10 only df
  all_cells <- as.character(umap_data$cellID)
  all_cells <- unique(all_cells)
  
  # Create base dataframe
  result <- data.frame(cellID = all_cells)
  
  # Add UMAP data
  if (!is.null(umap_data)) {
    umap_processed <- umap_data %>%
      select(-V2, -sample)
    result <- left_join(result, umap_processed, by = "cellID")
  }
  
  # Add GEX data
  if (!is.null(gex_data)) {
    gex_processed <- gex_data %>%
      mutate(cellID = as.character(cellID)) %>%
      select(cellID, total, US)
    result <- left_join(result, gex_processed, by = "cellID") %>%
      mutate(GEX = log1p(total)) %>%
      mutate(GEX.US = log1p(US)) %>%
      select(-total, -US)
  } else {
    result$GEX <- 0
  }
  
  # Add SALVE data
  if (!is.null(salve_data)) {
    salve_processed <- salve_data %>%
      #group_by(cellID) %>%
      rename(SALVE = total) %>%
      mutate(cellID = as.character(cellID)) %>%
      select(cellID, SALVE, US, SS, MS)
    result <- left_join(result, salve_processed, by = "cellID") %>%
      mutate(SALVE = log1p(SALVE)) %>%
      mutate(USprop = US / (US+MS+SS)) %>%
      mutate(SSprop = SS / (US+MS+SS)) %>%
      mutate(MSprop = MS / (US+MS+SS)) %>%
      mutate(US = log1p(US)) %>%
      mutate(SS = log1p(SS)) %>%
      mutate(MS = log1p(MS))
  } else {
    result$SALVE <- 0
  }
  
  # Replace NAs with 0
  result[is.na(result)] <- 0
  
  # Cell type filtering
  cd4_data <- cd4_data %>%
    mutate(CD3 = ifelse(CD3D > 0 | CD3E > 0 | CD3G > 0, 1, 0)) %>%
    mutate(CD4 = ifelse(CD4 > 0 , 1, 0)) %>%
    mutate(CD8 = ifelse(CD8A > 0 , 1, 0)) %>%
    select(cellID, CD4, CD3, CD8)
  result <- left_join(result, cd4_data, by = "cellID")
  result <- result %>%
    mutate(across(any_of(c("GEX", "GEX.US")), 
              ~ ifelse(CD4 > 0 & CD3 > 0, .x, 0))) %>% 
    mutate(across(any_of(c("SALVE", "US", "MS", "SS")), 
              ~ ifelse(CD4 > 0 & CD3 > 0, .x, 0))) %>%
    filter(CD4 == 0 | CD8 == 0) %>% # exclude cells CD4+ and CD8+
    select(-CD8, -CD3, -CD4)
  
  joint_data[[sample_name]] <- result
  cat("Processed", sample_name, "\n")
}

```

Saving
```{r}
output_dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/joint/dataframes/"
#cat("\nSample\tTotal_cells\tboth\t10X_only\tSALVE_only\n")

# Process each dataset in the joint_data list
for (sample_name in names(joint_data)) {
  current_df <- joint_data[[sample_name]]
  
  both <- current_df %>% filter(GEX != 0 & SALVE != 0)
  either <- current_df %>% filter(GEX != 0 | SALVE != 0)
  onlySingleCell <- current_df %>% 
    filter(GEX != 0) %>%
    filter(!(cellID %in% both$cellID))
  onlySALVE <- current_df %>% 
    filter(SALVE != 0) %>%
    filter(!(cellID %in% both$cellID))

  cat(sample_name, "\t",
      nrow(current_df), "\t",
      nrow(both), "\t",
      nrow(onlySingleCell), "\t",
      nrow(onlySALVE), "\n")

  # cat(sample_name, "\tCorrelation:", cor(current_df$GEX, current_df$SALVE), "\n",
  #     "\tMax SALVE:", max(current_df$SALVE), "\n",
  #     "\tMax GEX:", max(current_df$GEX), "\n\n")


  # p <- ggplot(data = current_df, aes(x = SALVE, y = GEX)) +
  #   geom_jitter(alpha = 0.6, size = 1.5, width = 0.1, height = 0.1) +
  #   labs(
  #     x = "SALVE",
  #     y = "GEX",
  #     title = paste0(sample_name, " SIV total Correlation")
  #   ) +
  #   theme_minimal() +
  #   coord_fixed(ratio = 1) +
  #   xlim(0, 9.1) +
  #   ylim(0, 9.1)
  # print(p)
  # p_exclude_zero <- ggplot(data = subset(current_df, SALVE > 0 | GEX > 0), aes(x = SALVE, y = GEX)) +
  #   #geom_abline(intercept = 0, slope = 1, color = "red", linewidth = 1) +
  #   geom_smooth(method = "lm", se = TRUE, color = "white", alpha = 0.7) +
  #   geom_bin2d(bins = 30, alpha = 0.8) +
  #   labs(
  #     x = "SALVE",
  #     y = "GEX",
  #     title = paste0(sample_name, " SIV total 2D Binning (Excluding 0,0)")
  #   ) +
  #   theme_minimal() +
  #   coord_cartesian(xlim = c(0, 9.1), ylim = c(0, 9.1)) +
  #   scale_fill_viridis_c(name = "Count") +
  #   theme(aspect.ratio = 1) +
  #   guides(fill = guide_colorbar(raster = TRUE))
  # print(p_exclude_zero)
  # output.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/joint/plots/"
  # ggsave(paste0(output.dir, sample_name, "_CD4_correlation_jitter.svg"),
  #        plot = p)
  # ggsave(paste0(output.dir, sample_name, "_CD4_correlation_2Dbinning.svg"),
  #        plot = p_exclude_zero)
  # 
  # 
  # output.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/joint/plots/"
  # plotUMAP(current_df, colorby = GEX,
  #        paste0(sample_name, " joint SIV total: GEX (bamsort)"),
  #        output.dir,
  #        paste0(sample_name, "_CD4_GEX_bamsort.pdf"),
  #        comparison = FALSE, color = "red")
  # plotUMAP(current_df, colorby = SALVE,
  #        paste0(sample_name, " joint SIV total: SALVE (bamsort)"),
  #        output.dir,
  #        paste0(sample_name, "_CD4_SALVE_bamsort.pdf"),
  #        comparison = FALSE, color = "red")

  # SKIP this chunk
  # pearson_cor <- cor(current_df$all_spliced, current_df$spliced.x, use = "complete.obs")
  # r_squared <- pearson_cor^2
  # lm_model <- lm(spliced.x ~ all_spliced, data = current_df)
  # slope <- coef(lm_model)[2]
  # p <- ggplot(data = current_df, aes(x = all_spliced, y = spliced.x)) +
  #     geom_point() +
  #   geom_smooth(method = "lm", se = TRUE, color = "red") +
  #     labs(
  #       x = "SALVE spliced",
  #       y = "GEX spliced",
  #       title = paste("Correlation for ", sample_name)
  #     ) +
  #     theme_minimal()# +
  #     # annotate("text",
  #     #      x = -Inf, y = Inf,
  #     #      label = paste("Slope =", round(slope, 3), "\n",
  #     #                   "R² =", round(r_squared, 3), "\n",
  #     #                   "r =", round(pearson_cor, 3)),
  #     #      hjust = -0.1, vjust = 1.2,
  #     #      size = 4, color = "black")
  # print(p)

  
  #Save CSVs
  # write.csv(both, paste0(output_dir, sample_name, "_both.csv"), row.names = FALSE)
  # write.csv(onlySingleCell, paste0(output_dir, sample_name, "_onlySingleCell.csv"), row.names = FALSE)
  # write.csv(onlySALVE, paste0(output_dir, sample_name, "_onlySALVE.csv"), row.names = FALSE)
  # write.csv(current_df, paste0(output_dir, sample_name, "_all.csv"), row.names = FALSE)
}

```

Splicing proportions
```{r proportions}
output_dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/joint/plots/proportions/"
for (sample_name in names(joint_data)) {
  df <- joint_data[[sample_name]]
  
  df_sorted <- df %>%
    filter(SALVE > 0) %>%
    arrange(desc(USprop), desc(SSprop)) %>%
    mutate(cellID_ordered = factor(cellID, levels = cellID))
  df_long <- df_sorted %>%
    pivot_longer(cols = contains("prop"), 
                 names_to = "component", 
                 values_to = "proportion")
  
  p1 <- ggplot(df_long, aes(x = cellID_ordered, y = proportion, fill = component)) +
    geom_bar(stat = "identity", width = 1) +
    scale_y_continuous(expand = c(0, 0)) +
    labs(title = paste0(sample_name, " Isoforms: Stacked Bar Chart"),
         x = "Cell ID",
         y = "Proportion",
         fill = "Component") +
    theme_minimal() +
    theme(axis.text.x = element_blank(),  # Remove x-axis labels for clarity
          axis.ticks.x = element_blank(),
          panel.grid.major.x = element_blank())
  
  ggsave(paste0(output_dir, sample_name, "_proportions_stacked.pdf"), p1, width = 10, height = 6, dpi = 300)
  
  
  
  p4 <- ggplot(df_long, aes(x = component, y = proportion, fill = component)) +
    geom_violin(alpha = 0.6) +
    geom_boxplot(width = 0.3, alpha = 0.8, outlier.alpha = 0.5) +
    labs(title = paste0(sample_name, " Isoforms: Violins"),
         x = "Isoform Proportion",
         y = "Proportion",
         fill = "Component") +
    theme_minimal() +
    theme(legend.position = "none")
  
  ggsave(paste0(output_dir, sample_name, "_proportions_violin.pdf"), p4, width = 8, height = 6, dpi = 300)
}

```

Splicing UMAPS (not updated)
```{r}
## for lightning talk figures
  output.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS016/joint/plots/v5/subset/"
  plotUMAP(current_df, colorby = US,
         paste0(sample_name, " joint SALVE US"),
         output.dir,
         paste0(sample_name, "_umap_SALVE_US.svg"),
         comparison = FALSE, color = "royalblue1")
  plotUMAP(current_df, colorby = spliced,
         paste0(sample_name, " joint SALVE spliced"),
         output.dir,
         paste0(sample_name, "_umap_SALVE_spliced.svg"),
         comparison = FALSE, color = "red")
  plotUMAP(current_df, colorby = SALVE,
         paste0(sample_name, " joint SALVE"),
         output.dir,
         paste0(sample_name, "_umap_SALVE.svg"),
         comparison = FALSE, color = "purple4")
    plotUMAP(current_df, colorby = GEX,
         paste0(sample_name, " joint GEX"),
         output.dir,
         paste0(sample_name, "_umap_GEX.svg"),
         comparison = FALSE, color = "purple4")
```

### combined counts
```{r}
# Loading
samples <-  c(
    "D0",
    "D195",
    "Invitro",
    "Pacute",
    "W0",
    "W2"
)

SALVE.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/SALVE/bamsort/reads/v6/"
GEX.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS016/singleCell/bamsort/reads/v6/"

combined_data <- list()

for (i in 1:length(samples)) {
  sample_name <- samples[i]
  cat("Processing", sample_name, "\n")
  
  # Read all data files
  salve_file <- paste0(SALVE.dir, sample_name, "_UMI_read_counts_full.csv")
  gex_file <- paste0(GEX.dir, sample_name, "_UMI_read_counts_full.csv")
  umap_file <- paste0("/projects/b1042/GoyalLab/egrody/extractedData/EGS016/singleCell/Seurat/UMAPcoords/", 
                      sample_name, "UMAP_coords.csv")
  cd4_file <- paste0("/projects/b1042/GoyalLab/egrody/extractedData/EGS016/singleCell/Seurat/expressionDF/", 
                     sample_name, "_CD4T.csv")
  salve_umi_file <- paste0(SALVE.dir, "minimum/valid_UMI/", sample_name, "_SALVE_valid_cellID_UMI.csv")
  gex_umi_file <- paste0(GEX.dir, "minimum/valid_UMI/", sample_name, "_GEX_valid_cellID_UMI.csv")
  
  salve_data <- if(file.exists(salve_file)) read.csv(salve_file) else NULL
  gex_data <- if(file.exists(gex_file)) read.csv(gex_file) else NULL
  salve_umi <- if(file.exists(salve_umi_file)) read.csv(salve_umi_file) else NULL
  gex_umi <- if(file.exists(gex_umi_file)) read.csv(gex_umi_file) else NULL
  umap_data <- if(file.exists(umap_file)) read.csv(umap_file, row.names = "X") else NULL
  cd4_data <- if(file.exists(cd4_file)) read.csv(cd4_file, row.names = "X") else NULL
  
  # Skip if no data
  if (is.null(salve_data) | is.null(gex_data) | is.null(umap_data)) {
    cat("Skipping", sample_name, ": no data\n")
    next
  }
  
  full_preprocess <- NULL
  
  # Join GEX and SALVE and filter by minimums-applied UMI
  if (!is.null(gex_data) && !is.null(gex_umi)) {
    full_preprocess <- gex_data %>%
      mutate(cellID = as.character(cellID)) %>%
      semi_join(gex_umi, by = c("cellID", "UMI"))
  }
  if (!is.null(salve_data) && !is.null(salve_umi)) {
    salve_processed <- salve_data %>%
      mutate(cellID = as.character(cellID)) %>%
      semi_join(salve_umi, by = c("cellID", "UMI"))
    full_preprocess <- full_join(full_preprocess, salve_processed, by = c("cellID", "UMI"))
  }
  
  if (is.null(full_preprocess)) {
    cat("Skipping", sample_name, ": no GEX or SALVE data\n")
    next
  }

  full_processed <- resolve_multimap_gs(full_preprocess)
  full_processed <- full_processed %>%
    count(cellID, category) %>%
    pivot_wider(names_from = category, values_from = n, values_fill = 0) %>%
    mutate(total = log1p(rowSums(select(., -cellID), na.rm = TRUE)))
  
  
  # Get cellIDs from Mmul_10 only df
  all_cells <- as.character(umap_data$cellID)
  all_cells <- unique(all_cells)
  
  # Create base dataframe
  result <- data.frame(cellID = all_cells)
  
  # Add UMAP data
  if (!is.null(umap_data)) {
    umap_processed <- umap_data %>%
      select(-V2, -sample)
    result <- left_join(result, umap_processed, by = "cellID")
  }
  
  result <- left_join(result, full_processed, by = "cellID")
  result[is.na(result)] <- 0
  
  # Cell type filtering
  cd4_data <- cd4_data %>%
    mutate(CD3 = ifelse(CD3D > 0 | CD3E > 0 | CD3G > 0, 1, 0)) %>%
    mutate(CD4 = ifelse(CD4 > 0 , 1, 0)) %>%
    mutate(CD8 = ifelse(CD8A > 0 , 1, 0)) %>%
    select(cellID, CD4, CD3, CD8)
  result <- left_join(result, cd4_data, by = "cellID")
  result <- result %>%
    mutate(across(any_of(c("total", "US", "MS", "SS", "any")), 
              ~ ifelse(CD4 > 0 & CD3 > 0, .x, 0))) %>%
    filter(CD4 == 0 | CD8 == 0) %>% # exclude cells CD4+ and CD8+
    select(-size, -CD8, -CD3, -CD4)
  
  combined_data[[sample_name]] <- result
  cat("Processed", sample_name, "\n")
}
```

```{r}
output_dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/combined/dataframes/"
cat("\nSample\tTotal_cells\tcombined\n")

# Process each dataset in the joint_data list
for (sample_name in names(combined_data)) {
  current_df <- combined_data[[sample_name]]
  
  both <- current_df %>% filter(total != 0)

  cat(sample_name, "\t",
      nrow(current_df), "\t",
      nrow(both), "\n")

  # cat(sample_name, "\tCorrelation:", cor(current_df$GEX, current_df$SALVE), "\n",
  #     "\tMax SALVE:", max(current_df$SALVE), "\n",
  #     "\tMax GEX:", max(current_df$GEX), "\n\n")
  
  output.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/combined/plots/"
  plotUMAP(current_df, colorby = total,
         paste0(sample_name, " joint SIV total: combined"),
         output.dir,
         paste0(sample_name, "_CD4_combined_bamsort.pdf"),
         comparison = FALSE, color = "red")

  
  #Save CSVs
  write.csv(both, paste0(output_dir, sample_name, "_both.csv"), row.names = FALSE)
  write.csv(current_df, paste0(output_dir, sample_name, "_all.csv"), row.names = FALSE)
}
```


## Other Joint Analyses
### GEX only - R^2 and JS Divergence

```{r R^2}
samples <- c("Invitro", "Pacute", "KLRB1")
joint.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/joint/"

for (sample in samples) {
  # Load data
  if (sample == "KLRB1") {
    seurat_obj <- readRDS(
    "/projects/b1042/GoyalLab/egrody/extractedData/EGS016/singleCell/Seurat/RDS/W0.rds")
    cellsList_GEXonly <- read.csv(
    paste0("/projects/b1042/GoyalLab/egrody/extractedData/EGS016/joint/dataframes/", sample, "_onlySingleCell.csv"))
    cellsList_SALVEonly <- read.csv(
    paste0("/projects/b1042/GoyalLab/egrody/extractedData/EGS016/joint/dataframes/", sample, "_onlySALVE.csv"))
    cellsList_both <- read.csv(
    paste0("/projects/b1042/GoyalLab/egrody/extractedData/EGS016/joint/dataframes/", sample, "_both.csv"))
  }
  else {
    seurat_obj <- readRDS(paste0(
    "/projects/b1042/GoyalLab/egrody/extractedData/EGS016/singleCell/Seurat/RDS/", 
    sample, ".rds"))
    cellsList_GEXonly <- read.csv(
    paste0(joint.dir, "dataframes/", sample, "_onlySingleCell.csv"))
    cellsList_SALVEonly <- read.csv(
    paste0(joint.dir, "dataframes/", sample, "_onlySALVE.csv"))
    cellsList_both <- read.csv(
    paste0(joint.dir, "dataframes/", sample, "_both.csv"))
  }
  
  # Extract expression data once for variable features
  var_features <- VariableFeatures(seurat_obj)
  cells_set1 <- cellsList_GEXonly$cellID
  cells_set2 <- c(cellsList_SALVEonly$cellID, cellsList_both$cellID)
  cells_both <- cellsList_both$cellID
  
  expr1 <- as.matrix(GetAssayData(seurat_obj[, cells_set1], slot = "data")[var_features, ])
  expr2 <- as.matrix(GetAssayData(seurat_obj[, cells_set2], slot = "data")[var_features, ])
  expr_both <- as.matrix(GetAssayData(seurat_obj[, cells_both], slot = "data")[var_features, ])
  
  # GEX vs SALVE
  r_squared_matrix <- cor(expr1, expr2, method = "pearson")^2
  max_r2_set2 <- apply(r_squared_matrix, 1, max)
  
  # SALVE vs SALVE
  r_squared_set2_internal <- cor(expr2, expr2, method = "pearson")^2
  diag(r_squared_set2_internal) <- NA  # Exclude self-correlation
  max_r2_set2_internal <- apply(r_squared_set2_internal, 1, max, na.rm = TRUE)
  
  # GEX vs GEX
  r_squared_gex_internal <- cor(expr1, expr1, method = "pearson")^2
  diag(r_squared_gex_internal) <- NA
  max_r2_gex_internal <- apply(r_squared_gex_internal, 1, max, na.rm = TRUE)
  
  # both vs both
  if (length(cells_both) > 1) {
    r_squared_both_internal <- cor(expr_both, expr_both, method = "pearson")^2
    diag(r_squared_both_internal) <- NA
    max_r2_both_internal <- apply(r_squared_both_internal, 1, max, na.rm = TRUE)
  } else {
    max_r2_both_internal <- numeric(0)
  }
  
  # GEX vs both
  if (length(cells_both) > 0) {
    r_squared_gex_vs_both <- cor(expr1, expr_both, method = "pearson")^2
    max_r2_gex_vs_both <- apply(r_squared_gex_vs_both, 1, max)
  } else {
    max_r2_gex_vs_both <- numeric(0)
  }
  
  # GEX vs random
  n_iterations <- 5
  n_cells_set2 <- length(cells_set1)
  available_cells <- setdiff(
    colnames(seurat_obj), 
    c(cells_set1, cellsList_SALVEonly$cellID, cellsList_both$cellID))
  
  set.seed(123)
  max_r2_per_iteration <- matrix(0, nrow = length(cells_set1), ncol = n_iterations)
  
  for (i in 1:n_iterations) {
    random_cells <- sample(available_cells, n_cells_set2)
    expr_random <- as.matrix(
      GetAssayData(seurat_obj[, random_cells], slot = "data")[var_features, ])
    r2_random <- cor(expr1, expr_random, method = "pearson")^2
    max_r2_per_iteration[, i] <- apply(r2_random, 1, max)
  }
  
  max_r2_random_single <- max_r2_per_iteration[, 1]
  median_r2_random <- apply(max_r2_per_iteration, 1, median)
  
  # Statistical tests comparing each group to GEX vs SALVE (the reference)
  cat("\n\n========== ", sample, " ==========\n")
  cat("\n--- Comparisons vs GEX vs SALVE (reference) ---\n")
  
  # Test 1: GEX vs Random1 (median) vs GEX vs SALVE
  w_rand_med <- wilcox.test(median_r2_random, max_r2_set2, paired = TRUE)
  cat("GEX vs Random1 (median) vs GEX vs SALVE:\t", w_rand_med$p.value, "\n")
  
  # Test 2: GEX vs Random1 (max) vs GEX vs SALVE
  w_rand_max <- wilcox.test(max_r2_random_single, max_r2_set2, paired = TRUE)
  cat("GEX vs Random1 (max) vs GEX vs SALVE:\t\t", w_rand_max$p.value, "\n")
  
  # Test 3: GEX vs GEX vs GEX vs SALVE
  w_gex <- wilcox.test(max_r2_gex_internal, max_r2_set2, paired = TRUE)
  cat("GEX vs GEX vs GEX vs SALVE:\t\t\t", w_gex$p.value, "\n")
  
  # Test 4: GEX vs both vs GEX vs SALVE
  if (length(max_r2_gex_vs_both) > 0) {
    w_both <- wilcox.test(max_r2_gex_vs_both, max_r2_set2, paired = TRUE)
    cat("GEX vs both vs GEX vs SALVE:\t\t\t", w_both$p.value, "\n")
  } else {
    w_both <- list(p.value = NA)
  }
  
  # Test 5: SALVE vs SALVE vs GEX vs SALVE
  w_salve <- wilcox.test(max_r2_set2_internal, max_r2_set2, paired = FALSE)
  cat("SALVE vs SALVE vs GEX vs SALVE:\t\t\t", w_salve$p.value, "\n")
  
  # Test 6: both vs both vs GEX vs SALVE
  if (length(max_r2_both_internal) > 0) {
    w_both_int <- wilcox.test(max_r2_both_internal, max_r2_set2, paired = FALSE)
    cat("both vs both vs GEX vs SALVE:\t\t\t", w_both_int$p.value, "\n")
  } else {
    w_both_int <- list(p.value = NA)
  }
  
  # Function to assign significance stars based on p-value
  get_significance_stars <- function(p_value) {
    if (is.na(p_value)) return("")
    if (p_value < 0.001) return("***")
    if (p_value < 0.01) return("**")
    if (p_value < 0.05) return("*")
    if (p_value < 0.1) return(".")
    return("")
  }
  
  # Store p-values and calculate significance stars
  p_values <- c(w_rand_med$p.value, w_rand_max$p.value, w_gex$p.value,
                w_both$p.value, NA, w_salve$p.value, w_both_int$p.value)
  
  significance_stars <- sapply(p_values, get_significance_stars)
  
  # Extended comparison plot with significance stars
  comparison_extended <- data.frame(
    Comparison = factor(
      rep(c("GEX vs Random1 (median)", "GEX vs Random1 (max)", "GEX vs GEX", 
            "GEX vs both", "GEX vs SALVE", "SALVE vs SALVE", "both vs both"),
          c(length(median_r2_random), length(max_r2_random_single), 
            length(max_r2_gex_internal), length(max_r2_gex_vs_both),
            length(max_r2_set2), length(max_r2_set2_internal), 
            length(max_r2_both_internal))),
      levels = c("GEX vs Random1 (median)", "GEX vs Random1 (max)", "GEX vs GEX",
                 "GEX vs both", "GEX vs SALVE", "SALVE vs SALVE", "both vs both")
    ),
    Max_R2 = c(median_r2_random, max_r2_random_single, max_r2_gex_internal,
               max_r2_gex_vs_both, max_r2_set2, max_r2_set2_internal, 
               max_r2_both_internal)
  )
  
  # Create annotation dataframe for significance stars
  # Position stars above each boxplot at y = 0.58
  star_positions <- data.frame(
    Comparison = levels(comparison_extended$Comparison),
    label = significance_stars,
    y_pos = rep(0.58, 7)
  )
  
  p_extended <- ggplot(comparison_extended, aes(x = Comparison, y = Max_R2, fill = Comparison)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.2, alpha = 0.5, size = 1.5) +
    scale_fill_manual(values = rep("lightgray", 7)) +
    geom_text(data = star_positions, aes(x = Comparison, y = y_pos, label = label),
              inherit.aes = FALSE, size = 6, vjust = 0) +
    theme_minimal() +
    theme(legend.position = "none",
          axis.text.x = element_text(size = 10, angle = 25, hjust = 1),
          plot.title = element_text(hjust = 0.5)) +
    labs(title = paste0(sample, " Extended Correlation Comparison"),
         x = "", y = "Maximum R²") +
    ylim(0, 0.6)
  
  ggsave(paste0(joint.dir, "plots/R2/", sample, "_R2_comparison_extended.svg"),
         p_extended, width = 10, height = 6)
}


```

```{r JS divergence}
library(philentropy)

samples <- c("Invitro", "Pacute", "KLRB1")
joint.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/joint/"

for (sample in samples) {
  # Load data
  if (sample == "KLRB1") {
    seurat_obj <- readRDS(
    "/projects/b1042/GoyalLab/egrody/extractedData/EGS016/singleCell/Seurat/RDS/W0.rds")
    cellsList_GEXonly <- read.csv(
    paste0("/projects/b1042/GoyalLab/egrody/extractedData/EGS016/joint/dataframes/", sample, "_onlySingleCell.csv"))
    cellsList_SALVEonly <- read.csv(
    paste0("/projects/b1042/GoyalLab/egrody/extractedData/EGS016/joint/dataframes/", sample, "_onlySALVE.csv"))
    cellsList_both <- read.csv(
    paste0("/projects/b1042/GoyalLab/egrody/extractedData/EGS016/joint/dataframes/", sample, "_both.csv"))
  }
  else {
    seurat_obj <- readRDS(paste0(
    "/projects/b1042/GoyalLab/egrody/extractedData/EGS016/singleCell/Seurat/RDS/", 
    sample, ".rds"))
    cellsList_GEXonly <- read.csv(
    paste0(joint.dir, "dataframes/", sample, "_onlySingleCell.csv"))
    cellsList_SALVEonly <- read.csv(
    paste0(joint.dir, "dataframes/", sample, "_onlySALVE.csv"))
    cellsList_both <- read.csv(
    paste0(joint.dir, "dataframes/", sample, "_both.csv"))
  }
  
  # Extract expression data for variable features
  var_features <- VariableFeatures(seurat_obj)
  cells_gex <- cellsList_GEXonly$cellID
  cells_salve <- c(cellsList_SALVEonly$cellID, cellsList_both$cellID)
  cells_both <- cellsList_both$cellID
  
  expr_gex <- as.matrix(GetAssayData(seurat_obj[, cells_gex], slot = "data")[var_features, ])
  expr_salve <- as.matrix(GetAssayData(seurat_obj[, cells_salve], slot = "data")[var_features, ])
  expr_both <- as.matrix(GetAssayData(seurat_obj[, cells_both], slot = "data")[var_features, ])
  
  # Function to calculate JS divergence between two expression profiles
  calculate_js <- function(profile1, profile2) {
    # Convert to probabilities by adding pseudocount and normalizing
    p1 <- (profile1 + 1e-10) / sum(profile1 + 1e-10)
    p2 <- (profile2 + 1e-10) / sum(profile2 + 1e-10)
    
    # Create matrix for philentropy package format
    prob_matrix <- rbind(p1, p2)
    
    # Calculate JS divergence
    js_div <- JSD(prob_matrix, unit = "log2", est.prob = "empirical")
    
    return(js_div)
  }
  
  # Calculate minimum JS divergence for GEX cells vs SALVE cells
  min_js_gex_vs_salve <- numeric(ncol(expr_gex))
  for (i in 1:ncol(expr_gex)) {
    js_values <- apply(expr_salve, 2, function(salve_cell) {
      calculate_js(expr_gex[, i], salve_cell)
    })
    min_js_gex_vs_salve[i] <- min(js_values)
  }
  
  # Calculate minimum JS divergence for GEX cells vs GEX cells (internal)
  min_js_gex_vs_gex <- numeric(ncol(expr_gex))
  for (i in 1:ncol(expr_gex)) {
    js_values <- apply(expr_gex[, -i], 2, function(gex_cell) {
      calculate_js(expr_gex[, i], gex_cell)
    })
    min_js_gex_vs_gex[i] <- min(js_values)
  }
  
  # Calculate minimum JS divergence for GEX cells vs both cells
  if (length(cells_both) > 0) {
    min_js_gex_vs_both <- numeric(ncol(expr_gex))
    for (i in 1:ncol(expr_gex)) {
      js_values <- apply(expr_both, 2, function(both_cell) {
        calculate_js(expr_gex[, i], both_cell)
      })
      min_js_gex_vs_both[i] <- min(js_values)
    }
  } else {
    min_js_gex_vs_both <- numeric(0)
  }
  
  # Calculate minimum JS divergence for SALVE cells vs SALVE cells (internal)
  if (ncol(expr_salve) > 1) {
    min_js_salve_vs_salve <- numeric(ncol(expr_salve))
    for (i in 1:ncol(expr_salve)) {
      js_values <- apply(expr_salve[, -i], 2, function(salve_cell) {
        calculate_js(expr_salve[, i], salve_cell)
      })
      min_js_salve_vs_salve[i] <- min(js_values)
    }
  } else {
    min_js_salve_vs_salve <- numeric(0)
  }
  
  # Calculate minimum JS divergence for both cells vs both cells (internal)
  if (length(cells_both) > 1) {
    min_js_both_vs_both <- numeric(ncol(expr_both))
    for (i in 1:ncol(expr_both)) {
      js_values <- apply(expr_both[, -i], 2, function(both_cell) {
        calculate_js(expr_both[, i], both_cell)
      })
      min_js_both_vs_both[i] <- min(js_values)
    }
  } else {
    min_js_both_vs_both <- numeric(0)
  }
  
  # Random sampling comparison
  n_iterations <- 5
  n_cells_sample <- length(cells_gex)
  available_cells <- setdiff(
    colnames(seurat_obj), 
    c(cells_gex, cellsList_SALVEonly$cellID, cellsList_both$cellID))
  
  set.seed(123)
  min_js_per_iteration <- matrix(0, nrow = length(cells_gex), ncol = n_iterations)
  
  for (iter in 1:n_iterations) {
    random_cells <- sample(available_cells, n_cells_sample)
    expr_random <- as.matrix(
      GetAssayData(seurat_obj[, random_cells], slot = "data")[var_features, ])
    
    for (i in 1:ncol(expr_gex)) {
      js_values <- apply(expr_random, 2, function(random_cell) {
        calculate_js(expr_gex[, i], random_cell)
      })
      min_js_per_iteration[i, iter] <- min(js_values)
    }
  }
  
  min_js_random_single <- min_js_per_iteration[, 1]
  min_js_random_pooled <- apply(min_js_per_iteration, 1, min)
  median_js_random <- apply(min_js_per_iteration, 1, median)
  
  # Statistical tests - comparing all groups to GEX vs both and GEX vs SALVE
  cat("\n\n========== ", sample, " JS Divergence ==========\n")
  
  cat("\n--- Comparisons vs GEX vs both ---\n")
  
  if (length(min_js_gex_vs_both) > 0) {
    # GEX vs Random1 median vs GEX vs both
    w_rand_med_vs_both <- wilcox.test(median_js_random, min_js_gex_vs_both, paired = TRUE)
    cat("GEX vs Random1 (median) vs GEX vs both:\t", w_rand_med_vs_both$p.value, "\n")
    
    # GEX vs Random1 min vs GEX vs both
    w_rand_min_vs_both <- wilcox.test(min_js_random_single, min_js_gex_vs_both, paired = TRUE)
    cat("GEX vs Random1 (min) vs GEX vs both:\t", w_rand_min_vs_both$p.value, "\n")
    
    # GEX vs GEX vs GEX vs both
    w_gex_vs_both <- wilcox.test(min_js_gex_vs_gex, min_js_gex_vs_both, paired = TRUE)
    cat("GEX vs GEX vs GEX vs both:\t\t", w_gex_vs_both$p.value, "\n")
    
    # GEX vs SALVE vs GEX vs both
    w_salve_vs_both <- wilcox.test(min_js_gex_vs_salve, min_js_gex_vs_both, paired = TRUE)
    cat("GEX vs SALVE vs GEX vs both:\t\t", w_salve_vs_both$p.value, "\n")
    
    # SALVE vs SALVE vs GEX vs both
    if (length(min_js_salve_vs_salve) > 0) {
      w_salve_int_vs_both <- wilcox.test(min_js_salve_vs_salve, min_js_gex_vs_both, paired = FALSE)
      cat("SALVE vs SALVE vs GEX vs both:\t\t", w_salve_int_vs_both$p.value, "\n")
    }
    
    # both vs both vs GEX vs both
    if (length(min_js_both_vs_both) > 0) {
      w_both_int_vs_both <- wilcox.test(min_js_both_vs_both, min_js_gex_vs_both, paired = FALSE)
      cat("both vs both vs GEX vs both:\t\t", w_both_int_vs_both$p.value, "\n")
    }
  }
  
  cat("\n--- Comparisons vs GEX vs SALVE ---\n")
  
  # GEX vs Random1 median vs GEX vs SALVE
  w_rand_med_vs_salve <- wilcox.test(median_js_random, min_js_gex_vs_salve, paired = TRUE)
  cat("GEX vs Random1 (median) vs GEX vs SALVE:\t", w_rand_med_vs_salve$p.value, "\n")
  
  # GEX vs Random1 min vs GEX vs SALVE
  w_rand_min_vs_salve <- wilcox.test(min_js_random_single, min_js_gex_vs_salve, paired = TRUE)
  cat("GEX vs Random1 (min) vs GEX vs SALVE:\t\t", w_rand_min_vs_salve$p.value, "\n")
  
  # GEX vs GEX vs GEX vs SALVE
  w_gex_vs_salve <- wilcox.test(min_js_gex_vs_gex, min_js_gex_vs_salve, paired = TRUE)
  cat("GEX vs GEX vs GEX vs SALVE:\t\t\t", w_gex_vs_salve$p.value, "\n")
  
  # GEX vs both vs GEX vs SALVE
  if (length(min_js_gex_vs_both) > 0) {
    w_both_vs_salve <- wilcox.test(min_js_gex_vs_both, min_js_gex_vs_salve, paired = TRUE)
    cat("GEX vs both vs GEX vs SALVE:\t\t\t", w_both_vs_salve$p.value, "\n")
  }
  
  # SALVE vs SALVE vs GEX vs SALVE
  if (length(min_js_salve_vs_salve) > 0) {
    w_salve_int_vs_salve <- wilcox.test(min_js_salve_vs_salve, min_js_gex_vs_salve, paired = FALSE)
    cat("SALVE vs SALVE vs GEX vs SALVE:\t\t\t", w_salve_int_vs_salve$p.value, "\n")
  }
  
  # both vs both vs GEX vs SALVE
  if (length(min_js_both_vs_both) > 0) {
    w_both_int_vs_salve <- wilcox.test(min_js_both_vs_both, min_js_gex_vs_salve, paired = FALSE)
    cat("both vs both vs GEX vs SALVE:\t\t\t", w_both_int_vs_salve$p.value, "\n")
  }
  
  # Create extended comparison plot
  comparison_extended <- data.frame(
    Comparison = factor(
      rep(c("GEX vs Random1 (median)", "GEX vs Random1 (min)", "GEX vs GEX", 
            "GEX vs both", "GEX vs SALVE", "SALVE vs SALVE", "both vs both"),
          c(length(median_js_random), length(min_js_random_single), 
            length(min_js_gex_vs_gex), length(min_js_gex_vs_both),
            length(min_js_gex_vs_salve), length(min_js_salve_vs_salve), 
            length(min_js_both_vs_both))),
      levels = c("GEX vs Random1 (median)", "GEX vs Random1 (min)", "GEX vs GEX",
                 "GEX vs both", "GEX vs SALVE", "SALVE vs SALVE", "both vs both")
    ),
    Min_JS = c(median_js_random, min_js_random_single, min_js_gex_vs_gex,
               min_js_gex_vs_both, min_js_gex_vs_salve, min_js_salve_vs_salve, 
               min_js_both_vs_both)
  )
  
  p_js <- ggplot(comparison_extended, aes(x = Comparison, y = Min_JS, fill = Comparison)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.2, alpha = 0.5, size = 1.5) +
    scale_fill_manual(values = rep("lightgray", 7)) +
    theme_minimal() +
    theme(legend.position = "none",
          axis.text.x = element_text(size = 10, angle = 25, hjust = 1),
          plot.title = element_text(hjust = 0.5)) +
    labs(title = paste0(sample, " JS Divergence Comparison"),
         x = "", y = "Minimum JS Divergence") +
    coord_cartesian(ylim = c(0, NA))
  
  ggsave(paste0(joint.dir, "plots/JSdivergence/", sample, "_JS_comparison_extended.svg"),
         p_js, width = 10, height = 6)
  
  # Create histogram of JS divergence distributions
  p_hist_salve <- ggplot(data.frame(JS = min_js_gex_vs_salve), aes(x = JS)) +
    geom_histogram(bins = 30, fill = "gray", color = "black") +
    theme_minimal() +
    labs(title = paste0(sample, " Distribution of Minimum JS Divergence: GEX vs SALVE"),
         x = "Minimum JS Divergence", y = "Count") +
    theme(plot.title = element_text(hjust = 0.5))
  ggsave(paste0(joint.dir, "plots/JSdivergence/", sample, "_JS_GEX_vs_SALVE_distribution.svg"), 
         p_hist_salve, width = 8, height = 6)
  
  if (length(min_js_salve_vs_salve) > 0) {
    p_hist_salve_int <- ggplot(data.frame(JS = min_js_salve_vs_salve), aes(x = JS)) +
      geom_histogram(bins = 30, fill = "gray", color = "black") +
      theme_minimal() +
      labs(title = paste0(sample, " Distribution of Minimum JS Divergence: SALVE vs SALVE"),
           x = "Minimum JS Divergence", y = "Count") +
      theme(plot.title = element_text(hjust = 0.5))
    ggsave(paste0(joint.dir, "plots/JSdivergence/", sample, "_JS_SALVE_vs_SALVE_distribution.svg"), 
           p_hist_salve_int, width = 8, height = 6)
  }
}
```

### Direct splicing
Quantifying splicing from actual reads instead of inference
```{r}
# Define sample names and files
sample_configs <- list(
  list(
    name = "Invitro_SALVE",
    files = c("Invitro_D1_PL", "Invitro_env_PL", "Invitro_pol_PL", "Invitro_SSenv_PL", "Invitro_tat"),
    input_dir = "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/SALVE/bamsort/splice/",
    umi_type = "SALVE",
    umi_sample_name = "Invitro"
  ),
  list(
    name = "Pacute_SALVE",
    files = c("Pacute_D1_PL", "Pacute_env_PL", "Pacute_pol_PL", "Pacute_SSenv_PL", "Pacute_tat"),
    input_dir = "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/SALVE/bamsort/splice/",
    umi_type = "SALVE",
    umi_sample_name = "Pacute"
  ),
  list(
    name = "W2_SALVE",
    files = c("W2_D1_PL", "W2_env_PL", "W2_pol_PL", "W2_SSenv_PL", "W2_tat"),
    input_dir = "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/SALVE/bamsort/splice/",
    umi_type = "SALVE",
    umi_sample_name = "W2"
  ),
  list(
    name = "Invitro_GEX",
    files = c("invitro"),
    input_dir = "/projects/b1042/GoyalLab/egrody/extractedData/EGS002/singleCell/bamsort/splice/",
    umi_type = "GEX",
    umi_sample_name = "Invitro"
  ),
  list(
    name = "Pacute_GEX",
    files = c("P_acute_combined"),
    input_dir = "/projects/b1042/GoyalLab/egrody/extractedData/EGS009/singleCell/bamsort/splice/",
    umi_type = "GEX",
    umi_sample_name = "Pacute"
  ),
  list(
    name = "W2_GEX",
    files = c("W2"),
    input_dir = "/projects/b1042/GoyalLab/egrody/extractedData/EGS002/singleCell/bamsort/splice/",
    umi_type = "GEX",
    umi_sample_name = "W2"
  )#,
  # list(
  #   name = "InvitrocDNA",
  #   files = c("cDNA"),
  #   input_dir = "/projects/b1042/GoyalLab/egrody/extractedData/EGS007/longRead/bamsort/splice/",
  #   umi_type = "GEX",
  #   umi_sample_name = "cDNA"
  # )
)

# UMI directories
SALVE.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/SALVE/bamsort/reads/v6/"
GEX.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS016/singleCell/bamsort/reads/v6/"

# Main processing loop
for (config in sample_configs) {
  sample_name <- config$name
  input_dir <- config$input_dir
  
  cat("\nProcessing sample:", sample_name, "\n")
  
  # Step 1: Combine data from all files in this sample
  combined_data <- data.frame()
  
  for (file in config$files) {
    cat("  Reading:", file, "\n")
    data <- read.csv(paste0(input_dir, file, "_splicesites.csv"))
    cat("    Raw rows in CSV:", nrow(data), "\n")
    
    # Load appropriate UMI file based on type
    if (config$umi_type == "SALVE") {
      umi_file <- paste0(SALVE.dir, "minimum/valid_UMI/", config$umi_sample_name, "_SALVE_valid_cellID_UMI.csv")
      umi_data <- if(file.exists(umi_file)) read.csv(umi_file) else NULL
      cat("    SALVE UMI file:", basename(umi_file), "\n")
    } else if (config$umi_type == "GEX") {
      umi_file <- paste0(GEX.dir, "minimum/valid_UMI/", config$umi_sample_name, "_GEX_valid_cellID_UMI.csv")
      umi_data <- if(file.exists(umi_file)) read.csv(umi_file) else NULL
      cat("    GEX UMI file:", basename(umi_file), "\n")
    } else {
      stop("Invalid umi_type: must be 'SALVE' or 'GEX'")
    }
    
    if (is.null(umi_data)) {
      cat("    WARNING: UMI file not found, skipping filtering\n")
      file_data <- data %>%
        count(Donor, Acceptor) %>%
        mutate(file = file)
    } else {
      cat("    UMI rows loaded:", nrow(umi_data), "\n")
      
      # Diagnostic information
      cat("    Unique CBs in data:", length(unique(data$CB)), "\n")
      cat("    Unique UBs in data:", length(unique(data$UB)), "\n")
      cat("    CB matches:", sum(data$CB %in% umi_data$cellID), "of", nrow(data), "\n")
      cat("    UB matches:", sum(data$UB %in% umi_data$UMI), "of", nrow(data), "\n")
      cat("    Both CB AND UB match:", sum(data$CB %in% umi_data$cellID & data$UB %in% umi_data$UMI), "\n")
      
      file_data <- data %>%
        filter(CB %in% umi_data$cellID & UB %in% umi_data$UMI) %>%
        count(Donor, Acceptor) %>%
        mutate(file = file)
      
      cat("    Rows after filtering:", nrow(file_data), "\n")
      cat("    Reads after filtering:", sum(file_data$n), "\n")
    }
    
    combined_data <- rbind(combined_data, file_data)
  }
  
  # Step 2: Sum counts across files for same donor-acceptor pairs
  plotdata <- combined_data %>%
    group_by(Donor, Acceptor) %>%
    summarise(
      raw_count = sum(n),
      files = paste(unique(file), collapse = ", "),
      .groups = 'drop'
    ) %>%
    arrange(Donor) %>%
    mutate(
      y_position = row_number(),
      log_count = log1p(raw_count)
    )
  
  cat("  Combined junctions:", nrow(plotdata), "\n")
  cat("  Total reads:", sum(plotdata$raw_count), "\n")
  
  # Step 3: Create log-normalized plot
  MAX_LOG <- 5.5
  p_log <- ggplot(plotdata) +
    geom_segment(aes(x = Donor, xend = Acceptor, 
                     y = y_position, yend = y_position,
                     color = log_count, size = log_count), 
                 alpha = 0.7) +
    scale_color_viridis_c(name = "log(Count+1)",
                          trans = "identity",
                          limits = c(0, MAX_LOG)) +
    scale_size_continuous(name = "log(Count+1)", 
                         range = c(0.5, 3),
                         limits = c(0, MAX_LOG)) +
    xlim(0, 10279) +
    labs(title = paste0("Splice Junctions: ", sample_name, " (Log Scale)"),
         subtitle = paste0("Combined from: ", paste(config$files, collapse = ", "), 
                          " | UMI: ", config$umi_type),
         x = "Genomic Position", 
         y = "Splice Junction") +
    theme_minimal() +
    theme(axis.text.y = element_blank(),
          axis.ticks.y = element_blank(),
          legend.position = "right")
  
  print(p_log)
  
  # Create output directory if it doesn't exist
  output_dir <- paste0(input_dir, "plots/")
  if (!dir.exists(output_dir)) {
    dir.create(output_dir, recursive = TRUE)
  }
  
  ggsave(paste0(output_dir, sample_name, "_combined_log.pdf"), 
         plot = p_log, width = 15, height = 8)
  
  # Step 4: Create linear (raw count) plot
  p_linear <- ggplot(plotdata) +
    geom_segment(aes(x = Donor, xend = Acceptor, 
                     y = y_position, yend = y_position,
                     color = raw_count, size = raw_count), 
                 alpha = 0.7) +
    scale_color_viridis_c(name = "Count",
                          trans = "identity") +
    scale_size_continuous(name = "Count", 
                         range = c(0.5, 3)) +
    xlim(0, 10279) +
    labs(title = paste0("Splice Junctions: ", sample_name, " (Linear Scale)"),
         subtitle = paste0("Combined from: ", paste(config$files, collapse = ", "),
                          " | UMI: ", config$umi_type),
         x = "Genomic Position", 
         y = "Splice Junction") +
    theme_minimal() +
    theme(axis.text.y = element_blank(),
          axis.ticks.y = element_blank(),
          legend.position = "right")
  
  ggsave(paste0(output_dir, sample_name, "_combined_linear.pdf"), 
         plot = p_linear, width = 15, height = 8)
  
  # Step 5: Create summary table for this sample
  if (nrow(plotdata) > 0) {
    summary_table <- plotdata %>%
      slice_max(raw_count, n = 10) %>%
      select(Donor, Acceptor, raw_count, log_count, files) %>%
      mutate(junction = paste0(Donor, "-", Acceptor))
    
    cat("  Top 10 junctions:\n")
    print(summary_table)
    
    # Save summary table
    write.csv(summary_table, 
              paste0(output_dir, sample_name, "_top_junctions.csv"),
              row.names = FALSE)
  } else {
    cat("  No junctions found for this sample\n")
  }
}

```

### Host factors

```{r restrict and depend}
factors_restrict <- c("APOBEC3G", "APOBEC3H", "TRIM5", "SAMHD1", 
                      "BST2", "SERINC5", "SERINC3", "MX2", "IFITM2")
factors_depend <- c("CD4", "CCR5", "PPIA", "PSIP1", "TNPO3", "CPSF6", "NUP153", "RANBP2", 
                    "TSG101", "PDCD6IP", "SP1", "NFKB1", "RELA", "CDK9", "CCNT1", "TNPO1",
                    "XPO1", "DDX3X", "RAN", "RANBP1", "RANGAP1")
cytotox <- c("TBX21", "EOMES", "RUNX3", "IFNG-AS1", "GZMK", "GZMA", "CX3CR1", "PRF1", 
                    "GZMB", "GZMH", 'GNLY', "ZNF683")

# Load all RDS
input.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS016/singleCell/Seurat/RDS/"
objects <- list()
for (sample_name in names(joint_data)) {
  seurat_obj <- readRDS(paste0(input.dir, sample_name, ".rds"))
  objects[[sample_name]] <- seurat_obj
}



# Run the enhanced sparse analysis
results_sparse <- host_virus_correlation_sparse(
    joint_salve_data = joint_data,
    joint_rds = objects,
    genes_list = factors_restrict,
    viral_col = "SALVE",
    min_coexpress = 10,
    n_permutations = 1000,
    output.dir = "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/joint/hostVirus/new_restrictionFactors/",
    outputs = TRUE
)

host_virus_correlation(joint_data, objects, factors_restrict, viral_col = "SALVE", 
            output.dir = "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/joint/hostVirus/new_restrictionFactors/")
host_virus_correlation(joint_data, objects, factors_restrict, viral_col = "USprop", 
            output.dir = "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/joint/hostVirus/restrictionFactors/")
host_virus_correlation(joint_data, objects, factors_restrict, viral_col = "SSprop", 
            output.dir = "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/joint/hostVirus/restrictionFactors/")
host_virus_correlation(joint_data, objects, factors_restrict, viral_col = "MSprop", 
            output.dir = "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/joint/hostVirus/restrictionFactors/")
host_virus_correlation(joint_data, objects, factors_restrict, viral_col = "GEX", 
            output.dir = "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/joint/hostVirus/restrictionFactors/")

host_virus_correlation(joint_data, objects, factors_depend, viral_col = "SALVE", 
            output.dir = "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/joint/hostVirus/dependencyFactors/")
host_virus_correlation(joint_data, objects, factors_depend, viral_col = "USprop", 
            output.dir = "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/joint/hostVirus/dependencyFactors/")
host_virus_correlation(joint_data, objects, factors_depend, viral_col = "SSprop", 
            output.dir = "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/joint/hostVirus/dependencyFactors/")
host_virus_correlation(joint_data, objects, factors_depend, viral_col = "MSprop", 
            output.dir = "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/joint/hostVirus/dependencyFactors/")
host_virus_correlation(joint_data, objects, factors_depend, viral_col = "GEX", 
            output.dir = "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/joint/hostVirus/dependencyFactors/")

# are the trends the same between SALVE and GEX? And the isoforms?
input.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/joint/hostVirus/dependencyFactors/"
for (sample_name in names(joint_data)) {
  gex_file <- paste0(input.dir, sample_name, "_GEX_correlations.csv")
  salve_file <- paste0(input.dir, sample_name, "_SALVE_correlations.csv")
  
  # Check if both files exist
  if (!file.exists(gex_file) || !file.exists(salve_file)) {
    cat("Skipping", sample_name, "- correlation files not found\n")
    next
  }
  
  file_GEX <- read.csv(gex_file) %>% 
    rename(GEXcor = correlation, GEXpadj = p_adj)
  file_SALVE <- read.csv(salve_file) %>% 
    rename(SALVEcor = correlation, SALVEpadj = p_adj)
  joint_corr <- full_join(file_GEX, file_SALVE, by = "gene") %>%
    select(gene, GEXcor, GEXpadj, SALVEcor, SALVEpadj)
  p1 <- ggplot(joint_corr, aes(x = GEXcor, y = SALVEcor)) +
        geom_point(alpha = 0.3, size = 2) +
        geom_text_repel(aes(label = gene)) +
        labs(title = paste0(sample_name, " GEX vs SALVE Correlations: Dependency Factors"),
             x = "GEX",
             y = "SALVE") +
        theme_minimal() +
        theme(plot.title = element_text(face = "bold", size = 14),
              axis.text = element_text(size = 10))
  ggsave(p1, file = paste0(input.dir, sample_name, "_compare_correlations.pdf"),
         device = "pdf")
}

```

```{r memory cytotox}
# Define cell type markers
memory_cd4_markers <- list(
  CD4 = "CD4",
  memory = c("IL7R", "CCR7", "SELL"),  # Central memory
  exclude_naive = c("CCR7", "SELL"),    # High in naive
  exclude_effector = c("GZMB", "PRF1")  # High in effector
)


subset_memory_cd4 <- function(seurat_obj, min_cd4_expr = 0.5) {
  cd4_expr <- GetAssayData(seurat_obj, slot = "data")["CD4", ]
  il7r_expr <- GetAssayData(seurat_obj, slot = "data")["IL7R", ]
  
  memory_cd4_cells <- names(cd4_expr)[cd4_expr > min_cd4_expr & il7r_expr > 0]
  seurat_subset <- subset(seurat_obj, cells = memory_cd4_cells)
  
  return(seurat_subset)
}

joint_data_memory <- list()
joint_rds_memory <- list()
# Subset both RDS and SALVE data
for (sample_name in names(objects)) {
  seurat_subset <- subset_memory_cd4(objects[[sample_name]])
  cells_keep <- colnames(seurat_subset)
  
  joint_rds_memory[[sample_name]] <- seurat_subset
  joint_data_memory[[sample_name]] <- joint_data[[sample_name]] %>%
    filter(cellID %in% cells_keep)
  
  cat("Sample:", sample_name, 
      "- Original cells:", ncol(seurat_obj),
      "- Memory CD4 cells:", ncol(joint_rds_memory[[sample_name]]), "\n")
}

# Run correlation on subset
host_virus_correlation(
  joint_salve_data = joint_data_memory,
  joint_rds = joint_rds_memory, 
  genes_list = cytotox,
  output.dir = "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/joint/hostVirus/memoryCytotoxic/",
  viral_col = "SALVE"
)
```

```{r metabolic}

calculate_metabolic_scores_macaque <- function(seurat_obj) {
  
  # Check if percent_mito already exists
  if (!"percent_mito" %in% colnames(seurat_obj@meta.data)) {
    stop("percent_mito not found. Run Add_Mito_Ribo() first.")
  }
  
  cat("Using existing percent_mito column\n")
  
  # Define metabolic genes (human names)
  glycolysis_human <- c("HK2", "PFKP", "ALDOA", "GAPDH", "PGK1", 
                        "PGAM1", "ENO1", "PKM")
  oxphos_human <- c("NDUFA1", "NDUFB1", "SDHB", "UQCRB", "COX5A", 
                    "ATP5A1", "ATP5B", "ATP5F1A", "ATP5PB")
  
  # Convert to match your data
  glycolysis_genes <- convert_gene_list(seurat_obj, glycolysis_human)
  oxphos_genes <- convert_gene_list(seurat_obj, oxphos_human)
  
  # Calculate scores
  if (length(glycolysis_genes) > 2) {
    seurat_obj <- AddModuleScore(seurat_obj,
                                 features = list(glycolysis_genes),
                                 name = "glycolysis_score")
    cat("Calculated glycolysis score with", length(glycolysis_genes), "genes\n")
  }
  
  if (length(oxphos_genes) > 2) {
    seurat_obj <- AddModuleScore(seurat_obj,
                                 features = list(oxphos_genes),
                                 name = "oxphos_score")
    cat("Calculated OXPHOS score with", length(oxphos_genes), "genes\n")
  }
  
  # Ribosomal genes (check if percent_ribo also exists from Add_Mito_Ribo)
  if (!"percent_ribo" %in% colnames(seurat_obj@meta.data)) {
    ribo_genes <- unique(c(
      grep("^RPL|^RPS", rownames(seurat_obj), value = TRUE),
      grep("^Rpl|^Rps", rownames(seurat_obj), value = TRUE)
    ))
    
    if (length(ribo_genes) > 10) {
      seurat_obj <- AddModuleScore(seurat_obj,
                                   features = list(ribo_genes),
                                   name = "translation_score")
      cat("Calculated translation score with", length(ribo_genes), "genes\n")
    }
  } else {
    cat("Using existing percent_ribo column\n")
  }
  
  return(seurat_obj)
}
# Function to test all metabolic correlations with viral load
test_metabolic_correlations <- function(joint_salve_data, joint_rds, 
                                       viral_col = "SALVE",
                                       output.dir = NULL) {
  
  all_results <- list()
  
  for (sample_name in names(joint_salve_data)) {
    cat("\n=== SAMPLE:", sample_name, "===\n")
    
    current_df <- joint_salve_data[[sample_name]]
    seurat_obj <- joint_rds[[sample_name]]
    
    # Extract viral measure
    if (!viral_col %in% colnames(current_df)) {
      cat("Column", viral_col, "not found. Skipping.\n")
      next
    }
    viral_measure <- current_df[[viral_col]]
    
    if (sd(viral_measure) == 0) {
      cat(viral_col, "has zero variance. Skipping.\n")
      next
    }
    
    # Calculate metabolic scores if not already done
    if (!"glycolysis_score1" %in% colnames(seurat_obj@meta.data)) {
      seurat_obj <- calculate_metabolic_scores_macaque(seurat_obj)
    }
    
    # Add total UMI/genes if not present
    if (!"total_UMI" %in% colnames(seurat_obj@meta.data)) {
      seurat_obj$total_UMI <- colSums(GetAssayData(seurat_obj, slot = "counts"))
      seurat_obj$total_genes <- colSums(GetAssayData(seurat_obj, slot = "counts") > 0)
    }
    
    # Extract all metabolic metrics
    metabolic_metrics <- c("percent_mito", "total_UMI", "total_genes",
                          "glycolysis_score1", "oxphos_score1")
    
    # Add translation score if exists
    if ("translation_score1" %in% colnames(seurat_obj@meta.data)) {
      metabolic_metrics <- c(metabolic_metrics, "translation_score1")
    }
    
    # Add ribosomal if exists
    if ("percent_ribo" %in% colnames(seurat_obj@meta.data)) {
      metabolic_metrics <- c(metabolic_metrics, "percent_ribo")
    }
    
    # Test correlations
    cat("\n=== METABOLIC CORRELATIONS WITH", viral_col, "===\n")
    cat(sprintf("%-25s %10s %12s %10s %8s\n", 
                "Metric", "Corr", "P-value", "N_cells", "Sig"))
    cat(strrep("-", 70), "\n")
    
    results <- data.frame()
    
    for (metric in metabolic_metrics) {
      if (metric %in% colnames(seurat_obj@meta.data)) {
        metric_values <- seurat_obj@meta.data[[metric]]
        
        # Filter out NA values
        valid_cells <- !is.na(metric_values) & !is.na(viral_measure)
        
        if (sum(valid_cells) > 10) {
          test <- cor.test(metric_values[valid_cells], 
                          viral_measure[valid_cells],
                          method = "spearman", 
                          exact = FALSE)
          
          sig <- if (test$p.value < 0.001) "***" else 
                 if (test$p.value < 0.01) "**" else 
                 if (test$p.value < 0.05) "*" else "ns"
          
          cat(sprintf("%-25s %10.4f %12.2e %10d %8s\n",
                     metric, test$estimate, test$p.value, sum(valid_cells), sig))
          
          results <- rbind(results, data.frame(
            sample = sample_name,
            metric = metric,
            correlation = as.numeric(test$estimate),
            p_value = test$p.value,
            n_cells = sum(valid_cells)
          ))
        }
      }
    }
    
    results$p_adj <- p.adjust(results$p_value, method = "BH")
    all_results[[sample_name]] <- results
    
    cat(strrep("-", 70), "\n\n")
  }
  
  # Combine all results
  combined_results <- do.call(rbind, all_results)
  
  # Save if output directory provided
  if (!is.null(output.dir)) {
    if (!dir.exists(output.dir)) {
      dir.create(output.dir, recursive = TRUE)
    }
    write.csv(combined_results, 
              file.path(output.dir, paste0(viral_col, "_metabolic_correlations.csv")),
              row.names = FALSE)
    cat("Saved results to:", output.dir, "\n")
  }
  
  return(combined_results)
}

# Run analysis
metabolic_results <- test_metabolic_correlations(
  joint_salve_data = joint_data,
  joint_rds = objects,
  viral_col = "SALVE",
  output.dir = "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/joint/hostVirus/metabolic/"
)

# View summary
metabolic_results %>%
  group_by(metric) %>%
  summarise(
    mean_corr = mean(correlation),
    median_corr = median(correlation),
    n_samples = n(),
    n_sig = sum(p_adj < 0.05)
  ) %>%
  arrange(desc(abs(mean_corr)))


```


### Correlation analysis (not updated)
Will do it custom for now and update correlation plot function later
```{r}
cat(
  "Correlation between 10X mac239 and SALVE total_lessLTR:\n",
    "  Pearson: ", round(cor(Pacute_paint_umap$mac239, Pacute_paint_umap$total_lessLTR, method="pearson"), 3),
    "\n  Spearman: ", round(cor(Pacute_paint_umap$mac239, Pacute_paint_umap$total_lessLTR, method="spearman"), 3),
    
    "\n\nCorrelation between 10X mac239 and SALVE absolute:\n",
    "  Pearson: ", round(cor(Pacute_paint_umap$mac239, Pacute_paint_umap$absolute, method="pearson"), 3),
    "\n  Spearman: ", round(cor(Pacute_paint_umap$mac239, Pacute_paint_umap$absolute, method="spearman"), 3)
)

# plotting
plot <- ggplot(data = Pacute_paint_umap, aes(x = total_lessLTR, y = mac239)) +
      geom_point() +
      labs(
        x = "SALVE total counts: D1 + tat + nef (filter > 2)",
        y = "SingleCell mac239 counts",
        title = "Pacute joint: total count") +
      theme_minimal() + 
    coord_fixed(ratio = 1)
plot <- ggplot(data = Pacute_paint_umap, aes(x = absolute, y = mac239)) +
      geom_point() +
      labs(
        x = "SALVE absolute (5' LTR) counts (filter > 0)",
        y = "SingleCell mac239 counts",
        title = "Pacute joint: absolute count") +
      theme_minimal() +
    coord_fixed(ratio = 1)
plot + theme(aspect.ratio = 1)
```


### Subsample reads
```{r Pacute}
subsets <- c("P_acute_150M", "P_acute_200M", "P_acute_300M", "P_acute_400M")

# Process
input.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS009/singleCell/bamsort/alignment/subsample/"
output.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS009/singleCell/bamsort/reads/subsample/"

raw_cellIDs_df <- read.csv("/projects/b1042/GoyalLab/egrody/extractedData/EGS016/singleCell/raw_cellIDs/all_raw_cellIDs.csv")
raw_cellIDs <- split(raw_cellIDs_df$cellID, raw_cellIDs_df$sample)
raw_cellID <- list(P_acute_150M = raw_cellIDs$Pacute,
                P_acute_200M = raw_cellIDs$Pacute,
                P_acute_300M = raw_cellIDs$Pacute,
                P_acute_400M = raw_cellIDs$Pacute)

process_bamsortv6(subsets, input.dir, output.dir, raw_cellID)

# Minimums
input_dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS009/singleCell/bamsort/reads/subsample/"
output_dir <- paste0(input_dir, "minimum/")

process_all_set_minimumsv6("GEX", subsets, input_dir, TRUE, output_dir, min_reads_cell = 2)
```

```{r joint}
subsets <- c("P_acute_150M", "P_acute_200M", "P_acute_300M", "P_acute_400M")

# Input directories
salve_input_dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/SALVE/bamsort/reads/v6/minimum/"
gex_input_dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS009/singleCell/bamsort/reads/subsample/minimum/"
umap_file <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS016/singleCell/Seurat/UMAPcoords/PacuteUMAP_coords.csv"
cd4_file <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS016/singleCell/Seurat/expressionDF/Pacute_CD4T.csv"

# Output directories
output_dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS009/joint/dataframes/"
plot_dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS009/joint/plots/"

# Create output directories if they don't exist
dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(plot_dir, showWarnings = FALSE, recursive = TRUE)

# Loading
salve <- read.csv(paste0(salve_input_dir, "Pacute_SALVE_filtered.csv"))
SALVE <- salve %>%
  rename(SALVE = total) %>%
  mutate(SALVE = log1p(SALVE))
cd4_data <- read.csv(cd4_file, row.names = "X")
cd4_data <- cd4_data %>%
    mutate(CD3 = ifelse(CD3D > 0 | CD3E > 0 | CD3G > 0, 1, 0)) %>%
    mutate(CD4 = ifelse(CD4 > 0 , 1, 0)) %>%
    mutate(CD8 = ifelse(CD8A > 0 , 1, 0)) %>%
    select(cellID, CD4, CD3, CD8)
GEX_UMAP <- read.csv(umap_file, row.names = "X")
GEX_UMAP <- GEX_UMAP %>% select(-sample, -V2)

# Print header for comparison table
cat("\n===========================================\n")
cat("COMPARISON STATISTICS\n")
cat("===========================================\n")
cat("Sample\tTotal_cells\tboth\t10X_only\tSALVE_only\tJaccard\tPearson_r\tSpearman_rho\n")

# Loop through each GEX subsample
for (sample in subsets) {
  
  # Load GEX data for this subsample
  gex_file <- paste0(gex_input_dir, sample, "_GEX_filtered.csv")
  
  if (!file.exists(gex_file)) {
    cat("WARNING: File not found:", gex_file, "\n")
    cat("Skipping sample", sample, "\n")
    next
  }
  
  GEX <- read.csv(gex_file, check.names = FALSE)
  if (colnames(GEX)[1] %in% c("X", "", "Unnamed: 0")) {
    rownames(GEX) <- GEX[,1]
    GEX <- GEX[,-1]
  }

  # Process GEX data
  gex_processed <- GEX %>%
     #mutate(cellID = as.character(cellID)) %>%
    rename(GEX = total) %>% 
    mutate(GEX = log1p(GEX)) %>%
      select(cellID, GEX)
      
  # Merge with UMAP coordinates
  GEX <- left_join(GEX_UMAP, gex_processed, by = "cellID")
  
  # Join GEX and SALVE data
  Pacute_joint <- left_join(GEX, SALVE, by = "cellID")
  
  # Replace NA with 0 (cells not detected in one method)
  Pacute_joint[is.na(Pacute_joint)] <- 0
  
  
  Pacute_joint <- left_join(Pacute_joint, cd4_data, by = "cellID")
  Pacute_joint <- Pacute_joint %>%
    mutate(across(any_of(c("GEX", "GEX.US")), 
              ~ ifelse(CD4 > 0 & CD3 > 0, .x, 0))) %>% 
    mutate(across(any_of(c("SALVE", "US", "MS", "SS")), 
              ~ ifelse(CD4 > 0 & CD3 > 0, .x, 0))) %>%
    filter(CD4 == 0 | CD8 == 0) %>% # exclude cells CD4+ and CD8+
    select(-CD8, -CD3, -CD4)
  
  # Calculate comparison statistics
  both <- Pacute_joint %>% filter(GEX != 0 & SALVE != 0)
  onlySingleCell <- Pacute_joint %>% 
    filter(GEX != 0) %>%
    filter(!(cellID %in% both$cellID))
  onlySALVE <- Pacute_joint %>% 
    filter(SALVE != 0) %>%
    filter(!(cellID %in% both$cellID))
  
  # Calculate Jaccard index: |intersection| / |union|
  gex_detected <- nrow(both) + nrow(onlySingleCell)
  salve_detected <- nrow(both) + nrow(onlySALVE)
  intersection <- nrow(both)
  union <- gex_detected + salve_detected - intersection  # A + B - (A ∩ B)
  
  if (union > 0) {
    jaccard_index <- intersection / union
  } else {
    jaccard_index <- NA
  }
  
  # Calculate correlations for cells detected in both methods
  if (nrow(both) > 2) {
    pearson_r <- cor(Pacute_joint$GEX, Pacute_joint$SALVE, method = "pearson")
    spearman_rho <- cor(Pacute_joint$GEX, Pacute_joint$SALVE, method = "spearman")
  } else {
    pearson_r <- NA
    spearman_rho <- NA
  }
  
  # Print statistics
    cat(sprintf("%s\t%d\t%d\t%d\t%d\t%.3f\t%.3f\t%.3f\n",
              sample,
              nrow(Pacute_joint),
              nrow(both),
              nrow(onlySingleCell),
              nrow(onlySALVE),
              jaccard_index,
              pearson_r,
              spearman_rho))
  
  # Generate plots
  plotUMAP(Pacute_joint, colorby = GEX,
           paste0("Pacute joint SIV: GEX (", sample, ")"),
           plot_dir,
           paste0("Pacute_subsample_GEX_", sample, ".pdf"),
           comparison = FALSE, color = "gray25")
  
  # Save dataframes for this subsample
  write.csv(both, paste0(output_dir, "Pacute_both_", sample, ".csv"), row.names = FALSE)
  write.csv(onlySingleCell, paste0(output_dir, "Pacute_onlySingleCell_", sample, ".csv"), row.names = FALSE)
  write.csv(onlySALVE, paste0(output_dir, "Pacute_onlySALVE_", sample, ".csv"), row.names = FALSE)
  write.csv(Pacute_joint, paste0(output_dir, "Pacute_all_", sample, ".csv"), row.names = FALSE)
}
plotUMAP(Pacute_joint, colorby = SALVE,
           paste0("Pacute joint SIV: SALVE (", sample, ")"),
           plot_dir,
           paste0("Pacute_subsample_SALVE.pdf"),
           comparison = FALSE, color = "gray25")
```

## Subset analysis

To load data, run first chunk of Joint bamsort (SALVE vs GEX) or Joint combined

### DEG analysis

```{r, warning=FALSE}
input.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS016/singleCell/Seurat/RDS/"
output.base <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/joint/subset/SALVE_new/"

samples <- c("Invitro",
             "D195",
             "Pacute",
             "W2")
for (i in 1:length(samples)) {
  sample_name <- samples[i]
  input_file <- paste0(input.dir, sample_name, ".rds")
  
  tryCatch({
    seurat_obj <- readRDS(input_file)
    genes_to_exclude <- c("CD4", "CD8A", "CD8B")  # remove from DEG analysis
    genes_to_keep <- rownames(seurat_obj)[!rownames(seurat_obj) %in% genes_to_exclude]
    seurat_obj <- subset(seurat_obj, features = genes_to_keep)

    
    salve <- joint_data[[sample_name]]
    list_of_cells1 <- salve %>% filter(SALVE != 0) %>% #update as desired
      select(cellID) %>% unlist()
    # list_of_cells2 <- salve %>% filter(total != 0) %>%
    #   select(cellID) %>% filter(!(cellID %in% list_of_cells1)) %>% unlist()
    results <- analyze_cell_subset(seurat_obj, list_of_cells1#, 
                                   #list_of_cells2, group_names = c("MS", "noMS")
                                   )
    
    output.dir <- paste0(output.base, sample_name, "/")
    if (!dir.exists(output.dir)) {
      dir.create(output.dir, recursive = TRUE)
    }
    #save_subset_analysis(results, output_dir = output.dir)
    plot_subset_analysis(seurat_obj, results, list_of_cells1, output_dir = paste0(output.dir, "plots/"))
    
    # results_uc <- analyze_cell_subset_unclustered(seurat_obj, list_of_cells1#, 
    #                                               #list_of_cells2, group_names = c("MS", "noMS")
    #                                               )
    # save_subset_analysis(results_uc, output_dir = output.dir)
    
    cat("Processed sample:", sample_name, "\n")
  }, error = function(e) {
    cat("Error processing sample", sample_name, ":", conditionMessage(e), "\n\n")
  })
}

```

Rewrote some functions
```{r}
library(gprofiler2)

#' Perform DEG analysis between two cell populations
#'
#' @param seurat_obj Seurat object containing single-cell data
#' @param cells_group1 Character vector of cell IDs for group 1 (numerator)
#' @param cells_group2 Character vector of cell IDs for group 2 (denominator)
#' @param assay Character, assay to use (default: "RNA")
#' @param test_use Character, statistical test method (default: "wilcox")
#' @param fc_threshold Numeric, log2FC threshold for filtering (default: 0.5)
#' @param pval_threshold Numeric, adjusted p-value threshold (default: 0.01)
#' @param output_prefix Character, prefix for output files (default: "DEG_analysis")
#' 
#' @return List containing DEG results, filtered DEGs, pathway results, and plot
#' 
#' @details
#' Statistical Test: Wilcoxon rank-sum test is recommended for single-cell data
#' due to its non-parametric nature and robustness to the zero-inflated, 
#' non-normal distributions typical of scRNA-seq data. Alternative tests include
#' "bimod" (likelihood-ratio test), "t" (Student's t-test), or "MAST" 
#' (Model-based Analysis of Single-cell Transcriptomics).
#' 
#' Multiple Testing Correction: Bonferroni correction is applied by default in
#' Seurat's FindMarkers function to control family-wise error rate.
#' 
#' Pathway Analysis: g:Profiler uses a hypergeometric test with g:SCS multiple
#' testing correction to identify significantly enriched pathways from GO, KEGG,
#' Reactome, and other databases.
perform_deg_pathway_analysis <- function(seurat_obj,
                                          cells_group1,
                                          cells_group2,
                                          assay = "RNA",
                                          test_use = "wilcox",
                                          fc_threshold = 0.5,
                                          pval_threshold = 0.01,
                                          output_prefix = "DEG_analysis") {
  
  # Validate inputs
  if (!inherits(seurat_obj, "Seurat")) {
    stop("seurat_obj must be a Seurat object")
  }
  
  all_cells <- colnames(seurat_obj)
  missing_cells1 <- setdiff(cells_group1, all_cells)
  missing_cells2 <- setdiff(cells_group2, all_cells)
  
  if (length(missing_cells1) > 0) {
    warning(paste(length(missing_cells1), "cells from group1 not found in Seurat object"))
    cells_group1 <- intersect(cells_group1, all_cells)
  }
  
  if (length(missing_cells2) > 0) {
    warning(paste(length(missing_cells2), "cells from group2 not found in Seurat object"))
    cells_group2 <- intersect(cells_group2, all_cells)
  }
  
  if (length(cells_group1) == 0 || length(cells_group2) == 0) {
    stop("One or both groups have no valid cells after filtering")
  }
  
  cat(sprintf("Group 1: %d cells\n", length(cells_group1)))
  cat(sprintf("Group 2: %d cells\n", length(cells_group2)))
  
  # Create temporary identity for DEG analysis
  seurat_obj$temp_ident <- "other"
  seurat_obj$temp_ident[colnames(seurat_obj) %in% cells_group1] <- "group1"
  seurat_obj$temp_ident[colnames(seurat_obj) %in% cells_group2] <- "group2"
  
  # Subset Seurat object to only include cells from both groups
  seurat_subset <- subset(seurat_obj, cells = c(cells_group1, cells_group2))
  Idents(seurat_subset) <- "temp_ident"
  
  cat("Performing differential expression analysis...\n")
  
  # Perform DEG analysis
  # group1 vs group2: positive log2FC means higher in group1
  deg_results <- FindMarkers(
    object = seurat_subset,
    ident.1 = "group1",
    ident.2 = "group2",
    assay = assay,
    test.use = test_use,
    logfc.threshold = 0,  # Get all genes, filter later
    min.pct = 0.1,  # Gene must be expressed in at least 10% of cells in one group
    verbose = TRUE
  )
  
  # Add gene names as column
  deg_results$gene <- rownames(deg_results)
  
  # Reorder columns
  deg_results <- deg_results[, c("gene", "avg_log2FC", "pct.1", "pct.2", 
                                  "p_val", "p_val_adj")]
  
  # Sort by adjusted p-value
  deg_results <- deg_results[order(deg_results$p_val_adj), ]
  
  # Save full results
  write.csv(deg_results, 
            file = paste0(output_prefix, "_all_genes.csv"),
            row.names = FALSE)
  cat(sprintf("Saved full results: %s_all_genes.csv\n", output_prefix))
  
  # Filter for significant DEGs
  deg_filtered <- deg_results %>%
    filter(abs(avg_log2FC) > fc_threshold & p_val_adj < pval_threshold)
  
  cat(sprintf("Significant DEGs: %d (|log2FC| > %.2f, adj p-val < %.2f)\n", 
              nrow(deg_filtered), fc_threshold, pval_threshold))
  cat(sprintf("  Upregulated in group1: %d\n", sum(deg_filtered$avg_log2FC > 0)))
  cat(sprintf("  Upregulated in group2: %d\n", sum(deg_filtered$avg_log2FC < 0)))
  
  # Save filtered results
  write.csv(deg_filtered,
            file = paste0(output_prefix, "_significant_genes.csv"),
            row.names = FALSE)
  cat(sprintf("Saved significant DEGs: %s_significant_genes.csv\n", output_prefix))
  
  # Pathway enrichment analysis using g:Profiler
  pathway_results <- NULL
  if (nrow(deg_filtered) > 0) {
    cat("\nPerforming pathway enrichment analysis with g:Profiler...\n")
    
    # Split into upregulated and downregulated genes
    up_genes <- deg_filtered %>% filter(avg_log2FC > 0) %>% pull(gene)
    down_genes <- deg_filtered %>% filter(avg_log2FC < 0) %>% pull(gene)
    
    pathway_results <- list()
    
    # Enrichment for upregulated genes (group1)
    if (length(up_genes) > 0) {
      cat(sprintf("  Analyzing %d upregulated genes (group1)...\n", length(up_genes)))
      gost_up <- gost(
        query = up_genes,
        organism = "mmulatta",  # Change to "mmusculus" for mouse
        significant = TRUE,
        correction_method = "g_SCS",  # g:Profiler's custom method
        sources = c("GO:BP", "GO:MF", "GO:CC", "KEGG", "REAC", "WP")
      )
      
      if (!is.null(gost_up$result)) {
        # Select available columns
        base_cols <- c("term_id", "term_name", "source", "p_value", "term_size", 
                       "query_size", "intersection_size")
        optional_cols <- c("intersection")
        available_cols <- base_cols[base_cols %in% colnames(gost_up$result)]
        optional_available <- optional_cols[optional_cols %in% colnames(gost_up$result)]
        
        pathway_results$upregulated <- gost_up$result %>%
          select(all_of(c(available_cols, optional_available))) %>%
          arrange(p_value)
        
        write.csv(pathway_results$upregulated,
                  file = paste0(output_prefix, "_pathways_group1_up.csv"),
                  row.names = FALSE)
        cat(sprintf("  Found %d enriched pathways for group1\n", 
                    nrow(pathway_results$upregulated)))
      } else {
        cat("  No significant pathways found for upregulated genes\n")
      }
    }
    
    # Enrichment for downregulated genes (group2)
    if (length(down_genes) > 0) {
      cat(sprintf("  Analyzing %d downregulated genes (group2)...\n", length(down_genes)))
      gost_down <- gost(
        query = down_genes,
        organism = "hsapiens",
        significant = TRUE,
        correction_method = "g_SCS",
        sources = c("GO:BP", "GO:MF", "GO:CC", "KEGG", "REAC", "WP")
      )
      
      if (!is.null(gost_down$result)) {
        # Select available columns
        base_cols <- c("term_id", "term_name", "source", "p_value", "term_size",
                       "query_size", "intersection_size")
        optional_cols <- c("intersection")
        available_cols <- base_cols[base_cols %in% colnames(gost_down$result)]
        optional_available <- optional_cols[optional_cols %in% colnames(gost_down$result)]
        
        pathway_results$downregulated <- gost_down$result %>%
          select(all_of(c(available_cols, optional_available))) %>%
          arrange(p_value)
        
        write.csv(pathway_results$downregulated,
                  file = paste0(output_prefix, "_pathways_group2_up.csv"),
                  row.names = FALSE)
        cat(sprintf("  Found %d enriched pathways for group2\n",
                    nrow(pathway_results$downregulated)))
      } else {
        cat("  No significant pathways found for downregulated genes\n")
      }
    }
  } else {
    cat("No significant DEGs found - skipping pathway analysis\n")
  }
  
  # Create volcano plot
  cat("\nGenerating volcano plot...\n")
  
  # Add significance category
  deg_results$significance <- "Not significant"
  deg_results$significance[deg_results$avg_log2FC > fc_threshold & 
                            deg_results$p_val_adj < pval_threshold] <- "Group1 enriched"
  deg_results$significance[deg_results$avg_log2FC < -fc_threshold & 
                            deg_results$p_val_adj < pval_threshold] <- "Group2 enriched"
  
  # Label top genes
  all_genes <- deg_results %>%
    filter(significance != "Not significant") %>%
    arrange(p_val_adj)
  
  volcano_plot <- ggplot(deg_results, aes(x = avg_log2FC, y = -log10(p_val_adj))) +
    geom_point(aes(color = significance), alpha = 0.6, size = 1.5) +
    scale_color_manual(
      values = c("Group1 enriched" = "#E64B35", 
                 "Not significant" = "grey70",
                 "Group2 enriched" = "#4DBBD5")
    ) +
    geom_vline(xintercept = c(-fc_threshold, fc_threshold), 
               linetype = "dashed", color = "grey30", linewidth = 0.5) +
    geom_hline(yintercept = -log10(pval_threshold), 
               linetype = "dashed", color = "grey30", linewidth = 0.5) +
    geom_text_repel(
      data = all_genes,
      aes(label = gene),
      size = 3,
      max.overlaps = 20,
      box.padding = 0.5,
      segment.size = 0.2
    ) +
    labs(
      title = "Differential Gene Expression: Group1 vs Group2",
      x = "log2 Fold Change",
      y = "-log10(Adjusted P-value)",
      color = "Significance"
    ) +
    theme_classic(base_size = 12) +
    theme(
      legend.position = "right",
      plot.title = element_text(hjust = 0.5, face = "bold"),
      axis.text = element_text(color = "black"),
      axis.title = element_text(face = "bold")
    )
  
  # Save plot
  ggsave(
    filename = paste0(output_prefix, "_volcano_plot.pdf"),
    plot = volcano_plot,
    width = 10,
    height = 8,
    units = "in"
  )
  
  ggsave(
    filename = paste0(output_prefix, "_volcano_plot.png"),
    plot = volcano_plot,
    width = 10,
    height = 8,
    units = "in",
    dpi = 300
  )
  
  cat(sprintf("Saved volcano plots: %s_volcano_plot.pdf/png\n", output_prefix))
  
  # Return results
  return(list(
    all_results = deg_results,
    significant_degs = deg_filtered,
    pathway_results = pathway_results,
    volcano_plot = volcano_plot,
    n_group1 = length(cells_group1),
    n_group2 = length(cells_group2),
    parameters = list(
      fc_threshold = fc_threshold,
      pval_threshold = pval_threshold,
      test_method = test_use
    )
  ))
}

```

```{r}
input.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS016/singleCell/Seurat/RDS/"
output.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/joint/subset/new/"

samples <- c("Invitro",
             "D195",
             "Pacute",
             "W2")
for (i in 1:length(samples)) {
  sample_name <- samples[i]
  input_file <- paste0(input.dir, sample_name, ".rds")
  
  tryCatch({
    seurat_obj <- readRDS(input_file)
    genes_to_exclude <- c("CD4", "CD8A", "CD8B")  # remove from DEG analysis
    genes_to_keep <- rownames(seurat_obj)[!rownames(seurat_obj) %in% genes_to_exclude]
    seurat_obj <- subset(seurat_obj, features = genes_to_keep)

    if (!dir.exists(output.dir)) {
      dir.create(output.dir, recursive = TRUE)
    }
    
    salve <- combined_data[[sample_name]]
    available_cols <- intersect(c("MS", "SS-MS", "SS"), colnames(salve))
    list_of_cells1 <- salve %>%
      filter(if_any(all_of(available_cols), ~ . != 0)) %>%
      select(cellID) %>%
      unlist()
    list_of_cells2 <- salve %>% filter(total != 0) %>%
      select(cellID) %>% filter(!(cellID %in% list_of_cells1)) %>% unlist()

    result <- analyze_cell_subset(
      seurat_obj,
      list_of_cells1,
      list_of_cells2,
      group_names = c("SIV_spliced", "SIV_other")
    )
    save_subset_analysis(result, paste0(output.dir, "combined_splicedvnot/", sample_name, "/"))
    
    
    cat("Processed sample:", sample_name, "\n")
  }, error = function(e) {
    cat("Error processing sample", sample_name, ":", conditionMessage(e), "\n\n")
  })
}
```



### CoGAPS
With all genes, takes 6hrs to run
I determined that 3 is not enough clusters
```{r all genes}
invitro <- readRDS("/projects/b1042/GoyalLab/egrody/extractedData/EGS016/singleCell/Seurat/RDS/Invitro.rds")

counts_matrix <- as.matrix(GetAssayData(invitro, assay = "RNA", layer = "counts"))
norm_counts_matrix <- log1p(counts_matrix)
invitro_params <- CogapsParams(nIterations=10000,
  seed=42,
  nPatterns=3,
  sparseOptimization=TRUE,
  distributed="genome-wide")
invitro_params <- setDistributedParams(invitro_params, nSets=5)
startTime <- Sys.time()
invitro_nmf_result <- CoGAPS(norm_counts_matrix, invitro_params)
endTime <- Sys.time()
output.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/joint/subset/CoGAPS/"
saveRDS(invitro_nmf_result, paste0(output.dir, "invitro_cogaps_result.rds"))

```

With highly variable genes, takes 4hrs to run
```{r hvg only}
invitro <- readRDS("/projects/b1042/GoyalLab/egrody/extractedData/EGS016/singleCell/Seurat/RDS/Invitro.rds")
invitro_var <- FindVariableFeatures(invitro, selection.method = "mvp", nfeatures = 3000) #mvp robust to sparse data
hvgs <- VariableFeatures(invitro_var)
counts_matrix <- as.matrix(GetAssayData(invitro_var, assay = "RNA", layer = "counts"))
counts_hvg <- counts_matrix[hvgs, ]
colnames(counts_hvg) <- colnames(invitro_var)
norm_counts_hvg <- log1p(counts_hvg)
invitro_params <- CogapsParams(
    nIterations = 15000,
    seed = 42,
    nPatterns = 8,
    sparseOptimization = TRUE,
    distributed = "genome-wide"
)
invitro_params <- setDistributedParams(invitro_params, nSets = 5)
startTime <- Sys.time()
cogapsresult_hvg <- CoGAPS(norm_counts_hvg, invitro_params)
endTime <- Sys.time()
output.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/joint/subset/CoGAPS/"
saveRDS(cogapsresult_hvg, paste0(output.dir, "invitro_cogaps_result_hvg.rds"))
```


Looking at results
```{r for Yuval}
# for Yuval
sample_factors <- getSampleFactors(cogapsresult)

# Compute the sum of pattern weights for each cell
pattern_weight_sums <- rowSums(sample_factors)

# Plot the distribution
ggplot(data.frame(weight_sum = pattern_weight_sums), aes(x = weight_sum)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "black") +
  labs(
    x = "Sum of Pattern Weights",
    y = "Number of Cells",
    title = "Distribution of Total Pattern Weights per Cell"
  ) +
  theme_minimal()
```


```{r NMF loading}
invitro <- readRDS("/projects/b1042/GoyalLab/egrody/extractedData/EGS016/singleCell/Seurat/RDS/Invitro.rds")
output.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/joint/subset/CoGAPS/"
#cogapsresult <- readRDS(paste0(output.dir, "invitro_cogaps_result.rds"))
cogapsresult <- readRDS(paste0(output.dir, "invitro_cogaps_result_hvg.rds"))
patterns_in_order <-t(cogapsresult@sampleFactors[colnames(invitro),])
invitro[["CoGAPS"]] <- CreateAssayObject(counts = patterns_in_order) 
DefaultAssay(invitro) <- "CoGAPS" 
pattern_names = rownames(invitro@assays$CoGAPS)

pm = patternMarkers(cogapsresult)

# plotting FeaturePlots
library(viridis)
color_palette <- viridis(n=10) 
plots <- FeaturePlot(
  invitro,
  features = pattern_names,
  cols = color_palette,
  reduction = "umap",
  combine = FALSE 
)
plots <- lapply(plots, function(p) {
  p + 
    coord_fixed(ratio = 1) +
    theme(
      aspect.ratio = 1,
      legend.position = "none"
    )
})
legend_plot <- FeaturePlot(
  invitro,
  features = pattern_names[1],
  cols = color_palette,
  reduction = "umap"
) +
  coord_fixed(ratio = 1) +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 8)
  ) +
  labs(color = "Weight")
library(cowplot)
shared_legend <- get_legend(legend_plot)
combined_plots <- wrap_plots(plots, ncol = 4) 
final_plot <- plot_grid(
  combined_plots,
  shared_legend,
  ncol = 2,
  rel_widths = c(1, 0.08)
)
print(final_plot)
save_plot(
  paste0(output.dir, "invitro_patterns_featureplot.pdf"),
  final_plot,
  base_width = 12,
  base_height = 10
)

```

```{r NMF QC}
# Check if patterns have meaningful variance or are dominated by a few cells
library(reshape2)

pattern_weights <- as.data.frame(t(patterns_in_order))
pattern_weights$cell <- rownames(pattern_weights)
weights_long <- melt(pattern_weights, id.vars = "cell", variable.name = "Pattern", value.name = "Weight")

ggplot(weights_long, aes(x = Weight)) +
    geom_histogram(bins = 50) +
    facet_wrap(~Pattern, scales = "free_y") +
    theme_minimal() +
    ggtitle("Distribution of pattern weights across cells")

# High correlation between patterns suggests redundancy (too many patterns)
# or that patterns aren't capturing distinct signals (too few iterations or patterns)
pattern_cor <- cor(t(patterns_in_order))
print(round(pattern_cor, 3))

# Examine what genes define each pattern
markers = patternMarkers(cogapsresult, threshold = "cut")
for (i in 1:ncol(cogapsresult@featureLoadings)) {
  pattern_name <- colnames(cogapsresult@featureLoadings)[i]
  cat("\nPattern", i, "markers:\n")
  
  # patternMarkers returns a list with PatternMarkers element
  pattern_markers <- markers$PatternMarkers[[pattern_name]]
  
  cat("Number of markers:", length(pattern_markers), "\n")
  cat("Genes:", paste(pattern_markers, collapse = ", "), "\n")
}

library(fgsea)
library(msigdbr)

#' Perform hallmark pathway enrichment for CoGAPS patterns
#' 
#' @param cogaps_result CoGAPS result object
#' @param threshold Threshold for patternMarkers ("cut" or "all")
#' @param species Species for msigdbr ("Homo sapiens" or "Mus musculus")
#' @param min_size Minimum gene set size for fgsea
#' @param max_size Maximum gene set size for fgsea
#' @return List containing enrichment results per pattern and combined data frame
pattern_hallmark_enrichment <- function(
    cogaps_result,
    threshold = "all",
    species = "Homo sapiens",
    min_size = 15,
    max_size = 500
) {
  # Get pattern markers
  markers <- patternMarkers(cogaps_result, threshold = threshold)
  # Get feature loadings for ranking (used as statistic for fgsea)
  feature_loadings <- cogaps_result@featureLoadings
  # Retrieve MSigDB hallmark gene sets
  hallmark_sets <- msigdbr(species = species, category = "H") %>%
    split(x = .$gene_symbol, f = .$gs_name)
  cat("Retrieved", length(hallmark_sets), "hallmark gene sets\n")
  cat("Total genes in CoGAPS result:", nrow(feature_loadings), "\n")
  # Run enrichment for each pattern
  pattern_names <- colnames(feature_loadings)
  results_list <- list()
  for (i in seq_along(pattern_names)) {
    pattern <- pattern_names[i]
    cat("\nProcessing", pattern, "...\n")
    # Create ranked gene list from feature loadings
    # fgsea uses the ranking to compute enrichment
    gene_ranks <- feature_loadings[, i]
    names(gene_ranks) <- rownames(feature_loadings)
    gene_ranks <- sort(gene_ranks, decreasing = TRUE)
    # Remove NA or zero variance
    gene_ranks <- gene_ranks[!is.na(gene_ranks)]
    # Run fgsea
    fgsea_res <- fgsea(
      pathways = hallmark_sets,
      stats = gene_ranks,
      minSize = min_size,
      maxSize = max_size
    )
    # Add pattern identifier and sort by adjusted p-value
    fgsea_res$pattern <- pattern
    fgsea_res <- fgsea_res[order(fgsea_res$padj), ]
    # Report significant results
    sig_pathways <- sum(fgsea_res$padj < 0.05, na.rm = TRUE)
    cat("  Significant pathways (padj < 0.05):", sig_pathways, "\n")
    
    results_list[[pattern]] <- fgsea_res
  }
  # Combine into single data frame
  combined_df <- do.call(rbind, results_list)
  rownames(combined_df) <- NULL
  
  # Create summary of top pathway per pattern
  top_per_pattern <- combined_df %>%
    group_by(pattern) %>%
    slice_min(padj, n = 5, with_ties = FALSE) %>%
    select(pattern, pathway, NES, padj, size) %>%
    as.data.frame()
  
  return(list(
    by_pattern = results_list,
    combined = combined_df,
    top_summary = top_per_pattern
  ))
}

# Run the analysis
hallmark_results <- pattern_hallmark_enrichment(
  cogaps_result = cogapsresult,
  #threshold = "cut",
  species = "Macaca mulatta"
)

# View top pathways per pattern
print(hallmark_results$top_summary)

# Export full results
# write.csv(
#   hallmark_results$combined,
#   "cogaps_hallmark_enrichment.csv",
#   row.names = FALSE
# )




#' Plot hallmark enrichment as horizontal bar chart
#' 
#' @param hallmark_results Output from pattern_hallmark_enrichment()
#' @param pattern_name Which pattern to plot (e.g., "Pattern_1")
#' @param top_n Maximum number of pathways to display
plot_hallmark_bars <- function(
    hallmark_results,
    pattern_name,
    top_n = 20
) {
  
  # Extract results for specified pattern - no filtering by significance
  plot_data <- hallmark_results$by_pattern[[pattern_name]] %>%
    arrange(padj) %>%
    head(top_n) %>%
    mutate(
      pathway = gsub("HALLMARK_", "", pathway),
      neg_log10_padj = -10 * log10(padj),
      pathway = factor(pathway, levels = rev(pathway))
    )
  
  if (nrow(plot_data) == 0) {
    cat("No pathways found for", pattern_name, "\n")
    return(NULL)
  }
  
  # Significance threshold line: -10 * log10(0.05) = 13.01
  sig_line <- -10 * log10(0.05)
  
  # Create plot
  p <- ggplot(plot_data, aes(x = pathway, y = neg_log10_padj)) +
    geom_col(fill = "gray50") +
    geom_hline(yintercept = sig_line, linetype = "dashed", color = "red") +
    geom_text(
      aes(label = sprintf("%.3e", padj)),
      hjust = -0.1,
      size = 3
    ) +
    coord_flip(clip = "off") +
    theme_minimal() +
    theme(
      axis.text.y = element_text(size = 9),
      plot.title = element_text(hjust = 0.5, face = "bold"),
      plot.margin = margin(5, 50, 5, 5)
    ) +
    labs(
      x = "MsigDB Hallmark",
      y = expression("-10" %*% "log10(FDR q-value)"),
      title = paste("Overrepresented msigDB hallmarks in", pattern_name)
    ) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.15)))
  
  return(p)
}

# Plot for a specific pattern
p <- plot_hallmark_bars(
  hallmark_results,
  pattern_name = "Pattern_7",
  top_n = 20
)
print(p)

# Generate plots for all patterns
pattern_names <- names(hallmark_results$by_pattern)

pdf(paste0(output.dir, "invitro_hallmarks.pdf"), width = 10, height = 8)
for (pat in pattern_names) {
  p <- plot_hallmark_bars(
    hallmark_results,
    pattern_name = pat,
    top_n = 20
  )
  if (!is.null(p)) print(p)
}
dev.off()
```

Subsetting for infected vs uninfected
```{r NMF SIV+ v SIV-}
joint_dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/joint/dataframes/"
invitro_both <- read.csv(paste0(joint_dir, "Invitro_both.csv"))
invitro_GEXonly <- read.csv(paste0(joint_dir, "Invitro_onlySingleCell.csv"))
invitro_SALVEonly <- read.csv(paste0(joint_dir, "Invitro_onlySALVE.csv"))
invitro_all <- read.csv(paste0(joint_dir, "Invitro_all.csv"))

# first by SALVE
infected <- rbind(invitro_SALVEonly, invitro_both)
pattern_matrix <- getSampleFactors(cogapsresult)
infected_cellids <- infected$cellID
all_cellids <- rownames(pattern_matrix)
uninfected_cellids <- setdiff(all_cellids, infected_cellids)
infected_patterns <- pattern_matrix[rownames(pattern_matrix) %in% infected_cellids, ]
uninfected_patterns <- pattern_matrix[rownames(pattern_matrix) %in% uninfected_cellids, ]

# Compare pattern usage between subsets
infected_means <- colMeans(infected_patterns)
uninfected_means <- colMeans(uninfected_patterns)

comparison_df <- data.frame(
  pattern = colnames(pattern_matrix),
  infected_mean = infected_means,
  uninfected_mean = uninfected_means,
  log2_fold_change = log2(infected_means / uninfected_means)
)
print(comparison_df)

stat_results <- lapply(colnames(pattern_matrix), function(pat) {
  test <- wilcox.test(
    infected_patterns[, pat],
    uninfected_patterns[, pat]
  )
  data.frame(
    pattern = pat,
    p_value = test$p.value,
    infected_median = median(infected_patterns[, pat]),
    uninfected_median = median(uninfected_patterns[, pat])
  )
})
stat_df <- do.call(rbind, stat_results)
stat_df$p_adj <- p.adjust(stat_df$p_value, method = "BH")
print(stat_df)



# plot pattern weight
plot_data <- rbind(
  data.frame(
    cellID = rownames(infected_patterns),
    group = "infected",
    as.data.frame(infected_patterns)
  ),
  data.frame(
    cellID = rownames(uninfected_patterns),
    group = "uninfected",
    as.data.frame(uninfected_patterns)
  )
)
plot_long <- pivot_longer(
  plot_data,
  cols = -c(cellID, group),
  names_to = "pattern",
  values_to = "weight"
)
max_weights <- plot_long %>%
  group_by(pattern) %>%
  summarise(ymax = max(weight), .groups = "drop")
sig_data <- merge(stat_df, max_weights, by = "pattern")
sig_data$y_pos <- sig_data$ymax * 1.05
sig_data$label <- case_when(
  sig_data$p_adj < 0.001 ~ "***",
  sig_data$p_adj < 0.01 ~ "**",
  sig_data$p_adj < 0.05 ~ "*",
  TRUE ~ "ns"
)
ggplot(plot_long, aes(x = pattern, y = weight, fill = group)) +
  geom_boxplot(outlier.size = 0.5) +
  geom_text(
    data = sig_data,
    aes(x = pattern, y = y_pos, label = label),
    inherit.aes = FALSE,
    size = 4
  ) +
  scale_fill_manual(values = c("infected" = "red", "uninfected" = "gray")) +
  theme_minimal() +
  labs(y = "Pattern Weight", x = "Pattern") +
  theme(legend.position = "right")
ggsave(paste0(output.dir, "invitro_SIVposneg_patternweight.pdf"))


# plot log2FC
ggplot(comparison_df, aes(x = pattern, y = log2_fold_change, fill = log2_fold_change > 0)) +
  geom_col() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_fill_manual(
    values = c("TRUE" = "red", "FALSE" = "gray"),
    labels = c("TRUE" = "Higher in infected", "FALSE" = "Higher in uninfected"),
    name = "Direction"
  ) +
  theme_minimal() +
  labs(y = "Log2 Fold Change (infected/uninfected)", x = "Pattern") +
  # Add significance markers
  geom_text(
    data = merge(comparison_df, stat_df[, c("pattern", "p_adj")], by = "pattern"),
    aes(label = case_when(
      p_adj < 0.001 ~ "***",
      p_adj < 0.01 ~ "**",
      p_adj < 0.05 ~ "*",
      TRUE ~ ""
    )),
    vjust = ifelse(comparison_df$log2_fold_change > 0, -0.5, 1.5),
    size = 4
  )
ggsave(paste0(output.dir, "invitro_SIVposneg_log2fc.pdf"))
```

Subsetting for splice+SIV+ vs splice-SIV+
```{r NMF splice+ v splice-}
joint_dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/joint/dataframes/"
invitro_both <- read.csv(paste0(joint_dir, "Invitro_both.csv"))
invitro_GEXonly <- read.csv(paste0(joint_dir, "Invitro_onlySingleCell.csv"))
invitro_SALVEonly <- read.csv(paste0(joint_dir, "Invitro_onlySALVE.csv"))
invitro_all <- read.csv(paste0(joint_dir, "Invitro_all.csv"))

# first by SALVE
infected <- rbind(invitro_SALVEonly, invitro_both)
splicers <- infected %>%
  filter(SS > 0 | MS > 0)
nonsplicers <- anti_join(infected, splicers, by = "cellID") #setdiff also works
pattern_matrix <- getSampleFactors(cogapsresult)
splicers_cellids <- splicers$cellID
nonsplicers_cellids <- nonsplicers$cellID
splicers_patterns <- pattern_matrix[rownames(pattern_matrix) %in% splicers_cellids, ]
nonsplicers_patterns <- pattern_matrix[rownames(pattern_matrix) %in% nonsplicers_cellids, ]

# Compare pattern usage between subsets
splicers_means <- colMeans(splicers_patterns)
nonsplicers_means <- colMeans(nonsplicers_patterns)

comparison_df <- data.frame(
  pattern = colnames(pattern_matrix),
  splicers_mean = splicers_means,
  nonsplicers_mean = nonsplicers_means,
  log2_fold_change = log2(splicers_means / nonsplicers_means)
)
#print(comparison_df)

stat_results <- lapply(colnames(pattern_matrix), function(pat) {
  test <- wilcox.test(
    splicers_patterns[, pat],
    nonsplicers_patterns[, pat]
  )
  data.frame(
    pattern = pat,
    p_value = test$p.value,
    splicers_median = median(splicers_patterns[, pat]),
    nonsplicers_median = median(nonsplicers_patterns[, pat])
  )
})
stat_df <- do.call(rbind, stat_results)
stat_df$p_adj <- p.adjust(stat_df$p_value, method = "BH")
#print(stat_df)



# plot pattern weight
plot_data <- rbind(
  data.frame(
    cellID = rownames(splicers_patterns),
    group = "splicers",
    as.data.frame(splicers_patterns)
  ),
  data.frame(
    cellID = rownames(nonsplicers_patterns),
    group = "nonsplicers",
    as.data.frame(nonsplicers_patterns)
  )
)
plot_long <- pivot_longer(
  plot_data,
  cols = -c(cellID, group),
  names_to = "pattern",
  values_to = "weight"
)
max_weights <- plot_long %>%
  group_by(pattern) %>%
  summarise(ymax = max(weight), .groups = "drop")
sig_data <- merge(stat_df, max_weights, by = "pattern")
sig_data$y_pos <- sig_data$ymax * 1.05
sig_data$label <- case_when(
  sig_data$p_adj < 0.001 ~ "***",
  sig_data$p_adj < 0.01 ~ "**",
  sig_data$p_adj < 0.05 ~ "*",
  TRUE ~ "ns"
)
ggplot(plot_long, aes(x = pattern, y = weight, fill = group)) +
  geom_boxplot(outlier.size = 0.5) +
  geom_text(
    data = sig_data,
    aes(x = pattern, y = y_pos, label = label),
    inherit.aes = FALSE,
    size = 4
  ) +
  scale_fill_manual(values = c("splicers" = "orange", "nonsplicers" = "purple")) +
  theme_minimal() +
  labs(y = "Pattern Weight", x = "Pattern") +
  theme(legend.position = "right")
ggsave(paste0(output.dir, "invitro_spliceNonsplice_patternweight.pdf"))


# plot log2FC
ggplot(comparison_df, aes(x = pattern, y = log2_fold_change, fill = log2_fold_change > 0)) +
  geom_col() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_fill_manual(
    values = c("TRUE" = "orange", "FALSE" = "purple"),
    labels = c("TRUE" = "Higher in splicers", "FALSE" = "Higher in nonsplicers"),
    name = "Direction"
  ) +
  theme_minimal() +
  labs(y = "Log2 Fold Change (splicers/nonsplicers)", x = "Pattern") +
  # Add significance markers
  geom_text(
    data = merge(comparison_df, stat_df[, c("pattern", "p_adj")], by = "pattern"),
    aes(label = case_when(
      p_adj < 0.001 ~ "***",
      p_adj < 0.01 ~ "**",
      p_adj < 0.05 ~ "*",
      TRUE ~ ""
    )),
    vjust = ifelse(comparison_df$log2_fold_change > 0, -0.5, 1.5),
    size = 4
  )
ggsave(paste0(output.dir, "invitro_spliceNonsplice_log2fc.pdf"))
```



## Saturation analysis

Why don't we see those 10X only cells also in SALVE?

### cellID sampling without reads

```{r cellID sampling}
input.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS014/SALVE/bamsort/split/"
output.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS014/SALVE/saturation/sample_cellID/"

# Process each sample
for (sample in samples$samples) {
  cat("Processing sample:", sample, "\n")
  
  # Read the data files for this sample
  reads_D1_nef <- read.csv(paste0(input.dir, sample, "_D1_nef_bamsort_split_inside.csv"))
  reads_LTR_tat <- read.csv(paste0(input.dir, sample, "_LTR_tat_bamsort_split_inside.csv"))
  
  # Combine all reads and calculate total
  reads_total <- bind_rows(reads_D1_nef, reads_LTR_tat) %>%
    group_by(cellID, UMI) %>%
    summarize(reads = sum(reads), .groups = "drop")
  
  # Create molecule datasets (remove reads column)
  molecules_D1_nef <- reads_D1_nef %>% select(-reads)
  molecules_LTR_tat <- reads_LTR_tat %>% select(-reads)
  molecules_total <- reads_total %>% select(-reads)
  
  # Create samples list
  sample_datasets <- list(
    D1_nef = molecules_D1_nef,
    LTR_tat = molecules_LTR_tat,
    total = molecules_total
  )
  plot_cumulative_distribution(reads_total$reads, sample)
  
  # Process each dataset type
  # for (dataset_name in names(sample_datasets)) {
  #   dataset <- sample_datasets[[dataset_name]]
  #   
  #   results <- sample_cellID(dataset)
  #   summary <- analyze_cellID_sampling(results, title = paste0("cellID Saturation: ", sample, "_", dataset_name))
  #   
  #   # Create output filenames with sample prefix
  #   results_file <- paste0(output.dir, sample, "_", dataset_name, "_sample_cellID_results.csv")
  #   summary_file <- paste0(output.dir, sample, "_", dataset_name, "_sample_cellID_summary.csv")
  #   
  #   write.csv(results, results_file, row.names = FALSE)
  #   write.csv(summary, summary_file, row.names = FALSE)
  # 
  # }
}





plot_cumulative_distribution <- function(data_vector, sample_name) {
    df_plot <- data.frame(values = data_vector)
    
    # Plot 2: Cumulative distribution
    df_sorted <- data.frame(
      rank = 1:length(data_vector),
      value = sort(data_vector),
      cumulative_pct = (1:length(data_vector)) / length(data_vector) * 100
    )
    
    # Calculate percentage above 3
    pct_above_3 <- sum(data_vector >= 3) / length(data_vector) * 100
    
    p <- ggplot(df_sorted, aes(x = value, y = cumulative_pct)) +
      geom_line(color = "red", linewidth = 1) +
      geom_vline(xintercept = 3, color = "blue", linetype = "dashed", linewidth = 1) +
      annotate("text", x = 3, y = 50, 
               label = paste0(round(pct_above_3, 1), "% ≥ 3 reads"), 
               color = "blue", hjust = -0.1, vjust = 0.5, size = 10) +
      labs(title = paste0("Reads/UMI Distribution:", sample_name),
           x = "Reads/UMI", y = "Cumulative Percentage") +
      theme_minimal() +
      theme(
        plot.title = element_text(size = 18),
        axis.title = element_text(size = 16),
        axis.text = element_text(size = 14),
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 16)
      ) +
      scale_x_log10(
        breaks = c(1, 10, 100, 1000, 10000),
        labels = c("1", "10", "100", "1,000", "10,000")
      )
    
    print(p)
  
  return(p)
}


plot_cumulative_distribution <- function(data_vector, sample_name) {
    df_plot <- data.frame(values = data_vector)
    
    # Plot 2: Cumulative distribution
    df_sorted <- data.frame(
      rank = 1:length(data_vector),
      value = sort(data_vector),
      cumulative_pct = (1:length(data_vector)) / length(data_vector) * 100
    )
    
    # Calculate percentage and count above 3
    count_above_3 <- sum(data_vector >= 3)
    pct_above_3 <- count_above_3 / length(data_vector) * 100
    
    p <- ggplot(df_sorted, aes(x = value, y = cumulative_pct)) +
      geom_line(color = "red", linewidth = 1) +
      geom_vline(xintercept = 3, color = "blue", linetype = "dashed", linewidth = 1) +
      annotate("text", x = 3, y = 50, 
               label = paste0(round(pct_above_3, 1), "% ≥ 3 reads\n(", format(count_above_3, big.mark = ","), " UMIs)"), 
               color = "blue", hjust = -0.1, vjust = 0.5, size = 8) +
      labs(title = paste0("Reads/UMI Distribution:", sample_name),
           x = "Reads/UMI", y = "Cumulative Percentage") +
      theme_minimal() +
      theme(
        plot.title = element_text(size = 18),
        axis.title = element_text(size = 16),
        axis.text = element_text(size = 14),
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 16)
      ) +
      scale_x_log10(
        breaks = c(1, 10, 100, 1000, 10000),
        labels = c("1", "10", "100", "1,000", "10,000")
      )
  
    print (p)
  return(p)
}

# Example usage:
# freq_results <- plot_cumulative_distribution(reads_total$reads, "Uninfected")
```

```{r model fitting}
model_results <- fit_models(results, target_coverage = 95)
model_results <- fit_models(results, target_coverage = 99)
model_results <- fit_models(results, target_coverage = 100)
model_results <- fit_models(results, target_coverage = 120)
```

### UMI sampling with reads

```{r}
input.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS014/SALVE/bamsort/split/"
output.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS014/SALVE/saturation/sample_UMI/"

# Process each sample
for (sample in samples$samples) {
  cat("Processing sample:", sample, "\n")
  
  # Read the data files for this sample
  reads_D1_nef <- read.csv(paste0(input.dir, sample, "_D1_nef_bamsort_split_inside.csv"))
  reads_LTR_tat <- read.csv(paste0(input.dir, sample, "_LTR_tat_bamsort_split_inside.csv"))
  
  # Combine all reads and calculate total
  reads_total <- bind_rows(reads_D1_nef, reads_LTR_tat) %>%
    group_by(cellID, UMI) %>%
    summarize(reads = sum(reads), .groups = "drop")
  
  # Create samples list
  sample_datasets <- list(
    D1_nef = reads_D1_nef,
    LTR_tat = reads_LTR_tat,
    total = reads_total
  )
  
  # Process each dataset type for this sample
  for (dataset_name in names(sample_datasets)) {
    dataset <- sample_datasets[[dataset_name]]
    
    results <- sample_UMI_weighted(dataset)
    summary <- analyze_UMI_sampling(results, title = paste0("Weighted UMI Saturation: ", sample))
  
    results_file <- paste0(output.dir, sample, "_sample_UMI_results.csv")
    summary_file <- paste0(output.dir, sample, "_sample_UMI_summary.csv")
    write.csv(results, results_file, row.names = FALSE)
    write.csv(summary, summary_file, row.names = FALSE)

  }
}
```

```{r}
model_results <- fit_models(results, target_coverage = 95, coverage_column = "pair_coverage")
model_results <- fit_models(results, target_coverage = 98.9, coverage_column = "pair_coverage")
model_results <- fit_models(results, target_coverage = 100, coverage_column = "pair_coverage")
model_results <- fit_models(results, target_coverage = 120)
```

### SALVE Summary: mac239 vs Mmul_10
```{r loading}
input.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS013/SALVE/bamsort/split/"
reads_summary <- read.csv("/projects/b1042/GoyalLab/egrody/extractedData/EGS013/SALVE/bamsort/split/all_samples_summary.txt", header = TRUE)

sample_names <- reads_summary$Sample
df_numeric <- reads_summary[, -which(names(reads_summary) == "Sample")]
df_transposed <- as.data.frame(t(df_numeric))
colnames(df_transposed) <- sample_names
rownames(df_transposed) <- c("Inside_Rows", "Inside_Total_Reads", "Outside_Rows", "Outside_Total_Reads")
df_transposed <- rbind(df_transposed, Inside_Fraction = round(df_transposed["Inside_Total_Reads",] / df_transposed["Outside_Total_Reads",], 2), Avg_Inside_Reads = round(df_transposed["Inside_Total_Reads",] / df_transposed["Inside_Rows",], 2))

```


# HCR
Comparing EGS016 sequencing to EGS023 FISH
```{r}
# Loading
samples <-  c(
    "Invitro"
)

SALVE.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/SALVE/bamsort/reads/v6/minimum/"
GEX.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS016/singleCell/bamsort/reads/v6/minimum/"

joint_data <- list()

for (i in 1:length(samples)) {
  sample_name <- samples[i]
  
  # Read all data files
  salve_file <- paste0(SALVE.dir, sample_name, "_SALVE_filtered.csv")
  gex_file <- paste0(GEX.dir, sample_name, "_GEX_filtered.csv")
  umap_file <- paste0("/projects/b1042/GoyalLab/egrody/extractedData/EGS016/singleCell/Seurat/UMAPcoords/", 
                      sample_name, "UMAP_coords.csv")
  cd4_file <- paste0("/projects/b1042/GoyalLab/egrody/extractedData/EGS016/singleCell/Seurat/expressionDF/", 
                     sample_name, "_CD4T.csv")
  
  salve_data <- if(file.exists(salve_file)) read.csv(salve_file) else NULL
  gex_data <- if(file.exists(gex_file)) read.csv(gex_file) else NULL
  umap_data <- if(file.exists(umap_file)) read.csv(umap_file, row.names = "X") else NULL
  cd4_data <- if(file.exists(cd4_file)) read.csv(cd4_file, row.names = "X") else NULL
  
  # Skip if no data at all
  if (is.null(salve_data) && is.null(gex_data) && is.null(umap_data)) {
    cat("Skipping", sample_name, ": no data\n")
    next
  }
  
  # Get cellIDs from Mmul_10 only df
  all_cells <- as.character(umap_data$cellID)
  all_cells <- unique(all_cells)
  
  # Create base dataframe
  result <- data.frame(cellID = all_cells)
  
  # Add UMAP data
  if (!is.null(umap_data)) {
    umap_processed <- umap_data %>%
      select(-V2, -sample)
    result <- left_join(result, umap_processed, by = "cellID")
  }
  
  # Add GEX data
  if (!is.null(gex_data)) {
    gex_processed <- gex_data %>%
      mutate(cellID = as.character(cellID)) %>%
      select(cellID, total, US)
    result <- left_join(result, gex_processed, by = "cellID") %>%
      mutate(GEX = log1p(total)) %>%
      mutate(GEX.US = US) %>%
      select(-total, -US)
  } else {
    result$GEX <- 0
  }
  
  # Add SALVE data
  if (!is.null(salve_data)) {
    salve_processed <- salve_data %>%
      #group_by(cellID) %>%
      rename(SALVE = total) %>%
      mutate(cellID = as.character(cellID)) %>%
      select(cellID, SALVE, US, SS, MS, SS.MS)
    result <- left_join(result, salve_processed, by = "cellID") %>%
      mutate(SALVE = log1p(SALVE)) %>%
      mutate(spliced = SS + MS + SS.MS)
  } else {
    result$SALVE <- 0
  }
  
  # Replace NAs with 0
  result[is.na(result)] <- 0
  
  # Cell type filtering
  cd4_data <- cd4_data %>%
    mutate(CD3 = ifelse(CD3D > 0 | CD3E > 0 | CD3G > 0, 1, 0)) %>%
    mutate(CD4 = ifelse(CD4 > 0 , 1, 0)) %>%
    mutate(CD8 = ifelse(CD8A > 0 , 1, 0)) %>%
    select(cellID, CD4, CD3, CD8)
  result <- left_join(result, cd4_data, by = "cellID")
  result <- result %>%
    mutate(across(any_of(c("GEX", "GEX.US")), 
              ~ ifelse(CD4 > 0 & CD3 > 0, .x, 0))) %>% 
    mutate(across(any_of(c("SALVE", "US", "MS", "SS")), 
              ~ ifelse(CD4 > 0 & CD3 > 0, .x, 0))) %>%
    filter(CD4 == 0 | CD8 == 0) %>% # exclude cells CD4+ and CD8+
    select(-CD8, -CD3, -CD4)
  
  joint_data[[sample_name]] <- result
  cat("Processed", sample_name, "\n")
}

df <- joint_data[["Invitro"]]

exp1 <- df %>%
  filter(SALVE > 0) %>%
  select(US, SS, MS, spliced, SALVE) %>%
  rename(total = SALVE)
exp2 <- df %>%
  filter(GEX > 0) %>%
  select(GEX.US, GEX) %>%
  mutate(SS = 0, MS = 0, spliced = 0) %>%
  rename(US = GEX.US, total = GEX)

HCR.data <- read.csv("/projects/b1042/GoyalLab/egrody/extractedData/EGS023/big-FISH/manual/all_cells_stats.csv")
fish <- HCR.data %>%
  filter(n_total_spots > 1) %>%
  select(n_US, n_SS, n_MS, n_total_spots) %>%
  rename(US = n_US, SS = n_SS, MS = n_MS, total = n_total_spots) %>%
  mutate(spliced = SS + MS) 

input_dir="/projects/b1042/GoyalLab/egrody/extractedData/EGS007/longRead/bamsort/isoform/"
longread <- read.csv(paste0(input_dir, "cDNA_isoforms.csv"))
longread <- longread %>%
  count(cellID, isoform) %>%
  pivot_wider(names_from = isoform, values_from = n, values_fill = 0) %>%
  mutate(spliced = MS + SS, total = US + MS + SS + any) %>%
  select(-cellID, -any)

#run combined section under Joint to get the combined_data df
# combined <- combined_data[["Invitro"]]
# combined <- combined %>%
#   mutate(spliced = SS + MS + SS-MS) %>%
#   select(US, SS, MS, spliced, total) %>%
#   filter(total > 0)

```

ECDF
```{r ECDF}
#for plotting
fishlog <- fish %>%
  mutate(across(everything(), log1p))
exp1log <- exp1 %>%
  mutate(across(everything(), log1p))
exp2log <- exp2 %>%
  mutate(across(everything(), log1p))
longreadlog <- longread %>%
  mutate(across(everything(), log1p))


variables <- c("US", "SS", "MS", "spliced", "total")

compute_wasserstein_with_context <- function(exp_vec, ref_vec, n_boot = 1000, seed = 123) {
  set.seed(seed)
  
  # Observed Wasserstein
  w_obs <- wasserstein1d(exp_vec, ref_vec)
  
  # Reference distribution statistics for normalization
  ref_sd <- sd(ref_vec)
  ref_iqr <- IQR(ref_vec)
  ref_mean <- mean(ref_vec)
  
  # Normalized Wasserstein (in units of reference SD)
  w_normalized <- w_obs / ref_sd
  
  # Bootstrap CI for Wasserstein
  n_exp <- length(exp_vec)
  n_ref <- length(ref_vec)
  boot_w <- numeric(n_boot)
  
  for (i in seq_len(n_boot)) {
    boot_exp <- sample(exp_vec, n_exp, replace = TRUE)
    boot_ref <- sample(ref_vec, n_ref, replace = TRUE)
    boot_w[i] <- wasserstein1d(boot_exp, boot_ref)
  }
  
  boot_ci <- quantile(boot_w, probs = c(0.025, 0.975))
  boot_se <- sd(boot_w)
  
  # Permutation test: H0 = distributions are identical
  pooled <- c(exp_vec, ref_vec)
  n_perm <- 1000
  perm_w <- numeric(n_perm)
  
  for (i in seq_len(n_perm)) {
    perm_idx <- sample(length(pooled))
    perm_exp <- pooled[perm_idx[1:n_exp]]
    perm_ref <- pooled[perm_idx[(n_exp + 1):length(pooled)]]
    perm_w[i] <- wasserstein1d(perm_exp, perm_ref)
  }
  
  # One-sided p-value: P(W_perm >= W_obs)
  p_value <- mean(perm_w >= w_obs)
  
  list(
    wasserstein = w_obs,
    wasserstein_normalized = w_normalized,
    ci_lower = boot_ci[1],
    ci_upper = boot_ci[2],
    boot_se = boot_se,
    p_value = p_value,
    ref_sd = ref_sd,
    ref_mean = ref_mean
  )
}

# --- Compute full statistics ---
compute_comparison_stats <- function(exp_raw, exp_log, ref_raw, ref_log, variables,
                                      n_boot = 1000, seed = 123) {
  
  results <- lapply(variables, function(var) {
    # KS test on log scale
    ks_test <- suppressWarnings(ks.test(exp_log[[var]], ref_log[[var]]))
    
    # Wasserstein with context on raw scale
    w_context <- compute_wasserstein_with_context(
      exp_raw[[var]], ref_raw[[var]], n_boot = n_boot, seed = seed
    )
    
    data.frame(
      variable = var,
      ks_d = ks_test$statistic,
      ks_p = ks_test$p.value,
      wasserstein = w_context$wasserstein,
      w_normalized = w_context$wasserstein_normalized,
      w_ci_lower = w_context$ci_lower,
      w_ci_upper = w_context$ci_upper,
      w_se = w_context$boot_se,
      w_pvalue = w_context$p_value,
      ref_sd = w_context$ref_sd,
      ref_mean = w_context$ref_mean
    )
  })
  
  bind_rows(results)
}

results_salve <- compute_comparison_stats(exp1, exp1log, fish, fishlog, variables)
results_gex <- compute_comparison_stats(exp2, exp2log, fish, fishlog, variables)
results_longread <- compute_comparison_stats(longread, longreadlog, fish, fishlog, variables)

print(results_salve)
print(results_gex)
print(results_longread)

# --- Plotting function ---
plot_ecdf_comparison <- function(exp_raw, exp_log, ref_raw, ref_log,
                                  variables = c("US", "SS", "MS", "spliced", "total"),
                                  exp_name = "SALVE", ref_name = "HCR",
                                  n_boot = 1000, seed = 123) {
  
  plot_list <- lapply(variables, function(var) {
    
    # KS test
    ks_test <- suppressWarnings(ks.test(exp_log[[var]], ref_log[[var]]))
    
    # Wasserstein with context
    w_ctx <- compute_wasserstein_with_context(
      exp_raw[[var]], ref_raw[[var]], n_boot = n_boot, seed = seed
    )
    
    # Data for plotting
    plot_df <- data.frame(
      value = c(exp_log[[var]], ref_log[[var]]),
      method = factor(
        rep(c(exp_name, ref_name), times = c(nrow(exp_log), nrow(ref_log))),
        levels = c(exp_name, ref_name)
      )
    )
    
    # Format annotation with context
    # Show normalized W (in SDs) and significance
    sig_symbol <- ifelse(w_ctx$p_value < 0.001, "***",
                  ifelse(w_ctx$p_value < 0.01, "**",
                  ifelse(w_ctx$p_value < 0.05, "*", "ns")))
    
    annotation <- sprintf(
      "D = %.2f\nW = %.1f (%.1f SD)%s",
      ks_test$statistic,
      w_ctx$wasserstein,
      w_ctx$wasserstein_normalized,
      sig_symbol
    )
    
    ggplot(plot_df, aes(x = value, color = method)) +
      stat_ecdf(linewidth = 0.8) +
      scale_color_manual(values = c("blue", "red")) +
      annotate("text", x = -Inf, y = Inf, label = annotation,
               hjust = -0.1, vjust = 1.3, size = 2.8) +
      theme_bw() +
      theme(
        aspect.ratio = 1,
        legend.position = "bottom",
        legend.title = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold", size = 10)
      ) +
      labs(
        title = var,
        x = "Counts (log1p)",
        y = "Cumulative Probability"
      )
  })
  
  wrap_plots(plot_list, nrow = 1) +
    plot_layout(guides = "collect") &
    theme(legend.position = "bottom")
}

p_salve <- plot_ecdf_comparison(exp1, exp1log, fish, fishlog,
                                 exp_name = "SALVE", ref_name = "HCR")
p_gex <- plot_ecdf_comparison(exp2, exp2log, fish, fishlog,
                               exp_name = "GEX", ref_name = "HCR")
p_long <- plot_ecdf_comparison(longread, longreadlog, fish, fishlog,
                               exp_name = "longread", ref_name = "HCR")

output.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS023/plots/"
ggsave(paste0(output.dir, "ecdf_SALVE.pdf"), p_salve, width = 15, height = 4)
ggsave(paste0(output.dir, "ecdf_GEX.pdf"), p_gex, width = 15, height = 4)
ggsave(paste0(output.dir, "ecdf_longread.pdf"), p_long, width = 15, height = 4)


p_salve
p_gex
p_long
```

New three tests for all three methods vs HCR
```{r}
compare_methods_to_reference <- function(
    method_list,
    ref_vec,
    var_name = "Variable",
    n_boot = 2000,
    n_perm = 2000,
    seed = 123
) {
  set.seed(seed)
  
  method_names <- names(method_list)
  n_methods <- length(method_list)
  
  for (nm in method_names) {
    stopifnot(all(method_list[[nm]] >= 0))
  }
  stopifnot(all(ref_vec >= 0))
  
  # Normalize to probability distribution with Laplace smoothing
  normalize <- function(x, max_val) {
    tab <- table(factor(x, levels = 0:max_val))
    (as.numeric(tab) + 1e-10) / (sum(tab) + 1e-10 * (max_val + 1))
  }
  
  # Compute distance metrics between two vectors
  compute_distances <- function(vec1, vec2) {
    max_val <- max(c(vec1, vec2))
    p1 <- normalize(vec1, max_val)
    p2 <- normalize(vec2, max_val)
    
    ks_stat <- suppressWarnings(ks.test(vec1, vec2)$statistic)
    js_val <- sqrt(philentropy::JSD(rbind(p1, p2), unit = "log2", est.prob = NULL))
    
    result <- c(as.numeric(ks_stat), as.numeric(js_val), transport::wasserstein1d(vec1, vec2))
    names(result) <- c("ks", "js", "wasserstein")
    result
  }
  
  # Permutation test for two-sample comparison
  # Tests H0: two samples come from same distribution
  perm_test_distribution <- function(vec1, vec2, obs_dist, n_perm) {
    pooled <- c(vec1, vec2)
    n1 <- length(vec1)
    n2 <- length(vec2)
    
    perm_stats <- matrix(NA, nrow = n_perm, ncol = 3)
    
    for (i in seq_len(n_perm)) {
      perm_idx <- sample(length(pooled))
      perm1 <- pooled[perm_idx[1:n1]]
      perm2 <- pooled[perm_idx[(n1 + 1):(n1 + n2)]]
      perm_stats[i, ] <- compute_distances(perm1, perm2)
    }
    
    # One-sided p-value: proportion >= observed (Phipson & Smyth 2010)
    sapply(seq_len(3), function(j) {
      (sum(perm_stats[, j] >= obs_dist[j]) + 1) / (n_perm + 1)
    })
  }
  
  # Permutation test for difference in distances to reference
  # Tests H0: method1 and method2 are equally distant from reference
  # Returns: p-value and direction (which method is closer)
  perm_test_distance_difference <- function(vec1, vec2, ref_vec, n_perm) {
    n1 <- length(vec1)
    n2 <- length(vec2)
    
    # Observed distances to reference
    obs_d1 <- compute_distances(vec1, ref_vec)
    obs_d2 <- compute_distances(vec2, ref_vec)
    obs_diff <- obs_d1 - obs_d2  # Negative = vec1 closer to ref
    
    # Pool the two method vectors and permute labels
    pooled <- c(vec1, vec2)
    perm_diffs <- matrix(NA, nrow = n_perm, ncol = 3)
    
    for (i in seq_len(n_perm)) {
      perm_idx <- sample(length(pooled))
      perm1 <- pooled[perm_idx[1:n1]]
      perm2 <- pooled[perm_idx[(n1 + 1):(n1 + n2)]]
      
      d1 <- compute_distances(perm1, ref_vec)
      d2 <- compute_distances(perm2, ref_vec)
      perm_diffs[i, ] <- d1 - d2
    }
    
    # Two-sided p-value
    pvals <- sapply(seq_len(3), function(j) {
      (sum(abs(perm_diffs[, j]) >= abs(obs_diff[j])) + 1) / (n_perm + 1)
    })
    names(pvals) <- c("ks", "js", "wasserstein")
    
    list(
      obs_diff = obs_diff,
      pval = pvals,
      perm_diffs = perm_diffs
    )
  }
  
  # Bootstrap CI for distance estimate
  boot_ci <- function(vec1, vec2, n_boot) {
    n1 <- length(vec1)
    n2 <- length(vec2)
    
    boot_stats <- matrix(NA, nrow = n_boot, ncol = 3)
    
    for (i in seq_len(n_boot)) {
      b1 <- sample(vec1, n1, replace = TRUE)
      b2 <- sample(vec2, n2, replace = TRUE)
      boot_stats[i, ] <- compute_distances(b1, b2)
    }
    
    list(
      ci_lower = apply(boot_stats, 2, quantile, probs = 0.025),
      ci_upper = apply(boot_stats, 2, quantile, probs = 0.975),
      se = apply(boot_stats, 2, sd)
    )
  }
  
  # Compute distances from each method to reference
  dist_to_ref <- list()
  ci_to_ref <- list()
  pval_to_ref <- list()
  
  for (nm in method_names) {
    dist_to_ref[[nm]] <- compute_distances(method_list[[nm]], ref_vec)
    ci_to_ref[[nm]] <- boot_ci(method_list[[nm]], ref_vec, n_boot)
    pval_to_ref[[nm]] <- perm_test_distribution(method_list[[nm]], ref_vec, dist_to_ref[[nm]], n_perm)
    names(pval_to_ref[[nm]]) <- c("ks", "js", "wasserstein")
  }
  
  # Compute paired-distance tests: is method A closer to ref than method B?
  paired_distance_tests <- list()
  
  for (i in seq_len(n_methods - 1)) {
    for (j in (i + 1):n_methods) {
      nm1 <- method_names[i]
      nm2 <- method_names[j]
      pair_key <- paste0(nm1, "_vs_", nm2)
      
      test_result <- perm_test_distance_difference(
        method_list[[nm1]], 
        method_list[[nm2]], 
        ref_vec, 
        n_perm
      )
      
      # Determine which method is closer for each metric
      closer <- ifelse(test_result$obs_diff < 0, nm1, nm2)
      names(closer) <- c("ks", "js", "wasserstein")
      
      paired_distance_tests[[pair_key]] <- list(
        obs_diff = test_result$obs_diff,
        pval = test_result$pval,
        closer = closer
      )
    }
  }
  
  # Compile results into data frame format
  metrics <- c("KS", "JS", "Wasserstein")
  
  distance_df <- data.frame(
    variable = var_name,
    metric = rep(metrics, each = n_methods + 1),
    method = rep(c(method_names, "HCR"), 3),
    distance = NA_real_,
    ci_lower = NA_real_,
    ci_upper = NA_real_,
    pval_vs_ref = NA_real_
  )
  
  for (m in seq_along(metrics)) {
    metric_key <- c("ks", "js", "wasserstein")[m]
    for (i in seq_along(method_names)) {
      nm <- method_names[i]
      row_idx <- (m - 1) * (n_methods + 1) + i
      distance_df$distance[row_idx] <- dist_to_ref[[nm]][metric_key]
      distance_df$ci_lower[row_idx] <- ci_to_ref[[nm]]$ci_lower[metric_key]
      distance_df$ci_upper[row_idx] <- ci_to_ref[[nm]]$ci_upper[metric_key]
      distance_df$pval_vs_ref[row_idx] <- pval_to_ref[[nm]][metric_key]
    }
    ref_row_idx <- (m - 1) * (n_methods + 1) + n_methods + 1
    distance_df$distance[ref_row_idx] <- 0
    distance_df$ci_lower[ref_row_idx] <- 0
    distance_df$ci_upper[ref_row_idx] <- 0
    distance_df$pval_vs_ref[ref_row_idx] <- NA
  }
  
  list(
    variable = var_name,
    method_names = method_names,
    distance_df = distance_df,
    dist_to_ref = dist_to_ref,
    pval_to_ref = pval_to_ref,
    paired_distance_tests = paired_distance_tests
  )
}

plot_barplot_comparison <- function(
    results_list,
    var_names = c("US", "spliced", "SS", "MS", "total"),
    method_names = c("SALVE", "GEX", "longread"),
    ref_name = "HCR",
    output_file = NULL,
    width = 16,
    height = 14
) {
  
  format_sig <- function(p) {
    if (is.na(p)) return("")
    if (p < 0.001) return("***")
    if (p < 0.01) return("**")
    if (p < 0.05) return("*")
    return("n.s.")
  }
  
  format_pval <- function(p) {
    if (is.na(p)) return("")
    if (p < 0.0001) return("p<1e-4")
    if (p < 0.001) return(sprintf("p=%.0e", p))
    return(sprintf("p=%.3f", p))
  }
  
  all_dist <- do.call(rbind, lapply(results_list, function(r) r$distance_df))
  
  all_methods <- c(method_names, ref_name)
  all_dist$method <- factor(all_dist$method, levels = all_methods)
  all_dist$variable <- factor(all_dist$variable, levels = var_names)
  
  method_colors <- c("#E41A1C", "#377EB8", "#984EA3", "#4DAF4A")
  names(method_colors) <- all_methods
  
  # Extract paired-distance p-values (tests if one method is closer to ref than another)
  get_paired_distance_pval <- function(result, metric_key, m1, m2) {
    pair_key <- paste0(m1, "_vs_", m2)
    alt_key <- paste0(m2, "_vs_", m1)
    
    if (pair_key %in% names(result$paired_distance_tests)) {
      return(result$paired_distance_tests[[pair_key]]$pval[metric_key])
    } else if (alt_key %in% names(result$paired_distance_tests)) {
      return(result$paired_distance_tests[[alt_key]]$pval[metric_key])
    }
    return(NA)
  }
  
  create_metric_plot <- function(metric_name, metric_label) {
    
    df <- all_dist[all_dist$metric == metric_name, ]
    metric_key <- tolower(metric_name)
    
    y_max_per_var <- tapply(df$ci_upper, df$variable, max, na.rm = TRUE)
    dist_max <- tapply(df$distance, df$variable, max, na.rm = TRUE)
    y_max_per_var[is.na(y_max_per_var) | y_max_per_var == 0] <- dist_max[is.na(y_max_per_var) | y_max_per_var == 0]
    
    p <- ggplot(df, aes(x = variable, y = distance, fill = method)) +
      geom_bar(stat = "identity", position = position_dodge(width = 0.85), width = 0.75) +
      geom_errorbar(
        aes(ymin = ci_lower, ymax = ci_upper),
        position = position_dodge(width = 0.85),
        width = 0.2,
        linewidth = 0.4
      ) +
      scale_fill_manual(values = method_colors) +
      labs(
        title = metric_label,
        x = NULL,
        y = "Distance to HCR Reference",
        fill = "Method"
      ) +
      theme_bw() +
      theme(
        plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
        legend.position = "bottom",
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        axis.text.x = element_text(size = 11, face = "bold")
      )
    
    n_methods_total <- length(all_methods)
    dodge_width <- 0.85
    bar_positions <- seq(-1.5, 1.5, length.out = n_methods_total) * (dodge_width / n_methods_total)
    
    for (i in seq_along(var_names)) {
      var <- var_names[i]
      result <- results_list[[i]]
      
      y_base <- y_max_per_var[var] * 1.08
      y_step <- y_max_per_var[var] * 0.10
      
      bracket_count <- 0
      
      # Comparisons to reference (methods vs HCR): tests if distributions differ
      for (m_idx in seq_along(method_names)) {
        m <- method_names[m_idx]
        pval <- result$pval_to_ref[[m]][metric_key]
        
        x_method <- i + bar_positions[m_idx]
        x_ref <- i + bar_positions[n_methods_total]
        y_bracket <- y_base + bracket_count * y_step
        
        sig <- format_sig(pval)
        label <- paste0(format_pval(pval), " ", sig)
        
        p <- p +
          annotate("segment", x = x_method, xend = x_method, 
                   y = y_bracket, yend = y_bracket + y_step * 0.25, linewidth = 0.3) +
          annotate("segment", x = x_method, xend = x_ref, 
                   y = y_bracket + y_step * 0.25, yend = y_bracket + y_step * 0.25, linewidth = 0.3) +
          annotate("segment", x = x_ref, xend = x_ref, 
                   y = y_bracket, yend = y_bracket + y_step * 0.25, linewidth = 0.3) +
          annotate("text", x = (x_method + x_ref) / 2, y = y_bracket + y_step * 0.45, 
                   label = label, size = 2.0)
        
        bracket_count <- bracket_count + 1
      }
      
      # Paired-distance comparisons: tests if one method is closer to ref than another
      for (m1_idx in seq_len(length(method_names) - 1)) {
        for (m2_idx in (m1_idx + 1):length(method_names)) {
          m1 <- method_names[m1_idx]
          m2 <- method_names[m2_idx]
          
          pval <- get_paired_distance_pval(result, metric_key, m1, m2)
          
          x1 <- i + bar_positions[m1_idx]
          x2 <- i + bar_positions[m2_idx]
          y_bracket <- y_base + bracket_count * y_step
          
          sig <- format_sig(pval)
          label <- paste0(format_pval(pval), " ", sig)
          
          p <- p +
            annotate("segment", x = x1, xend = x1, 
                     y = y_bracket, yend = y_bracket + y_step * 0.25, linewidth = 0.3) +
            annotate("segment", x = x1, xend = x2, 
                     y = y_bracket + y_step * 0.25, yend = y_bracket + y_step * 0.25, linewidth = 0.3) +
            annotate("segment", x = x2, xend = x2, 
                     y = y_bracket, yend = y_bracket + y_step * 0.25, linewidth = 0.3) +
            annotate("text", x = (x1 + x2) / 2, y = y_bracket + y_step * 0.45, 
                     label = label, size = 3.0)
          
          bracket_count <- bracket_count + 1
        }
      }
    }
    
    max_y <- max(y_max_per_var, na.rm = TRUE) * 2.0
    p <- p + coord_cartesian(ylim = c(0, max_y))
    
    p
  }
  
  p_ks <- create_metric_plot("KS", "Kolmogorov-Smirnov Distance")
  p_js <- create_metric_plot("JS", "Jensen-Shannon Divergence")
  p_w <- create_metric_plot("Wasserstein", "Wasserstein Distance")
  
  combined <- gridExtra::arrangeGrob(p_ks, p_js, p_w, nrow = 3)
  
  if (!is.null(output_file)) {
    ggplot2::ggsave(output_file, combined, width = width, height = height)
    message("Saved: ", output_file)
  }
  
  grid::grid.draw(combined)
  invisible(list(plot = combined, ks = p_ks, js = p_js, wasserstein = p_w))
}



# Run comparison for each variable:
results_US <- compare_methods_to_reference(
  method_list = list(
    SALVE = exp1$US,
    GEX = exp2$US,
    longread = longread$US
  ),
  ref_vec = fish$US,
  var_name = "US"
)

results_SS <- compare_methods_to_reference(
  method_list = list(
    SALVE = exp1$SS,
    GEX = exp2$SS,
    longread = longread$SS
  ),
  ref_vec = fish$SS,
  var_name = "SS"
)

results_MS <- compare_methods_to_reference(
  method_list = list(
    SALVE = exp1$MS,
    GEX = exp2$MS,
    longread = longread$MS
  ),
  ref_vec = fish$MS,
  var_name = "MS"
)

results_spliced <- compare_methods_to_reference(
  method_list = list(
    SALVE = exp1$spliced,
    GEX = exp2$spliced,
    longread = longread$spliced
  ),
  ref_vec = fish$spliced,
  var_name = "spliced"
)


# Compile and plot
results_list <- list(results_US, results_spliced, results_SS, results_MS)
output.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS023/plots/"

plot_barplot_comparison(
  results_list = results_list,
  var_names = c("US", "spliced", "SS", "MS"),
  method_names = c("SALVE", "GEX", "longread"),
  ref_name = "HCR",
  output_file = paste0(output.dir, "four_method_comparison.pdf"),
  width = 16,
  height = 14
)
```

```{r}
results_US <- compare_methods_to_reference(
  method_list = list(
    SALVE = exp1$US,
    GEX = exp2$US,
    combined = combined$US
  ),
  ref_vec = fish$US,
  var_name = "US"
)

results_SS <- compare_methods_to_reference(
  method_list = list(
    SALVE = exp1$SS,
    GEX = exp2$SS,
    combined = combined$SS
  ),
  ref_vec = fish$SS,
  var_name = "SS"
)

results_MS <- compare_methods_to_reference(
  method_list = list(
    SALVE = exp1$MS,
    GEX = exp2$MS,
    combined = combined$MS
  ),
  ref_vec = fish$MS,
  var_name = "MS"
)

results_spliced <- compare_methods_to_reference(
  method_list = list(
    SALVE = exp1$spliced,
    GEX = exp2$spliced,
    combined = combined$spliced
  ),
  ref_vec = fish$spliced,
  var_name = "spliced"
)

results_US$distance_df
results_SS$distance_df
results_MS$distance_df
results_spliced$distance_df

results_list <- list(results_US, results_spliced, results_SS, results_MS)
output.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS023/plots/"

plot_barplot_comparison(
  results_list = results_list,
  var_names = c("US", "spliced", "SS", "MS"),
  method_names = c("SALVE", "GEX", "combined"),
  ref_name = "HCR",
  output_file = paste0(output.dir, "four_method_comparison_wcombined.pdf"),
  width = 16,
  height = 14
)
```



QQ plot
```{r QQ plot}
plot_qq_comparison <- function(
    method_list,
    ref_vec,
    var_name = "Variable",
    n_boot = 1000,
    seed = 123
) {
  set.seed(seed)
  
  method_names <- names(method_list)
  n_methods <- length(method_list)
  
  probs <- seq(0.01, 0.99, by = 0.01)
  q_ref <- quantile(ref_vec, probs)
  
  # Compute quantiles for each method
  q_methods <- lapply(method_list, function(vec) quantile(vec, probs))
  
  # Bootstrap CI for reference (null expectation)
  n_boot_size <- max(sapply(method_list, length))
  boot_q <- matrix(NA, nrow = n_boot, ncol = length(probs))
  
  for (i in seq_len(n_boot)) {
    boot_samp <- sample(ref_vec, n_boot_size, replace = TRUE)
    boot_q[i, ] <- quantile(boot_samp, probs)
  }
  
  ci_lower <- apply(boot_q, 2, quantile, probs = 0.025)
  ci_upper <- apply(boot_q, 2, quantile, probs = 0.975)
  
  # Compute deviation statistics for each method
  dev_stats <- lapply(method_names, function(nm) {
    dev <- q_methods[[nm]] - q_ref
    c(mad = mean(abs(dev)), max_dev = max(abs(dev)))
  })
  names(dev_stats) <- method_names
  
  # Build subtitle with deviation stats
  subtitle_parts <- sapply(method_names, function(nm) {
    sprintf("%s: MAD=%.1f, Max=%.1f", nm, dev_stats[[nm]]["mad"], dev_stats[[nm]]["max_dev"])
  })
  subtitle_text <- paste(subtitle_parts, collapse = " | ")
  
  # Combine into long-format data frame
  df_list <- lapply(method_names, function(nm) {
    data.frame(
      ref = q_ref,
      exp = q_methods[[nm]],
      method = nm
    )
  })
  df <- do.call(rbind, df_list)
  df$method <- factor(df$method, levels = method_names)
  
  ci_df <- data.frame(ref = q_ref, ci_lower = ci_lower, ci_upper = ci_upper)
  
  # Color palette matching barplot
  method_colors <- c("#4DAF4A", "#E41A1C", "#377EB8", "#984EA3")
  names(method_colors) <- c("SALVE", "GEX", "longread", "HCR")
  plot_colors <- method_colors[method_names]
  
  # Axis limits
  all_exp <- unlist(q_methods)
  
  ggplot() +
    geom_ribbon(data = ci_df, aes(x = ref, ymin = ci_lower, ymax = ci_upper),
                fill = "grey70", alpha = 0.4) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey40") +
    geom_line(data = df, aes(x = ref, y = exp, color = method), linewidth = 1) +
    scale_color_manual(values = plot_colors) +
    coord_cartesian(
      xlim = c(0, max(q_ref) * 1.05),
      ylim = c(0, max(all_exp) * 1.05)
    ) +
    theme_bw() +
    theme(
      legend.position = "top",
      plot.subtitle = element_text(size = 9)
    ) +
    labs(
      x = "Reference quantiles (HCR)",
      y = "Method quantiles",
      title = paste0(var_name, ": Q-Q plot vs HCR reference"),
      subtitle = subtitle_text,
      color = NULL
    )
}


plot_qq_grid <- function(
    method_data,
    ref_data,
    var_names = c("US", "spliced", "SS", "MS"),
    method_names = c("SALVE", "GEX", "longread"),
    output_file = NULL,
    width = 12,
    height = 10,
    n_boot = 1000,
    seed = 123
) {
  
  plot_list <- list()
  
  for (var in var_names) {
    method_list <- setNames(
      lapply(method_names, function(m) method_data[[m]][[var]]),
      method_names
    )
    
    p <- plot_qq_comparison(
      method_list = method_list,
      ref_vec = ref_data[[var]],
      var_name = var,
      n_boot = n_boot,
      seed = seed
    )
    
    plot_list[[var]] <- p
  }
  
  combined <- gridExtra::arrangeGrob(grobs = plot_list, ncol = 2)
  
  if (!is.null(output_file)) {
    ggplot2::ggsave(output_file, combined, width = width, height = height)
    message("Saved: ", output_file)
  }
  
  grid::grid.draw(combined)
  invisible(plot_list)
}


# Single plot:
# plot_qq_comparison(
#   method_list = list(
#     SALVE = df_SALVE$US,
#     GEX = df_GEX$GEX.US,
#     longread = df_longread$US
#   ),
#   ref_vec = sample2$US,
#   var_name = "US"
# )

# Grid of plots:
method_data <- list(
  SALVE = list(US = exp1$US, SS = exp1$SS, MS = exp1$MS, spliced = exp1$spliced),
  GEX = list(US = exp2$US, SS = exp2$SS, MS = exp2$MS, spliced = exp2$spliced),
  longread = list(US = longread$US, SS = longread$SS, MS = longread$MS, spliced = longread$spliced)
)
ref_data <- list(US = fish$US, SS = fish$SS, MS = fish$MS, spliced = fish$spliced)

plot_qq_grid(
  method_data = method_data,
  ref_data = ref_data,
  var_names = c("US", "spliced", "SS", "MS"),
  output_file = paste0(output.dir, "QQ_four_grid.pdf")
)

```


For Elena's RPPR
```{r}
D195 <- current_df %>%
  mutate(spliceform = if_else(US > 0 & SALVE > 0, "US", "any")) %>%
  mutate(spliceform = if_else(SALVE > 0, spliceform, "none")) %>%
  arrange(spliceform != "none")
output.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/joint/plots/"

umap <- ggplot(D195, aes(x = UMAP1, y = UMAP2)) +
  geom_point(aes(color = spliceform), size = 1, shape = 16) +
  theme_classic() +
  scale_color_manual(values = c("none" = "gray93", "US" = "purple", "any" = "red")) +
  theme(legend.position = "bottom",
        legend.title = element_text(size = rel(0.6)),
        legend.text = element_text(size = rel(0.6), angle = 30),
        axis.text = element_blank(),
        axis.ticks = element_blank()) +
  labs(title = paste0("D195 joint SIV spliceforms: SALVE"), color = "")
print(umap)

ggsave(umap, file = paste0(output.dir, "D195_SALVE_spliceforms.pdf"),
       device = "pdf")


invitro <- df %>%
  mutate(spliceform = case_when(
    SALVE == 0 ~ "none",
    spliced > 0 ~ "spliced",
    spliced == 0 & US > 0 ~ "nospliced",
    spliced == 0 & US == 0 ~ "any"
  )) %>%
  mutate(spliceform = factor(spliceform, levels = c("none", "any", "nospliced", "spliced"))) %>%
  arrange(spliceform)
output.dir <- "/projects/b1042/GoyalLab/egrody/extractedData/EGS018/joint/plots/"

umap <- ggplot(invitro, aes(x = UMAP1, y = UMAP2)) +
  geom_point(aes(color = spliceform), size = 1, shape = 16) +
  theme_classic() +
  scale_color_manual(values = c("none" = "gray93", "spliced" = "blue", "nospliced" = "purple", "any" = "red")) +
  theme(legend.position = "bottom",
        legend.title = element_text(size = rel(0.6)),
        legend.text = element_text(size = rel(0.6), angle = 30),
        axis.text = element_blank(),
        axis.ticks = element_blank()) +
  labs(title = paste0("Invitro joint SIV spliceforms: SALVE"), color = "")
print(umap)

ggsave(umap, file = paste0(output.dir, "Invitro_SALVE_spliceforms.pdf"),
       device = "pdf")

```




